---
title: "Lecture 03 — Describing Our Data"
subtitle: "Wrangling, summaries, statistics, and our first real comparisons"
author: "Bill Perry"
date: today
format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 02)

- **Setup** — R and Positron installed; project folders created (`data/`, `figures/`, `scripts/`)
- **R as a calculator** — arithmetic, order of operations, assignment with `<-`
- **Vectors** — `c()`, data types (`numeric`, `character`, `logical`)
- **Functions** — calling built-in functions, reading help with `?`
- **Loading data** — `read_excel()` and `read_csv()`; inspecting with `glimpse()`, `head()`, `dim()`
- **Pipe** — `%>%` reads as "then"; chains steps together
- **First ggplot** — boxplot + `geom_jitter()`, saved as PNG

::: callout-note
**✅ Key idea from Lecture 02**

You can get data into R and make a plot. Today we learn to *wrangle* and *describe* what we actually have.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
**Part 1 — Tidyverse wrangling:**

- `filter()` — pick rows by a condition
- `select()` — pick columns by name
- `mutate()` — create new columns
- `arrange()` — sort rows
- Chain them all into a **pipeline**

**Part 2 — Descriptive statistics:**

- mean, median, variance, SD, SE
- NA handling — `sum(!is.na())`
- `group_by()` + `summarize()`
- `skimr` for a fast full overview
- Boxplot + mean ± SE plots

::: callout-tip
**🖐 Try it yourself**

By the end you will wrangle and describe our leaf data in one connected pipeline.
:::
::::

::: {.column width="40%"}
**Our tools today:**

- `readxl`
- `tidyverse`
- `skimr`

**References:**

- 📖 [R4DS Ch 3](https://r4ds.hadley.nz/data-transform) — Data Transformation
- 📖 [R4DS Ch 13](https://r4ds.hadley.nz/numbers) — Numbers
- 📖 [R4DS Ch 18](https://r4ds.hadley.nz/missing-values) — Missing Values
- 🌐 [Data Carpentry](https://datacarpentry.github.io/R-ecology-lesson/)

**Naming conventions:**

- data frames → `_df`
- plots → `_plot`
- models → `_model`
:::
::::::

# Load Libraries and Data

```{r load-libs-03}
# Load all packages at the top of every script ----------
library(readxl)      # reading Excel files
library(tidyverse)   # data wrangling + ggplot2
library(skimr)       # fast descriptive summaries
```

```{r load-data-03}
# Read the leaf data from the data folder ---------------
tree_df <- read_excel("data/2026_06_25_tree_experiment_raw_data.xlsx")
```

::: callout-note
**Install once — load every session**

- Run in the Console one time only: `install.packages("skimr")`
- Then every session: `library(skimr)` activates it.
:::

# Inspect the Data

::::: columns
::: {.column width="60%"}
```{r inspect-head-03}
# Preview the first rows ----------------------------
head(tree_df)
```
:::

::: {.column width="40%"}
```{r inspect-glimpse-03}
# Column names, types, first values -----------------
glimpse(tree_df)
```

```{r inspect-dim-03}
# Rows × columns
dim(tree_df)
```

- `<chr>` = character
- `<dbl>` = numeric
:::
:::::

# Part 1 · The Core Tidyverse Verbs

:::::: columns
::: {.column width="60%"}
Five functions do most of the work in data wrangling:

| Function      | What it does                       |
|---------------|------------------------------------|
| `filter()`    | keep **rows** matching a condition |
| `select()`    | keep **columns** by name           |
| `mutate()`    | **add or change** a column         |
| `arrange()`   | **sort** rows                      |
| `summarize()` | **collapse** rows to a summary     |

All take a data frame as first input (via `%>%`) and return a data frame.

📖 [R4DS §3 — Data Transformation](https://r4ds.hadley.nz/data-transform)
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

These five verbs are the tidyverse grammar for data. Learn them once and you can wrangle almost any dataset.
:::

Think of them as building blocks:

```         
filter()   → rows you want
select()   → columns you want
mutate()   → new columns you need
arrange()  → order the rows
summarize()→ collapse to summaries
```
::::
::::::

# `filter()` — Picking Rows

:::::: columns
::: {.column width="60%"}
```{r filter-demo}
# filter() keeps rows where the condition is TRUE -------

# Keep only sunny leaves
tree_df %>% filter(side == "sunny") %>% head()   

# Keep only leaves heavier than 5 g
tree_df %>% filter(weight_g > 5) %>% head()

# Combine two conditions with & (AND)
tree_df %>% filter(side == "shady", weight_g > 7)%>% head()
```
:::

:::: {.column width="40%"}
**Common comparison operators:**

| Symbol    | Meaning                             |
|-----------|-------------------------------------|
| `==`      | exactly equal                       |
| `!=`      | not equal                           |
| `>` `<`   | greater / less than                 |
| `>=` `<=` | greater / less than or equal        |
| is.na     | return all NA rows                  |
| !is.na    | return all rows that are **NOT NA** |

::: callout-warning
**⚠️ Watch out!**

- `=` assigns a value.
- `==` tests equality.
- `filter(side = "sunny")` → error.
- `filter(side == "sunny")` → correct.
:::
::::
::::::

# `select()` — Picking Columns

::::: columns
::: {.column width="60%"}
```{r select-demo}
# select() keeps the columns you name ------------------

# Keep only side and weight
tree_df %>% select(side, weight_g)%>% head()

# Drop the index column (use minus sign)
tree_df %>% select(-index)%>% head()

# Keep columns that start with a word
tree_df %>% select(starts_with("weight")) %>% head()
```
:::

::: {.column width="40%"}
- **Why `select()` matters:**
  - Real datasets often have 50–100+ columns
  - You rarely need all of them
  - `select()` keeps your workspace clean
- **Useful helpers:**
  - `starts_with("x")` — columns starting with "x"
  - `ends_with("_g")` — columns ending with "\_g"
  - `contains("cm")` — columns containing "cm"

📖 [R4DS §3.3](https://r4ds.hadley.nz/data-transform#select)
:::
:::::

# `mutate()` — Creating New Columns

:::::: columns
::: {.column width="60%"}
```{r mutate-demo}
# mutate() adds a new column to the data frame ---------

# Convert grams to milligrams
tree_df %>%
  mutate(weight_mg = weight_g * 1000) %>% head()

# Add a size category based on weight
tree_df %>%
  mutate(size_class = if_else(weight_g > 5, "large", "small")) %>% head()

# mutate() can reference multiple existing columns
tree_df %>%
  mutate(
    weight_mg  = weight_g * 1000,
    area_approx = width_cm * height_cm   # rough area estimate
  ) %>% head()
```
:::

:::: {.column width="40%"}
- `mutate()` keeps **all existing columns** and adds new ones
- **Use `if_else(condition, value_if_true, value_if_false)` to create categories**

::: callout-note
**✅ Key idea**

`mutate()` never removes columns — it adds to the right of your data frame.

To overwrite a column, just use the same name on the left: `mutate(weight_g = round(weight_g, 1))`
:::

📖 [R4DS §3.3](https://r4ds.hadley.nz/data-transform#sec-mutate)
::::
::::::

# `arrange()` — Sorting Rows

:::::: columns
::: {.column width="60%"}
```{r arrange-demo}
# arrange() sorts rows by one or more columns ----------

# Lightest leaves first
tree_df %>% arrange(weight_g)

# Heaviest leaves first (use desc() to reverse)
tree_df %>% arrange(desc(weight_g))

# Sort by side, then by weight within each side
tree_df %>% arrange(side, desc(weight_g))
```
:::

:::: {.column width="40%"}
- Default: **ascending** order (smallest first)
- `desc(column)` → **descending** order (largest first)
- Very useful before printing a table or checking outliers

::: callout-tip
`arrange()` is rarely used in the middle of an analysis pipeline — it is most useful at the **end** when you want to inspect your results in a specific order.
:::
::::
::::::

# The Full Pipeline

:::::: columns
::: {.column width="60%"}
```{r full-pipeline}
# Chain all the verbs together -------------------------
# Read it top to bottom like a recipe

wrangled_df <- tree_df %>%
  filter(weight_g > 0) %>%             # remove any zeros
  select(side, weight_g, height_cm) %>% # keep only needed cols
  mutate(
    weight_mg    = weight_g * 1000,
    size_class   = if_else(weight_g > 5, "large", "small")
  ) %>%
  arrange(side, desc(weight_g))        # sort by side, then weight

head(wrangled_df)
```
:::

:::: {.column width="40%"}
**Read the pipeline out loud:**

> - Take `tree_df`,
>   - **then** filter to non-zero weights,
>     - **then** select three columns,
>       - **then** add two new columns,
>         - **then** sort by side and weight.

::: callout-note
**✅ Key idea**

The pipeline `%>%` is what makes tidyverse readable. You can see every step in order — unlike nested function calls that read inside-out.
:::
::::
::::::

# Part 2 · Our Scientific Question

::::: columns
::: {.column width="60%"}
**Biological prediction:** Leaves on the *shady* side of a tree will be **larger and heavier** to capture more of the limited light.

|   | Hypothesis |
|------------------------------------|------------------------------------|
| **H₀ — null** | No difference in leaf size between sunny and shady sides |
| **Hₐ — alternate** | Leaf size differs between sunny and shady sides |
:::

::: {.column width="40%"}
**Plan for this unit:**

1.  **Today** — *describe* the data
2.  **Lecture 04** — formally *test* the hypotheses with a t-test

Statistics helps us decide which hypothesis is better supported by the data.
:::
:::::

# What Is the Mean?

::::: columns
::: {.column width="60%"}
$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

- Add up all values, divide by *n*
- The **balance point** of the distribution
- **Sensitive to outliers** — one very large value pulls the mean up

**Example:**

Values: 5, 6, 7, 8, **50** → mean = 15.2

*Is 15.2 a fair summary? Not really — the outlier dominates!*
:::

::: {.column width="40%"}
**Why it matters:**

- The most commonly reported statistic
- Used in every t-test, ANOVA, and regression
- Always pair it with a measure of **spread** (SD or SE)

📖 [R4DS §13 — Numbers](https://r4ds.hadley.nz/numbers)
:::
:::::

# Calculating the Mean — filter() and pull()

::::: columns
::: {.column width="60%"}
```{r filter-pull-sunny}
# Use filter() + pull() to extract one group's values --
sunny_wt <- tree_df %>%
  filter(side == "sunny") %>%
  pull(weight_g)

sunny_wt   # look at the raw values
```

```{r mean-sunny}
# Calculate the mean -----------------------------------
mean_sunny <- mean(sunny_wt, na.rm = TRUE)
cat("Mean sunny weight:", round(mean_sunny, 2), "g\n")
```
:::

::: {.column width="40%"}
- `filter(side == "sunny")` — keeps only sunny rows
- `pull(weight_g)` — extracts that column as a **vector**
- `na.rm = TRUE` — removes any `NA` before computing

**Key pattern:**

`filter()` → picks **rows**

`pull()` → extracts a **column** as a vector

📖 [R4DS §3 — Data Transformation](https://r4ds.hadley.nz/data-transform)
:::
:::::

# What Is the Median?

::::: columns
::: {.column width="60%"}
**The middle value when data are sorted**

| Sorted data            | Median                |
|------------------------|-----------------------|
| 5, 6, **7**, 8, 50     | **7**                 |
| 5, 6, **7, 8**, 50, 51 | **(7 + 8) / 2 = 7.5** |

- Not pulled by outliers → **robust**
- If **mean ≫ median** → right-skewed
- If **mean ≈ median** → roughly symmetric
:::

::: {.column width="40%"}
```{r median-calc}
# Median for each side ---------------------------------
shady_wt <- tree_df %>%
  filter(side == "shady") %>%
  pull(weight_g)

med_sunny <- median(sunny_wt, na.rm = TRUE)
med_shady <- median(shady_wt, na.rm = TRUE)

cat("Median sunny:", round(med_sunny, 2), "\n")
cat("Median shady:", round(med_shady, 2), "\n")
```
:::
:::::

# What Is Variance?

:::::: columns
::: {.column width="60%"}
$$s^2 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n - 1}$$

1.  For each value, compute distance from mean: $(x_i - \bar{x})$
2.  **Square** those distances → all positive
3.  Sum them up, divide by $n - 1$

Dividing by $n - 1$ gives an **unbiased** estimate of population variance.

**Larger variance = more spread = more uncertainty**
:::

:::: {.column width="40%"}
```{r variance-calc}
# Variance for each side ------------------------------
var_sunny <- var(sunny_wt, na.rm = TRUE)
var_shady <- var(shady_wt, na.rm = TRUE)

cat("Variance sunny:", round(var_sunny, 2), "\n")
cat("Variance shady:", round(var_shady, 2), "\n")
```

::: callout-note
Variance is in **squared units** (g²) — hard to interpret. That is why we use SD.
:::

📖 [R4DS §13.3](https://r4ds.hadley.nz/numbers#numeric-summaries)
::::
::::::

# What Is Standard Deviation?

::::: columns
::: {.column width="60%"}
$$s = \sqrt{s^2} = \sqrt{\frac{\sum(x_i - \bar{x})^2}{n-1}}$$

- The **square root of variance** → back in original units (g) ✓
- For a normal distribution:
  - \~**68%** of values fall within mean ± 1 SD
  - \~**95%** fall within mean ± 2 SD
- Reports how spread out *individual* data points are
:::

::: {.column width="40%"}
```{r sd-calc}
# Standard deviation for each side -------------------
sd_sunny <- sd(sunny_wt, na.rm = TRUE)
sd_shady <- sd(shady_wt, na.rm = TRUE)

cat("SD sunny:", round(sd_sunny, 2), "g\n")
cat("SD shady:", round(sd_shady, 2), "g\n")
```
:::
:::::

# What Is Standard Error?

::::: columns
::: {.column width="60%"}
$$SE = \frac{s}{\sqrt{n}}$$

- Measures how precisely we know the **mean**
- Not the spread of raw data — the **uncertainty of the mean itself**
- Gets **smaller** as *n* increases → more data = more precise estimate

|        | Describes                          |
|--------|------------------------------------|
| **SD** | spread of *individual* data points |
| **SE** | precision of the *sample mean*     |

We report SE when making claims about group means.
:::

::: {.column width="40%"}
```{r se-calc}
# SE = SD / sqrt(n) — need correct n -----------------
n_sunny <- sum(!is.na(sunny_wt))
n_shady <- sum(!is.na(shady_wt))

se_sunny <- sd_sunny / sqrt(n_sunny)
se_shady <- sd_shady / sqrt(n_shady)

cat("n sunny  =", n_sunny, "\n")
cat("SE sunny =", round(se_sunny, 2), "g\n\n")
cat("n shady  =", n_shady, "\n")
cat("SE shady =", round(se_shady, 2), "g\n")
```
:::
:::::

# The Problem with `length()`

::::: columns
::: {.column width="60%"}
```{r length-problem-demo}
# length() counts ALL positions — including NAs -------
x <- c(4, 3, NA, 7, NA)   # a vector with missing values

length(x)              # returns 5 — but 2 are NA!
sum(is.na(x))          # how many NAs are there?
```

```{r length-wrong-mean}
# Using length() in SE gives the WRONG answer ---------
wrong_n  <- length(x)
wrong_se <- sd(x, na.rm = TRUE) / sqrt(wrong_n)

cat("Wrong n  =", wrong_n,              "\n")
cat("Wrong SE =", round(wrong_se, 3),   "\n")
```
:::

::: {.column width="40%"}
- `length(x)` counts **all positions** in the vector — including `NA`
- This is a **silent error** — R will not warn you!
- Using n = 5 in SE = SD / √n when n should be 3 gives the wrong answer

📖 [R4DS §18 — Missing Values](https://r4ds.hadley.nz/missing-values)
:::
:::::

# Fix — `sum(!is.na())`

:::::: columns
::: {.column width="60%"}
```{r na-fix-step-by-step}
# Break it down step by step -------------------------
x <- c(4, 3, NA, 7, NA)

is.na(x)          # TRUE where NA, FALSE where real
!is.na(x)         # FLIP: TRUE where real, FALSE where NA
sum(!is.na(x))    # count the TRUEs = number of real values
```

```{r na-fix-real}
# Apply to our real leaf data ------------------------
n_sun_check <- sum(!is.na(sunny_wt))
cat("Non-missing sunny weights =", n_sun_check, "\n")
```
:::

:::: {.column width="40%"}
**How it works step by step:**

1.  `is.na(x)` → `TRUE` for each `NA`
2.  `!is.na(x)` → flips it: `TRUE` for **real** values
3.  `sum(...)` → counts `TRUE`s

::: callout-tip
**Best practice:** always use `sum(!is.na(x))` when you need *n* — even when you think there are no NAs.
:::
::::
::::::

# All Stats — Base R, One Group at a Time

::::: columns
::: {.column width="60%"}
```{r basestats-manual}
# All stats for the sunny side -----------------------
n_s    <- sum(!is.na(sunny_wt))
mean_s <- mean(sunny_wt,   na.rm = TRUE)
med_s  <- median(sunny_wt, na.rm = TRUE)
sd_s   <- sd(sunny_wt,     na.rm = TRUE)
se_s   <- sd_s / sqrt(n_s)

cat("--- Sunny Leaves ---\n")
cat("n      =", n_s,             "\n")
cat("Mean   =", round(mean_s, 2), "\n")
cat("Median =", round(med_s, 2),  "\n")
cat("SD     =", round(sd_s, 2),   "\n")
cat("SE     =", round(se_s, 2),   "\n")
```
:::

::: {.column width="40%"}
- Every function takes `na.rm = TRUE`
- `cat()` prints clean labelled output
- This is **clear** but repetitive for many groups → that is why we use `group_by()` next
:::
:::::

# Stats the Tidy Way — `group_by()` + `summarize()`

::::: columns
::: {.column width="60%"}
```{r groupby-stats}
# Calculate all stats for both sides at once ----------
stats_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    n       = sum(!is.na(weight_g)),
    mean_wt = round(mean(weight_g,   na.rm = TRUE), 2),
    med_wt  = round(median(weight_g, na.rm = TRUE), 2),
    sd_wt   = round(sd(weight_g,     na.rm = TRUE), 2),
    se_wt   = round(sd_wt / sqrt(n), 2)
  )

stats_df
```
:::

::: {.column width="40%"}
- `group_by(side)` — splits the data by the `side` column
- `summarize()` — collapses each group to one row of stats
- You get **both groups in one clean table**
- You can reference a column just created (`sd_wt` → `se_wt`)

📖 [R4DS §3.5 — `summarize()`](https://r4ds.hadley.nz/data-transform#sec-summarize)
:::
:::::

# Multiple Variables at Once

::::: columns
::: {.column width="60%"}
```{r groupby-multi}
# One group_by() pipe summarizes all measurements -----
size_stats_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    n       = sum(!is.na(weight_g)),
    mean_wt = round(mean(weight_g,  na.rm = TRUE), 2),
    mean_ht = round(mean(height_cm, na.rm = TRUE), 2),
    mean_wd = round(mean(width_cm,  na.rm = TRUE), 2)
  )

size_stats_df
```
:::

::: {.column width="40%"}
**With one `group_by()` pipe** you summarize **all three measurements** across both sides.

**Do the numbers support our hypothesis?**

- Shady heavier? → check `mean_wt`
- Shady taller? → check `mean_ht`
- Shady wider? → check `mean_wd`
:::
:::::

# Fast Summaries with `skimr`

```{r skimr-skim}
# skim() grouped by side — full column summaries ------
tree_df %>%
  group_by(side) %>%
  skim()
```

:::::: columns
::: {.column width="60%"}
Per column and per group: `n_missing`, `mean`, `sd`, percentiles (`p0` to `p100`), and a mini histogram.
:::

:::: {.column width="40%"}
::: callout-tip
`skim()` is your fastest **first look** at any dataset. Pair it with `group_by()` to compare groups instantly.
:::
::::
::::::

# Boxplot — Basic Structure

::::: columns
::: {.column width="60%"}
```{r boxplot-basic}
#| fig-format: png
# Basic boxplot — data is already long format ---------
tree_box_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(
    title = "Leaf Weight by Tree Side",
    x     = "Side of Tree",
    y     = "Leaf Weight (g)",
    fill  = "Side"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

tree_box_plot
```
:::

::: {.column width="40%"}
**Boxplot anatomy:**

- Middle line = **median**
- Box edges = 25th and 75th percentile (**IQR**)
- Whiskers = 1.5 × IQR
- Dots beyond whiskers = **outliers**

Because our data is in **long format**, `ggplot2` reads `side` straight off the x-axis — no reshaping needed.

📖 [R4DS §9 — Layers](https://r4ds.hadley.nz/layers)
:::
:::::

# Boxplot — Overlay Individual Points

:::::: columns
::: {.column width="60%"}
```{r boxplot-jitter}
#| fig-format: png
# Add raw data points over the boxplot ----------------
tree_jitter_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha    = 0.6,
    size     = 2.5
  ) +
  labs(
    title = "Leaf Weight by Tree Side",
    x     = "Side of Tree",
    y     = "Leaf Weight (g)",
    fill  = "Side"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

tree_jitter_plot
```
:::

:::: {.column width="40%"}
**Key arguments:**

- `position_jitter(width = 0.15)` — spreads points left and right
- `seed = 42` — same jitter layout every render
- `alpha = 0.6` — semi-transparent to show overlap
- `size = 2.5` — point size

::: callout-tip
Always show raw points with a boxplot — with only 10 leaves per group, every point matters!
:::
::::
::::::

# Mean ± SE Plot — `stat_summary()`

::::: columns
::: {.column width="60%"}
```{r mean-se-plot}
#| fig-format: png
# stat_summary() computes and plots mean and SE -------
tree_mean_se_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, color = side)) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha    = 0.35, size = 2
  ) +
  stat_summary(fun      = mean,    geom = "point",
               size = 4) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.15, linewidth = 0.9) +
  labs(
    title = "Mean ± SE Leaf Weight by Tree Side",
    x     = "Side of Tree",
    y     = "Leaf Weight (g)",
    color = "Side"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

tree_mean_se_plot
```
:::

::: {.column width="40%"}
**Two `stat_summary()` layers:**

| call                 | what it draws           |
|----------------------|-------------------------|
| `fun = mean`         | large point at the mean |
| `fun.data = mean_se` | error bars for ± 1 SE   |

- Individual points in the **background** (`alpha = 0.35`)
- Mean shown as a **larger foreground point**
- Error bars = **± 1 SE**
:::
:::::

# What We Learned Today

::::: columns
::: {.column width="60%"}
- **R wrangling skills:**
  - `filter()` — keep rows by condition
  - `select()` — keep columns by name
  - `mutate()` — add new calculated columns
  - `arrange()` — sort rows
  - Full pipeline: chain all verbs with `%>%`
- **Statistics concepts:**
  - **Mean** — balance point; sensitive to outliers
  - **Median** — middle value; robust
  - **Variance / SD** — spread of individual data
  - **SE** — precision of the sample mean
- **More R skills:**
  - `group_by()` + `summarize()` — all stats for all groups at once
  - `sum(!is.na())` — the right way to count *n*
  - `skim()` — instant full dataset overview
  - `geom_boxplot()` + `geom_point()`, `stat_summary()`
:::

::: {.column width="40%"}
**References:**

- 📖 [R4DS §3](https://r4ds.hadley.nz/data-transform) — Data Transformation
- 📖 [R4DS §13](https://r4ds.hadley.nz/numbers) — Numbers
- 📖 [R4DS §18](https://r4ds.hadley.nz/missing-values) — Missing Values
- 🌐 [Data Carpentry](https://datacarpentry.github.io/R-ecology-lesson/)

**Up next — Lecture 04:**

- Two-sample t-test
- Checking assumptions (normality, equal variance)
- Interpreting and reporting results
:::
:::::