---
title: "Lecture 03 — Describing Our Data"
subtitle: "Summaries, statistics, and our first real comparisons"
author: "Bill Perry"
date: today

format:
  html:
    downloads: [docx, pptx, typst]  # This creates download links for all three
  revealjs:
    output-ext: "slides.html"
  docx: default
  pptx: default
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 02)

- **Setup** — R and Positron installed; project folders created (`documents/`, `data/`, `output/`, `figures/`, `scripts/`)
- **R as a calculator** — arithmetic, order of operations
- **Assignment** — storing values with `<-`
- **Vectors** — `c()`, indexing, data types (`numeric`, `character`, `logical`)
- **Functions** — calling built-in functions, passing arguments
- **First look** — loaded our leaf data with `read_excel()` and made a first `ggplot`

::: callout-note
**✅ Key idea from Lecture 02**

You can get data into R and make a plot. Today we learn to *describe* and *summarize* what we actually have.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- **Review** what descriptive statistics really mean
  - mean, median, variance, SD, SE
  - a first look at *p*-values
- **Count observations** correctly — the NA problem
- **Compute stats** three ways:
  - base R, `tidyverse`, and `skimr`
- **Visualize** distributions:
  - boxplots with individual points overlaid
  - mean ± SE summary plots

::: callout-tip
**🖐 Try it yourself**

By the end you will compute and plot descriptive stats on our own leaf data.
:::
::::

::: {.column width="40%"}
**Our tools today:**

- `readxl`
- `tidyverse`
- `skimr`

**References:**

- 📖 [R4DS Ch 3, 13, 18, 27](https://r4ds.hadley.nz/)
- 🌐 [Data Carpentry](https://datacarpentry.github.io/R-ecology-lesson/)

**Naming conventions:**

- data frames → `_df`
- plots → `_plot`
- models → `_model`
:::
::::::

# Our Scientific Question

::::: columns
::: {.column width="60%"}
**Biological prediction:** Leaves on the *shady* side of a tree will be **larger and heavier** to capture more of the limited light.

|   | Hypothesis |
|------------------------------------|------------------------------------|
| **H₀ — null** | No difference in leaf size between sunny and shady sides |
| **Hₐ — alternate** | Leaf size differs between sunny and shady sides |
:::

::: {.column width="40%"}
**Plan for this unit:**

1.  **Today** — *describe* the data
2.  **Lecture 04** — formally *test* the hypotheses with a t-test

Statistics helps us decide which hypothesis is better supported by the data.
:::
:::::

# Our Data — Long Format

::::: columns
::: {.column width="60%"}
**Each row = one leaf**

| Column      | Description            |
|-------------|------------------------|
| `index`     | leaf number (1–20)     |
| `side`      | `"sunny"` or `"shady"` |
| `weight_g`  | leaf weight (g)        |
| `width_cm`  | leaf width (cm)        |
| `height_cm` | leaf height (cm)       |

This is called **long format** — one measurement per row, group label in a column.
:::

::: {.column width="40%"}
**Long vs Wide format:**

| Shape | Layout |
|------------------------------------|------------------------------------|
| **Long** | one row per leaf; `side` in a column |
| **Wide** | one row per leaf *pair*; `sunny_wt`, `shady_wt` as separate columns |

R and `ggplot2` work best with **long** format. We will convert between the two in a later lecture.

📖 [R4DS §5 — Data Tidying](https://r4ds.hadley.nz/data-tidy)
:::
:::::

# Load Libraries

```{r load-libraries}
# Load all packages — always at the top of every script --------
library(readxl)      # reading Excel files (.xlsx)
library(tidyverse)   # data manipulation + ggplot2
library(skimr)       # fast descriptive column summaries
```

:::::: columns
:::: {.column width="60%"}
::: callout-note
**Install once — load every session**

Run in the Console **one time only:**

``` r
install.packages("readxl")
install.packages("skimr")
```

Then every session: `library(readxl)` activates it.
:::
::::

::: {.column width="40%"}
**Why `readxl`?**

`read_csv()` reads `.csv` files. `read_excel()` reads `.xlsx` files. Both land the data in an R data frame.

📖 [R4DS §7 — Data Import](https://r4ds.hadley.nz/data-import)
:::
::::::

# Load Our Leaf Data

::::: columns
::: {.column width="60%"}
```{r load-data}
# Read the Excel file from the data folder -----------
tree_df <- read_excel(
  "data/2026_06_25_tree_experiment_raw_data.xlsx"
)
```

```{r inspect-head}
# Preview the first rows of the data -----------------
head(tree_df)
```
:::

::: {.column width="40%"}
- Path is **relative** to your project folder
- `"data/..."` means: look in the `data/` sub-folder
- This path works on **any** computer, not just yours
- The file stays untouched — R reads; it does not edit
:::
:::::

# Inspect the Data

::::: columns
::: {.column width="60%"}
```{r inspect-glimpse}
# Column names, types, and first values --------------
glimpse(tree_df)
```
:::

::: {.column width="40%"}
```{r inspect-dim}
# Rows × columns
dim(tree_df)
```

- `<chr>` = character (text)
- `<dbl>` = numeric (decimal)

**10 sunny leaves, 10 shady leaves** — we have a **balanced** design.
:::
:::::

# What Is the Mean?

::::: columns
::: {.column width="60%"}
$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

- Add up all values, divide by *n*
- The **balance point** of the distribution
- **Sensitive to outliers** — one very large value pulls the mean up

**Example:**

Values: 5, 6, 7, 8, **50** → mean = 15.2

*Is 15.2 a fair summary? Not really — the outlier dominates!*
:::

::: {.column width="40%"}
**Why it matters:**

- The most commonly reported statistic
- Used in every t-test, ANOVA, and regression
- Always pair it with a measure of **spread** (SD or SE)

📖 [R4DS §13 — Numbers](https://r4ds.hadley.nz/numbers)
:::
:::::

# Calculating the Mean — filter() and pull()

::::: columns
::: {.column width="60%"}
```{r filter-pull-sunny}
# Extract sunny leaf weights as a plain vector --------
sunny_wt <- tree_df %>%
  filter(side == "sunny") %>%
  pull(weight_g)

sunny_wt   # look at the raw values
```

```{r mean-sunny}
# Now calculate the mean of that vector --------------
mean_sunny <- mean(sunny_wt, na.rm = TRUE)
cat("Mean sunny weight:", round(mean_sunny, 2), "g\n")
```
:::

::: {.column width="40%"}
- `filter(side == "sunny")` — keeps only sunny rows
- `pull(weight_g)` — extracts that column as a **vector**
- `na.rm = TRUE` — removes any `NA` before computing

**Key pattern:**

`filter()` → picks **rows**

`pull()` → extracts a **column** as a vector

📖 [R4DS §3 — Data Transformation](https://r4ds.hadley.nz/data-transform)
:::
:::::

# What Is the Median?

::::: columns
::: {.column width="60%"}
**The middle value when data are sorted**

| Sorted data            | Median                |
|------------------------|-----------------------|
| 5, 6, **7**, 8, 50     | **7**                 |
| 5, 6, **7, 8**, 50, 51 | **(7 + 8) / 2 = 7.5** |

- Not pulled by outliers → **robust**
- If **mean ≫ median** → right-skewed
- If **mean ≈ median** → roughly symmetric
:::

::: {.column width="40%"}
```{r median-calc}
# Median for each side via filter + pull -------------
shady_wt <- tree_df %>%
  filter(side == "shady") %>%
  pull(weight_g)

med_sunny <- median(sunny_wt, na.rm = TRUE)
med_shady <- median(shady_wt, na.rm = TRUE)

cat("Median sunny:", round(med_sunny, 2), "\n")
cat("Median shady:", round(med_shady, 2), "\n")
```
:::
:::::

# What Is Variance?

:::::: columns
::: {.column width="60%"}
$$s^2 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n - 1}$$

1.  For each value, compute distance from mean: $(x_i - \bar{x})$
2.  **Square** those distances → all positive
3.  Sum them up, divide by $n - 1$

Dividing by $n - 1$ gives an **unbiased** estimate of population variance.

**Larger variance = more spread = more uncertainty**
:::

:::: {.column width="40%"}
```{r variance-calc}
# Variance for each side ----------------------------
var_sunny <- var(sunny_wt, na.rm = TRUE)
var_shady <- var(shady_wt, na.rm = TRUE)

cat("Variance sunny:", round(var_sunny, 2), "\n")
cat("Variance shady:", round(var_shady, 2), "\n")
```

::: callout-note
Variance is in **squared units** (g²) — hard to interpret. That is why we use SD.
:::

📖 [R4DS §13.3](https://r4ds.hadley.nz/numbers#numeric-summaries)
::::
::::::

# What Is Standard Deviation?

::::: columns
::: {.column width="60%"}
$$s = \sqrt{s^2} = \sqrt{\frac{\sum(x_i - \bar{x})^2}{n-1}}$$

- The **square root of variance** → back in original units (g) ✓
- For a normal distribution:
  - \~**68%** of values fall within mean ± 1 SD
  - \~**95%** fall within mean ± 2 SD
- Reports how spread out *individual* data points are
:::

::: {.column width="40%"}
```{r sd-calc}
# Standard deviation for each side ------------------
sd_sunny <- sd(sunny_wt, na.rm = TRUE)
sd_shady <- sd(shady_wt, na.rm = TRUE)

cat("SD sunny:", round(sd_sunny, 2), "g\n")
cat("SD shady:", round(sd_shady, 2), "g\n")
```

Most sunny leaves are within ±`r round(sd_sunny, 2)` g of the mean.
:::
:::::

# What Is Standard Error?

::::: columns
::: {.column width="60%"}
$$SE = \frac{s}{\sqrt{n}}$$

- Measures how precisely we know the **mean**
- Not the spread of raw data — the **uncertainty of the mean itself**
- Gets **smaller** as *n* increases → more data = more precise estimate

|        | Describes                          |
|--------|------------------------------------|
| **SD** | spread of *individual* data points |
| **SE** | precision of the *sample mean*     |

We report SE when making claims about group means.
:::

::: {.column width="40%"}
```{r se-calc}
# SE = SD / sqrt(n) — need correct n ----------------
n_sunny <- sum(!is.na(sunny_wt))
n_shady <- sum(!is.na(shady_wt))

se_sunny <- sd_sunny / sqrt(n_sunny)
se_shady <- sd_shady / sqrt(n_shady)

cat("n sunny  =", n_sunny, "\n")
cat("SE sunny =", round(se_sunny, 2), "g\n\n")
cat("n shady  =", n_shady, "\n")
cat("SE shady =", round(se_shady, 2), "g\n")
```
:::
:::::

# A First Look at the *p*-value

:::::: columns
::: {.column width="60%"}
**The key question:** Could the difference in means be due to *random chance alone?*

A ***p*****-value** answers:

> *"If H₀ were true (no real difference), how likely is it that we would see a difference this large, just by chance?"*

- ***p*** **\< 0.05** → data are **unlikely** under H₀ → **reject H₀**
- ***p*** **≥ 0.05** → data are consistent with H₀ → **fail to reject H₀**

We will **calculate** a *p*-value in Lecture 04.
:::

:::: {.column width="40%"}
::: callout-important
**Common misconception**

A *p*-value is **not** the probability that H₀ is true.

It is the probability of observing *this data (or more extreme)* **given** that H₀ is true.
:::

Our shady leaves weigh \~`r round(mean(shady_wt),1)` g vs sunny at \~`r round(mean(sunny_wt),1)` g.

Could that gap be random chance? We will find out next lecture.
::::
::::::

# The Problem with `length()`

::::: columns
::: {.column width="60%"}
```{r length-problem-demo}
# length() counts ALL positions — including NAs ------
x <- c(4, 3, NA, 7, NA)   # a vector with missing values

length(x)              # returns 5 — but 2 are NA!
sum(is.na(x))          # how many NAs are there?
```

```{r length-wrong-mean}
# Using length() in SE gives the WRONG n and SE ------
wrong_n  <- length(x)
wrong_se <- sd(x, na.rm = TRUE) / sqrt(wrong_n)

cat("Wrong n  =", wrong_n,              "\n")
cat("Wrong SE =", round(wrong_se, 3),   "\n")
```
:::

::: {.column width="40%"}
- `length(x)` counts **all positions** in the vector — including `NA`
- This is a **silent error** — R will not warn you!
- Using n = 5 in SE = SD / √n when n should be 3 gives the wrong answer

📖 [R4DS §18 — Missing Values](https://r4ds.hadley.nz/missing-values)
:::
:::::

# Fix — `sum(!is.na())`

:::::: columns
::: {.column width="60%"}
```{r na-fix-step-by-step}
# Break it down step by step --------------------------
x <- c(4, 3, NA, 7, NA)

is.na(x)          # TRUE where NA, FALSE where real
!is.na(x)         # FLIP: TRUE where real, FALSE where NA
sum(!is.na(x))    # count the TRUEs = number of real values
```

```{r na-fix-real}
# Apply to our real leaf data ------------------------
# (Our data has no NAs — but always use this pattern!)
n_sun_check <- sum(!is.na(sunny_wt))
cat("Non-missing sunny weights =", n_sun_check, "\n")
```
:::

:::: {.column width="40%"}
**How it works step by step:**

1.  `is.na(x)` → `TRUE` for each `NA`
2.  `!is.na(x)` → flips it: `TRUE` for **real** values
3.  `sum(...)` → counts `TRUE`s

::: callout-tip
**Best practice:** always use `sum(!is.na(x))` when you need *n* — even when you think there are no NAs.
:::
::::
::::::

# All Stats — Base R, One Group at a Time

::::: columns
::: {.column width="60%"}
```{r basestats-manual}
# All stats for the sunny side using our vectors ------
n_s    <- sum(!is.na(sunny_wt))
mean_s <- mean(sunny_wt,   na.rm = TRUE)
med_s  <- median(sunny_wt, na.rm = TRUE)
sd_s   <- sd(sunny_wt,     na.rm = TRUE)
se_s   <- sd_s / sqrt(n_s)

cat("--- Sunny Leaves ---\n")
cat("n      =", n_s,             "\n")
cat("Mean   =", round(mean_s, 2), "\n")
cat("Median =", round(med_s, 2),  "\n")
cat("SD     =", round(sd_s, 2),   "\n")
cat("SE     =", round(se_s, 2),   "\n")
```
:::

::: {.column width="40%"}
- Every function takes `na.rm = TRUE`
- `cat()` prints clean labelled output
- This is **clear** but repetitive for many groups → that is why we use `group_by()` next
:::
:::::

# Stats the Tidy Way — `group_by()` + `summarize()`

::::: columns
::: {.column width="60%"}
```{r groupby-stats}
# Calculate all stats for both sides at once ---------
stats_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    n       = sum(!is.na(weight_g)),
    mean_wt = round(mean(weight_g,   na.rm = TRUE), 2),
    med_wt  = round(median(weight_g, na.rm = TRUE), 2),
    sd_wt   = round(sd(weight_g,     na.rm = TRUE), 2),
    se_wt   = round(sd_wt / sqrt(n), 2)
  )

stats_df
```
:::

::: {.column width="40%"}
- `group_by(side)` — splits the data by the `side` column
- `summarize()` — collapses each group to one row of stats
- You get **both groups in one clean table**
- You can reference a column you just created (`sd_wt` → `se_wt`)

📖 [R4DS §3.5 — `summarize()`](https://r4ds.hadley.nz/data-transform#sec-summarize)
:::
:::::

# Multiple Variables at Once

::::: columns
::: {.column width="60%"}
```{r groupby-multi}
# group_by() works on any numeric column you choose --
size_stats_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    n          = sum(!is.na(weight_g)),
    mean_wt    = round(mean(weight_g,   na.rm = TRUE), 2),
    mean_ht    = round(mean(height_cm,  na.rm = TRUE), 2),
    mean_wd    = round(mean(width_cm,   na.rm = TRUE), 2)
  )

size_stats_df
```
:::

::: {.column width="40%"}
**With one `group_by()` pipe** you can summarize **all three measurements** across both sides.

**Do the numbers support our hypothesis?**

- Shady heavier? → check `mean_wt`
- Shady taller? → check `mean_ht`
- Shady wider? → check `mean_wd`

📖 [R4DS §3.3 — `mutate()`](https://r4ds.hadley.nz/data-transform#sec-mutate)
:::
:::::

# Fast Summaries with `skimr`

```{r skimr-skim}
#| echo: true
# skim() grouped by side — full column summaries ------
tree_df %>%
  group_by(side) %>%
  skim()
```

:::::: columns
::: {.column width="60%"}
Per column and per group: `n_missing`, `mean`, `sd`, percentiles (`p0` to `p100`), and a mini histogram.
:::

:::: {.column width="40%"}
::: callout-tip
`skim()` is your fastest **first look** at any dataset. Pair it with `group_by()` to compare groups instantly.
:::
::::
::::::

# The `[[ ]]` Notation

::::: columns
::: {.column width="60%"}
```{r bracket-demo}
# Three ways to extract the same column as a vector --

# 1. Dollar sign notation
tree_df$weight_g

# 2. Double bracket with name in quotes
tree_df[["weight_g"]]

# 3. Double bracket with a variable holding the name
col_name <- "weight_g"
tree_df[[col_name]]      # works!
# tree_df$col_name        # does NOT work
```
:::

::: {.column width="40%"}
- All three return an identical **vector** of all 20 weights
- `$` is great for interactive typing
- `[[ ]]` is more **flexible** — accepts names stored in variables
- Essential for writing functions and loops later

📖 [R4DS §27 — A Field Guide to Base R](https://r4ds.hadley.nz/base-r)
:::
:::::

# Using `[[ ]]` with `filter()`

::::: columns
::: {.column width="60%"}
```{r bracket-stats-sunny}
# Combine filter() and [[ ]] to get one group --------
sunny_df <- tree_df %>% filter(side == "sunny")
sun_wt   <- sunny_df[["weight_g"]]

n_sun  <- sum(!is.na(sun_wt))
m_sun  <- mean(sun_wt, na.rm = TRUE)
se_sun <- sd(sun_wt,   na.rm = TRUE) / sqrt(n_sun)

cat("Sunny — n:", n_sun,
    "| mean:", round(m_sun, 2),
    "| SE:",   round(se_sun, 2), "g\n")
```

```{r bracket-stats-shady}
shady_df <- tree_df %>% filter(side == "shady")
sha_wt   <- shady_df[["weight_g"]]

n_sha  <- sum(!is.na(sha_wt))
m_sha  <- mean(sha_wt, na.rm = TRUE)
se_sha <- sd(sha_wt,   na.rm = TRUE) / sqrt(n_sha)

cat("Shady — n:", n_sha,
    "| mean:", round(m_sha, 2),
    "| SE:",   round(se_sha, 2), "g\n")
```
:::

::: {.column width="40%"}
**Pattern:**

1.  `filter()` → subset to one group
2.  `[[ ]]` → extract the column as a vector
3.  Apply `mean()`, `sd()`, `sum(!is.na())`

**Do the means match your prediction?**

- Sunny: \~`r round(m_sun, 1)` g
- Shady: \~`r round(m_sha, 1)` g

Shady leaves appear heavier — but is the difference *statistically significant?*
:::
:::::

# Extension: A Simple Stats Function

::::: columns
::: {.column width="60%"}
```{r describe-function}
# A reusable function using filter() + [[ ]] --------
# We will build functions properly in a later lecture!

describe_side <- function(data, grp, col_name) {
  x  <- data %>% filter(side == grp) %>% pull(col_name)
  n  <- sum(!is.na(x))
  m  <- mean(x, na.rm = TRUE)
  s  <- sd(x,   na.rm = TRUE)
  cat(grp, "|", col_name,
      "→ n:", n,
      "| mean:", round(m, 2),
      "| SE:",  round(s/sqrt(n), 2), "\n")
}

describe_side(tree_df, "sunny", "weight_g")
describe_side(tree_df, "shady", "weight_g")
describe_side(tree_df, "sunny", "height_cm")
describe_side(tree_df, "shady", "height_cm")
```
:::

::: {.column width="40%"}
**What this function does:**

1.  Accepts data frame, group name, and column name
2.  Filters to that group with `filter()`
3.  Extracts the column with `pull()`
4.  Computes n, mean, SE
5.  Prints a compact labelled summary

**For grins 🎉** — we will build these properly in Lecture 08!

📖 [R4DS §25 — Functions](https://r4ds.hadley.nz/functions)
:::
:::::

# Boxplot — Basic Structure

::::: columns
::: {.column width="60%"}
```{r boxplot-basic}
#| fig-format: png
# Basic boxplot — data is ALREADY long format --------
# No reshaping needed!
tree_box_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(
    title = "Leaf Weight by Tree Side",
    x     = "Side of Tree",
    y     = "Leaf Weight (g)",
    fill  = "Side"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

tree_box_plot
```
:::

::: {.column width="40%"}
**Boxplot anatomy:**

- Middle line = **median**
- Box edges = 25th and 75th percentile (**IQR**)
- Whiskers = 1.5 × IQR
- Dots beyond whiskers = **outliers**

Because our data is already in **long format**, we skip the `pivot_longer()` step entirely — `ggplot2` reads `side` straight off the x-axis.

📖 [R4DS §9 — Layers](https://r4ds.hadley.nz/layers)
:::
:::::

# Boxplot — Overlay Individual Points

:::::: columns
::: {.column width="60%"}
```{r boxplot-jitter}
#| fig-format: png
# Add raw data points over the boxplot ---------------
tree_jitter_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha    = 0.6,
    size     = 2.5
  ) +
  labs(
    title = "Leaf Weight by Tree Side",
    x     = "Side of Tree",
    y     = "Leaf Weight (g)",
    fill  = "Side"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

tree_jitter_plot
```
:::

:::: {.column width="40%"}
**Key arguments:**

- `position_jitter(width = 0.15)` — spreads points a little left and right
- `seed = 42` — same jitter layout every render
- `alpha = 0.6` — semi-transparent to show overlap
- `size = 2.5` — point size

::: callout-tip
Always show raw points with a boxplot — with only 10 leaves per group, every point matters!
:::
::::
::::::

# Boxplot — Using `position_dodge()`

::::: columns
::: {.column width="60%"}
```{r boxplot-dodge}
#| fig-format: png
# position_dodge() spreads groups apart horizontally -
tree_dodge_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g,
             color = side, fill = side)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +
  geom_point(
    position = position_dodge(width = 0.3),
    alpha    = 0.7,
    size     = 2.5
  ) +
  labs(
    title = "Leaf Weight by Tree Side",
    x     = "Side of Tree",
    y     = "Leaf Weight (g)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

tree_dodge_plot
```
:::

::: {.column width="40%"}
**`position_dodge(width = 0.3)`**

- Spreads overlapping geoms **horizontally**
- Most useful when you have **two grouping variables** (e.g., `x = species`, `fill = treatment`)
- `width` controls how far apart groups are spread

We will use `position_dodge()` extensively in future lectures when we have **factorial designs**.
:::
:::::

# Mean ± SE Plot — `stat_summary()`

::::: columns
::: {.column width="60%"}
```{r mean-se-plot}
#| fig-format: png
# stat_summary() computes and plots mean and SE ------
tree_mean_se_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, color = side)) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha    = 0.35, size = 2
  ) +
  stat_summary(fun      = mean,    geom = "point",
               size = 4) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.15, linewidth = 0.9) +
  labs(
    title = "Mean ± SE Leaf Weight by Tree Side",
    x     = "Side of Tree",
    y     = "Leaf Weight (g)",
    color = "Side"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

tree_mean_se_plot
```
:::

::: {.column width="40%"}
**Two `stat_summary()` layers:**

| call                 | what it draws           |
|----------------------|-------------------------|
| `fun = mean`         | large point at the mean |
| `fun.data = mean_se` | error bars for ± 1 SE   |

- Individual points in the **background** (`alpha = 0.35`)
- Mean shown as a **larger foreground point**
- Error bars = **± 1 SE**

This is the standard comparison figure in biology.
:::
:::::

# Comparing the Two Plot Types

::::: columns
::: {.column width="60%"}
```{r side-by-side-plots}
#| fig-format: png
#| fig-width: 8
#| fig-height: 3.5
# Arrange both plots side by side --------------------
library(patchwork)

tree_jitter_plot + tree_mean_se_plot
```
:::

::: {.column width="40%"}
**When to use each:**

| Plot type        | Use when                      |
|------------------|-------------------------------|
| Boxplot + points | showing full **distribution** |
| Mean ± SE        | making **group comparisons**  |

With only n = 10 per group, showing every point is essential — the reader needs to see the raw data.

Boxplots show *where* data fall. Mean ± SE shows *how confident* you are in the mean.
:::
:::::

# What We Learned Today

::::: columns
::: {.column width="60%"}
**Statistics concepts:**

- **Mean** — balance point; sensitive to outliers
- **Median** — middle value; robust
- **Variance / SD** — spread of individual data
- **SE** — precision of the sample mean
- ***p*****-value** — how unlikely are the data if H₀ is true?

**R skills:**

- `read_excel()` — load real Excel data
- `filter() %>% pull()` — extract a group's values as a vector
- `group_by() %>% summarize()` — all stats for all groups at once
- `sum(!is.na())` — the right way to count *n*
- `skim()` — instant full dataset overview
- `[[ ]]` — flexible column extraction
- `geom_boxplot()` + `geom_point()` — no pivot needed with long data!
- `stat_summary()` — mean ± SE
:::

::: {.column width="40%"}
**References:**

- 📖 [R4DS §3](https://r4ds.hadley.nz/data-transform) — Data Transformation
- 📖 [R4DS §13](https://r4ds.hadley.nz/numbers) — Numbers
- 📖 [R4DS §18](https://r4ds.hadley.nz/missing-values) — Missing Values
- 📖 [R4DS §27](https://r4ds.hadley.nz/base-r) — Base R
- 🌐 [Data Carpentry](https://datacarpentry.github.io/R-ecology-lesson/)

**Up next — Lecture 04:**

- Two-sample t-test
- Checking assumptions (normality, equal variance)
- Interpreting and reporting results
:::
:::::