---
title: "Lecture 05 — Wrangling Data"
subtitle: "filter, select, mutate, arrange — and chaining them into a pipeline"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# These are the ONLY things you need to edit to move this lecture around.
# The folder name and filename never have to change.
order: 5
week: 3
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Filter, select, mutate, and arrange with the tidyverse pipe — building
  one clean wrangling pipeline on the leaf data.
categories: [tidyverse]

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---



```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 04 — GGPlot II)

- **Setup** — R and Positron installed; project folders created (`data/`, `figures/`, `scripts/`)
- **Loading data** — `read_excel()` and `read_csv()`; a first look with `glimpse()`
- **Pipe** — `%>%` reads as "then"; chains steps together
- **GGPlot I & II** — geoms, facets, `stat_summary()`, `scale_color_manual()`, coordinates, and themes

::: callout-note
**✅ Key idea from Lecture 04**

You can already load data and plot it beautifully. Today we learn to *reshape and clean* the data itself, before it ever reaches `ggplot()`.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
**The core tidyverse verbs:**

- `filter()` — pick rows by a condition
- `select()` — pick columns by name
- `mutate()` — create new columns
- `arrange()` — sort rows
- Chain them all into a **pipeline**

::: callout-tip
**🖐 Try it yourself**

By the end you will wrangle our leaf data into exactly the columns and rows you need, in one connected pipeline.
:::
::::

::: {.column width="40%"}
**Our tools today:**

- `readxl`
- `tidyverse`

**References:**

- 📖 [R4DS Ch 3](https://r4ds.hadley.nz/data-transform) — Data Transformation
- 📖 [R4DS Ch 5](https://r4ds.hadley.nz/data-tidy) — Data Tidying
- 🌐 [Data Carpentry](https://datacarpentry.github.io/R-ecology-lesson/)

**Naming conventions:**

- data frames → `_df`
- plots → `_plot`
- models → `_model`
:::
::::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This is a **shorter lecture than usual** — one focused idea, wrangling, done properly. You will switch to the activity once, at the end, and type the code yourself into your R script there.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

-   `filter()`, `select()`, `mutate()`, and `arrange()` look interchangeable on a slide — guessing the output first is what actually forces you to tell them apart.
-   You will type `%>%` dozens of times today; typing it now is what makes it automatic later, when you're chaining five verbs under a deadline instead of reading one line calmly.
:::
::::
::::::

# Load Libraries and Data

```{r load-libs-05}
# Load all packages at the top of every script ----------
library(readxl) # reading Excel files
library(tidyverse) # data wrangling + ggplot2
library(janitor) # clean_names()
```

```{r load-data-05}
# Read the leaf data from the data folder ---------------
leaf_df <- read_excel("data/2026_09_03_data_sci_leaf_area.xlsx") %>%
  clean_names()
```

# Take a First Look

```{r inspect-glimpse-05}
# glimpse(): one row per column — name, type, first values
glimpse(leaf_df)
```

- `<chr>` = character (`shade`, `teams`)
- `<dbl>` = numeric (`mass_g`, `petiole_mm`, `thickness_mm`, `paper_mass_g`)

# The Core Tidyverse Verbs

:::::: columns
::: {.column width="60%"}
Five functions do most of the work in data wrangling:

| Function      | What it does                       |
|---------------|------------------------------------|
| `filter()`    | keep **rows** matching a condition |
| `select()`    | keep **columns** by name           |
| `mutate()`    | **add or change** a column         |
| `arrange()`   | **sort** rows                      |
| `summarize()` | **collapse** rows to a summary     |

All take a data frame as first input (via `%>%`) and return a data frame. We'll wrangle with the first four today — `summarize()` is next lecture's topic.

📖 [R4DS §3 — Data Transformation](https://r4ds.hadley.nz/data-transform)
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why these five?**

Almost every wrangling step you'll write this semester decomposes into some combination of picking rows, picking columns, adding a column, sorting, or collapsing to a summary — that's `filter`, `select`, `mutate`, `arrange`, and `summarize`.
:::

Think of them as building blocks:

```
filter()   → rows you want
select()   → columns you want
mutate()   → new columns you need
arrange()  → order the rows
summarize()→ collapse to summaries (next lecture)
```
::::
::::::

# `filter()` — Picking Rows

::: callout-note
**🔮 Predict first:** We have 53 leaves. Before you run it, guess how many rows `filter(shade == "sunny")` returns. Write your number down, then run it and check.
:::

:::::: columns
::: {.column width="60%"}
```{r filter-demo}
# filter() keeps rows where the condition is TRUE -------

# Keep only sunny leaves
leaf_df %>% filter(shade == "sunny") %>% head()

# Keep only leaves heavier than 0.5 g
leaf_df %>% filter(mass_g > 0.5) %>% head()

# Combine two conditions with , (AND)
leaf_df %>% filter(shade == "shady", mass_g > 0.5) %>% head()
```
:::

:::: {.column width="40%"}
**Common comparison operators:**

| Symbol    | Meaning                             |
|-----------|-------------------------------------|
| `==`      | exactly equal                       |
| `!=`      | not equal                           |
| `>` `<`   | greater / less than                 |
| `>=` `<=` | greater / less than or equal        |
| is.na     | return all NA rows                  |
| !is.na    | return all rows that are **NOT NA** |

::: callout-warning
**⚠️ Watch out!**

- `=` assigns a value.
- `==` tests equality.
- `filter(shade = "sunny")` → error.
- `filter(shade == "sunny")` → correct.
:::
::::
::::::

# Live Demo — Watch It Break (on purpose)

:::::: columns
::: {.column width="60%"}
I will type this **wrong** on purpose:

``` r
# One equals sign — a very common mistake
leaf_df %>% filter(shade = "sunny")
```

R stops and tells us:

```
Error in `filter()`:
! We detected a named input.
i This usually means that you've used `=` instead of `==`.
```

Now the fix — two equals signs:

``` r
leaf_df %>% filter(shade == "sunny")
```
:::

:::: {.column width="40%"}
::: callout-tip
**✅ Why show a broken `filter()`?**

`filter(shade = "sunny")` is the single most common typo in this whole unit, because `=` and `==` look almost identical and only one of them tests equality. dplyr's error message actually names the mistake — "you've used `=` instead of `==`" — so once you've seen it fixed here, the next time R prints it at you it's a fix, not a mystery.
:::
::::
::::::

# `select()` — Picking Columns

::::: columns
::: {.column width="60%"}
```{r select-demo}
# select() keeps the columns you name ------------------

# Keep only shade and mass
leaf_df %>% select(shade, mass_g) %>% head()

# Drop the paper_mass_g column (use minus sign)
leaf_df %>% select(-paper_mass_g) %>% head()

# Keep columns that start with a word
leaf_df %>% select(starts_with("mass")) %>% head()
```
:::

::: {.column width="40%"}
- **Why `select()` matters:**
  - Real datasets often have 50–100+ columns
  - You rarely need all of them
  - `select()` keeps your workspace clean
- **Useful helpers:**
  - `starts_with("x")` — columns starting with "x"
  - `ends_with("_mm")` — columns ending with "_mm"
  - `contains("mass")` — columns containing "mass"

📖 [R4DS §3.3](https://r4ds.hadley.nz/data-transform#select)
:::
:::::

# `mutate()` — Creating New Columns

::: callout-note
**🔮 Predict first:** `mass_g * 1000` converts grams to milligrams. Before running: will `mutate()` **replace** `mass_g` or **add** a new column? How many columns will the result have?
:::

:::::: columns
::: {.column width="60%"}
```{r mutate-demo}
# mutate() adds a new column to the data frame ---------

# Convert grams to milligrams
leaf_df %>%
  mutate(mass_mg = mass_g * 1000) %>%
  head()

# Add a size category based on mass
leaf_df %>%
  mutate(size_class = if_else(mass_g > 0.5, "large", "small")) %>%
  head()

# mutate() can create several new columns at once
leaf_df %>%
  mutate(
    mass_mg    = mass_g * 1000,
    petiole_cm = petiole_mm / 10 # petiole length in cm
  ) %>%
  head()
```
:::

:::: {.column width="40%"}
- `mutate()` keeps **all existing columns** and adds new ones
- **Use `if_else(condition, value_if_true, value_if_false)` to create categories**

::: callout-note
**✅ Watch the column count**

`mutate()` never removes columns — it adds to the right of your data frame, so `leaf_df` gains one column per new variable you create.

To overwrite a column instead of adding one, reuse the same name on the left: `mutate(mass_g = round(mass_g, 2))`
:::

📖 [R4DS §3.3](https://r4ds.hadley.nz/data-transform#sec-mutate)
::::
::::::

# `arrange()` — Sorting Rows

:::::: columns
::: {.column width="60%"}
```{r arrange-demo}
# arrange() sorts rows by one or more columns ----------

# Lightest leaves first
leaf_df %>% arrange(mass_g)

# Heaviest leaves first (use desc() to reverse)
leaf_df %>% arrange(desc(mass_g))

# Sort by shade, then by mass within each shade
leaf_df %>% arrange(shade, desc(mass_g))
```
:::

:::: {.column width="40%"}
- Default: **ascending** order (smallest first)
- `desc(column)` → **descending** order (largest first)
- Very useful before printing a table or checking outliers

::: callout-tip
`arrange()` is rarely used in the middle of an analysis pipeline — it is most useful at the **end** when you want to inspect your results in a specific order.
:::
::::
::::::

# The Full Pipeline

:::::: columns
::: {.column width="60%"}
```{r full-pipeline}
# Chain all the verbs together -------------------------
# Read it top to bottom like a recipe

wrangled_df <- leaf_df %>%
  filter(mass_g > 0) %>% # remove any zeros
  select(shade, mass_g, thickness_mm) %>% # keep only needed cols
  mutate(
    mass_mg = mass_g * 1000,
    size_class = if_else(mass_g > 0.5, "large", "small")
  ) %>%
  arrange(shade, desc(mass_g)) # sort by shade, then mass

head(wrangled_df)
```
:::

:::: {.column width="40%"}
**Read the pipeline out loud:**

> - Take `leaf_df`,
>   - **then** filter to non-zero masses,
>     - **then** select three columns,
>       - **then** add two new columns,
>         - **then** sort by shade and mass.

::: callout-note
**✅ Compare it to nested calls**

Write the same pipeline as nested functions — `arrange(mutate(select(filter(leaf_df, mass_g > 0), shade, mass_g, thickness_mm), mass_mg = mass_g * 1000, size_class = if_else(mass_g > 0.5, "large", "small")), shade, desc(mass_g))` — and the *first* thing that runs is buried in the middle. The pipe lets you read top to bottom in the order R actually executes.
:::
::::
::::::

# → ACTIVITY 5 starts now

::: {.activitybanner}
**🛑 Go to Activity 5 — Wrangling**

Close the slides. Open the Activity 5 and work through `filter`, `select`, `mutate`, `arrange`, and the full pipeline. Type every line yourself into your script; predict before you run.
:::

# What We Learned Today

::::: columns
::: {.column width="60%"}
- `filter()` — keep rows by condition
- `select()` — keep columns by name
- `mutate()` — add new calculated columns
- `arrange()` — sort rows
- Full pipeline: chain all verbs with `%>%`
:::

::: {.column width="40%"}
**References:**

- 📖 [R4DS §3](https://r4ds.hadley.nz/data-transform) — Data Transformation
- 📖 [R4DS §5](https://r4ds.hadley.nz/data-tidy) — Data Tidying
- 🌐 [Data Carpentry](https://datacarpentry.github.io/R-ecology-lesson/)

**Up next — Lecture 06, Summary Statistics:**

- Mean, median, variance, SD, SE
- Counting correctly with `sum(!is.na())`
- `group_by()` + `summarize()`, and `skimr`
:::
:::::
