---
title: "Lecture 06 — Real Climate Data in R"
subtitle: "Downloading, summarizing, and modeling Duluth weather station data"
author: "Bill Perry"
date: today
format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 05)

- **Linear regression** — fit a calibration curve (area ~ mass) with `lm()`
- **Reading `summary()`** — slope *b*, intercept *a*, R², p-value for the slope
- **Least squares** — the line that minimizes the sum of squared residuals
- **Assumption checks** — residual vs. fitted plot, QQ plot, Shapiro-Wilk on **residuals**
- **`predict()`** — confidence intervals (mean Y) vs. prediction intervals (one new Y)
- **Applied it** — converted paper tracing mass into predicted leaf area (cm²)

::: callout-note
**✅ Transition**

So far every regression we have run used a dataset you were handed. Today we **download real data from inside R** for the first time — long-term weather records for Duluth, MN — and use that same regression toolkit to ask a much bigger question: *is it getting warmer?*
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- Download real weather station data **from inside R** (no Excel step!)
- Explore raw daily data — and see why raw data alone is hard to read
- **Summarize** the same data by day, week, month, and year — and watch the story get clearer
- Reuse `lm()` to estimate the **rate of change** (slope) in temperature over time
- Use `case_when()` to build a new categorical variable — **summer vs. winter**
- Compare how fast each season is warming

::: callout-tip
**🖐 Try it yourself**

By the end you'll have downloaded a real NOAA climate record and estimated how many degrees per decade Duluth has warmed.
:::
::::

::: {.column width="40%"}
**Tools today:**

- `GSODR`, `tidyverse`, `lubridate`

**Textbook:**

- 📖 R for Data Science 2e — Hadley Wickham
  - [Ch. 3 Data transformation](https://r4ds.hadley.nz/data-transform)
  - [Ch. 17 Dates and times](https://r4ds.hadley.nz/datetimes)
- 📖 Data Carpentry — [R for Ecologists](https://datacarpentry.github.io/R-ecology-lesson/)

**Naming conventions:**

- data frames → `_df`
- plots → `_plot`
- models → `_model`
:::
::::::

# Where Does This Data Come From?

::::: columns
::: {.column width="60%"}
**GSOD** = Global Surface Summary of the Day, maintained by NOAA. It contains daily weather observations from thousands of stations worldwide, some going back to the **1940s**.

The `GSODR` package lets us pull this data **directly into R** — no manual download, no Excel, no copy-paste.

**The station we'll use:**

- Duluth International Airport
- Station ID `727450-14913`
- Daily records back to **1948** — 78 years of data
:::

::: {.column width="40%"}
**Why this matters:**

- This is the *same* kind of messy, real-world data you'll meet in your own research
- Today's workflow — download → explore → summarize → model — is the same workflow you'll use for your final project

📖 R4DS Ch. 3 — every analysis starts with getting data into R
:::
:::::

# Load Libraries and Download the Data

::::: columns
::: {.column width="60%"}
```{r load-libs-06}
# Load packages at the top — always -------------------
library(tidyverse)   # data manipulation + ggplot2
library(GSODR)        # download NOAA weather station data
```

```{r get-duluth-06}
# Download daily Duluth weather data, 1948-2025 --------
# this is DAILY data - rows = days, so expect ~28,000 rows
duluth_df <- get_GSOD(years = 1948:2025,
                       station = "727450-14913")
```

```{r glimpse-duluth-06}
# Look at what we got -----------------------------------
glimpse(duluth_df)
```
:::

::: {.column width="40%"}
**What just happened:**

- `get_GSOD()` reached out to NOAA's servers and pulled every daily record for this station
- `years = 1948:2025` is a vector — shorthand for "every year in this range"
- This download can take a minute — daily data for 78 years is a lot of rows!

::: callout-important
Don't panic at the row count. **Daily** data for 78 years ≈ 28,000 rows. We will fix that with summarizing.
:::
:::
:::::

# Trim to What We Need

::::: columns
::: {.column width="60%"}
```{r select-cols-06}
# Keep only the columns we actually need -----------------
dlh_temp_df <- duluth_df %>%
  select(
    STNID, NAME, YEARMODA,
    YEAR, MONTH, DAY,
    TEMP, MXSPD, PRCP, I_SNOW_ICE
  )

head(dlh_temp_df)
```
:::

::: {.column width="40%"}
**Why `select()` here:**

- `get_GSOD()` returns 40+ columns — station metadata, multiple temperature scales, attribute flags
- We only need a handful for this question
- This is the same `select()` skill from Lecture 02 — applied to a much wider real-world file

📖 R4DS Ch. 3.2 — `select()` for picking columns
:::
:::::

# First Look — Plot the Raw Daily Data

::::: columns
::: {.column width="60%"}
```{r raw-plot-06}
#| fig-format: png
# Plot every single daily temperature ---------------------
raw_temp_plot <- dlh_temp_df %>%
  ggplot(aes(x = YEARMODA, y = TEMP)) +
  geom_point(alpha = 0.2, size = 0.5) +
  geom_line(alpha = 0.3) +
  labs(
    title = "Duluth Daily Mean Temperature, 1948–2025",
    x     = "Date",
    y     = "Temperature (°C)"
  ) +
  theme_minimal()

raw_temp_plot
```
:::

::: {.column width="40%"}
**What do you see?**

- A dense, fuzzy band — the seasonal cycle (winter ↔ summer) dominates the plot
- ~28,000 points overlapping each other
- Any long-term *trend* is completely hidden inside the seasonal noise

**The problem:** raw daily data shows you *everything* — which means it can hide the *one thing* you're looking for.
:::
:::::

# The Fix — Summarize the Story Out of the Noise

::::: columns
::: {.column width="60%"}
**The idea:** group the data into bigger time chunks, then take the **mean** within each chunk.

| Summary level | `group_by()` | Rows after summary |
|----------------|--------------|---------------------|
| Daily (raw)    | none         | ~28,000             |
| Monthly        | `YEAR, MONTH`| ~936                |
| Yearly         | `YEAR`       | 78                  |

Each step **trades detail for clarity**. We lose day-to-day weather noise, but the climate signal becomes visible.
:::

::: {.column width="40%"}
**This is the same logic as Lecture 03's `group_by()` + `summarize()`** — just applied to time instead of to `side` (sunny/shady).

::: callout-tip
**🖐 Try it yourself**

Watch the *same* underlying data tell three different stories as we change the summary level.
:::
:::
:::::

# Summarize by Month

::::: columns
::: {.column width="60%"}
```{r monthly-summary-06}
# Mean temperature per year-month -----------------------
dlh_month_df <- dlh_temp_df %>%
  group_by(YEAR, MONTH) %>%
  summarize(
    YEARMODA = first(YEARMODA),
    TEMP     = mean(TEMP, na.rm = TRUE)
  )
```

```{r monthly-plot-06}
#| fig-format: png
# Plot the monthly means ----------------------------------
month_temp_plot <- dlh_month_df %>%
  ggplot(aes(x = YEARMODA, y = TEMP)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_line(alpha = 0.4) +
  geom_smooth(method = "lm", color = "steelblue") +
  labs(
    title = "Duluth Monthly Mean Temperature",
    x     = "Date", y = "Temperature (°C)"
  ) +
  theme_minimal()

month_temp_plot
```
:::

::: {.column width="40%"}
**Better — but still busy.**

- The seasonal cycle (summer peaks, winter dips) is now a clean, regular wave
- A faint regression line is visible, but the seasonal swing still dominates the y-axis range

`first(YEARMODA)` keeps one representative date per month so we can still plot on a time axis.
:::
:::::

# Summarize by Year — The Story Gets Clear

::::: columns
::: {.column width="60%"}
```{r yearly-summary-06}
# Mean temperature per year -------------------------------
dlh_year_df <- dlh_temp_df %>%
  group_by(YEAR) %>%
  summarize(
    YEARMODA = first(YEARMODA),
    TEMP     = mean(TEMP, na.rm = TRUE)
  )

head(dlh_year_df)
```

```{r yearly-plot-06}
#| fig-format: png
yearly_temp_plot <- dlh_year_df %>%
  ggplot(aes(x = YEAR, y = TEMP)) +
  geom_point(size = 2) +
  geom_line() +
  geom_smooth(method = "lm", color = "firebrick") +
  labs(
    title = "Duluth Annual Mean Temperature, 1948–2025",
    x     = "Year", y = "Mean Temp (°C)"
  ) +
  theme_minimal()

yearly_temp_plot
```
:::

::: {.column width="40%"}
**Now the trend is obvious.**

- The seasonal noise is completely averaged out
- 78 points instead of 28,000 — and the story is *clearer*, not weaker
- The blue/red trend line is the same kind of `geom_smooth(method = "lm")` line from Lecture 03

This is the power of `group_by() %>% summarize()`: the right level of summary reveals the signal.
:::
:::::

# Same Data, Three Levels — Side by Side

::::: columns
::: {.column width="60%"}
| Level | What it shows | What it hides |
|-------|---------------|----------------|
| **Daily** | every weather event | the long-term trend |
| **Monthly** | the seasonal cycle | day-to-day noise |
| **Yearly** | the climate trend | season, weather events |

**Key idea:** there is no single "correct" level — it depends on your *question*. Want to know about a heat wave? Use daily. Want to know about climate change? Use yearly.
:::

::: {.column width="40%"}
::: callout-important
**Summarizing is not "losing information" — it's choosing which question you're answering.**
:::

📖 R4DS Ch. 3.5 — `group_by()` + `summarize()` workflow
:::
:::::

# What Is the Rate of Change? — Reusing `lm()`

::::: columns
::: {.column width="60%"}
```{r yearly-lm-06}
# Fit a regression: TEMP predicted by YEAR ---------------
yearly_temp_model <- lm(TEMP ~ YEAR, data = dlh_year_df)

summary(yearly_temp_model)
```
:::

::: {.column width="40%"}
**This is exactly Lecture 05's `lm()` syntax** — only now:

- `X` = `YEAR` (instead of paper mass)
- `Y` = `TEMP` (instead of paper area)
- The **slope** is no longer "cm² per gram" — it's **°C per year**

**H₀:** β = 0 (no warming trend)

**Hₐ:** β ≠ 0 (temperature is changing over time)
:::
:::::

# Extracting and Interpreting the Slope

::::: columns
::: {.column width="60%"}
```{r yearly-slope-06}
# Pull the slope and convert to °C per decade ------------
yearly_intercept <- coef(yearly_temp_model)[1]
yearly_slope     <- coef(yearly_temp_model)[2]

cat("Slope (b)         =", round(yearly_slope, 4), "°C/year\n")
cat("Warming rate       =",
    round(yearly_slope * 10, 3), "°C per decade\n")
```
:::

::: {.column width="40%"}
**Why multiply by 10?**

- The raw slope is in °C **per year** — a small, hard-to-feel number
- Multiplying by 10 converts to °C **per decade** — easier to communicate
- This is the same `coef()` extraction skill from Lecture 05

> *b* = 0.02 °C/year → 0.2 °C/decade — small year to year, but adds up over 78 years
:::
:::::

# Does the Same Pattern Hold at the Monthly Level?

::::: columns
::: {.column width="60%"}
```{r monthly-lm-06}
# Fit the same model on the monthly summary ---------------
monthly_temp_model <- lm(TEMP ~ YEARMODA, data = dlh_month_df)

summary(monthly_temp_model)
```
:::

::: {.column width="40%"}
**Compare to the yearly model:**

- Same overall warming direction
- More residual scatter, because each point still carries some seasonal "memory"
- The yearly summary gives the cleanest, most interpretable slope for a *long-term* trend question

This is why we chose the **yearly** summary for our headline result.
:::
:::::

# A New Question — Is Every Season Warming the Same?

::::: columns
::: {.column width="60%"}
**So far we've asked:** is Duluth warming overall?

**New question:** is **winter** warming at the same rate as **summer**?

This matters biologically — earlier springs, shorter ice seasons, and changing growing seasons don't depend on the *annual average*. They depend on which **season** is changing fastest.

**Plan:**

1.  Label every day as summer or winter using `case_when()`
2.  Summarize mean temperature by year **within each season**
3.  Fit a *separate* `lm()` for each season
4.  Compare the two slopes
:::

::: {.column width="40%"}
::: callout-tip
**🖐 Try it yourself**

Make a prediction now: which season do you think is warming faster — winter or summer? We'll find out with real data.
:::
:::
:::::

# Building a Season Variable with `case_when()`

::::: columns
::: {.column width="60%"}
```{r season-case-when-06}
# Label each day as summer, winter, or shoulder season ----
dlh_season_df <- dlh_temp_df %>%
  mutate(
    season = case_when(
      MONTH %in% c(6, 7, 8)   ~ "summer",
      MONTH %in% c(12, 1, 2)  ~ "winter",
      TRUE                     ~ "shoulder"
    )
  )

# Quick check - did it work? -------------------------------
dlh_season_df %>%
  count(season)
```
:::

::: {.column width="40%"}
**Reading `case_when()`:**

- Each line is read top to bottom: *condition* `~` *result*
- `MONTH %in% c(6, 7, 8)` — "is MONTH one of 6, 7, or 8?"
- `TRUE ~ "shoulder"` — the catch-all for everything not matched above (spring/fall)

This is a new, more powerful cousin of the `ifelse()` you've used before — `case_when()` handles **many** conditions cleanly.

📖 R4DS Ch. 3.4 — `case_when()` for multi-way `mutate()`
:::
:::::

# Summarize Each Season by Year

::::: columns
::: {.column width="60%"}
```{r season-summary-06}
# Keep only summer and winter, then summarize by year -----
season_year_df <- dlh_season_df %>%
  filter(season %in% c("summer", "winter")) %>%
  group_by(YEAR, season) %>%
  summarize(
    TEMP = mean(TEMP, na.rm = TRUE)
  )

head(season_year_df)
```
:::

::: {.column width="40%"}
**What changed:**

- `filter()` drops shoulder-season days — we only want the two extremes
- `group_by(YEAR, season)` — grouping by **two** variables gives us one mean temperature per year **per season**

This produces a tidy long-format data frame — perfect for both plotting and modeling.
:::
:::::

# Plot Both Seasons Together

::::: columns
::: {.column width="60%"}
```{r season-plot-06}
#| fig-format: png
# Plot both seasons with separate trend lines --------------
season_temp_plot <- season_year_df %>%
  ggplot(aes(x = YEAR, y = TEMP, color = season)) +
  geom_point(size = 1.8, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c(
    "summer" = "firebrick",
    "winter" = "steelblue"
  )) +
  labs(
    title = "Duluth Summer vs. Winter Temperature Trends",
    x     = "Year", y = "Mean Temp (°C)",
    color = "Season"
  ) +
  theme_minimal()

season_temp_plot
```
:::

::: {.column width="40%"}
**What to look for:**

- Two clearly separated bands — winter is much colder than summer, as expected
- Compare the **steepness** of each trend line, not just the starting height
- A steeper slope means a faster rate of warming for that season
:::
:::::

# Fit a Separate Model for Each Season

::::: columns
::: {.column width="60%"}
```{r summer-model-06}
# Summer-only model -----------------------------------------
summer_df <- season_year_df %>% filter(season == "summer")
summer_model <- lm(TEMP ~ YEAR, data = summer_df)
summer_slope <- coef(summer_model)[2]
```

```{r winter-model-06}
# Winter-only model -----------------------------------------
winter_df <- season_year_df %>% filter(season == "winter")
winter_model <- lm(TEMP ~ YEAR, data = winter_df)
winter_slope <- coef(winter_model)[2]
```

```{r compare-slopes-06}
cat("Summer warming rate:",
    round(summer_slope * 10, 3), "°C/decade\n")
cat("Winter warming rate:",
    round(winter_slope * 10, 3), "°C/decade\n")
```
:::

::: {.column width="40%"}
**Why fit two separate models** instead of one model with `season` as a predictor?

- Two simple models are easier to interpret at this stage
- Each gives us a clean, separate slope (°C/decade) to directly compare

**Spoiler from the climate science literature:** in most northern locations, **winter warms faster than summer** — fewer extreme-cold days, shorter ice seasons. Does Duluth's data agree?
:::
:::::

# What We Learned Today

::::: columns
::: {.column width="60%"}
**Concepts:**

- Real climate data can be downloaded **directly into R** with `GSODR`
- Raw daily data hides long-term trends inside seasonal/day-to-day noise
- `group_by() %>% summarize()` at different time scales reveals different stories
- The same `lm()` workflow from Lecture 05 works on *any* X ~ Y relationship — including time
- `case_when()` builds a new categorical variable from multiple conditions
- Splitting data by group (season) before modeling lets us compare **rates of change**

**R skills:**

- `get_GSOD()` — download NOAA station data
- `group_by(YEAR, MONTH)` / `group_by(YEAR)` — multi-level time summaries
- `case_when()` — multi-condition `mutate()`
- `filter(x %in% c(...))` — keep only specific categories
- Two-group regression comparison with separate `lm()` calls
:::

::: {.column width="40%"}
**References:**

- 📖 R4DS Ch. 3 — Data transformation
- 📖 R4DS Ch. 17 — Dates and times
- 📖 Data Carpentry — R for Ecologists

**Up next — Homework:**

- Pick **your own city** (anywhere with a long GSOD record)
- Repeat this entire workflow: download → raw plot → summarize → model the trend
- Build your own summer vs. winter `case_when()` comparison
- Write a short results paragraph comparing your city's warming rate to Duluth's
:::
:::::
