---
order: 5
week: 3
# =====================================================================
#  t_test_report.qmd
#  A WORKED EXAMPLE for the Quarto intro (Lecture 4.5).
#  This is the analysis from Worksheet 04 (the Welch's t-test) written
#  up as a Quarto report. Open it after the Quarto lecture to SEE why we
#  bother: the same t-test you ran in a script becomes a finished Word
#  document, with the t-statistic, p-value, and means dropped straight
#  into the sentences — nothing copied by hand.
# =====================================================================
title: "Do Shady Leaves Weigh More? A Two-Sample t-Test"
subtitle: "A reproducible write-up of the Worksheet 04 analysis"
author: "Your Name Here"
date: today

format:
  docx:
    toc: true
    number-sections: true
    fig-width: 6
    fig-height: 4
    fig-format: png
  html: default

# Hide the code so the report reads like a report.
# Set echo: true if you want to show the code as a teaching handout.
execute:
  echo: false
  warning: false
  message: false
---

```{r libraries}
# Load all packages in one chunk at the top ----------------------------
library(readxl)
library(tidyverse)
```

```{r load-data}
# Read the leaf data ---------------------------------------------------
tree_df <- read_excel("data/2026_06_25_tree_experiment_raw_data.xlsx")
```

# Introduction

Leaves on the **shady** side of a tree may grow larger and heavier to capture
more of the limited light beneath the canopy. We measured the weight (g) of ten
leaves from each side of a tree and tested whether mean leaf weight differs
between sides.

- **Null hypothesis (H₀):** $\mu_{shady} = \mu_{sunny}$ — no difference in mean weight
- **Alternate hypothesis (Hₐ):** $\mu_{shady} \neq \mu_{sunny}$ — the means differ

We used a two-tailed test at $\alpha = 0.05$.

# Methods

We compared the two sides with a **Welch's two-sample t-test**
(`var.equal = FALSE`), which does not assume equal variances. Before testing we
checked normality within each group with a Shapiro–Wilk test. All analysis was
done in R with the tidyverse, and this report was produced in Quarto so that
every value below is computed directly from the raw data.

# Results

```{r summary-table}
# Descriptive statistics per side --------------------------------------
summary_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    n       = sum(!is.na(weight_g)),
    mean_wt = round(mean(weight_g, na.rm = TRUE), 2),
    sd_wt   = round(sd(weight_g,   na.rm = TRUE), 2),
    se_wt   = round(sd_wt / sqrt(n), 2)
  )
```

```{r print-table}
knitr::kable(summary_df,
             caption = "Table 1. Leaf weight (g) by side of tree (mean, SD, SE).")
```

```{r run-test}
# Welch's two-sample t-test --------------------------------------------
leaf_ttest_model <- t.test(
  weight_g ~ side,
  data        = tree_df,
  var.equal   = FALSE,
  alternative = "two.sided"
)

# Pull the pieces we want to write into the sentences below
t_val   <- round(leaf_ttest_model$statistic, 2)
df_val  <- round(leaf_ttest_model$parameter, 1)
p_val   <- signif(leaf_ttest_model$p.value, 2)
mean_sh <- round(leaf_ttest_model$estimate[1], 2)   # shady (alphabetical first)
mean_su <- round(leaf_ttest_model$estimate[2], 2)   # sunny
se_sh   <- summary_df$se_wt[summary_df$side == "shady"]
se_su   <- summary_df$se_wt[summary_df$side == "sunny"]
```

Shady-side leaves were significantly **heavier** than sunny-side leaves
(Welch's two-sample t-test: t(`r df_val`) = `r t_val`, *p* = `r p_val`).
Mean ± SE weight was `r mean_sh` ± `r se_sh` g for shady leaves and
`r mean_su` ± `r se_su` g for sunny leaves.

::: {.callout-note appearance="simple"}
Every statistic in that sentence — t, df, *p*, and both means ± SE — was inserted
with inline code such as `` `r t_val` ``. If a leaf measurement changed, you
would fix the data file, press **Render**, and the sentence would rewrite itself.
In a plain script you would retype all of these by hand.
:::

```{r boxplot}
#| fig-cap: "Figure 1. Leaf weight by side of tree. Points show individual leaves; the test result is printed in the subtitle."
leaf_box_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.15, seed = 42),
             alpha = 0.6, size = 2.5) +
  labs(
    subtitle = paste0("Welch's t-test: t = ", t_val, ", p = ", p_val),
    x = "Side of tree",
    y = "Leaf weight (g)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

leaf_box_plot
```

# Conclusion

At $\alpha = 0.05$ we **reject the null hypothesis**: mean leaf weight differs
between the sunny and shady sides of the tree, with shady leaves heavier. This
matches the biological prediction that shade leaves grow larger to capture more
of the limited light. *(Add a sentence of your own interpretation here.)*

# References

- Whitlock, M. & Schluter, D. *The Analysis of Biological Data* (2nd ed.).
- Wickham, H. *R for Data Science (2nd ed.)* — <https://r4ds.hadley.nz/>
