---
title: "Lecture 05 — Linear Regression"
subtitle: "Predicting leaf area from paper tracing mass (Whitlock & Schluter Ch. 17)"
author: "Bill Perry"
date: today
format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 04)

- **Welch's two-sample t-test** — compared mean leaf weight between sunny and shady sides (`var.equal = FALSE`, two-tailed)
- **Five-step framework** — hypotheses → normality → variance → run test → report
  - Normality: histogram, QQ plot, Shapiro-Wilk on each group
  - Variance: Levene's test; Welch's is safe regardless of the result
- **Reading `t.test()` output** — t-statistic, Welch-Satterthwaite df, *p*-value, 95% CI, group means
- **Result** — shady leaves were significantly *heavier* than sunny leaves (Welch's t-test, *p* \< 0.001)

::: callout-note
**✅ Transition**

We confirmed that shady and sunny leaves *differ in weight*. Today we go further — we build a **calibration curve** to convert that weight into a more biologically meaningful measurement: **leaf surface area** (cm²).
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- Understand what linear regression does and when to use it
- Learn the equation: $\hat{Y} = a + bX$
  - slope *b*, intercept *a*, residuals, R²
- Understand **least squares** — what the line minimizes
- Fit a model in R with `lm()` and read every line of `summary()`
- **Apply it:** build a calibration curve to predict leaf area from paper tracing mass
- Check all four assumptions of regression
- Use `predict()` to make predictions with intervals
- Use `mutate()` to add model results to a data frame

::: callout-tip
**🖐 Try it yourself**

By the end you will predict leaf area from a tracing weight.
:::
::::

::: {.column width="40%"}
**Tools today:**

- `readxl`, `tidyverse`

**Textbook:**

- 📖 Whitlock & Schluter *Analysis of Biological Data* 2nd ed.
- **Chapter 17 — Regression** (pp. 539–575)
  - §17.1 Linear regression
  - §17.3 Testing slope hypotheses
  - §17.4 R²
  - §17.5 Assumptions

**Naming conventions:**

- models → `_model`
- plots → `_plot`
:::
::::::

# Our Calibration Idea

::::: columns
::: {.column width="60%"}
**The problem:** measuring leaf surface area without a leaf area meter.

**The solution — a calibration curve:**

1.  Cut paper squares of **known area** (1, 4, 16 … 567 cm²)
2.  **Weigh** each square on a precision balance
3.  Fit a **regression line**: area \~ mass
4.  Trace a leaf on the *same paper*, weigh the tracing
5.  **Plug the tracing mass into the equation** → predicted leaf area

**Why does this work?**

Paper of uniform stock has constant *area density* (cm² per gram). Mass and area are therefore **perfectly linearly related** within a sheet.
:::

::: {.column width="40%"}
| Known area (cm²) | n replicates |
|------------------|--------------|
| 1                | 25           |
| 4                | 32           |
| 16               | 24           |
| 36               | 3            |
| 64               | 7            |
| 100              | 14           |
| 225–567          | 12           |

**Total:** 118 paper squares weighed

📖 W&S §17.1, Example 17.1 (the same logic: measure X to predict Y)
:::
:::::

# What Is Linear Regression?

::::: columns
::: {.column width="60%"}
**Linear regression** draws the *best straight line* through a scatter of points to **predict** a response variable (Y) from an explanatory variable (X).

**Two key roles:**

| Role            | Variable                                | Our example |
|-----------------|-----------------------------------------|-------------|
| Explanatory (X) | on the x-axis; what you measure         | `mass_g`    |
| Response (Y)    | on the y-axis; what you want to predict | `area_cm2`  |

X is often something easy to measure; Y is what you actually want to know. Regression gives you the equation to convert X → Y.

📖 W&S §17.1 p. 540–541
:::

::: {.column width="40%"}
**When to use regression:**

- You have **two numeric variables**
- One variable can reasonably **predict or cause** the other
- You want an **equation** to make predictions for new observations

**Regression vs. our t-test (Lecture 04):**

| Test       | Question                        |
|------------|---------------------------------|
| t-test     | Do two group *means* differ?    |
| Regression | Does X *predict* Y numerically? |

Regression produces a **predictive equation**. The t-test produces a decision about group means.
:::
:::::

# The Regression Equation

::::: columns
::: {.column width="60%"}
**Population model** (W&S §17.1):

$$Y = \alpha + \beta X + \varepsilon$$

**Sample estimate** (what R calculates):

$$\hat{Y} = a + bX$$

| Symbol        | Name        | Meaning                              |
|---------------|-------------|--------------------------------------|
| $\hat{Y}$     | predicted Y | estimated mean area at a given mass  |
| *a*           | intercept   | predicted Y when X = 0               |
| *b*           | slope       | change in Y per 1-unit increase in X |
| $\varepsilon$ | residual    | observed − predicted                 |

**Our equation will be:** $$\widehat{\text{area}} = a + b \times \text{mass\_g}$$
:::

::: {.column width="40%"}
**Interpreting the slope:**

> *b* = the change in area (cm²) for every additional 1 gram of paper

Since 1 g ≈ 130 cm² of paper, the slope is the **area density** of the paper stock.

**Interpreting the intercept:**

> *a* = predicted area when mass = 0

Physically this should be 0 (no paper, no area). Small non-zero values reflect measurement error.

📖 W&S §17.1 pp. 543–545
:::
:::::

# Residuals and Least Squares

::::: columns
::: {.column width="60%"}
**Residual** for observation *i*:

$$e_i = Y_i - \hat{Y}_i = \text{observed} - \text{predicted}$$

The **least-squares line** minimizes the sum of *squared* residuals:

$$\text{minimize} \sum_{i=1}^{n}(Y_i - \hat{Y}_i)^2$$

- Squaring serves two purposes:
  - \- Makes all deviations **positive** (no cancellation)

  - \- Penalizes **large** deviations more than small ones
- There is only one line that achieves this minimum — it is unique.
- 📖 W&S §17.1 p. 542 (Figure 17.1-2)
:::

::: {.column width="40%"}
**Visual intuition:**

Each data point has a **vertical residual** — the distance from the point to the regression line.

```         
observed Y │     ●  ← residual (positive)
           │     |
predicted Ŷ│-----●--- regression line
           │
           └──────────── X
```

A point *above* the line has a **positive** residual. A point *below* has a **negative** residual. The sum of all residuals = 0.
:::
:::::

# R² — How Much Does X Explain?

::::: columns
::: {.column width="60%"}
$$R^2 = \frac{SS_\text{regression}}{SS_\text{total}} = 1 - \frac{SS_\text{residual}}{SS_\text{total}}$$

- **SS_total** = total variation in Y around its mean
- **SS_regression** = variation explained by X
- **SS_residual** = variation *not* explained (scatter around the line)

R² ranges from 0 to 1:

| R²   | Interpretation                |
|------|-------------------------------|
| 1.00 | X explains all variation in Y |
| 0.85 | X explains 85% of variation   |
| 0.00 | X explains nothing            |

📖 W&S §17.4 p. 555
:::

::: {.column width="40%"}
**Our calibration data:**

Because paper has uniform density, mass and area are nearly perfectly linearly related.

We expect R² ≈ 1.00.

**Important:** R² tells you *how well the line fits* the data you have. It does not by itself tell you whether the relationship is statistically significant — that requires a test of the slope.

A significant slope with a low R² is possible (weak but real relationship). In biology, R² values of 0.3–0.7 are common and meaningful.
:::
:::::

# Assumptions of Regression

:::::: columns
::: {.column width="60%"}
Four assumptions must hold for the p-value and confidence intervals to be valid (W&S §17.5):

| \# | Assumption | How to check |
|------------------------|------------------------|------------------------|
| 1 | **Linearity** — Y changes linearly with X | Residual vs. fitted plot |
| 2 | **Independence** — observations are independent | Study design |
| 3 | **Equal variance** — spread of residuals is constant | Residual vs. fitted plot |
| 4 | **Normality** — residuals are normally distributed | QQ plot + Shapiro-Wilk |
:::

:::: {.column width="40%"}
::: callout-important
**Check residuals, not raw data.**

The normality assumption is about the **residuals** (observed − predicted), not the original Y values.

A well-fitted residual plot looks like a random horizontal cloud of points centered at zero — no curves, no funnel shapes.
:::

📖 W&S §17.5 pp. 557–560 (Figures 17.5-1, 17.5-4)
::::
::::::

# Load Libraries and Data

```{r load-libs-05}
# Load packages at the top of the script ---------------
library(readxl)      # reading Excel files
library(tidyverse)   # data manipulation + ggplot2
library(skimr)       # fast data overview
```

```{r load-data-05}
# Load the paper calibration data ----------------------
paper_df <- read_excel("data/paper_area_weights.xlsx")
```

::: callout-note
The file `paper_area_weights.xlsx` contains 118 paper squares of **known area** (1–567 cm²) with their measured masses. This is the calibration dataset we use to predict leaf area from a tracing weight.
:::

# Inspect the Data

::::: columns
::: {.column width="60%"}
```{r skim-data-05}
# Quick overview of the calibration data ---------------
skim(paper_df)
```
:::

::: {.column width="40%"}
```{r scatter-summary-05}
# How many squares per area group? --------------------
paper_df %>%
  group_by(area_cm2) %>%
  summarize(
    n       = n(),
    mean_wt = round(mean(mass_g), 5)
  )
```
:::
:::::

# First Plot — Scatter

::::: columns
::: {.column width="60%"}
```{r scatter-raw-05}
#| fig-format: png
# Scatter plot of area vs mass to check linearity -----
scatter_05_plot <- paper_df %>%
  ggplot(aes(x = mass_g, y = area_cm2)) +
  geom_point(alpha = 0.5, size = 2) +
  labs(
    title = "Paper Mass vs. Known Area",
    x     = "Mass (g)",
    y     = "Area (cm²)"
  ) +
  theme_minimal()
scatter_05_plot
```
:::

::: {.column width="40%"}
**What to look for before fitting:**

- Is the relationship **linear**? → points should follow a straight line
- Is the variance **constant**? → spread around the trend should look similar at all X values
- Any obvious **outliers**?

Both variables are continuous and numeric → linear regression is appropriate.

📖 W&S §17.1 p. 541
:::
:::::

# Fit the Model with `lm() - Linear Model`

::::: columns
::: {.column width="60%"}
```{r fit-lm-05}
# Fit the least-squares regression line ---------------
# formula: response ~ predictor
paper_lm_model <- lm(area_cm2 ~ mass_g, data = paper_df)
```

```{r summary-lm-05}
# Full model summary — every number matters -----------
summary(paper_lm_model)
```
:::

::: {.column width="40%"}
**The `lm()` formula syntax:**

- `lm(Y ~ X, data = df)`
  - `Y` goes on the **left** of `~`
  - `X` goes on the **right**
  - `~` reads as "is predicted by"
- So `area_cm2 ~ mass_g` means: "area is predicted by mass"
- The result is a **model object** — we can extract everything from it.
- 📖 W&S §17.1 p. 543
:::
:::::

# Reading the `summary()` Output

::::: columns
::: {.column width="60%"}
```{r coef-extract-05}
# Extract the key numbers from the model --------------
a_intercept <- coef(paper_lm_model)[1]   # intercept
b_slope     <- coef(paper_lm_model)[2]   # slope

cat("Intercept (a) =", round(a_intercept, 4), "cm²\n")
cat("Slope (b)     =", round(b_slope, 2),    "cm²/g\n")
```
:::

::: {.column width="40%"}
**Decode the Coefficients table:**

| Row           | What it is                         |
|---------------|------------------------------------|
| `(Intercept)` | *a* — predicted area when mass = 0 |
| `mass_g`      | *b* — cm² per gram of paper        |
| `Std. Error`  | uncertainty in each estimate       |
| `t value`     | *b* / SE — tests if slope ≠ 0      |
| `Pr(>|t|)`    | *p*-value for that t-test          |

**H₀:** β = 0 (mass does not predict area)

**Hₐ:** β ≠ 0 — our slope tests this hypothesis

📖 W&S §17.3 p. 551–553
:::
:::::

# The Calibration Equation and R²

::::: columns
::: {.column width="60%"}
```{r r2-compute-05}
# Extract R-squared and build the equation label ------
r2_val <- summary(paper_lm_model)$r.squared

cat("R² =", round(r2_val, 6), "\n\n")
cat("Calibration equation:\n")
cat("  area (cm²) =", round(b_slope, 2), "× mass (g) +",
    round(a_intercept, 4), "\n\n")
cat("Interpretation of slope:\n")
cat("  Every 1 g of paper =", round(b_slope, 1), "cm² of area\n")
```
:::

::: {.column width="40%"}
- **What R² = 0.9999 means:**
  - 99.99% of the variation in paper area is explained by mass. The remaining 0.01% is measurement noise from the balance.

  - This is expected — the relationship between mass and area in uniform paper is essentially perfect.

  - In biological data, R² values of 0.3–0.7 are common and meaningful.

  - 📖 W&S §17.4 p. 555
:::
:::::

# Plot the Calibration Curve

::::: columns
::: {.column width="60%"}
```{r calibration-line-plot-05}
#| fig-format: png
# Calibration curve with regression line + 95% CI band
calibration_plot <- paper_df %>%
  ggplot(aes(x = mass_g, y = area_cm2)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE,
              color = "steelblue", fill = "lightblue") +
  labs(
    title    = "Paper Calibration Curve",
    subtitle = paste0("area = ", round(b_slope, 2),
                      " × mass + ", round(a_intercept, 3),
                      "  |  R² = ", round(r2_val, 4)),
    x        = "Paper Mass (g)",
    y        = "Area (cm²)"
  ) +
  theme_minimal()

calibration_plot
```
:::

::: {.column width="40%"}
**The shaded band** = 95% confidence band for the *mean* predicted area at each mass value.

**Key observation:** the band is narrowest in the middle of the data range and widens at the extremes — predictions are most precise near $\bar{X}$.

This is exactly Figure 17.2-1 from W&S (p. 549).

📖 W&S §17.2 p. 548–550
:::
:::::

# Check Assumptions — Residual Plot

:::::: columns
::: {.column width="60%"}
```{r add-resid-df-05}
# mutate() adds new columns to the data frame ----------
# Here we add the model's fitted values and residuals
# so we can plot them — recall mutate() from Lecture 03!
paper_resid_df <- paper_df %>%
  mutate(
    fitted    = fitted(paper_lm_model),
    residuals = residuals(paper_lm_model)
  )
```

```{r resid-vs-fitted-05}
#| fig-format: png
# Residuals vs Fitted — checks linearity & equal variance
resid_plot <- paper_resid_df %>%
  ggplot(aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Fitted Values (cm²)", y = "Residuals (cm²)") +
  theme_minimal()

resid_plot
```
:::

:::: {.column width="40%"}
**What to look for:**

- ✅ **Good:** random cloud of points centered at zero; constant spread across all fitted values
- ❌ **Bad:** - Curved pattern → linearity violated - Fan shape (funnel) → unequal variance
- 📖 W&S §17.5 p. 559 (Figure 17.5-4)

::: callout-tip
The residual plot is the most important diagnostic. Always check it before interpreting your results.
:::
::::
::::::

# Check Assumptions — QQ Plot and Shapiro-Wilk

::::: columns
::: {.column width="60%"}
```{r qq-resid-05}
#| fig-format: png
# QQ plot of residuals — checks normality assumption --
qq_resid_plot <- paper_resid_df %>%
  ggplot(aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red", linewidth = 0.8) +
  labs(
    title = "Normal QQ Plot of Residuals",
    x     = "Theoretical Quantiles",
    y     = "Sample Quantiles (residuals)"
  ) +
  theme_minimal()

qq_resid_plot
```
:::

::: {.column width="40%"}
```{r shapiro-resid-05}
# Shapiro-Wilk test on residuals — NOT on raw data ---
shapiro.test(residuals(paper_lm_model))
```

**Remember:**

- The normality assumption is about **residuals**, not raw data
- H₀ (Shapiro-Wilk): residuals are normally distributed
- *p* \> 0.05 → proceed with confidence intervals and *p*-values

📖 W&S §17.5 p. 559
:::
:::::

# Predict Leaf Area from Tracing Mass

::::: columns
::: {.column width="60%"}
```{r predict-point-05}
# Plug tracing masses into the calibration line -------
# Suppose sunny leaf tracing weighs 0.092 g
# and shady leaf tracing weighs 0.138 g

mass_sunny <- 0.092
mass_shady <- 0.138

area_sunny <- b_slope * mass_sunny + a_intercept
area_shady <- b_slope * mass_shady + a_intercept

cat("Sunny tracing mass:", mass_sunny, "g\n")
cat("Predicted area:    ", round(area_sunny, 2), "cm²\n\n")
cat("Shady tracing mass:", mass_shady, "g\n")
cat("Predicted area:    ", round(area_shady, 2), "cm²\n")
```
:::

::: {.column width="40%"}
**The equation in action:**

$$\widehat{\text{area}} = 130.5 \times \text{mass} + 0.28$$

For a sunny tracing at 0.092 g:

$130.5 \times 0.092 + 0.28 = 12.3$ cm²

For a shady tracing at 0.138 g:

$130.5 \times 0.138 + 0.28 = 18.3$ cm²

This matches our observation from Lecture 03 — shady leaves are larger!
:::
:::::

# Using `predict()` — with Prediction Intervals

::::: columns
::: {.column width="60%"}
```{r predict-interval-05}
# predict() gives uncertainty around predictions ------
new_masses <- tibble(mass_g = c(0.092, 0.138))

# Confidence interval: for the MEAN area at this mass
predict(paper_lm_model,
        newdata  = new_masses,
        interval = "confidence") %>%
  round(3)

# Prediction interval: for a SINGLE new observation
predict(paper_lm_model,
        newdata  = new_masses,
        interval = "prediction") %>%
  round(3)
```
:::

::: {.column width="40%"}
**Two types of interval (W&S §17.2 p. 549):**

| Type           | Predicts                             | Width    |
|----------------|--------------------------------------|----------|
| **Confidence** | mean area for *all* leaves this mass | narrower |
| **Prediction** | area of *one specific* leaf          | wider    |

For predicting a single leaf, use the **prediction interval** — it captures individual-to-individual variation within the same mass class.

📖 W&S §17.2 pp. 548–550
:::
:::::

# How to Report Regression Results

::::: columns
::: {.column width="60%"}
```{r report-values-05}
# Pull reporting values from the model ----------------
f_stat   <- summary(paper_lm_model)$fstatistic
f_val    <- round(f_stat[1], 1)
df1      <- f_stat[2]
df2      <- f_stat[3]
p_val    <- pf(f_stat[1], df1, df2, lower.tail = FALSE)

cat("--- Results paragraph ---\n")
cat("Paper mass was a strong predictor of paper area\n")
cat("(linear regression: F(", df1, ",", df2, ") =", f_val, ",\n")
cat("p <", signif(p_val, 1), ", R² =", round(r2_val, 4), ").\n")
cat("The calibration equation is:\n")
cat("  area =", round(b_slope, 2), "× mass +",
    round(a_intercept, 3), "cm²\n")
cat("(slope 95% CI:", round(confint(paper_lm_model)[2,1], 2),
    "to", round(confint(paper_lm_model)[2,2], 2), "cm²/g)\n")
```
:::

::: {.column width="40%"}
**Standard report format:**

> *"Paper area was accurately predicted from paper mass (linear regression: F(df₁, df₂) = F, p \< 0.001, R² = 0.9999). The regression equation was area = 130.5 × mass + 0.28 (slope 95% CI: X to Y cm²/g)."*

**Always include:** - F-statistic and both df - *p*-value (never "p = 0.000" — write "p \< 0.001") - R² - The actual equation - 95% CI for the slope

📖 W&S §17.3 p. 553
:::
:::::

# What We Learned Today

::::: columns
::: {.column width="60%"}
**Concepts:**

- **Regression** predicts Y from X using a straight line equation
- Equation: $\hat{Y} = a + bX$ — slope *b*, intercept *a*
- **Least squares** — minimizes sum of squared residuals
- **R²** — fraction of Y variation explained by X
- **Four assumptions** — linearity, independence, equal variance, normality of residuals
- **Two prediction types** — confidence vs. prediction intervals

**R skills:**

- `lm(Y ~ X, data)` — fits the model
- `summary()` — extracts slope, intercept, R², p-values
- `coef()`, `fitted()`, `residuals()` — pull model components
- `mutate()` with `fitted()` and `residuals()` — add model output to a data frame
- `predict(model, newdata, interval = "prediction")` — predictions with uncertainty
- `geom_smooth(method = "lm", se = TRUE)` — plot the line with CI band
:::

::: {.column width="40%"}
**Chapter references:**

- 📖 W&S §17.1 — Linear regression
- 📖 W&S §17.2 — Confidence in predictions
- 📖 W&S §17.3 — Testing slope hypotheses
- 📖 W&S §17.4 — R²
- 📖 W&S §17.5 — Assumptions

**Up next — Lecture 06:**

- Applying the calibration: predicting and comparing leaf area between sides
- Revisiting the t-test on *predicted area* (cm²) rather than raw weight (g)
- Discussion: which measurement is more biologically meaningful?
:::
:::::