---
order: 6
week: 3
# =====================================================================
#  regression_analysis_report.qmd
#  THE "MARKDOWN WAY" of doing the SAME regression analysis.
#  This is the twin of regression_analysis_script.R — identical code,
#  but wrapped in writing so that pressing Render produces a finished
#  Word report with the numbers and figures already in place.
# =====================================================================
title: "Paper Calibration Curve: Predicting Leaf Area from Mass"
subtitle: "The same analysis as regression_analysis_script.R — but as a report"
author: "Your Name Here"
date: today

format:
  docx:
    toc: true
    number-sections: true
    fig-width: 6
    fig-height: 4
    fig-format: png
  html: default          # bonus: the same file also makes a web page

# Hide the code by default so the report reads like a report.
# Flip echo to true if you want a teaching handout that shows the code.
execute:
  echo: false
  warning: false
  message: false
---

```{r libraries}
# Load all packages in one chunk at the top ----------------------------
library(readxl)
library(tidyverse)
library(skimr)
```

```{r load-data}
# Read the paper calibration data --------------------------------------
paper_df <- read_excel("data/paper_area_weights.xlsx")
```

<!--
HOW TO READ THIS WORKED EXAMPLE (active reading)
Don't just skim. Before you press Render, PREDICT what each computed value and
figure will show, then render and compare. Better still, retype a chunk or two by
hand. The PREDICT comments below mark good places to pause. None of these comment
lines appear in the rendered Word document.
-->

# Introduction

To estimate leaf surface area without a leaf-area meter, we built a
**calibration curve**. Paper of uniform stock has a constant area-per-gram, so
the **mass** of a paper square is a near-perfect linear predictor of its
**area**. Once we know that relationship, we can trace a leaf onto the same
paper, weigh the tracing, and convert its mass to area.

- **Explanatory variable (X):** paper mass (g) — easy to measure on a balance
- **Response variable (Y):** paper area (cm²) — what we want to predict

# Methods

We weighed 118 paper squares of known area (1–567 cm²) and fit a simple linear
regression, `area_cm2 ~ mass_g`, with `lm()`. We checked the four regression
assumptions using a residuals-vs-fitted plot and a normal Q–Q plot with a
Shapiro–Wilk test on the residuals. All work was done in R with the tidyverse;
this report was written in Quarto so every number and figure below is computed
directly from the data.

# Results

```{r fit-model}
# Fit the least-squares regression line --------------------------------
paper_lm_model <- lm(area_cm2 ~ mass_g, data = paper_df)
```

```{r pull-values}
# Pull the values we want to write into the text below -----------------
a_intercept <- coef(paper_lm_model)[1]
b_slope <- coef(paper_lm_model)[2]
r2_val <- summary(paper_lm_model)$r.squared

f_stat <- summary(paper_lm_model)$fstatistic
f_val <- round(f_stat[1], 1)
df1 <- f_stat[2]
df2 <- f_stat[3]
p_val <- pf(f_stat[1], df1, df2, lower.tail = FALSE)
```

<!-- 🔮 PREDICT before you Render: what F, p, and R² will fill the sentence below, and what slope (cm² per gram)? Write your guesses, then render and check. -->

Paper mass was an extremely strong predictor of paper area
(linear regression: F(`r df1`, `r df2`) = `r f_val`,
*p* < `r signif(p_val, 1)`, R² = `r round(r2_val, 4)`). The calibration
equation was:

$$\widehat{\text{area}} = `r round(b_slope, 2)` \times \text{mass} + `r round(a_intercept, 3)`$$

In words, every additional gram of paper corresponds to about
**`r round(b_slope, 1)` cm²** of area.

::: {.callout-note appearance="simple"}
Notice that **F, *p*, R², the slope, and the intercept above were never typed by
hand** — they are inserted with inline code like `` `r round(b_slope, 2)` ``.
Change the data file and re-render, and every one of these updates itself. That
is the difference from the `.R` script, where you would copy each number into
Word yourself.
:::

## The calibration curve

```{r calibration-plot}
#| fig-cap: "Figure 1. Paper area vs. mass with the least-squares line and 95% confidence band."
calibration_plot <- paper_df %>%
  ggplot(aes(x = mass_g, y = area_cm2)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "steelblue",
    fill = "lightblue"
  ) +
  labs(
    subtitle = paste0(
      "area = ",
      round(b_slope, 2),
      " × mass + ",
      round(a_intercept, 3),
      "  |  R² = ",
      round(r2_val, 4)
    ),
    x = "Paper Mass (g)",
    y = "Area (cm²)"
  ) +
  theme_minimal()

calibration_plot
```

## Assumption checks

```{r resid-df}
paper_resid_df <- paper_df %>%
  mutate(
    fitted = fitted(paper_lm_model),
    residuals = residuals(paper_lm_model)
  )
```

```{r resid-plot}
#| fig-cap: "Figure 2. Residuals vs. fitted values — a random band with no funnel or curve supports linearity and equal variance."
paper_resid_df %>%
  ggplot(aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Fitted Values (cm²)", y = "Residuals (cm²)") +
  theme_minimal()
```

```{r shapiro}
# Shapiro-Wilk on the RESIDUALS (not the raw data)
shapiro_out <- shapiro.test(residuals(paper_lm_model))
```

A Shapiro–Wilk test on the residuals gave W = `r round(shapiro_out$statistic, 3)`,
*p* = `r signif(shapiro_out$p.value, 2)`.

## Predicting leaf area

```{r predict}
# Predict area for two example leaf tracings ---------------------------
mass_sunny <- 0.092
mass_shady <- 0.138

area_sunny <- b_slope * mass_sunny + a_intercept
area_shady <- b_slope * mass_shady + a_intercept
```

<!-- 🔮 PREDICT before you Render: which tracing gives the larger predicted area — the 0.092 g sunny or the 0.138 g shady? Guess the two areas, then render. -->

Applying the equation to two example tracings, a sunny-side tracing of
`r mass_sunny` g predicts an area of **`r round(area_sunny, 1)` cm²**, while a
heavier shady-side tracing of `r mass_shady` g predicts
**`r round(area_shady, 1)` cm²** — consistent with our earlier finding that
shady leaves are larger.

# Discussion

*Write one short paragraph here.* Did the calibration fit well (look at R² and
Figure 1)? Did the residual plot and Shapiro–Wilk support the assumptions? Why
is working in **area** (cm²) more biologically meaningful than raw **mass** (g)
when comparing sunny and shady leaves? This is where you paste the results
paragraph you drafted in the worksheet.

# References

- Whitlock, M. & Schluter, D. *The Analysis of Biological Data* (2nd ed.), Ch. 17.
- Wickham, H. *R for Data Science (2nd ed.)* — <https://r4ds.hadley.nz/>
