---
title: "Lecture 10 — T-Tests II: Running & Reporting the Test"
subtitle: "Running Welch's t-test, reading the output, and writing it up"
author: "Bill Perry"
date: today

order: 10
week: 5
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Running Welch's two-sample t-test, decoding every line of the output,
  making a formal decision, and writing a proper scientific results
  sentence.
categories: [tidyverse, descriptive-stats]

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 09 — T-Tests I)

- **H₀:** $\mu_{shady} = \mu_{sunny}$ · **Hₐ:** $\mu_{shady} \neq \mu_{sunny}$ (two-tailed, α = 0.05)
- **Normality checked** — histogram + Shapiro-Wilk, both sides roughly bell-shaped for n = 10
- **Variance checked** — Levene's test run (we use Welch's either way)
- **Decision:** proceed with Welch's two-sample t-test

::: callout-note
**✅ Key idea from Lecture 09**

Every box is checked. Today we actually run the test, decode what R hands back, and turn it into a sentence a reader can trust.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- **Run** Welch's two-sample t-test in R
- **Read** every line of the `t.test()` output
- **Decide**: reject or fail to reject H₀
- **Visualize** the result with the test statistics embedded
- **Report** results in proper scientific writing format

::: callout-tip
**🖐 Try it yourself**

By the end you will have a complete, reportable t-test result and a publication-style figure to go with it.
:::
::::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 4 — *Estimating with Uncertainty*
- 📖 [R4DS §3](https://r4ds.hadley.nz/data-transform)

Both are in the course `readings/` folder.

**Today's naming:**

- test objects → `_model`
- plots → `_plot`
:::
::::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This lecture runs in **two short chunks**. After each chunk you switch to the **activity worksheet** and type the code yourself.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

Guessing "above or below 0.05" before you see `t.test()` output forces you to reason about the test instead of just reading a number off the screen — that habit is what makes a p-value meaningful instead of magic.
:::
::::
::::::

# Load Libraries and Data

```{r load-libs-10}
# Load all packages at the top ----------------------------
library(readxl) # reading Excel files
library(tidyverse) # data manipulation + ggplot2
library(car) # Levene's test for variance equality
```

```{r load-data-10}
# Load the leaf data from the data folder -----------------
tree_df <- read_excel("data/2026_06_25_tree_experiment_raw_data.xlsx")
```

# Quick Recap of Our Assumption Checks

::::: columns
::: {.column width="60%"}
```{r recap-checks}
# Rerun the assumption checks quickly for reference -----
recap_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    n = sum(!is.na(weight_g)),
    mean_wt = round(mean(weight_g, na.rm = TRUE), 2),
    sd_wt = round(sd(weight_g, na.rm = TRUE), 2),
    se_wt = round(sd_wt / sqrt(n), 2)
  )
recap_df

leveneTest(weight_g ~ side, data = tree_df)
```
:::

::: {.column width="40%"}
- Both groups: roughly normal (Lecture 09)
- Levene's: recorded, but **does not change our test choice**
- We proceed with **Welch's** (`var.equal = FALSE`)

::: callout-tip
If your Worksheet 09 numbers don't match exactly, that's fine — small rounding differences don't change the conclusion.
:::
:::
:::::

# 🧩 Chunk 1 of 2 · Run & Interpret

**We will cover:** running Welch's `t.test()`, reading every line of the output, and making a formal reject / fail-to-reject decision.

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 1–3** (run, decode, decide).
:::

# Step 4 — Run the Welch's t-Test

::: callout-note
**🔮 Predict first:** Write down a guess for the *p*-value before you run `t.test()` — above or below 0.05? Then see how close you were.
:::

::::: columns
::: {.column width="60%"}
```{r welch-ttest}
# Welch's two-sample t-test — unequal variance ---------
leaf_ttest_model <- t.test(
  weight_g ~ side, # formula: response ~ grouping variable
  data = tree_df,
  var.equal = FALSE, # Welch's — no pooled variance
  alternative = "two.sided" # two-tailed test
)

leaf_ttest_model
```
:::

::: {.column width="40%"}
**Key arguments:**

| argument                    | what it does                                |
|-----------------------------|---------------------------------------------|
| `weight_g ~ side`           | compare `weight_g` between levels of `side` |
| `var.equal = FALSE`         | use Welch's correction                      |
| `alternative = "two.sided"` | two-tailed test                             |

The result is a model object. We decode each line of the output on the next slide.
:::
:::::

# Reading the `t.test()` Output

::::: columns
::: {.column width="60%"}
```{r ttest-parse}
# Extract individual values from the model object ------
cat("t-statistic  :", round(leaf_ttest_model$statistic, 3), "\n")
cat("df (Welch)   :", round(leaf_ttest_model$parameter, 2), "\n")
cat("p-value      :", signif(leaf_ttest_model$p.value, 3), "\n")
cat(
  "95% CI       :",
  round(leaf_ttest_model$conf.int[1], 2),
  "to",
  round(leaf_ttest_model$conf.int[2], 2),
  "g\n"
)
cat("Mean shady   :", round(leaf_ttest_model$estimate[1], 2), "g\n")
cat("Mean sunny   :", round(leaf_ttest_model$estimate[2], 2), "g\n")
```
:::

::: {.column width="40%"}
**Line by line:**

| output                | meaning                                          |
|-----------------------|--------------------------------------------------|
| `t`                   | our calculated t-statistic                       |
| `df`                  | Welch-Satterthwaite df (non-integer is normal!)  |
| `p-value`             | probability of this result if H₀ is true         |
| `95% CI`              | plausible range for the true difference in means |
| `mean of shady/sunny` | the two group means                              |

The **95% CI** is the range we could expect to see if we repeated this study many times.
:::
:::::

# Step 5 — Make a Decision

:::::: columns
::: {.column width="60%"}
```{r decision-output}
# State the decision formally --------------------------
# if/else: R checks the condition and runs one of two blocks
# We will learn these formally in the functions lecture
p_val <- leaf_ttest_model$p.value
alpha <- 0.05

if (p_val < alpha) {
  cat("p =", signif(p_val, 3), "< α =", alpha, "\n")
  cat("Decision: REJECT H₀\n")
  cat("Conclusion: leaf weight differs between sides.\n")
} else {
  cat("p =", signif(p_val, 3), ">= α =", alpha, "\n")
  cat("Decision: FAIL TO REJECT H₀\n")
  cat("Conclusion: insufficient evidence of a difference.\n")
}
```
:::

:::: {.column width="40%"}
::: callout-important
**What a small p-value means:**

If H₀ were true (no real difference), we would get a t-statistic this large or larger less than `p` × 100% of the time just by chance.

We set α = 0.05 *before* the test. If *p* \< α, we reject H₀.
:::

**What a small p-value does NOT mean:**

- It is not the probability that H₀ is true
- Statistical significance ≠ biological importance
- Report the effect size (difference in means) alongside *p*
::::
::::::

# → ACTIVITY 10 Parts 1–3 now

::: {.activitybanner}
**🛑 Do Activity Parts 1–3 now**

Run the test, decode t / df / p / CI, and state your decision in formal language. Predict the p-value first.
:::

# 🧩 Chunk 2 of 2 · Visualize & Report

**We will cover:** embedding the test result in a boxplot and writing a scientific results sentence.

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 4–5** (plot and the results sentence).
:::

# Visualize the Result

::::: columns
::: {.column width="60%"}
```{r boxplot-10}
#| fig-format: png
#| fig-alt: "Boxplot of leaf weight in grams by tree side with jittered individual points overlaid, subtitled with the Welch's t-test statistic and p-value showing shady leaves heavier than sunny leaves."
# Final boxplot with all points -----------------------
leaf_box_10_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.6,
    size = 2.5
  ) +
  labs(
    title = "Leaf Weight by Tree Side",
    subtitle = paste0(
      "Welch's t-test: t = ",
      round(leaf_ttest_model$statistic, 2),
      ", p = ",
      signif(leaf_ttest_model$p.value, 2)
    ),
    x = "Side of Tree",
    y = "Leaf Weight (g)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
leaf_box_10_plot
```
:::

:::: {.column width="40%"}
**Including the test result in the figure:**

- `subtitle` pulls t and p directly from the model object
- R fills in the actual numbers automatically
- If you re-run with new data, the subtitle updates

::: callout-tip
Embedding test statistics in the plot subtitle saves you from copying numbers by hand — and avoids typos in your report.
:::
::::
::::::

# How to Report Results in Science

::::: columns
::: {.column width="60%"}
```{r report-stats}
# Compute reporting values from the model object -------
t_val <- round(leaf_ttest_model$statistic, 2)
df_val <- round(leaf_ttest_model$parameter, 1)
p_val2 <- signif(leaf_ttest_model$p.value, 2)

cat("--- Results sentence ---\n")
cat("Shady-side leaves were significantly heavier than\n")
cat(
  "sunny-side leaves (Welch's t-test: t(",
  df_val,
  ") =",
  t_val,
  ", p =",
  p_val2,
  ").\n"
)
cat(
  "Mean ± SE: shady =",
  round(leaf_ttest_model$estimate[1], 2),
  "±",
  round(recap_df$se_wt[recap_df$side == "shady"], 2),
  "g,\n"
)
cat(
  "           sunny =",
  round(leaf_ttest_model$estimate[2], 2),
  "±",
  round(recap_df$se_wt[recap_df$side == "sunny"], 2),
  "g.\n"
)
```
:::

::: {.column width="40%"}
**Standard format:**

> *"\[Finding\] (Welch's t-test: t(df) = X.XX, p = X.XXX; mean ± SE: group1 = X.X ± X.X units, group2 = X.X ± X.X units)."*

**Always include:**

- Test type (Welch's t-test)
- t-statistic and df
- Exact *p*-value (or \< 0.001)
- Group means ± SE
- Units

**Never report** "p = 0.000" — write "p \< 0.001" instead.
:::
:::::

# Save Your Plot

```{r save-plots-10}
# Save the plot to the figures folder ------------------
ggsave(
  "figures/leaf_weight_boxplot_ttest.png",
  plot = leaf_box_10_plot,
  width = 5,
  height = 5,
  units = "in",
  dpi = 300
)
```

:::::: columns
::: {.column width="60%"}
Your plot is now in your `figures/` folder with the test statistics embedded in the subtitle — ready for a lab report or paper.
:::

:::: {.column width="40%"}
::: callout-tip
`dpi = 300` is publication quality. Use it for any figure that will appear in a paper, poster, or formal report.
:::
::::
::::::

# → ACTIVITY 10 Parts 4–5 now

::: {.activitybanner}
**🛑 Go to Activity 10 — Parts 4–5**

Build the figure with the statistics embedded in the subtitle, then write your results sentence. Predict what the subtitle will say before you run it.
:::

# What We Learned Today

::::: columns
::: {.column width="60%"}
**Concepts:**

- Steps 4–5 of the five-step framework — run, interpret, report
- **Reading output** — t, df, *p*, CI, group means
- A results sentence needs: test name, t, df, *p*, means ± SE, units

**R skills:**

- `t.test(y ~ group, data, var.equal = FALSE, alternative = "two.sided")`
- `model$statistic`, `model$p.value`, `model$conf.int` — extract results
- `paste0(...)` in `labs(subtitle = )` — embed test statistics in figures
:::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 4 — *Estimating with Uncertainty*
- 📖 [R4DS §3](https://r4ds.hadley.nz/data-transform)

**Up next — Lecture 11:**

- Linear regression — predicting leaf area from paper tracing mass
- Building a calibration curve with `lm()`
- R², residuals, and assumption checking
:::
:::::
