---
title: "Lecture 04 — GGPlot II"
subtitle: "New data, safe zooming, distributions, themes, and patchwork"
author: "Bill Perry"
date: today

order: 4
week: 2
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Practicing the plots you already know on a brand-new dataset, zooming
  safely with coord_cartesian() vs. the ylim()/scale_y_continuous(limits=)
  trap, histograms and density plots, a reusable custom theme file, and
  combining panels with patchwork.
categories: [ggplot, tidyverse]

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(width = 65)   # printed R output wraps at 65 too
```

# Where we left off (Lecture 03 — GGPlot I)

- Piping straight into a plot: `leaf_df %>% ggplot(...)`
- `scale_color_manual()` / `scale_fill_manual()` — colors you choose
- `facet_wrap()` and `facet_grid()` — one panel per group
- `stat_summary()` — mean and mean ± SE, drawn from raw data

::: callout-note
**✅ Key idea from Lecture 03**

You already know how to build a grouped, colored, faceted figure — but only on `leaf_df`. Today's real test: can you do it on data you've never seen before?
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- Practice your plotting skills on a **brand-new dataset**, and see where you'd find one yourself
- Zoom into a plot **without silently dropping data** — `coord_cartesian()`
- Build a **histogram** and a **density plot** — two views of one distribution
- Apply a **reusable theme file**, and combine panels with **`patchwork`**

::: callout-tip
**🖐 Try it yourself**

By the end you will recognize the single most common ggplot mistake — zooming with `ylim()` / `scale_y_continuous(limits = )` — before you ever make it yourself.
:::
::::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 2 — *Displaying Data*
- 📖 [R4DS Ch 1](https://r4ds.hadley.nz/data-visualize) — Data Visualization

Both PDFs are in the course `readings/` folder.
:::
::::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This lecture runs in **five short chunks**. After each chunk you switch to the activity and type the code yourself into your R script.
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

`ylim()` and `coord_cartesian()` produce plots that look *identical* for a simple scatter — the difference only shows up once you compute something (a boxplot, a smooth). Typing both yourself and comparing outputs is the only way this sticks.
:::
::::
::::::

# 🧩 Chunk 1 of 5 · New Data, Same Skills

**We will cover:** loading `palmerpenguins`, and rebuilding the two plot types you already know on data you've never seen.

::: activitybanner
**🛑 After this chunk you will do Activity Parts 1–2.**
:::

# A Dataset You Didn't Collect

::::: columns
::: {.column width="60%"}
```{r load-libs-04}
# Load all packages at the top of every script ----------
library(tidyverse)      # data wrangling + ggplot2
library(palmerpenguins) # penguins data, Palmer Station LTER
```

```{r load-data-04}
# palmerpenguins ships the data already clean -----------
penguins_df <- penguins %>%
  drop_na(species, bill_length_mm, bill_depth_mm,
          flipper_length_mm, body_mass_g)

glimpse(penguins_df)
```
:::

::: {.column width="40%"}
- `penguins` comes **inside the package** — no `read_excel()` needed, just `library(palmerpenguins)`
- 344 penguins, 3 species, Palmer Station, Antarctica
- `drop_na()` removes the handful of rows missing a measurement — same idea as the `NA`s in `leaf_df`
:::
:::::

# The Same Simple XY Plot — On New Data

::: callout-note
**🔮 Predict first:** flipper length and body mass should move together — longer flippers, heavier birds. Will three species separate into three clouds, or overlap into one?
:::

::::: columns
::: {.column width="60%"}
```{r xy-demo}
#| fig-format: png
#| fig-alt: "Scatter plot of penguin body mass in grams against flipper length in millimeters, colored by species, showing three roughly separated clouds of points."
# The exact same recipe you used on leaf_df ------------
penguins_df %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             color = species)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Body Mass vs. Flipper Length",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  ) +
  theme_minimal()
```
:::

::: {.column width="40%"}
- Same `aes()` + `geom_point()` + `labs()` + `theme_minimal()` you already typed for leaves
- The **dataset changed**, the **recipe didn't** — that's the point
- Nothing here is new syntax; it's proof you can already do this on any data
:::
:::::

# The Same Simple Categorical Plot — On New Data

::::: columns
::: {.column width="60%"}
```{r box-demo}
#| fig-format: png
#| fig-alt: "Boxplot of penguin body mass in grams by species, with jittered raw points overlaid."
# boxplot + jitter, exactly like the leaf_df pattern ---
penguins_df %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4) +
  labs(x = "Species", y = "Body Mass (g)",
       title = "Body Mass by Species") +
  theme_minimal() +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
::: callout-tip
**💡 Where would you find a dataset like this on your own?**

- `library(palmerpenguins)` or `library(FSA)` (fish data, Derek Ogle) — real data bundled inside an R package, zero downloading
- [Kaggle — search "UFO sightings"](https://www.kaggle.com/datasets?search=ufo+sightings) or [Kaggle — search "Bigfoot sightings"](https://www.kaggle.com/datasets?search=bigfoot+sightings) — download a `.csv`, load it with `read_csv()` like any other file

Both a simple XY numeric plot and a simple categorical plot (like today's two) work on almost any tidy dataset you find.
:::
::::
:::::

# → ACTIVITY 4 Parts 1–2 now

::: activitybanner
**🛑 Do Activity Parts 1–2 now**

Load `palmerpenguins`, then rebuild the XY scatter and the boxplot+jitter — same recipe, new data. Predict, type, run.
:::

# 🧩 Chunk 2 of 5 · Zooming In Safely

**We will cover:** `coord_cartesian()` to zoom — and the `ylim()` / `scale_y_continuous(limits = )` trap to avoid.

::: activitybanner
**🛑 After this chunk you will do Activity Part 3.**
:::

# `coord_cartesian()` — Zoom the View, Keep the Data

::::: columns
::: {.column width="60%"}
```{r coord-cartesian-demo}
#| fig-format: png
#| fig-alt: "Boxplot of penguin body mass by species zoomed to a narrower y-axis window using coord_cartesian, with all 342 penguins still contributing to each box."
# coord_cartesian() zooms the VIEW, keeps all the data -
penguins_df %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  coord_cartesian(ylim = c(3000, 5000)) +
  labs(x = "Species", y = "Body Mass (g)",
       title = "Zoomed with coord_cartesian()") +
  theme_minimal() +
  theme(legend.position = "none")
```
:::

::: {.column width="40%"}
- Zooms the **view** — every penguin still contributes to each box's median and quartiles
- The window just crops what you can *see* afterward
- This is the one to reach for whenever you just want a closer look
:::
:::::

# The Trap — `ylim()` / `scale_y_continuous(limits = )` Delete Rows First

::: callout-note
**🔮 Predict first:** Gentoo penguins run heavier than the other two species — many sit above 5000 g. If we drop every penguin outside `3000`–`5000` g *before* drawing the boxes, what happens to the Gentoo box specifically?
:::

::::: columns
::: {.column width="60%"}
```{r ylim-demo}
#| fig-format: png
#| fig-alt: "Boxplot of penguin body mass by species using ylim, with the Gentoo box recomputed from only the penguins remaining after out-of-range rows were silently dropped."
# ylim() DELETES rows outside the range first ----------
penguins_df %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  ylim(3000, 5000) +
  labs(x = "Species", y = "Body Mass (g)",
       title = "ylim() — boxes recomputed on fewer penguins") +
  theme_minimal() +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

`ylim()` — and its long form `scale_y_continuous(limits = c(3000, 5000))` — **silently delete every row** outside the range *before* anything is computed. Compare the Gentoo box above to the `coord_cartesian()` version: same window, different box.
:::

**Rule:** zoom with `coord_cartesian()`. Only use `ylim()` / `scale_y_continuous(limits = )` when you actually want those rows gone from the analysis.
::::
:::::

# → ACTIVITY 4 Part 3 now

::: activitybanner
**🛑 Do Activity Part 3 now**

Zoom with `coord_cartesian()`, then run the `ylim()` version and count exactly how many penguins it drops. Predict, type, run.
:::

# 🧩 Chunk 3 of 5 · Two Views of One Distribution

**We will cover:** `geom_histogram()` and `geom_density()` — the two standard ways to show how one numeric variable is distributed.

::: activitybanner
**🛑 After this chunk you will do Activity Part 4.**
:::

# `geom_histogram()` — Bars of Counts

::::: columns
::: {.column width="60%"}
```{r histogram-demo}
#| fig-format: png
#| fig-alt: "Histogram of penguin body mass in grams, showing a somewhat bimodal distribution."
# geom_histogram() bins the data and counts rows per bin
penguins_df %>%
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200, fill = "steelblue",
                 color = "white") +
  labs(x = "Body Mass (g)", y = "Count",
       title = "Distribution of Body Mass") +
  theme_minimal()
```
:::

::: {.column width="40%"}
- `binwidth` sets the width of each bar in the **units of `x`** — here, 200 g per bin
- Too narrow → noisy, spiky bars. Too wide → real structure gets smoothed away
- Try a few `binwidth` values before settling on one
:::
:::::

# `geom_density()` — a Smooth Curve Instead of Bars

::::: columns
::: {.column width="60%"}
```{r density-demo}
#| fig-format: png
#| fig-alt: "Smoothed density curve of penguin body mass in grams, covering the same range as the histogram above."
# geom_density() smooths the same shape into a curve ---
penguins_df %>%
  ggplot(aes(x = body_mass_g)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(x = "Body Mass (g)", y = "Density",
       title = "Density of Body Mass") +
  theme_minimal()
```
:::

:::: {.column width="40%"}
- Same variable, same shape — a density curve is a **smoothed histogram**, area under the curve sums to 1
- The y-axis is no longer a count — it's a density, which is why the numbers look small
- 📖 Whitlock & Schluter, Ch. 2 — *Displaying Data*

::: callout-tip
**Coming soon:** in the Summary Statistics lecture we'll come back to this exact curve and shade in the **95% range** underneath it. Today, just get comfortable building both plot types.
:::
::::
:::::

# → ACTIVITY 4 Part 4 now

::: activitybanner
**🛑 Do Activity Part 4 now**

Build a histogram and a density plot of the same variable. Predict, type, run.
:::

# 🧩 Chunk 4 of 5 · A Reusable Theme File

**We will cover:** sourcing a custom theme file and applying it — no need to build one from scratch today.

::: activitybanner
**🛑 After this chunk you will do Activity Part 5.**
:::

# `source()` a Theme File Instead of Repeating `theme()`

::::: columns
::: {.column width="60%"}
```{r themes-source-demo}
#| fig-format: png
#| fig-alt: "Boxplot of penguin body mass by species rendered with the custom theme_regular function sourced from an external file, with a clean white background and bold axis titles."
# source() loads functions from another .R file -------
source("themes/r_themes_for_3_sizes.R")

penguins_df %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(x = "Species", y = "Body Mass (g)",
       title = "Body Mass by Species") +
  theme_regular() +
  theme(legend.position = "none")
```
:::

::: {.column width="40%"}
- `source("themes/r_themes_for_3_sizes.R")` runs that file once and loads `theme_small()`, `theme_regular()`, `theme_large()` into your session
- You don't need to understand every line **inside** the file today — just how to call it
- Pick the size that matches where the figure is going: slides/report → `theme_regular()`
:::
:::::

# Tweaking Fonts and Colors on Top

::::: columns
::: {.column width="60%"}
```{r theme-tweak-demo}
#| fig-format: png
#| fig-alt: "The same boxplot with a larger bold title and custom fill colors layered on top of the sourced theme."
# theme() and scale_fill_manual() still work as usual --
penguins_df %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  scale_fill_manual(values = c(
    "Adelie" = "darkorange", "Chinstrap" = "purple",
    "Gentoo" = "cyan4"
  )) +
  labs(x = "Species", y = "Body Mass (g)",
       title = "Body Mass by Species") +
  theme_regular() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = rel(1.4), face = "bold")
  )
```
:::

:::: {.column width="40%"}
- A sourced theme is still just a `theme()` call — `+ theme(...)` **after** it overrides individual pieces, same as any built-in theme
- `scale_fill_manual()` still controls colors; the theme controls everything else (fonts, lines, margins)

::: callout-note
**✅ Key idea**

One theme file, sourced at the top of every script, keeps every figure in a project looking consistent — change the file once, every plot updates.
:::
::::
:::::

# → ACTIVITY 4 Part 5 now

::: activitybanner
**🛑 Do Activity Part 5 now**

Download the theme file, source it, and apply `theme_regular()` to one of your plots. Then tweak a font size and a color. Predict, type, run.
:::

# 🧩 Chunk 5 of 5 · Combining Panels with `patchwork`

**We will cover:** `patchwork` — putting two ggplot objects side by side as one figure.

::: activitybanner
**🛑 After this chunk you will do Activity Part 6.**
:::

# `patchwork` — Two Plots, One Figure

::::: columns
::: {.column width="60%"}
```{r patchwork-demo}
#| fig-format: png
#| fig-alt: "Two panels combined side by side with patchwork: a scatter plot of body mass against flipper length on the left, and a boxplot of body mass by species on the right, with the overall figure labeled panel A and B."
library(patchwork) # combine ggplot objects with + and /

xy_plot <- penguins_df %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             color = species)) +
  geom_point(alpha = 0.7) +
  labs(x = "Flipper Length (mm)", y = "Body Mass (g)",
       color = "Species") +
  theme_regular()

box_plot <- penguins_df %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(x = "Species", y = "Body Mass (g)") +
  theme_regular() +
  theme(legend.position = "none")

xy_plot + box_plot +
  plot_annotation(
    title = "Penguin Body Mass, Two Ways",
    tag_levels = "A"
  )
```
:::

::: {.column width="40%"}
- Each panel is an **ordinary saved ggplot object** — build them separately first, exactly like you already do
- `+` places two `patchwork`-ready plots **side by side**; `/` stacks them
- `plot_annotation()` adds one title and A/B/C tags across the **whole combined figure**, not any single panel
:::
:::::

# → ACTIVITY 4 Part 6 now

::: activitybanner
**🛑 Go to Activity 4 — Part 6**

Close the slides. Combine two of your plots with `patchwork` and `plot_annotation()`, then save the combined figure.
:::

# What We Learned Today

::::: columns
::: {.column width="60%"}
- The XY-scatter / boxplot+jitter recipe you already know works on **any** tidy dataset — `palmerpenguins`, `FSA`, or a `.csv` from Kaggle
- `coord_cartesian()` zooms safely; `ylim()` / `scale_y_continuous(limits = )` silently deletes rows first
- `geom_histogram()` and `geom_density()` — two views of one distribution
- `source()` a theme file once, then call `theme_regular()` (or `_small()` / `_large()`) everywhere
- `patchwork` (`+` / `/`) combines saved plots; `plot_annotation()` titles and tags the whole figure
:::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 2 — *Displaying Data*
- 📖 [R4DS Ch 1](https://r4ds.hadley.nz/data-visualize) — Data Visualization

**Up next — Lecture 05, Wrangling:**

- `filter()`, `select()`, `mutate()`, `arrange()`
- Chaining verbs into one pipeline
:::
:::::
