---
title: "Lecture 04 — GGPlot II"
subtitle: "More geoms, zooming safely, and picking a theme"
author: "Bill Perry"
date: today
 
order: 4
week: 2
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Mapping shape alongside color, adding a trend line with geom_smooth(),
  coord_cartesian() vs. xlim()/ylim(), and choosing a built-in theme.
categories: [ggplot, tidyverse]

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 03 — GGPlot I)

- Piping straight into a plot: `leaf_df %>% ggplot(...)`
- `scale_color_manual()` / `scale_fill_manual()` — colors you choose
- `facet_wrap()` and `facet_grid()` — one panel per group
- `stat_summary()` — mean and mean ± SE, drawn from raw data

::: callout-note
**✅ Key idea from Lecture 03**

You can already build a grouped, colored, faceted figure. Today: one more geom, a safer way to zoom, and picking a theme that looks intentional instead of default.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- Map **shape** alongside color, and add a trend line with `geom_smooth()`
- Zoom into a plot **without silently dropping data** — `coord_cartesian()`
- Pick a built-in **theme** and drop a redundant legend with `theme()`
- Save a figure in the right **format** for the job

::: callout-tip
**🖐 Try it yourself**

By the end you will recognize the single most common ggplot mistake — zooming with `xlim()`/`ylim()` — before you ever make it yourself.
:::
::::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 2 — *Displaying Data*
- 📖 [R4DS Ch 1](https://r4ds.hadley.nz/data-visualize) — Data Visualization

Both PDFs are in the course `readings/` folder.
:::
::::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This lecture runs in **three short chunks**. After each chunk you switch to the activity and type the code yourself into your R script.
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

`xlim()` and `coord_cartesian()` produce plots that look *identical* for simple scatter plots — the difference only shows up once you compute something (a smooth line, a boxplot). Typing both yourself and comparing outputs is the only way this sticks.
:::
::::
::::::

# Load Libraries and Data

```{r load-libs-04}
# Load all packages at the top of every script ----------
library(readxl) # reading Excel files
library(tidyverse) # data wrangling + ggplot2
library(janitor) # cleans up messy column names
```

```{r load-data-04}
# Read the leaf data from the data folder ---------------
leaf_df <- read_excel("data/2026_09_03_data_sci_leaf_area.xlsx") %>%
  clean_names()
```

# 🧩 Chunk 1 of 3 · Shape, Color, and a Trend Line

**We will cover:** mapping shape alongside color, and `geom_smooth()`.

::: activitybanner
**🛑 After this chunk you will do Activity Parts 1–2.**
:::

# Mapping Shape Alongside Color

::: callout-note
**🔮 Predict first:** We map both `color = shade` and `shape = shade` to the *same* variable. Will ggplot draw one legend or two?
:::

::::: columns
::: {.column width="60%"}
```{r shape-color-demo}
#| fig-format: png
#| fig-alt: "Scatter plot of leaf mass against petiole length, with sunny leaves shown as gold triangles and shady leaves as green circles."
# Map color AND shape to the same variable -------------
leaf_df %>%
  ggplot(aes(x = petiole_mm, y = mass_g, color = shade, shape = shade)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(
    values = c("sunny" = "goldenrod", "shady" = "forestgreen")
  ) +
  labs(
    title = "Leaf Mass vs. Petiole Length",
    x = "Petiole Length (mm)",
    y = "Leaf Mass (g)",
    color = "Shade",
    shape = "Shade"
  ) +
  theme_minimal()
```
:::

::: {.column width="40%"}
- Mapping the **same variable** to both `color` and `shape` merges them into **one legend**
- This makes the plot easier to read in black-and-white or for colorblind readers
- If `color` and `shape` legend titles differ in `labs()`, you get **two** legends instead — usually not what you want

📖 [R4DS §9 — Layers](https://r4ds.hadley.nz/layers)
:::
:::::

# Adding a Trend Line — `geom_smooth()`

:::::: columns
::: {.column width="60%"}
```{r geom-smooth-demo}
#| fig-format: png
#| fig-alt: "Scatter plot of leaf mass against petiole length with a linear trend line and shaded confidence ribbon added."
# geom_smooth() fits a line straight from the data -----
leaf_df %>%
  ggplot(aes(x = petiole_mm, y = mass_g)) +
  geom_point(alpha = 0.6, color = "grey40") +
  geom_smooth(method = "lm", color = "tomato", fill = "tomato", alpha = 0.15) +
  labs(
    title = "Leaf Mass vs. Petiole Length, with Trend Line",
    x = "Petiole Length (mm)",
    y = "Leaf Mass (g)"
  ) +
  theme_minimal()
```
:::

:::: {.column width="40%"}
- `method = "lm"` fits a straight line (linear model)
- The shaded ribbon is the **95% confidence band** around the line
- We'll fit this same relationship formally with `lm()` in the Regression unit later this semester

::: callout-tip
`geom_smooth()` is exploration, not a hypothesis test — it shows you *whether a line is worth fitting formally* before you commit to one.
:::
::::
::::::

# → ACTIVITY 4 Parts 1–2 now

::: activitybanner
**🛑 Do Activity Parts 1–2 now**

Load the data, map shape and color to `shade`, then add a `geom_smooth()` trend line. Predict, type, run.
:::

# 🧩 Chunk 2 of 3 · Zooming In Safely

**We will cover:** `coord_cartesian()` to zoom — and the `xlim()`/`ylim()` trap to avoid.

::: activitybanner
**🛑 After this chunk you will do Activity Part 3.**
:::

# `coord_cartesian()` — Zoom the View, Keep the Data

::::: columns
::: {.column width="60%"}
```{r coord-cartesian-demo}
#| fig-format: png
#| fig-alt: "Scatter plot of leaf mass against petiole length zoomed to a narrower window using coord_cartesian, with the same underlying data and trend line as the full plot."
# coord_cartesian() zooms the VIEW, keeps all the data -
leaf_df %>%
  ggplot(aes(x = petiole_mm, y = mass_g)) +
  geom_point(alpha = 0.6, color = "grey40") +
  geom_smooth(method = "lm", color = "tomato") +
  coord_cartesian(xlim = c(50, 70)) +
  labs(
    title = "Zoomed with coord_cartesian()",
    x = "Petiole Length (mm)",
    y = "Leaf Mass (g)"
  ) +
  theme_minimal()
```
:::

::: {.column width="40%"}
- Zooms the **view** — every point still contributes to the trend line
- The line is fit on **all 53 leaves**, then the window is cropped afterward
- This is the one to reach for whenever you just want a closer look
:::
:::::

# The Trap — `xlim()` / `ylim()` Delete Rows First

::: callout-note
**🔮 Predict first:** If we drop the leaves outside the window *before* fitting the trend line, will the line look the same, or different?
:::

:::::: columns
::: {.column width="60%"}
```{r xlim-demo}
#| fig-format: png
#| fig-alt: "Scatter plot of leaf mass against petiole length using xlim, showing a trend line refit only on the remaining visible points after out-of-range leaves were silently dropped."
# xlim() DELETES rows outside the range first ----------
leaf_df %>%
  ggplot(aes(x = petiole_mm, y = mass_g)) +
  geom_point(alpha = 0.6, color = "grey40") +
  geom_smooth(method = "lm", color = "tomato") +
  xlim(50, 70) +
  labs(
    title = "xlim() — trend line refit on fewer leaves",
    x = "Petiole Length (mm)",
    y = "Leaf Mass (g)"
  ) +
  theme_minimal()
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

`xlim()` / `ylim()` **silently delete every row** outside the range *before* anything is computed. The line above is fit only on the leaves that remain — a different line than `coord_cartesian()` gave, with no warning.
:::

**Rule:** zoom with `coord_cartesian()`. Only use `xlim()` / `ylim()` when you actually want those rows gone from the analysis.
::::
::::::

# → ACTIVITY 4 Part 3 now

::: activitybanner
**🛑 Do Activity Part 3 now**

Zoom with `coord_cartesian()`, then run the `xlim()` version once and watch the line move. Predict, type, run.
:::

# 🧩 Chunk 3 of 3 · Themes and Saving

**We will cover:** picking a built-in theme, `theme()` tweaks, and choosing a file format with `ggsave()`.

::: activitybanner
**🛑 After this chunk you will do Activity Parts 4–5.**
:::

# Picking a Theme

:::::: columns
::: {.column width="60%"}
```{r themes-demo}
#| fig-format: png
#| fig-alt: "Boxplot of leaf mass by shade rendered in theme_classic, with a white background and axis lines only."
# theme_classic() — clean, close to journal defaults ---
leaf_df %>%
  ggplot(aes(x = shade, y = mass_g, fill = shade)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  scale_fill_manual(
    values = c("sunny" = "goldenrod", "shady" = "forestgreen")
  ) +
  labs(x = "Shade", y = "Leaf Mass (g)", title = "Leaf Mass by Shade") +
  theme_classic() +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
**A few built-in themes worth knowing:**

| Theme             | Look                              |
|-------------------|-----------------------------------|
| `theme_grey()`    | the default — grey background     |
| `theme_minimal()` | no background, faint gridlines    |
| `theme_classic()` | white background, axis lines only |

::: callout-tip
`theme_classic()` is a safe default for scientific figures — clean, uncluttered, close to what most journals expect. Pick one you like and use it everywhere.
:::
::::
::::::

# Adjusting a Theme with `theme()`

:::::: columns
::: {.column width="60%"}
```{r theme-tweak-demo}
#| fig-format: png
#| fig-alt: "The same boxplot of leaf mass by shade, with the redundant fill legend removed via a theme adjustment."
# theme() overrides individual elements afterward -----
leaf_df %>%
  ggplot(aes(x = shade, y = mass_g, fill = shade)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  scale_fill_manual(
    values = c("sunny" = "goldenrod", "shady" = "forestgreen")
  ) +
  labs(x = "Shade", y = "Leaf Mass (g)", title = "Leaf Mass by Shade") +
  theme_classic() +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
- Pick a **built-in theme first** (`theme_classic()`), then add `theme()` afterward to tweak individual pieces
- `legend.position = "none"` — drop a legend that just repeats the x-axis, as `fill = shade` does here
- Other common tweaks: `plot.title = element_text(face = "bold")`, `axis.text = element_text(size = 12)`

::: callout-note
**✅ Key idea**

You rarely need to build a theme from scratch. Start from a built-in theme and override only what looks wrong.
:::
::::
::::::

# Saving in the Right Format

:::::: columns
::: {.column width="60%"}
```{r ggsave-formats}
#| eval: false
leaf_plot <- leaf_df %>%
  ggplot(aes(x = shade, y = mass_g, fill = shade)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  scale_fill_manual(
    values = c("sunny" = "goldenrod", "shady" = "forestgreen")
  ) +
  labs(x = "Shade", y = "Leaf Mass (g)") +
  theme_classic() +
  theme(legend.position = "none")

# PNG — good for Word docs, slides, web
ggsave(
  "figures/leaf_mass_theme.png",
  plot = leaf_plot,
  width = 5,
  height = 5,
  units = "in",
  dpi = 300
)

# PDF — vector, scales to any size, good for print
ggsave(
  "figures/leaf_mass_theme.pdf",
  plot = leaf_plot,
  width = 5,
  height = 5,
  units = "in"
)
```
:::

:::: {.column width="40%"}
| Format  | Best for               | Scales?         |
|---------|------------------------|-----------------|
| `.png`  | Word docs, slides, web | ❌ fixed pixels |
| `.pdf`  | Print, journals        | ✅ infinite     |
| `.tiff` | Journal submission     | ❌ fixed pixels |

::: callout-warning
**⚠️ Watch out!** `ggsave()` takes the **filename first**, then `plot =`. Always name the plot object explicitly — do not rely on ggsave grabbing "the last plot shown."
:::
::::
::::::

# → ACTIVITY 4 Parts 4–5 now

::: activitybanner
**🛑 Go to Activity 4 — Parts 4–5**

Close the slides. Pick a theme, tweak it, and save your plot as both PNG and PDF.
:::

# What We Learned Today

::::: columns
::: {.column width="60%"}
- Map `shape` alongside `color` to merge legends
- `geom_smooth(method = "lm")` — a quick trend line, not a hypothesis test
- `coord_cartesian()` zooms safely; `xlim()`/`ylim()` silently deletes rows first
- Pick one built-in theme (`theme_classic()`) and use `theme(legend.position = "none")` to drop a redundant legend
- `.png` for documents/slides, `.pdf` for print/journals
:::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 2 — *Displaying Data*
- 📖 [R4DS Ch 1](https://r4ds.hadley.nz/data-visualize) — Data Visualization

**Up next — Lecture 05, Wrangling:**

- `filter()`, `select()`, `mutate()`, `arrange()`
- Chaining verbs into one pipeline
:::
:::::
