---
title: "Lecture 04 — GGPlot II"
subtitle: "More geoms, zooming safely, and picking a theme"
author: "Bill Perry"
date: today

order: 4
week: 2
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Mapping shape alongside color, adding a trend line with geom_smooth(),
  coord_cartesian() vs. xlim()/ylim(), and choosing a built-in theme.
categories: [ggplot, tidyverse]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---



```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 03 — GGPlot I)

- Piping straight into a plot: `tree_df %>% ggplot(...)`
- `scale_color_manual()` / `scale_fill_manual()` — colors you choose
- `facet_wrap()` and `facet_grid()` — one panel per group
- `stat_summary()` — mean and mean ± SE, drawn from raw data

::: callout-note
**✅ Key idea from Lecture 03**

You can already build a grouped, colored, faceted figure. Today: one more geom, a safer way to zoom, and picking a theme that looks intentional instead of default.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- Map **shape** alongside color, and add a trend line with `geom_smooth()`
- Zoom into a plot **without silently dropping data** — `coord_cartesian()`
- Pick a built-in **theme** and adjust it with `theme()`
- Save a figure in the right **format** for the job

::: callout-tip
**🖐 Try it yourself**

By the end you will recognize the single most common ggplot mistake — zooming with `xlim()`/`ylim()` — before you ever make it yourself.
:::
::::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 2 — *Displaying Data*
- 📖 [R4DS Ch 1](https://r4ds.hadley.nz/data-visualize) — Data Visualization

Both PDFs are in the course `readings/` folder.
:::
::::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This lecture runs in **three short chunks**. After each chunk you switch to the activity worksheet and type the code yourself.
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

`xlim()` and `coord_cartesian()` produce plots that look *identical* for simple scatter plots — the difference only shows up once you compute something (a smooth line, a boxplot). Typing both yourself and comparing outputs is the only way this sticks.
:::
::::
::::::

# Load Libraries and Data

```{r load-libs-04}
# Load all packages at the top of every script ----------
library(readxl) # reading Excel files
library(tidyverse) # data wrangling + ggplot2
library(janitor) # cleans up messy column names
```

```{r load-data-04}
# Read the leaf data from the data folder ---------------
tree_df <- read_excel("data/2026_06_25_tree_experiment_raw_data.xlsx") %>%
  clean_names()
```

# 🧩 Chunk 1 of 3 · Shape, Color, and a Trend Line

**We will cover:** mapping shape alongside color, and `geom_smooth()`.

::: {.activitybanner}
**🛑 After this chunk you will do Activity Part 1.**
:::

# Mapping Shape Alongside Color

::: callout-note
**🔮 Predict first:** We map both `color = side` and `shape = side` to the *same* variable. Will ggplot draw one legend or two?
:::

:::::: columns
::: {.column width="60%"}
```{r shape-color-demo}
#| fig-format: png
#| fig-alt: "Scatter plot of leaf weight against leaf width, with sunny leaves shown as gold triangles and shady leaves as green circles."
# Map color AND shape to the same variable -------------
tree_df %>%
  ggplot(aes(x = width_cm, y = weight_g, color = side, shape = side)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("sunny" = "goldenrod", "shady" = "forestgreen")) +
  labs(
    title = "Leaf Weight vs. Width",
    x = "Width (cm)", y = "Weight (g)",
    color = "Side", shape = "Side"
  ) +
  theme_minimal()
```
:::

:::: {.column width="40%"}
- Mapping the **same variable** to both `color` and `shape` merges them into **one legend**
- This makes the plot easier to read in black-and-white or for colorblind readers
- If `color` and `shape` legend titles differ in `labs()`, you get **two** legends instead — usually not what you want

📖 [R4DS §9 — Layers](https://r4ds.hadley.nz/layers)
::::
::::::

# Adding a Trend Line — `geom_smooth()`

::::: columns
::: {.column width="60%"}
```{r geom-smooth-demo}
#| fig-format: png
#| fig-alt: "Scatter plot of leaf weight against width with a linear trend line and shaded confidence ribbon added."
# geom_smooth() fits a line straight from the data -----
tree_df %>%
  ggplot(aes(x = width_cm, y = weight_g)) +
  geom_point(alpha = 0.6, color = "grey40") +
  geom_smooth(method = "lm", color = "tomato", fill = "tomato", alpha = 0.15) +
  labs(
    title = "Leaf Weight vs. Width, with Trend Line",
    x = "Width (cm)", y = "Weight (g)"
  ) +
  theme_minimal()
```
:::

::: {.column width="40%"}
- `method = "lm"` fits a straight line (linear model)
- The shaded ribbon is the **95% confidence band** around the line
- We'll fit this same relationship formally with `lm()` in the Regression unit later this semester

::: callout-tip
`geom_smooth()` is exploration, not a hypothesis test — it shows you *whether a line is worth fitting formally* before you commit to one.
:::
::::
::::::

# → ACTIVITY 4 Part 1 now

::: {.activitybanner}
**🛑 Do Activity Part 1 now**

Map shape and color to `side`, then add a `geom_smooth()` trend line. Predict, type, run.
:::

# 🧩 Chunk 2 of 3 · Zooming Without Losing Data

**We will cover:** `coord_cartesian()` vs. `xlim()`/`ylim()` — they look the same until they don't.

::: {.activitybanner}
**🛑 After this chunk you will do Activity Part 2.**
:::

# `coord_cartesian()` — Zoom Without Removing Data

::::: columns
::: {.column width="60%"}
```{r coord-cartesian-demo}
#| fig-format: png
#| fig-alt: "Scatter plot of leaf weight against width zoomed to a narrower window using coord_cartesian, with the same underlying data and trend line as the full plot."
# coord_cartesian() zooms the VIEW, keeps all the data -
tree_df %>%
  ggplot(aes(x = width_cm, y = weight_g)) +
  geom_point(alpha = 0.6, color = "grey40") +
  geom_smooth(method = "lm", color = "tomato", fill = "tomato", alpha = 0.15) +
  coord_cartesian(xlim = c(12, 15), ylim = c(2, 5)) +
  labs(
    title = "Zoomed with coord_cartesian()",
    x = "Width (cm)", y = "Weight (g)"
  ) +
  theme_minimal()
```
:::

:::: {.column width="40%"}
- Zooms the **view** — every point still contributes to the trend line
- The trend line is fit on **all 20 leaves**, then the plot window is cropped afterward
- Safe to use anytime you just want a closer look
::::
::::::

# `xlim()` / `ylim()` — Removes Data First

::: callout-note
**🔮 Predict first:** If we drop the leaves outside our zoom window *before* fitting the trend line, will the line look the same as before, or different?
:::

:::::: columns
::: {.column width="60%"}
```{r xlim-demo}
#| fig-format: png
#| fig-alt: "Scatter plot of leaf weight against width using xlim and ylim, showing a trend line refit only on the remaining visible points after out-of-range leaves were silently dropped."
# xlim()/ylim() DELETE rows outside the range first ----
tree_df %>%
  ggplot(aes(x = width_cm, y = weight_g)) +
  geom_point(alpha = 0.6, color = "grey40") +
  geom_smooth(method = "lm", color = "tomato", fill = "tomato", alpha = 0.15) +
  xlim(12, 15) +
  ylim(2, 5) +
  labs(
    title = "Zoomed with xlim() / ylim() — different line!",
    x = "Width (cm)", y = "Weight (g)"
  ) +
  theme_minimal()
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

`xlim()`/`ylim()` **silently delete rows** outside the range *before* anything is computed. The trend line above is fit only on the leaves that remain — a different line than `coord_cartesian()` produced, with no warning that data went missing.
:::

**Rule of thumb:** use `coord_cartesian()` to zoom. Only use `xlim()`/`ylim()` when you actually want those rows excluded from the analysis, not just out of view.
::::
::::::

# → ACTIVITY 4 Part 2 now

::: {.activitybanner}
**🛑 Do Activity Part 2 now**

Zoom the same plot two ways and compare the trend lines. Predict, type, run.
:::

# 🧩 Chunk 3 of 3 · Themes and Saving

**We will cover:** picking a built-in theme, `theme()` tweaks, and choosing a file format with `ggsave()`.

::: {.activitybanner}
**🛑 After this chunk you will do Activity Parts 3–4.**
:::

# Picking a Theme

::::: columns
::: {.column width="60%"}
```{r themes-demo}
#| fig-format: png
#| fig-alt: "Boxplot of leaf weight by side rendered in theme_classic, with a white background and axis lines only."
# theme_classic() — clean, close to journal defaults ---
tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  scale_fill_manual(values = c("sunny" = "goldenrod", "shady" = "forestgreen")) +
  labs(x = "Side of Tree", y = "Leaf Weight (g)", title = "Leaf Weight by Side") +
  theme_classic() +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
**Five built-in themes to try on the same plot:**

| Theme | Look |
|---|---|
| `theme_grey()` | default — grey background |
| `theme_bw()` | white background, grey gridlines |
| `theme_minimal()` | no background, subtle gridlines |
| `theme_classic()` | white background, axis lines only |
| `theme_light()` | light grey lines and border |

::: callout-tip
`theme_classic()` is a safe default for scientific figures — clean, uncluttered, close to what most journals expect.
:::
::::
::::::

# Adjusting a Theme with `theme()`

::::: columns
::: {.column width="60%"}
```{r theme-tweak-demo}
#| fig-format: png
#| fig-alt: "The same boxplot of leaf weight by side, with legend removed and larger axis text via theme adjustments."
# theme() overrides individual elements afterward -----
tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  scale_fill_manual(values = c("sunny" = "goldenrod", "shady" = "forestgreen")) +
  labs(x = "Side of Tree", y = "Leaf Weight (g)", title = "Leaf Weight by Side") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13)
  )
```
:::

:::: {.column width="40%"}
- Pick a **built-in theme first** (`theme_classic()`), then add `theme()` afterward to tweak individual pieces
- `legend.position = "none"` — drop a redundant legend
- `axis.text` / `axis.title` — control tick labels and axis titles separately

::: callout-note
**✅ Key idea**

You rarely need to build a theme from scratch. Start from a built-in theme and override only what looks wrong.
:::
::::
::::::

# Saving in the Right Format

::::: columns
::: {.column width="60%"}
```{r ggsave-formats}
#| eval: false
leaf_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  scale_fill_manual(values = c("sunny" = "goldenrod", "shady" = "forestgreen")) +
  labs(x = "Side of Tree", y = "Leaf Weight (g)") +
  theme_classic() +
  theme(legend.position = "none")

# PNG — good for Word docs, slides, web
ggsave("figures/leaf_weight_theme.png",
       plot = leaf_plot, width = 5, height = 5,
       units = "in", dpi = 300)

# PDF — vector, scales to any size, good for print
ggsave("figures/leaf_weight_theme.pdf",
       plot = leaf_plot, width = 5, height = 5, units = "in")
```
:::

:::: {.column width="40%"}
| Format | Best for | Scales? |
|---|---|---|
| `.png` | Word docs, slides, web | ❌ fixed pixels |
| `.pdf` | Print, journals | ✅ infinite |
| `.tiff` | Journal submission | ❌ fixed pixels |

::: callout-warning
**⚠️ Watch out!** `ggsave()` takes the **filename first**, then `plot =`. Always name the plot object explicitly — do not rely on ggsave grabbing "the last plot shown."
:::
::::
::::::

# → ACTIVITY 4 Parts 3–4 now

::: {.activitybanner}
**🛑 Go to Activity 4 — Parts 3–4**

Close the slides. Pick a theme, tweak it, and save your plot as both PNG and PDF.
:::

# What We Learned Today

::::: columns
::: {.column width="60%"}
- Map `shape` alongside `color` to merge legends
- `geom_smooth(method = "lm")` — a quick trend line, not a hypothesis test
- `coord_cartesian()` zooms safely; `xlim()`/`ylim()` silently deletes rows first
- Built-in themes (`theme_classic()`, etc.) + `theme()` for small overrides
- `.png` for documents/slides, `.pdf` for print/journals
:::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 2 — *Displaying Data*
- 📖 [R4DS Ch 1](https://r4ds.hadley.nz/data-visualize) — Data Visualization

**Up next — Lecture 05, Wrangling:**

- `filter()`, `select()`, `mutate()`, `arrange()`
- Chaining verbs into one pipeline
:::
:::::
