---
order: 6
week: 3
# =====================================================================
#  regression_analysis_report.qmd
#  THE "MARKDOWN WAY" of doing the SAME regression analysis.
#  This is the twin of regression_analysis_script.R — identical code,
#  but wrapped in writing so that pressing Render produces a finished
#  Word report with the numbers and figures already in place.
# =====================================================================
title: "Paper Calibration Curve: Predicting Leaf Area from Mass"
subtitle: "The same analysis as regression_analysis_script.R — but as a report"
author: "Your Name Here"
date: today

format:
  docx:
    toc: true
    number-sections: true
    fig-width: 6
    fig-height: 4
    fig-format: png
  html: default          # bonus: the same file also makes a web page

# Hide the code by default so the report reads like a report.
# Flip echo to true if you want a teaching handout that shows the code.
execute:
  echo: false
  warning: false
  message: false
---

```{r libraries}
# Load all packages in one chunk at the top ----------------------------
library(readxl)
library(tidyverse)
library(skimr)
library(broom)   # tidy() and glance() turn model objects into data frames
```

```{r load-data}
# Read the paper calibration data --------------------------------------
paper_df <- read_excel("data/paper_area_weights.xlsx")
```

<!--
HOW TO READ THIS WORKED EXAMPLE (active reading)
Don't just skim. Before you press Render, PREDICT what each computed value and
figure will show, then render and compare. Better still, retype a chunk or two by
hand. The PREDICT comments below mark good places to pause. None of these comment
lines appear in the rendered Word document.
-->

# Introduction

To estimate leaf surface area without a leaf-area meter, we built a
**calibration curve**. Paper of uniform stock has a constant area-per-gram, so
the **mass** of a paper square is a near-perfect linear predictor of its
**area**. Once we know that relationship, we can trace a leaf onto the same
paper, weigh the tracing, and convert its mass to area.

- **Explanatory variable (X):** paper mass (g) — easy to measure on a balance
- **Response variable (Y):** paper area (cm²) — what we want to predict

# Methods

We weighed 118 paper squares of known area (1–567 cm²) and fit a simple linear
regression, `area_cm2 ~ mass_g`, with `lm()`. We checked the four regression
assumptions using base R's diagnostic plots for a fitted model, with a
Shapiro–Wilk test on the residuals as a formal companion. All work was done in
R with the tidyverse; this report was written in Quarto so every number and
figure below is computed directly from the data.

# Results

```{r fit-model}
# Fit the least-squares regression line --------------------------------
paper_lm_model <- lm(area_cm2 ~ mass_g, data = paper_df)

# Values used by the figure and prediction code below (not in the text) -
a_intercept <- coef(paper_lm_model)[1]
b_slope <- coef(paper_lm_model)[2]
r2_val <- summary(paper_lm_model)$r.squared
```

<!-- 🔮 PREDICT before you Render: what F, p, and R² will the model summary show, and what slope (cm² per gram)? Write your guesses, then render and check. -->

Paper mass was an extremely strong predictor of paper area. The overall model
fit, coefficients, and *p*-values are shown in the two tables below.

```{r model-summary}
# Overall model fit: F-statistic, R-squared, p-value ------------------
glance(paper_lm_model) |>
  select(r.squared, statistic, df, df.residual, p.value) |>
  knitr::kable(digits = c(4, 1, 0, 0, 20),
               col.names = c("R-squared", "F", "df (model)",
                             "df (residual)", "p-value"),
               caption = "Table 1. Overall fit of the calibration regression.")
```

```{r coef-table}
# The calibration equation: intercept and slope ----------------------
tidy(paper_lm_model) |>
  knitr::kable(digits = c(0, 3, 3, 1, 20),
               col.names = c("Term", "Estimate", "Std. error",
                             "t", "p-value"),
               caption = "Table 2. Calibration equation coefficients. The `mass_g` estimate is the area gained per additional gram of paper; `(Intercept)` is the fitted area at zero mass.")
```

Read the calibration equation off Table 2: predicted area equals the
`(Intercept)` estimate plus the `mass_g` estimate times paper mass. In words,
every additional gram of paper corresponds to about the `mass_g` estimate in
cm² of area.

::: {.callout-note appearance="simple"}
Every number in Tables 1 and 2 is computed from the data when the document
renders — nothing was typed by hand. Change the data file and re-render, and the
tables update themselves. That is the difference from the `.R` script, where you
would copy each number into Word yourself.
:::

## The calibration curve

```{r calibration-plot}
#| fig-cap: "Figure 1. Paper area vs. mass with the least-squares line and 95% confidence band."
#| fig-alt: "Scatter plot of paper area against mass with a dark blue least-squares line and light blue 95 percent confidence band; the subtitle reports the fitted equation and R-squared. Points cluster tightly around the line, narrowing the band toward the middle of the mass range."
calibration_plot <- paper_df %>%
  ggplot(aes(x = mass_g, y = area_cm2)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "darkblue",
    fill = "lightblue"
  ) +
  labs(
    subtitle = paste0(
      "area = ",
      round(b_slope, 2),
      " × mass + ",
      round(a_intercept, 3),
      "  |  R² = ",
      round(r2_val, 4)
    ),
    x = "Paper Mass (g)",
    y = "Area (cm²)"
  ) +
  theme_minimal()

calibration_plot
```

## Assumption checks

```{r diagnostic-plots}
#| fig-cap: "Figure 2. Base R diagnostic plots for the fitted model — residuals vs. fitted, normal Q-Q, scale-location, and residuals vs. leverage. A random band with no funnel or curve in panel 1 supports linearity and equal variance."
#| fig-alt: "Four base R diagnostic plots for the calibration linear model in a two by two grid: residuals versus fitted values, a normal Q-Q plot of the residuals, a scale-location plot, and a residuals versus leverage plot."
par(mfrow = c(2, 2))
plot(paper_lm_model)
```

```{r shapiro}
# Shapiro-Wilk on the RESIDUALS (not the raw data)
shapiro.test(residuals(paper_lm_model)) |>
  tidy() |>
  knitr::kable(digits = c(3, 20, 0),
               col.names = c("W", "p-value", "Method"),
               caption = "Table 3. Shapiro–Wilk normality test on the regression residuals.")
```

The Shapiro–Wilk result is in Table 3. Read *W* and its *p*-value there, then
compare that verdict with the normal Q-Q panel of Figure 2.

## Predicting leaf area

<!-- 🔮 PREDICT before you Render: which tracing gives the larger predicted area — the 0.092 g sunny or the 0.138 g shady? Guess the two areas, then render. -->

```{r predict}
# Predict area for two example leaf tracings --------------------------
new_tracings <- tibble(
  tracing = c("sunny-side", "shady-side"),
  mass_g  = c(0.092, 0.138)
)

new_tracings |>
  mutate(predicted_area_cm2 = predict(paper_lm_model, newdata = new_tracings)) |>
  knitr::kable(digits = c(0, 3, 1),
               col.names = c("Tracing", "Mass (g)", "Predicted area (cm²)"),
               caption = "Table 4. Predicted leaf area for two example paper tracings, from the calibration equation.")
```

Table 4 applies the calibration equation to two example tracings. The heavier
shady-side tracing has the larger predicted area — consistent with our earlier
finding that shady leaves are larger.

# Discussion

*Write one short paragraph here.* Did the calibration fit well (look at R² and
Figure 1)? Did the residual plot and Shapiro–Wilk support the assumptions? Why
is working in **area** (cm²) more biologically meaningful than raw **mass** (g)
when comparing sunny and shady leaves? This is where you paste the results
paragraph you drafted in the worksheet.

# References

- Whitlock, M. & Schluter, D. *The Analysis of Biological Data* (2nd ed.), Ch. 17.
- Wickham, H. *R for Data Science (2nd ed.)* — <https://r4ds.hadley.nz/>
