---
title: "Lecture 11 — Linear Regression"
subtitle: "Using one measurement to predict another (Whitlock & Schluter Ch. 17)"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# These are the ONLY things you need to edit to move this lecture around.
# The folder name and filename never have to change.
order: 11
week: 6
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Fitting a linear regression with lm(), reading the summary() output,
  checking the residuals, and using predict() to turn a mass into an area.
categories: [tidyverse, descriptive-stats]

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(width = 65) # printed R output wraps at 65 too
```

```{r hidden-libs}
#| echo: false
library(readxl)
library(tidyverse)
library(janitor)
library(patchwork)
```

# Where we left off (Lectures 09–10)

- **Five steps** — hypotheses → check normality → check variance → run the test → interpret and report
- **Welch's t-test** compares the **means of two groups**
- **Pseudoreplication** — average up to the thing you actually sampled before you test
- **Our leaves** — mean leaf **mass** did not differ between the shady and sunny sides
- **Our sugar maples** — shady-side leaves were significantly **larger in area** than sunny-side leaves on the same tree

::: callout-note
**✅ Where today comes in**

- A t-test compares two **groups**.
- Today we ask a different question:
  - can one **number** predict another number?
  - we weighed our leaf tracings, but what we actually want is **leaf area in cm²**
:::

# Goals for Today

::::: columns
::: {.column width="60%"}
- **Understand** what a linear regression does: fit a straight line, $\hat{Y} = a + bX$
- **See** what a **residual** is, in a real plot
- **Fit** a model with `lm()`
- **Read and interpret** the `summary()` output — slope, *p*-value, R²
- **Check** the residuals the way R shows them to you
- **Use** `predict()` to turn a tracing mass into a leaf area
- **Report** the result in a standard sentence
:::

::: {.column width="40%"}
- **Tools today:**
  - `lm()` — fit the line
  - `summary()` — read the result
  - `plot(model)` — check the residuals
  - `predict()` — make predictions
  - `geom_smooth(method = "lm")` — draw the line
- **Reading:**
  - 📖 Whitlock & Schluter, Ch. 17 — *Regression*
    - §17.1 Linear regression
    - §17.3 Testing the slope
    - §17.4 R²
    - §17.5 Assumptions
- **Today's naming:**
  - models → `_model`
  - plots → `_plot`
:::
:::::

# 🧩 Chunk 1 of 3 · What Regression Does

- **We will cover:**

- the question regression answers,

- the equation $\hat{Y} = a + bX$,

- what a **residual** is,

- what **R²** tells you,

- and the assumptions.

- No code in this chunk — just the idea.

# What Is Linear Regression?

::::: columns
::: {.column width="58%"}
**Linear regression** fits the best straight line through a cloud of points so you can **predict** one number (Y) from another number (X).

| Role                | Variable                       | On the plot |
|---------------------|--------------------------------|-------------|
| **Explanatory (X)** | what you can measure easily    | x-axis      |
| **Response (Y)**    | what you actually want to know | y-axis      |

X is usually the **easy** measurement. Y is the **expensive** one. Regression gives you an equation that converts X into Y.

📖 W&S §17.1 p. 540–541
:::

::: {.column width="42%"}
**Regression vs. the t-test:**

| Test       | The question                            |
|------------|-----------------------------------------|
| t-test     | Do two **groups** have different means? |
| Regression | Does X **predict** Y?                   |

**Use regression when:**

- both variables are **numeric**
- X could reasonably predict Y
- you want an **equation**, not just a yes/no
:::
:::::

# A Real Example from the Textbook

::::: columns
::: {.column width="55%"}
```{r lion-scatter}
#| echo: false
#| fig-format: png
#| fig-height: 3.8
#| fig-alt: "Scatter plot of lion age in years against the proportion of the nose that is black, for 32 lions, with an upward-sloping least-squares line through the points. The points scatter noticeably around the line."
lion_df <- read_csv("data/ws_lion_noses.csv", show_col_types = FALSE)

ggplot(lion_df, aes(x = proportionBlack, y = ageInYears)) +
     geom_point(size = 2.2, color = "sienna") +
     geom_smooth(method = "lm", se = FALSE, color = "black") +
     labs(x = "Proportion of nose that is black", y = "Age (years)") +
     theme_minimal(base_size = 13)
```

**W&S Example 17.1 — "the lion's nose."** 32 male lions of known age.
:::

::: {.column width="45%"}
Aging a wild lion is hard — you have to have watched it since it was a cub.

But a male lion's nose **darkens as it ages**, and you can photograph a nose from a truck.

- **X** = proportion of the nose that is black — *easy*
- **Y** = age in years — *what you want*

Fit the line once on lions of known age, and from then on you can **estimate** a lion's age from a photograph.

📖 W&S §17.1, Example 17.1
:::
:::::

# Our Version of the Same Problem

::::: columns
::: {.column width="58%"}
**The problem:** we want **leaf area in cm²**, but we don't have a leaf-area meter. We have a balance.

**The fix — a calibration curve:**

1.  Cut paper squares of **known area** (1, 4, 16 … 567 cm²)
2.  **Weigh** each square
3.  Fit the line: **area \~ mass**
4.  Trace a leaf onto the **same paper** and weigh the tracing
5.  Put the tracing mass into the equation → **predicted leaf area**

**Why it works:** paper from one stock has a constant **area per gram**, so mass and area move together almost perfectly.
:::

::: {.column width="42%"}
| Known area (cm²) | Squares weighed |
|------------------|-----------------|
| 1                | 25              |
| 4                | 32              |
| 16               | 24              |
| 36               | 3               |
| 64               | 7               |
| 100              | 14              |
| 225–567          | 13              |

**118 paper squares in all.**

Remember `paper_mass_g` in our leaf data? That is the tracing mass. By the end of today it becomes an **area**.
:::
:::::

# The Regression Equation

::::: columns
::: {.column width="55%"}
$$\hat{Y} = a + bX$$

| Symbol    | Name          | What it means                         |
|-----------|---------------|---------------------------------------|
| $\hat{Y}$ | predicted Y   | what the line says Y should be        |
| *a*       | **intercept** | Y when X = 0 — where the line crosses |
| *b*       | **slope**     | how much Y changes per 1 unit of X    |

**Ours will read:**

$$\widehat{\text{area}} = a + b \times \text{mass}$$

📖 W&S §17.1 pp. 543–545
:::

::: {.column width="45%"}
**The slope is the useful one.**

> *b* = the extra cm² you get for each extra gram of paper

That is just the **area density** of the paper stock — and it is the number that does the converting.

**The intercept is a bookkeeping number.**

> *a* = the predicted area of a piece of paper weighing nothing

Physically that should be 0. A small value near zero is fine; it is measurement noise, not biology.
:::
:::::

# What Is a Residual?

::::: columns
::: {.column width="55%"}
```{r lion-residuals}
#| echo: false
#| fig-format: png
#| fig-height: 4
#| fig-alt: "The same lion scatter plot with the least-squares line, plus a vertical red segment drawn from every point to the line. Segments above the line are longer for a few lions; most are short. The plot shows the residual as the vertical gap between the observed point and the line."
lion_fit_df <- lion_df %>%
  mutate(predicted = fitted(lm(ageInYears ~ proportionBlack, data = lion_df)))

ggplot(lion_fit_df, aes(x = proportionBlack, y = ageInYears)) +
  geom_segment(
    aes(xend = proportionBlack, yend = predicted),
    color = "firebrick",
    linewidth = 0.7
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  geom_point(size = 2.2, color = "sienna") +
  labs(x = "Proportion of nose that is black", y = "Age (years)") +
  theme_minimal(base_size = 13)
```

**Red = the residuals** for the lion data.
:::

::: {.column width="45%"}
A **residual** is one point's miss:

$$\text{residual} = \text{observed} - \text{predicted}$$

- A point **above** the line → **positive** residual (that lion was older than its nose suggested)
- A point **below** the line → **negative** residual
- A point **on** the line → residual of 0

**The line R gives you is the one that makes those red segments as small as possible overall.** That is all "least squares" means — and R does the work.

📖 W&S §17.1 p. 542
:::
:::::

# R² — How Tight Is the Fit?

::::: columns
::: {.column width="62%"}
```{r r2-comparison}
#| echo: false
#| fig-format: png
#| fig-width: 7
#| fig-height: 3.6
#| fig-alt: "Two scatter plots side by side. Left: lion age against nose blackness with a fitted line and clearly scattered points, labelled R-squared equals 0.62. Right: paper area against paper mass with a fitted line and points sitting almost exactly on it, labelled R-squared equals 0.9999."
paper_demo_df <- read_excel("data/paper_area_weights.xlsx")

lion_panel <- ggplot(lion_df, aes(x = proportionBlack, y = ageInYears)) +
     geom_point(size = 2, color = "sienna") +
     geom_smooth(method = "lm", se = FALSE, color = "black") +
     labs(
          subtitle = "Lions:  R² = 0.62",
          x = "Proportion black",
          y = "Age (years)"
     ) +
     theme_minimal(base_size = 12)

paper_panel <- ggplot(paper_demo_df, aes(x = mass_g, y = area_cm2)) +
     geom_point(size = 2, color = "blue4") +
     geom_smooth(method = "lm", se = FALSE, color = "black") +
     labs(subtitle = "Paper:  R² = 0.9999", x = "Mass (g)", y = "Area (cm²)") +
     theme_minimal(base_size = 12)

lion_panel + paper_panel
```
:::

::: {.column width="38%"}
**R² runs from 0 to 1.** It is the fraction of the variation in Y that the line accounts for.

| R²   | Reads as                            |
|------|-------------------------------------|
| 1.00 | points sit **on** the line          |
| 0.62 | X explains **62%** of the variation |
| 0.00 | the line tells you nothing          |

**Both of these are good regressions.** Nose blackness really does predict a lion's age — just with scatter, because lions vary. Paper is manufactured to be uniform, so it has almost none.

In biology, **R² of 0.3–0.7 is normal and useful**. Don't expect 0.99.

📖 W&S §17.4 p. 555
:::
:::::

# R² and the *p*-value Answer Different Questions

:::::: columns
::: {.column width="52%"}
|   | The question it answers |
|-------------------------|-----------------------------------------------|
| ***p*****-value** | Is there **really** a relationship, or could this slope be chance? |
| **R²** | **How tightly** do the points follow the line? |

You can have a **significant slope with a low R²** — a real relationship with a lot of scatter. That is most of ecology.
:::

:::: {.column width="48%"}
::: callout-important
**The slope is the hypothesis test.**

**H₀:** the slope is 0 — X does not predict Y

**Hₐ:** the slope is not 0 — X does predict Y

`summary()` gives you the *p*-value for exactly that test, and we will find it on the next few slides.
:::

📖 W&S §17.3 pp. 551–553
::::
::::::

# Assumptions of Regression

:::::: columns
::: {.column width="58%"}
Four things need to be roughly true for the *p*-value and the intervals to mean anything (W&S §17.5):

| \# | Assumption | How you check it |
|------------------|-----------------------------|-------------------------|
| 1 | **Linear** — Y really does follow a straight line | the scatter plot, then `plot(model)` panel 1 |
| 2 | **Independent** — one observation doesn't depend on another | your study design |
| 3 | **Equal variance** — the scatter is about the same all the way along | `plot(model)` panels 1 and 3 |
| 4 | **Normal residuals** — the misses are bell-shaped | `plot(model)` panel 2 |
:::

:::: {.column width="42%"}
::: callout-important
**Check the residuals, not the raw data.**

Normality here is about the **misses** — the red segments — not about Y itself. Y is *supposed* to go up with X; of course it isn't bell-shaped.

The good news: once you have a fitted model, `plot()` draws all of these for you. No extra code.
:::

Same habit as the t-test: **check → decide → report.**

📖 W&S §17.5 pp. 557–560
::::
::::::

# 🧩 Chunk 2 of 3 · Fit It and Read It

**We will cover:** loading the paper data, the scatter plot, fitting with `lm()`, and reading the `summary()` output line by line.

::: activitybanner
**🛑 After this chunk: Activity Parts 1–4** (load, plot, fit, read the summary).
:::

# Load Libraries and Data

```{r load-libs-11}
# Load packages at the top ------------------------------
library(readxl) # read Excel files
library(tidyverse) # dplyr + ggplot2
library(janitor) # clean_names()
```

```{r load-data-11}
# Load the paper calibration data ----------------------
paper_df <- read_excel("data/paper_area_weights.xlsx")
```

Two columns, 118 rows: `area_cm2` is the **known** area of each square, `mass_g` is what it weighed.

# Always Plot It First

:::::: columns
::: {.column width="55%"}
```{r scatter-11}
#| fig-format: png
#| fig-alt: "Scatter plot of paper area in square centimetres against paper mass in grams. The points form a tight straight line rising from the bottom left to the top right."
# Scatter plot of area against mass --------------------
scatter_plot <- paper_df %>%
  ggplot(aes(x = mass_g, y = area_cm2)) +
  geom_point() +
  geom_smooth(method = "lm")
scatter_plot
```
:::

:::: {.column width="45%"}
**Three plain lines.** `aes(x = , y = )` puts the explanatory variable on x and the response on y, and `geom_point()` draws the dots.

**What you are looking for:**

- **Straight?** → a line is the right tool
- **Scatter the same all the way along?**
- **Any point way off on its own?**

Here: straight, tight, nothing strange. Fit the line.

::: callout-note
Never fit a regression you haven't looked at. A curve, a funnel, or one wild point all change what you should do next.
:::
::::
::::::

# Fit the Model with `lm()`

:::::: columns
::: {.column width="55%"}
```{r fit-lm-11}
# Fit the regression line ------------------------------
paper_lm_model <- lm(area_cm2 ~ mass_g, data = paper_df)
```

Nothing prints. `lm()` hands back a **model object** that is holding the answer — `summary()` is what shows it to you.
:::

:::: {.column width="45%"}
**The formula is the whole trick:**

``` r
lm(Y ~ X, data = df)
```

- **Y** — the response — goes on the **left** of `~`
- **X** — the explanatory variable — goes on the **right**
- `~` reads as **"is predicted by"**

So `area_cm2 ~ mass_g` is "area is predicted by mass."

::: callout-warning
**⚠️ Get the order wrong and R will still answer.** `lm(mass_g ~ area_cm2)` runs fine — it just fits the question you didn't ask.
:::
::::
::::::

# `summary()` — the Whole Output

```{r summary-lm-11}
#| output-location: column
# Read the model ---------------------------------------
summary(paper_lm_model)
```

There are only **four numbers** in here that you need, and we will take them one at a time:

1.  the **slope** (`mass_g` → `Estimate`)
2.  the **intercept** (`(Intercept)` → `Estimate`)
3.  the slope's ***p*****-value** (`mass_g` → `Pr(>|t|)`)
4.  the **R²** (`Multiple R-squared`)

# Reading It — 1. The Coefficients Table

::::: columns
::: {.column width="52%"}
```         
Coefficients:
            Estimate Std. Error t value Pr(>|t|)
(Intercept)   0.2792     0.1812   1.541    0.126
mass_g      130.5234     0.1473 885.964   <2e-16
```

**The `Estimate` column is the equation.**

- `(Intercept)` = **0.28** → that is *a*
- `mass_g` = **130.52** → that is *b*, the slope

So the fitted line is:

$$\widehat{\text{area}} = 0.28 + 130.52 \times \text{mass}$$
:::

::: {.column width="48%"}
**Say the slope out loud, with units:**

> Every extra **1 gram** of this paper is another **130.5 cm²** of area.

That is the conversion factor we came for.

**The other three columns, in one line each:**

| Column       | What it is                         |
|--------------|------------------------------------|
| `Std. Error` | how uncertain that estimate is     |
| `t value`    | Estimate ÷ Std. Error              |
| `Pr(>|t|)`   | the ***p*****-value** for that row |

📖 W&S §17.3 p. 551
:::
:::::

# Reading It — 2. The Two *p*-Values

::::::: columns
::: {.column width="52%"}
```         
(Intercept)   0.2792   ...   0.126
mass_g      130.5234   ...  <2e-16
```

**The slope row is the one that matters.** `Pr(>|t|)` for `mass_g` is `<2e-16` — R's way of writing "smaller than I can print."

- **H₀:** the slope is 0 → mass tells you nothing about area
- *p* \< 0.001 → **reject H₀**. Mass predicts area.

**The intercept row:** *p* = 0.126, so the intercept is **not** distinguishable from 0 — exactly what physics says it should be. A piece of paper with no mass has no area.
:::

::::: {.column width="48%"}
::: callout-tip
**Which *p*-value do I report?**

The **slope's**. That is the biological question. Nobody reports the intercept's *p*-value unless zero means something special.
:::

::: callout-note
**Never write *p* = 0.** When R prints `<2e-16`, you write ***p*** **\< 0.001**.
:::
:::::
:::::::

# Reading It — 3. R² and the Bottom Line

:::::: columns
::: {.column width="52%"}
```         
Residual standard error: 1.764 on 116 degrees of freedom
Multiple R-squared:  0.9999, Adjusted R-squared:  0.9999
F-statistic: 7.849e+05 on 1 and 116 DF,  p-value: < 2.2e-16
```

- **Multiple R-squared = 0.9999** → mass explains **99.99%** of the variation in area. That is manufactured paper, not biology.
- **F-statistic** with **1 and 116 DF**, *p* \< 2.2e-16 → the overall test of the model. With one X it tells you the same thing as the slope's *p*-value.
- **Residual standard error = 1.76 cm²** → a typical miss is about 1.8 cm².
:::

:::: {.column width="48%"}
**Where the df come from:**

- **1** = one explanatory variable
- **116** = 118 squares − 2 numbers we estimated (*a* and *b*)

**Use `Adjusted R-squared`** once you have more than one X; with a single X, report `Multiple R-squared`.

::: callout-important
**R² = 0.9999 does not mean "everything is fine."** It means the points are close to the line. Whether the *assumptions* hold is a separate question — and we check it two slides from now.
:::
::::
::::::

# → ACTIVITY 11 Parts 1–4 now

::: activitybanner
**🛑 Do Activity Parts 1–4 now**

Load the paper data, make the scatter plot, fit `lm(area_cm2 ~ mass_g)`, and pull the slope, the *p*-value, and R² **off the summary output** — writing them in your answer boxes.
:::

# 🧩 Chunk 3 of 3 · Plot It, Check It, Use It

**We will cover:** drawing the fitted line, reading `plot(model)`, using `predict()` on one tracing and then on a whole column, and writing the result up.

::: activitybanner
**🛑 After this chunk: Activity Parts 5–9** (plot the line, check the residuals, predict, report).
:::

# Draw the Line

:::::: columns
::: {.column width="55%"}
```{r calibration-plain-11}
#| fig-format: png
#| fig-alt: "Scatter plot of paper area against paper mass with a straight fitted regression line and a narrow grey confidence band through the points."
# Scatter plot with the fitted line --------------------
calibration_plot <- paper_df %>%
  ggplot(aes(x = mass_g, y = area_cm2)) +
  geom_point() +
  geom_smooth(method = "lm")
calibration_plot
```
:::

:::: {.column width="45%"}
**One extra line.** `geom_smooth(method = "lm")` fits the same regression again, just for drawing, and adds:

- the **fitted line**
- a grey **95% confidence band** around it

**The band is narrowest in the middle** of the data and flares at the ends — the line is pinned down best where you have the most data. Here it is so narrow you can barely see it.

::: callout-note
`geom_smooth()` defaults to a wiggly curve. **`method = "lm"` is what makes it a straight regression line.**
:::
::::
::::::

# The Same Plot, Dressed Up

:::::: columns
::: {.column width="55%"}
``` r
# The publication version — for reference ---------------
calibration_plot <- paper_df %>%
  ggplot(aes(x = mass_g, y = area_cm2)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Paper mass (g)",
       y = "Paper area (cm²)") +
  theme_minimal()
calibration_plot
```

```{r calibration-dressed-11}
#| echo: false
#| fig-format: png
#| fig-height: 3.2
#| fig-alt: "The same paper calibration scatter plot with the fitted line, now with axis labels reading Paper mass in grams and Paper area in square centimetres, on a minimal theme."
paper_df %>%
     ggplot(aes(x = mass_g, y = area_cm2)) +
     geom_point() +
     geom_smooth(method = "lm") +
     labs(x = "Paper mass (g)", y = "Paper area (cm²)") +
     theme_minimal()
```
:::

:::: {.column width="45%"}
**This is the version you'd put in a report** — and it is only two additions to the plot you already made:

- `labs()` — real axis labels with units
- `theme_minimal()` — drop the grey background

::: callout-tip
**Don't type this today.** Get the plain `geom_point() + geom_smooth(method = "lm")` working first. Labels and themes are the last thing you add, once you know the plot is right.
:::
::::
::::::

# Check the Residuals — `plot(model)`

::::: columns
::: {.column width="50%"}
```{r diagnostic-plots-11}
#| fig-format: png
#| fig-width: 5.5
#| fig-height: 5
#| fig-alt: "Four base R diagnostic plots for the paper calibration model in a two by two grid: residuals versus fitted values showing a clear funnel that widens to the right, a normal Q-Q plot with points curving away at both ends, a scale-location plot rising to the right, and residuals versus leverage."
# All four residual plots in one call ------------------
par(mfrow = c(2, 2))
plot(paper_lm_model)
```
:::

::: {.column width="50%"}
`par(mfrow = c(2, 2))` just says "put the next four plots in a 2×2 grid."

| Panel | Name                  | Checks                         |
|-------|-----------------------|--------------------------------|
| 1     | Residuals vs Fitted   | linear? equal scatter?         |
| 2     | Normal Q-Q            | residuals bell-shaped?         |
| 3     | Scale-Location        | equal scatter again            |
| 4     | Residuals vs Leverage | any one point running the show |

**Panel 1 is the one to look at first.**

- ✅ **Good:** a flat, shapeless cloud around 0
- ❌ **Curved:** the relationship isn't straight
- ❌ **Funnel:** the scatter grows as Y grows
:::
:::::

# What Our Residual Plot Is Telling Us

:::::: columns
::: {.column width="52%"}
Look at panel 1 again. The points **fan out** — a 1 cm² square misses by hundredths of a cm², a 567 cm² square misses by several cm².

That is a **funnel**: assumption 3 (equal variance) does not hold.

```{r resid-by-size-11}
# How big are the misses at each square size? ---------
paper_df %>%
  mutate(residual = residuals(paper_lm_model)) %>%
  group_by(area_cm2) %>%
  summarize(biggest_miss = round(max(abs(residual)), 2))
```
:::

:::: {.column width="48%"}
::: callout-important
**A great R² and a bad residual plot can live together.** R² says the points are near the line. The residual plot says the *size of the miss* depends on how big the square is — which it does, because a balance has percentage error, not fixed error.
:::

**So what do we do?**

- For **predicting a leaf tracing** (0.1–0.4 g) we are down in the tight end of the line, where a typical miss is a fraction of a cm². The calibration is fine for the job.
- The **confidence intervals** at the big-square end are the part you shouldn't trust.
- The proper fix — fitting on a log scale, or weighting the points — is a **graduate** topic. Knowing to look, and saying what you saw, is the undergraduate job.
::::
::::::

# Predict One Tracing

:::::: columns
::: {.column width="55%"}
```{r predict-one-11}
# Predict the area of one leaf tracing ----------------
# The tracing weighed 0.138 g
one_tracing <- tibble(mass_g = 0.138)

predict(paper_lm_model, newdata = one_tracing)
```

An 0.138 g tracing is about **18.3 cm²** of leaf.
:::

:::: {.column width="45%"}
**`predict()` takes two things:**

- the **model** you fitted
- `newdata =` — a small data frame of the X values you want predictions for

::: callout-warning
**⚠️ The column in `newdata` must have the same name as in the model.** The model was built on `mass_g`, so `newdata` needs a column called `mass_g`. Call it `mass` and R stops with *"object 'mass_g' not found."*
:::

This beats doing the arithmetic yourself — no copying the slope, nothing to mistype.
::::
::::::

# Predict with an Interval

::::: columns
::: {.column width="55%"}
```{r predict-interval-11}
# Two tracings, with 95% prediction intervals ---------
two_tracings <- tibble(mass_g = c(0.092, 0.138))

predict(paper_lm_model, newdata = two_tracings, interval = "prediction")
```

The 0.138 g tracing is **18.3 cm²**, somewhere between about **14.8 and 21.8 cm²**.
:::

::: {.column width="45%"}
`interval = "prediction"` adds two columns:

- `fit` — the prediction
- `lwr` and `upr` — the 95% range

**Two kinds of interval:**

| `interval =`   | Range for                                       | Width    |
|---------------------|---------------------------------|------------------|
| `"confidence"` | the **average** area of all tracings this heavy | narrower |
| `"prediction"` | **one particular** tracing                      | wider    |

For a single real leaf, use **`"prediction"`**.

📖 W&S §17.2 pp. 548–550
:::
:::::

# Convert a Whole Column of Tracings

:::::: columns
::: {.column width="58%"}
```{r leaf-convert-11}
# Load our leaf data ----------------------------------
leaf_df <-
  read_excel("data/2026_09_03_data_sci_leaf_area.xlsx") %>%
  clean_names()

# Turn every tracing mass into an area ----------------
leaf_df <- leaf_df %>%
  mutate(
    leaf_area_cm2 = predict(
      paper_lm_model,
      newdata = tibble(mass_g = paper_mass_g)
    )
  )

leaf_df %>%
  select(teams, shade, mass_g, paper_mass_g, leaf_area_cm2)
```
:::

:::: {.column width="42%"}
**This is the payoff.** `paper_mass_g` was the mass of each leaf's tracing. Now every leaf has a **`leaf_area_cm2`**.

**The one fiddly bit** is the same rule as before: `predict()` wants a column called `mass_g`, so we hand it `tibble(mass_g = paper_mass_g)` — our tracing masses, under the name the model knows.

::: callout-note
**The `NA`s are honest.** One team never weighed their tracings, so those leaves have no area. `predict()` passes the missing value straight through instead of inventing a number.
:::
::::
::::::

# Was the Area Different Between Sides?

:::::: columns
::: {.column width="52%"}
```{r leaf-area-plot-11}
#| fig-format: png
#| fig-height: 3.4
#| fig-alt: "Box plot of predicted leaf area in square centimetres for shady and sunny leaves with the individual leaves jittered over the boxes. The two boxes overlap almost completely."
# Predicted leaf area by side -------------------------
leaf_area_plot <- leaf_df %>%
  ggplot(aes(x = shade, y = leaf_area_cm2)) +
  geom_boxplot() +
  geom_jitter(width = 0.15)
leaf_area_plot
```
:::

:::: {.column width="48%"}
```{r leaf-area-mean-11}
# Mean predicted area per side ------------------------
leaf_df %>%
  group_by(shade) %>%
  summarize(
    n = sum(!is.na(leaf_area_cm2)),
    mean_area = mean(leaf_area_cm2, na.rm = TRUE)
  )
```

About **37 cm²** on the shady side and **36 cm²** on the sunny side — the same story the t-test told us about mass, now in the units we actually wanted.

::: callout-note
**✅ Why this matters.** "Shady leaves weigh 0.02 g more" means little to a reader. "Shady leaves are about 1 cm² bigger" is a sentence an ecologist can use.
:::
::::
::::::

# How to Report a Regression

::::: columns
::: {.column width="52%"}
**The template — fill in from your `summary()` output:**

> *\[What predicted what\] (linear regression: F(df₁, df₂) = \_\_\_, p \_\_\_, R² = \_\_\_). The fitted equation was Y = \_\_\_ + \_\_\_ × X.*

**Ours:**

> *Paper mass was a very strong predictor of paper area (linear regression: F(1, 116) = 784,900, p \< 0.001, R² = 0.9999). The calibration equation was area = 0.28 + 130.52 × mass. Residual scatter increased with square size, so intervals for the largest squares should be treated with caution.*
:::

::: {.column width="48%"}
**Where every number comes from:**

| In the sentence | In `summary()`             |
|-----------------|----------------------------|
| F and both df   | the `F-statistic:` line    |
| *p*             | that same line             |
| R²              | `Multiple R-squared`       |
| intercept       | `(Intercept)` → `Estimate` |
| slope           | `mass_g` → `Estimate`      |

**Always include:** the test name, F with both df, *p*, R², and **the equation**. Add a line about the assumptions you checked.

📖 W&S §17.3 p. 553
:::
:::::

# → ACTIVITY 11 Parts 5–9 now

::: activitybanner
**🛑 Do Activity Parts 5–9 now**

- Draw the line and run `plot(model)`
- Predict one tracing, then convert the whole `paper_mass_g` column into areas
- Write your results sentence
- Your **answers** in the script are what you hand in
:::

# What We Learned Today

::::: columns
::: {.column width="58%"}
**The ideas:**

- Regression fits a line so you can **predict Y from X**
- $\hat{Y} = a + bX$ — the **slope** converts X into Y
- A **residual** is one point's miss: observed − predicted
- **R²** = how tight the fit is; ***p*** = whether the slope is real
- Check the **residuals** — and say what you saw

**The code:**

- `lm(Y ~ X, data = df)` — fit it
- `summary(model)` — slope, intercept, *p*, R²
- `geom_smooth(method = "lm")` — draw it
- `plot(model)` — check it
- `predict(model, newdata = ...)` — use it
:::

::: {.column width="42%"}
- **Hand in:** your finished `scripts/11_regression.R` — the code is given; your **answers** are graded

**📖 W&S Ch. 17:**

- §17.1 Linear regression · §17.2 Prediction
- §17.3 Testing the slope · §17.4 R²
- §17.5 Assumptions

**Up next — Lecture 12:**

- Pulling real climate data into R with `GSODR`
- Summarizing daily weather into yearly values
- The **same `lm()`** — temperature predicted by date
  - the slope becomes a rate of change per year
  - and you write that regression yourself
:::
:::::
