---
title: "Lecture 11 — Linear Regression"
subtitle: "Predicting leaf area from paper tracing mass (Whitlock & Schluter Ch. 17)"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# These are the ONLY things you need to edit to move this lecture around.
# The folder name and filename never have to change.
order: 11
week: 6
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  How to fit a linear regression, test its assumptions, interpret the model output, and calculate predictions.
categories: [tidyverse, descriptive-stats]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lectures 07, 09–10)

- **Quarto** — one `.qmd` file holds your **writing *and* your code**, and renders to Word, HTML, or PowerPoint
- **Three ingredients** — YAML front matter, Markdown text, and named code chunks
- **Reproducibility** — numbers and figures are *computed*, never copied by hand; change the data, press **Render**
- **Inline code** — drop a live result straight into a sentence with `` `r ` ``
- **Word reports** — `format: docx` with `toc` and `number-sections` gives a professional document
- **Carried forward (Lectures 09–10)** — shady leaves were significantly *heavier* than sunny leaves (Welch's t-test, *p* \< 0.001)

::: callout-note
**✅ Transition**

You can now build a reproducible report. Today's analysis — a **calibration curve** that converts leaf weight into a more meaningful measurement, **leaf surface area** (cm²) — is the perfect thing to write up in one. The activity gives you this same analysis as *both* a plain `.R` script **and** a Quarto report, so you can feel the difference.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- Understand what linear regression does and when to use it
- Learn the equation: $\hat{Y} = a + bX$
  - slope *b*, intercept *a*, residuals, R²
- Understand **least squares** — what the line minimizes
- Fit a model in R with `lm()` and read every line of `summary()`
- **Apply it:** build a calibration curve to predict leaf area from paper tracing mass
- Check all four assumptions of regression
- Use `predict()` to make predictions with intervals
- Use `mutate()` to add model results to a data frame

::: callout-tip
**🖐 Try it yourself**

By the end you will predict leaf area from a tracing weight.
:::
::::

::: {.column width="40%"}
**Tools today:**

- `readxl`, `tidyverse`

**Textbook:**

- 📖 Whitlock & Schluter *Analysis of Biological Data* 2nd ed.
- **Chapter 17 — Regression** (pp. 539–575)
  - §17.1 Linear regression
  - §17.3 Testing slope hypotheses
  - §17.4 R²
  - §17.5 Assumptions

**Naming conventions:**

- models → `_model`
- plots → `_plot`
:::
::::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This lecture runs in **four short chunks**. After each chunk you switch to the **activity** and type the code yourself.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother? (the evidence)**

-   A guess about the sign of the slope, or the size of R², is a claim you can be proven wrong about — and being wrong is what makes the right answer stick.
-   Typing `lm(area_cm2 ~ mass_g, data = paper_df)` yourself, instead of pasting it, is what makes you actually notice which variable sits on which side of the `~`.
-   Running each short chunk and then immediately applying it in the activity keeps the idea active in your head instead of letting it go stale between lecture and practice.
:::
::::
::::::

# 🧩 Chunk 1 of 4 · What Regression Is

**We will cover:** the calibration idea, the equation $\hat{Y}=a+bX$, least squares, R², and the four assumptions.

::: callout-tip
**🖐 After this chunk: Activity Parts 1–3** (load, inspect, and scatter-plot the paper data).
:::

# Our Calibration Idea

::::: columns
::: {.column width="60%"}
**The problem:** measuring leaf surface area without a leaf area meter.

**The solution — a calibration curve:**

1.  Cut paper squares of **known area** (1, 4, 16 … 567 cm²)
2.  **Weigh** each square on a precision balance
3.  Fit a **regression line**: area \~ mass
4.  Trace a leaf on the *same paper*, weigh the tracing
5.  **Plug the tracing mass into the equation** → predicted leaf area

**Why does this work?**

Paper of uniform stock has constant *area density* (cm² per gram). Mass and area are therefore **perfectly linearly related** within a sheet.
:::

::: {.column width="40%"}
| Known area (cm²) | n replicates |
|------------------|--------------|
| 1                | 25           |
| 4                | 32           |
| 16               | 24           |
| 36               | 3            |
| 64               | 7            |
| 100              | 14           |
| 225–567          | 12           |

**Total:** 118 paper squares weighed

📖 W&S §17.1, Example 17.1 (the same logic: measure X to predict Y)
:::
:::::

# What Is Linear Regression?

::::: columns
::: {.column width="60%"}
**Linear regression** draws the *best straight line* through a scatter of points to **predict** a response variable (Y) from an explanatory variable (X).

**Two key roles:**

| Role            | Variable                                | Our example |
|-----------------|-----------------------------------------|-------------|
| Explanatory (X) | on the x-axis; what you measure         | `mass_g`    |
| Response (Y)    | on the y-axis; what you want to predict | `area_cm2`  |

X is often something easy to measure; Y is what you actually want to know. Regression gives you the equation to convert X → Y.

📖 W&S §17.1 p. 540–541
:::

::: {.column width="40%"}
**When to use regression:**

- You have **two numeric variables**
- One variable can reasonably **predict or cause** the other
- You want an **equation** to make predictions for new observations

**Regression vs. our t-test (Lectures 09–10):**

| Test       | Question                        |
|------------|---------------------------------|
| t-test     | Do two group *means* differ?    |
| Regression | Does X *predict* Y numerically? |

Regression produces a **predictive equation**. The t-test produces a decision about group means.
:::
:::::

# The Regression Equation

::::: columns
::: {.column width="60%"}
**Population model** (W&S §17.1):

$$Y = \alpha + \beta X + \varepsilon$$

**Sample estimate** (what R calculates):

$$\hat{Y} = a + bX$$

| Symbol        | Name        | Meaning                              |
|---------------|-------------|--------------------------------------|
| $\hat{Y}$     | predicted Y | estimated mean area at a given mass  |
| *a*           | intercept   | predicted Y when X = 0               |
| *b*           | slope       | change in Y per 1-unit increase in X |
| $\varepsilon$ | residual    | observed − predicted                 |

**Our equation will be:** $$\widehat{\text{area}} = a + b \times \text{mass\_g}$$
:::

::: {.column width="40%"}
**Interpreting the slope:**

> *b* = the change in area (cm²) for every additional 1 gram of paper

Since 1 g ≈ 130 cm² of paper, the slope is the **area density** of the paper stock.

**Interpreting the intercept:**

> *a* = predicted area when mass = 0

Physically this should be 0 (no paper, no area). Small non-zero values reflect measurement error.

📖 W&S §17.1 pp. 543–545
:::
:::::

# Residuals and Least Squares

::::: columns
::: {.column width="60%"}
**Residual** for observation *i*:

$$e_i = Y_i - \hat{Y}_i = \text{observed} - \text{predicted}$$

The **least-squares line** minimizes the sum of *squared* residuals:

$$\text{minimize} \sum_{i=1}^{n}(Y_i - \hat{Y}_i)^2$$

- Squaring serves two purposes:
  - Makes all deviations **positive** (no cancellation)
  - Penalizes **large** deviations more than small ones
- There is only one line that achieves this minimum — it is unique.
- 📖 W&S §17.1 p. 542 (Figure 17.1-2)
:::

::: {.column width="40%"}
**Visual intuition:**

Each data point has a **vertical residual** — the distance from the point to the regression line.

```
observed Y │     ●  ← residual (positive)
           │     |
predicted Ŷ│-----●--- regression line
           │
           └──────────── X
```

A point *above* the line has a **positive** residual. A point *below* has a **negative** residual. The sum of all residuals = 0.
:::
:::::

# R² — How Much Does X Explain?

::: callout-note
**🔮 Predict first:** Paper of uniform stock has constant area-per-gram. Before you see any output, predict: will R² be near **0, 0.5, or 1**?
:::

::::: columns
::: {.column width="60%"}
$$R^2 = \frac{SS_\text{regression}}{SS_\text{total}} = 1 - \frac{SS_\text{residual}}{SS_\text{total}}$$

- **SS_total** = total variation in Y around its mean
- **SS_regression** = variation explained by X
- **SS_residual** = variation *not* explained (scatter around the line)

R² ranges from 0 to 1:

| R²   | Interpretation                |
|------|-------------------------------|
| 1.00 | X explains all variation in Y |
| 0.85 | X explains 85% of variation   |
| 0.00 | X explains nothing            |

📖 W&S §17.4 p. 555
:::

::: {.column width="40%"}
**Our calibration data:**

Because paper has uniform density, mass and area are nearly perfectly linearly related.

We expect R² ≈ 1.00.

**Important:** R² tells you *how well the line fits* the data you have. It does not by itself tell you whether the relationship is statistically significant — that requires a test of the slope.

A significant slope with a low R² is possible (weak but real relationship). In biology, R² values of 0.3–0.7 are common and meaningful.
:::
:::::

# Assumptions of Regression

:::::: columns
::: {.column width="60%"}
Four assumptions must hold for the p-value and confidence intervals to be valid (W&S §17.5):

| \# | Assumption | How to check |
|------------------------|------------------------|------------------------|
| 1 | **Linearity** — Y changes linearly with X | `plot(model)`, panel 1 |
| 2 | **Independence** — observations are independent | Study design |
| 3 | **Equal variance** — spread of residuals is constant | `plot(model)`, panels 1 & 3 |
| 4 | **Normality** — residuals are normally distributed | `plot(model)`, panel 2 + Shapiro-Wilk |
:::

:::: {.column width="40%"}
::: callout-important
**Check residuals, not raw data.**

The normality assumption is about the **residuals** (observed − predicted), not the original Y values.

R builds these diagnostic plots for you the instant you fit a model — `plot()` on any `lm` object gives you all four in one call.
:::

📖 W&S §17.5 pp. 557–560 (Figures 17.5-1, 17.5-4)
::::
::::::

# 🛑 Pause — Do Activity Parts 1–3 Now

Load `paper_area_weights.xlsx`, inspect it with `skim()`, and make the scatter of area vs mass. Predict the shape before you plot it.

# 🧩 Chunk 2 of 4 · Fit the Model & Read It

**We will cover:** loading the data, `lm()`, and reading every line of `summary()` — slope, intercept, and R².

::: callout-tip
**🖐 After this chunk: Activity Parts 4–7** (fit `lm()`, decode `summary()`, pull the equation and R²).
:::

# Load Libraries and Data

```{r load-libs-05}
# Load packages at the top of the script ---------------
library(readxl) # reading Excel files
library(tidyverse) # data manipulation + ggplot2
library(skimr) # fast data overview
```

```{r load-data-05}
# Load the paper calibration data ----------------------
paper_df <- read_excel("data/paper_area_weights.xlsx")
```

::: callout-note
The file `paper_area_weights.xlsx` contains 118 paper squares of **known area** (1–567 cm²) with their measured masses. This is the calibration dataset we use to predict leaf area from a tracing weight.
:::

# Inspect the Data

::::: columns
::: {.column width="60%"}
```{r skim-data-05}
# Quick overview of the calibration data ---------------
skim(paper_df)
```
:::

::: {.column width="40%"}
```{r scatter-summary-05}
# How many squares per area group? --------------------
paper_df %>%
  group_by(area_cm2) %>%
  summarize(
    n = n(),
    mean_wt = round(mean(mass_g), 5)
  )
```
:::
:::::

# First Plot — Scatter

::::: columns
::: {.column width="60%"}
```{r scatter-raw-05}
#| fig-format: png
#| fig-alt: "Scatter plot of paper area in square centimeters against paper mass in grams, showing a tight, linear, positively sloped relationship with no obvious curve or outliers."
# Scatter plot of area vs mass to check linearity -----
scatter_05_plot <- paper_df %>%
  ggplot(aes(x = mass_g, y = area_cm2)) +
  geom_point(alpha = 0.5, size = 2) +
  labs(
    title = "Paper Mass vs. Known Area",
    x = "Mass (g)",
    y = "Area (cm²)"
  ) +
  theme_minimal()
scatter_05_plot
```
:::

::: {.column width="40%"}
**What to look for before fitting:**

- Is the relationship **linear**? → points should follow a straight line
- Is the variance **constant**? → spread around the trend should look similar at all X values
- Any obvious **outliers**?

Both variables are continuous and numeric → linear regression is appropriate.

📖 W&S §17.1 p. 541
:::
:::::

# Fit the Model with `lm()` — the Linear Model Function

::: callout-note
**🔮 Predict first:** area rises with mass. Before you read the output — will the slope be **positive or negative**? Roughly how many cm² per gram of paper?
:::

::::: columns
::: {.column width="60%"}
```{r fit-lm-05}
# Fit the least-squares regression line ---------------
# formula: response ~ predictor
paper_lm_model <- lm(area_cm2 ~ mass_g, data = paper_df)
```

```{r summary-lm-05}
# Full model summary — every number matters -----------
summary(paper_lm_model)
```
:::

::: {.column width="40%"}
**The `lm()` formula syntax:**

- `lm(Y ~ X, data = df)`
  - `Y` goes on the **left** of `~`
  - `X` goes on the **right**
  - `~` reads as "is predicted by"
- So `area_cm2 ~ mass_g` means: "area is predicted by mass"
- The result is a **model object** — we can extract everything from it.
- 📖 W&S §17.1 p. 543
:::
:::::

# Reading the `summary()` Output

::::: columns
::: {.column width="60%"}
```{r coef-extract-05}
# Extract the key numbers from the model --------------
a_intercept <- coef(paper_lm_model)[1] # intercept
b_slope <- coef(paper_lm_model)[2] # slope

cat("Intercept (a) =", round(a_intercept, 4), "cm²\n")
cat("Slope (b)     =", round(b_slope, 2), "cm²/g\n")
```
:::

::: {.column width="40%"}
**Decode the Coefficients table:**

| Row           | What it is                         |
|---------------|------------------------------------|
| `(Intercept)` | *a* — predicted area when mass = 0 |
| `mass_g`      | *b* — cm² per gram of paper        |
| `Std. Error`  | uncertainty in each estimate       |
| `t value`     | *b* / SE — tests if slope ≠ 0      |
| `Pr(>|t|)`    | *p*-value for that t-test          |

**H₀:** β = 0 (mass does not predict area)

**Hₐ:** β ≠ 0 — our slope tests this hypothesis

📖 W&S §17.3 p. 551–553
:::
:::::

# The Calibration Equation and R²

::::: columns
::: {.column width="60%"}
```{r r2-compute-05}
# Extract R-squared and build the equation label ------
r2_val <- summary(paper_lm_model)$r.squared

cat("R² =", round(r2_val, 6), "\n\n")
cat("Calibration equation:\n")
cat(
  "  area (cm²) =",
  round(b_slope, 2),
  "× mass (g) +",
  round(a_intercept, 4),
  "\n\n"
)
cat("Interpretation of slope:\n")
cat("  Every 1 g of paper =", round(b_slope, 1), "cm² of area\n")
```
:::

::: {.column width="40%"}
- **What R² = 0.9999 means:**
  - 99.99% of the variation in paper area is explained by mass. The remaining 0.01% is measurement noise from the balance.

  - This is expected — the relationship between mass and area in uniform paper is essentially perfect.

  - In biological data, R² values of 0.3–0.7 are common and meaningful.

  - 📖 W&S §17.4 p. 555
:::
:::::

# 🛑 Pause — Do Activity Parts 4–7 Now

Fit `lm(area_cm2 ~ mass_g)`, decode the `summary()` table, and pull the slope, intercept, and R². Predict each number before it prints.

# 🧩 Chunk 3 of 4 · Visualize & Check Assumptions

**We will cover:** the calibration curve with its CI band, and the four base R diagnostic plots + Shapiro-Wilk on the residuals.

::: callout-tip
**🖐 After this chunk: Activity Parts 8–10** (plot the curve; check `plot(model)` and Shapiro-Wilk).
:::

# Plot the Calibration Curve

::::: columns
::: {.column width="60%"}
```{r calibration-line-plot-05}
#| fig-format: png
#| fig-alt: "Scatter plot of paper area versus mass with a dark blue least-squares regression line and a light blue 95 percent confidence band that narrows near the middle of the mass range and widens at the extremes."
# Calibration curve with regression line + 95% CI band
calibration_plot <- paper_df %>%
  ggplot(aes(x = mass_g, y = area_cm2)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "darkblue",
    fill = "lightblue"
  ) +
  labs(
    title = "Paper Calibration Curve",
    subtitle = paste0(
      "area = ",
      round(b_slope, 2),
      " × mass + ",
      round(a_intercept, 3),
      "  |  R² = ",
      round(r2_val, 4)
    ),
    x = "Paper Mass (g)",
    y = "Area (cm²)"
  ) +
  theme_minimal()

calibration_plot
```
:::

::: {.column width="40%"}
**The shaded band** = 95% confidence band for the *mean* predicted area at each mass value.

**Key observation:** the band is narrowest in the middle of the data range and widens at the extremes — predictions are most precise near $\bar{X}$.

This is exactly Figure 17.2-1 from W&S (p. 549).

📖 W&S §17.2 p. 548–550
:::
:::::

# Check Assumptions — the Base R Diagnostic Plots

::: callout-note
**🔮 Predict first:** `plot()` on a fitted model gives you four panels at once. One of them is a QQ plot — what do you predict the other three will show?
:::

:::::: columns
::: {.column width="60%"}
```{r diagnostic-plots-05}
#| fig-format: png
#| fig-width: 6
#| fig-height: 6
#| fig-alt: "Four base R diagnostic plots for the paper calibration linear model, arranged in a two by two grid: residuals versus fitted values, a normal Q-Q plot of the residuals, a scale-location plot, and a residuals versus leverage plot."
# plot() on an lm object gives all four diagnostics -----
par(mfrow = c(2, 2))
plot(paper_lm_model)
```
:::

:::: {.column width="40%"}
**No extra code needed.** Once you have a fitted `lm` object, `plot()` already knows how to diagnose it.

| Panel | Name | Checks |
|---|---|---|
| 1 | Residuals vs Fitted | linearity, equal variance |
| 2 | Normal Q-Q | normality of residuals |
| 3 | Scale-Location | equal variance (another view) |
| 4 | Residuals vs Leverage | influential points |

**What to look for:**

- ✅ **Good:** panel 1 & 3 show a random flat cloud; panel 2 tracks the dashed line
- ❌ **Bad:** a curve or funnel in panel 1, or points curving away at the ends of panel 2

📖 W&S §17.5 p. 557–560 (Figures 17.5-1, 17.5-4)
::::
::::::

# Confirming Normality — Shapiro-Wilk

::::: columns
::: {.column width="60%"}
```{r shapiro-resid-05}
# Shapiro-Wilk test on residuals — NOT on raw data ---
shapiro.test(residuals(paper_lm_model))
```
:::

::: {.column width="40%"}
**Remember:**

- The normality assumption is about **residuals**, not raw data
- H₀ (Shapiro-Wilk): residuals are normally distributed
- *p* \> 0.05 → proceed with confidence intervals and *p*-values
- Use this as a **formal companion** to panel 2 of the diagnostic plot — not a replacement for looking at it

📖 W&S §17.5 p. 559
:::
:::::

# 🛑 Pause — Do Activity Parts 8–10 Now

Plot the calibration curve, then run `plot(model)` and Shapiro-Wilk on the residuals. Predict each diagnostic before you run it.

# 🧩 Chunk 4 of 4 · Predict & Report

**We will cover:** turning a tracing mass into a predicted area, `predict()` with confidence vs prediction intervals, and writing the results paragraph.

::: callout-tip
**🖐 After this chunk: Activity Parts 11–12** (predict leaf area; write the results paragraph).
:::

# Predict Leaf Area from Tracing Mass

::: callout-note
**🔮 Predict first:** The sunny tracing weighs 0.092 g, the shady 0.138 g. Before running — which predicts the **larger** area, and does that match "shady leaves are bigger"?
:::

::::: columns
::: {.column width="60%"}
```{r predict-point-05}
# Plug tracing masses into the calibration line -------
# Suppose sunny leaf tracing weighs 0.092 g
# and shady leaf tracing weighs 0.138 g

mass_sunny <- 0.092
mass_shady <- 0.138

area_sunny <- b_slope * mass_sunny + a_intercept
area_shady <- b_slope * mass_shady + a_intercept

cat("Sunny tracing mass:", mass_sunny, "g\n")
cat("Predicted area:    ", round(area_sunny, 2), "cm²\n\n")
cat("Shady tracing mass:", mass_shady, "g\n")
cat("Predicted area:    ", round(area_shady, 2), "cm²\n")
```
:::

::: {.column width="40%"}
**The equation in action:**

$$\widehat{\text{area}} = 130.5 \times \text{mass} + 0.28$$

For a sunny tracing at 0.092 g:

$130.5 \times 0.092 + 0.28 = 12.3$ cm²

For a shady tracing at 0.138 g:

$130.5 \times 0.138 + 0.28 = 18.3$ cm²

This matches our confirmed result from Lectures 09–10 — shady leaves are significantly heavier and, now, larger in area too!
:::
:::::

# Live Demo — Watch It Break (on purpose)

:::::: columns
::: {.column width="60%"}
`predict()` needs the **exact** predictor column name. I'll get it wrong:

``` r
# The column is mass_g, but I typed mass
new <- tibble(mass = 0.092)
predict(paper_lm_model, newdata = new)
```

R errors:

``` 
Error in eval(predvars, data, env) :
  object 'mass_g' not found
```

The fix — match the model's variable name exactly:

``` r
new <- tibble(mass_g = 0.092)
predict(paper_lm_model, newdata = new)
```
:::

:::: {.column width="40%"}
::: callout-tip
**✅ Why show a broken run?**

`predict()` looks for the same column names the model was built on. *"object 'mass_g' not found"* means your `newdata` is missing that column — almost always a typo. This is the #1 `predict()` mistake.
:::
::::
::::::

# Using `predict()` — with Prediction Intervals

::::: columns
::: {.column width="60%"}
```{r predict-interval-05}
# predict() gives uncertainty around predictions ------
new_masses <- tibble(mass_g = c(0.092, 0.138))

# Confidence interval: for the MEAN area at this mass
predict(paper_lm_model, newdata = new_masses, interval = "confidence") %>%
  round(3)

# Prediction interval: for a SINGLE new observation
predict(paper_lm_model, newdata = new_masses, interval = "prediction") %>%
  round(3)
```
:::

::: {.column width="40%"}
**Two types of interval (W&S §17.2 p. 549):**

| Type           | Predicts                             | Width    |
|----------------|--------------------------------------|----------|
| **Confidence** | mean area for *all* leaves this mass | narrower |
| **Prediction** | area of *one specific* leaf          | wider    |

For predicting a single leaf, use the **prediction interval** — it captures individual-to-individual variation within the same mass class.

📖 W&S §17.2 pp. 548–550
:::
:::::

# How to Report Regression Results

::::: columns
::: {.column width="60%"}
```{r report-values-05}
# Pull reporting values from the model ----------------
f_stat <- summary(paper_lm_model)$fstatistic
f_val <- round(f_stat[1], 1)
df1 <- f_stat[2]
df2 <- f_stat[3]
p_val <- pf(f_stat[1], df1, df2, lower.tail = FALSE)

cat("--- Results paragraph ---\n")
cat("Paper mass was a strong predictor of paper area\n")
cat("(linear regression: F(", df1, ",", df2, ") =", f_val, ",\n")
cat("p <", signif(p_val, 1), ", R² =", round(r2_val, 4), ").\n")
cat("The calibration equation is:\n")
cat("  area =", round(b_slope, 2), "× mass +", round(a_intercept, 3), "cm²\n")
cat(
  "(slope 95% CI:",
  round(confint(paper_lm_model)[2, 1], 2),
  "to",
  round(confint(paper_lm_model)[2, 2], 2),
  "cm²/g)\n"
)
```
:::

::: {.column width="40%"}
**Standard report format:**

> *"Paper area was accurately predicted from paper mass (linear regression: F(df₁, df₂) = F, p \< 0.001, R² = 0.9999). The regression equation was area = 130.5 × mass + 0.28 (slope 95% CI: X to Y cm²/g)."*

**Always include:**

- F-statistic and both df
- *p*-value (never "p = 0.000" — write "p \< 0.001")
- R²
- The actual equation
- 95% CI for the slope

📖 W&S §17.3 p. 553
:::
:::::

# 🛑 Pause — Do Activity Parts 11–12 Now

Predict leaf area from tracing mass, compare confidence vs prediction intervals, and write your results paragraph. Predict the numbers before they print.

# What We Learned Today

::::: columns
::: {.column width="60%"}
**Concepts:**

- **Regression** predicts Y from X using a straight line equation
- Equation: $\hat{Y} = a + bX$ — slope *b*, intercept *a*
- **Least squares** — minimizes sum of squared residuals
- **R²** — fraction of Y variation explained by X
- **Four assumptions** — linearity, independence, equal variance, normality of residuals
- **Two prediction types** — confidence vs. prediction intervals

**R skills:**

- `lm(Y ~ X, data)` — fits the model
- `summary()` — extracts slope, intercept, R², p-values
- `coef()`, `fitted()`, `residuals()` — pull model components
- `plot(model)` — all four diagnostic plots in one call
- `predict(model, newdata, interval = "prediction")` — predictions with uncertainty
- `geom_smooth(method = "lm", se = TRUE)` — plot the line with CI band
:::

::: {.column width="40%"}
**Chapter references:**

- 📖 W&S §17.1 — Linear regression
- 📖 W&S §17.2 — Confidence in predictions
- 📖 W&S §17.3 — Testing slope hypotheses
- 📖 W&S §17.4 — R²
- 📖 W&S §17.5 — Assumptions

**Up next — Lecture 12:**

- Downloading real climate data straight into R with `GSODR`
- Summarizing daily data to reveal long-term trends
- Reusing `lm()` to estimate a rate of change over time
:::
:::::
