---
order: 11
week: 6
# =====================================================================
#  regression_analysis_report.qmd
#  OPTIONAL / SECONDARY. The same regression as the activity skeleton
#  (scripts/11_regression_skeleton.R), wrapped in writing so that
#  pressing Render produces a finished Word report with the numbers and
#  figures already in place. The .R script is what you hand in; this is
#  here if you would rather hand in a report.
# =====================================================================
title: "Paper Calibration Curve: Predicting Leaf Area from Mass"
subtitle: "The regression from Activity 11, written up as a report"
author: "Your Name Here"
date: today

format:
  docx:
    toc: true
    number-sections: true
    fig-width: 6
    fig-height: 4
    fig-format: png
  html: default          # the same file also makes a web page

# Hide the code by default so the report reads like a report.
execute:
  echo: false
  warning: false
  message: false
---

```{r libraries}
# Load all packages in one chunk at the top ----------------------------
library(readxl)
library(tidyverse)
library(janitor)
library(broom)   # tidy() and glance() turn a model into a table
```

```{r load-data}
# Read the paper calibration data --------------------------------------
paper_df <- read_excel("data/paper_area_weights.xlsx")
```

# Introduction

To estimate leaf surface area without a leaf-area meter, we built a
**calibration curve**. Paper from one stock has a constant area per gram, so the
**mass** of a paper square is a near-perfect linear predictor of its **area**.
Once we know that relationship, we can trace a leaf onto the same paper, weigh
the tracing, and convert its mass to an area.

- **Explanatory variable (X):** paper mass (g) — easy to measure on a balance
- **Response variable (Y):** paper area (cm²) — what we actually want

# Methods

We weighed 118 paper squares of known area (1–567 cm²) and fit a simple linear
regression, `area_cm2 ~ mass_g`, with `lm()`. We checked the residuals with base
R's diagnostic plots for the fitted model. All work was done in R with the
tidyverse; this report was written in Quarto, so every number and figure below is
computed from the data rather than typed in by hand.

# Results

```{r fit-model}
# Fit the regression line ----------------------------------------------
paper_lm_model <- lm(area_cm2 ~ mass_g, data = paper_df)
```

Paper mass was an extremely strong predictor of paper area. The overall fit is in
Table 1 and the fitted equation is in Table 2.

```{r model-summary}
# Overall model fit: R-squared, F, p -----------------------------------
glance(paper_lm_model) |>
  select(r.squared, statistic, df, df.residual, p.value) |>
  knitr::kable(digits = c(4, 1, 0, 0, 20),
               col.names = c("R-squared", "F", "df (model)",
                             "df (residual)", "p-value"),
               caption = "Table 1. Overall fit of the calibration regression.")
```

```{r coef-table}
# The calibration equation: intercept and slope ------------------------
tidy(paper_lm_model) |>
  knitr::kable(digits = c(0, 3, 3, 1, 20),
               col.names = c("Term", "Estimate", "Std. error",
                             "t", "p-value"),
               caption = "Table 2. Coefficients of the calibration equation. The `mass_g` estimate is the area gained per additional gram of paper; `(Intercept)` is the fitted area at zero mass.")
```

Read the equation off Table 2: predicted area equals the `(Intercept)` estimate
plus the `mass_g` estimate times the mass. In words, each additional gram of this
paper stock adds about the `mass_g` estimate in cm² of area. The slope's
*p*-value tests whether mass predicts area at all; the intercept is not
distinguishable from zero, which is what physics says it should be — paper with
no mass has no area.

## The calibration curve

```{r calibration-plot}
#| fig-cap: "Figure 1. Paper area against mass, with the fitted regression line and its 95% confidence band."
#| fig-alt: "Scatter plot of paper area against paper mass with a straight fitted regression line and a narrow confidence band. The points sit almost exactly on the line."
paper_df %>%
  ggplot(aes(x = mass_g, y = area_cm2)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Paper mass (g)",
       y = "Paper area (cm²)") +
  theme_minimal()
```

## Residual checks

```{r diagnostic-plots}
#| fig-cap: "Figure 2. Base R diagnostic plots for the fitted model — residuals vs. fitted, normal Q-Q, scale-location, and residuals vs. leverage."
#| fig-alt: "Four base R diagnostic plots for the paper calibration model in a two by two grid: residuals versus fitted values showing a funnel that widens to the right, a normal Q-Q plot with points curving away at both ends, a scale-location plot rising to the right, and residuals versus leverage."
par(mfrow = c(2, 2))
plot(paper_lm_model)
```

Panel 1 of Figure 2 shows a funnel rather than a flat cloud: the residuals get
larger as the squares get larger. Table 3 puts numbers on it.

```{r residual-table}
# How big are the misses at each square size? --------------------------
paper_df %>%
  mutate(residual = residuals(paper_lm_model)) %>%
  group_by(area_cm2) %>%
  summarize(biggest_miss = max(abs(residual))) %>%
  knitr::kable(digits = c(0, 2),
               col.names = c("Known area (cm²)", "Largest residual (cm²)"),
               caption = "Table 3. The largest miss at each square size. The equal-variance assumption does not hold: a balance has percentage error, not fixed error, so bigger squares miss by more.")
```

The equal-variance assumption is therefore violated at the large-square end of
the range, and confidence intervals there should be treated with caution. The
leaf tracings we apply the curve to below weigh 0.1–0.4 g, at the tight end of
the line, where a typical miss is a fraction of a cm².

## Predicting leaf area

```{r predict-two}
# Predict area for two example leaf tracings ---------------------------
new_tracings <- tibble(
  tracing = c("sunny-side", "shady-side"),
  mass_g  = c(0.092, 0.138)
)

new_tracings |>
  mutate(predicted_area_cm2 =
           predict(paper_lm_model, newdata = new_tracings)) |>
  knitr::kable(digits = c(0, 3, 1),
               col.names = c("Tracing", "Mass (g)", "Predicted area (cm²)"),
               caption = "Table 4. Predicted leaf area for two example paper tracings.")
```

The same `predict()` call converts a whole column at once. Table 5 applies the
calibration to every leaf in our class data set, then summarizes the predicted
areas by side of the tree.

```{r predict-column}
# Turn every tracing mass in the leaf data into an area ----------------
leaf_df <-
  read_excel("data/2026_09_03_data_sci_leaf_area.xlsx") %>%
  clean_names() %>%
  mutate(leaf_area_cm2 =
           predict(paper_lm_model,
                   newdata = tibble(mass_g = paper_mass_g)))

leaf_df %>%
  group_by(shade) %>%
  summarize(
    n         = sum(!is.na(leaf_area_cm2)),
    mean_area = mean(leaf_area_cm2, na.rm = TRUE),
    sd_area   = sd(leaf_area_cm2, na.rm = TRUE)
  ) %>%
  knitr::kable(digits = c(0, 0, 1, 1),
               col.names = c("Side", "n", "Mean area (cm²)", "SD (cm²)"),
               caption = "Table 5. Predicted leaf area by side of the tree. Leaves whose tracings were never weighed are excluded.")
```

Table 5 reports the predicted areas in the units we wanted. The two sides differ
by about a square centimetre — the same story the Welch's t-test told us about
leaf mass.

# Discussion

*Write one short paragraph here.* Did the calibration fit well — what do R² in
Table 1 and Figure 1 say? What did the residual checks in Figure 2 and Table 3
show, and what does that mean for how far you would trust the curve? Why is
working in **area** (cm²) more useful to a reader than raw tracing **mass** (g)
when comparing sunny and shady leaves? This is where you paste the results
paragraph you wrote in the activity.

# References

- Whitlock, M. & Schluter, D. *The Analysis of Biological Data* (2nd ed.), Ch. 17.
- Wickham, H. *R for Data Science (2nd ed.)* — <https://r4ds.hadley.nz/>
