---
title: "Lecture 12 — Real Climate Data in R"
subtitle: "How fast is your city warming? (regression on time)"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# These are the ONLY things you need to edit to move this lecture around.
# The folder name and filename never have to change.
order: 12
week: 6
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Download real weather station data from inside R, summarize it to yearly
  means, and use lm() to estimate a warming rate in degrees per decade — for
  Duluth, for each season, and for a city you choose.
categories: [tidyverse, descriptive-stats]

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(width = 65)   # printed R output wraps at 65 too
```

# Where we left off (Lecture 11 — Regression)

- **`lm(Y ~ X, data = df)`** fits a straight line through a cloud of points
- **`summary(model)`** gives you:
  - the **slope** — how much Y changes per 1 unit of X
  - its ***p*-value** — is that slope real, or could it be chance?
  - **R²** — how tightly the points follow the line
- **`geom_smooth(method = "lm")`** draws the line
- **`plot(model)`** checks the residuals

::: callout-note
**✅ Where today comes in**

- Last time X was **paper mass** and the slope converted grams into cm²
- Today X is **time**
  - the slope becomes a **rate of change** — degrees per year
  - and that is exactly what a "warming rate" is
:::

# Goals for Today

:::::: columns
:::: {.column width="58%"}
- **Download** real weather station data from inside R — no Excel, no copy-paste
- **Clean** the names with `janitor` so the code stays easy to type
- **Summarize** ~25,000 daily records into one mean per year
- **Fit** `lm(temp ~ year)` and read the warming rate straight off `summary()`
- **Split by season** — is winter warming faster than summer?
- **Find a station for a city you choose** and run the whole thing again

::: callout-tip
**🖐 Where this is going**

- The analysis is the same four lines every time
- Change the station, and it is a new city
- That is your out-of-class extension
:::
::::

::: {.column width="42%"}
**Tools today:**

- `GSODR` — download NOAA station data
- `janitor` — `clean_names()`
- `tidyverse` — summarize and plot
- `lm()` — the slope is the answer

**Reading:**

- 📖 R for Data Science 2e
  - [Ch. 3 Data transformation](https://r4ds.hadley.nz/data-transform)
- 📖 Whitlock & Schluter Ch. 17 — Regression

**Naming conventions:**

- data frames → `_df`
- plots → `_plot`
- models → `_model`
:::
::::::

# 🧩 Chunk 1 of 3 · The Question

**We will cover:** where the data comes from, why the slope of a regression on time *is* a warming rate, and why you have to summarize before you can see it. No code in this chunk.

# Where Does This Data Come From?

::::: columns
::: {.column width="58%"}
- **GSOD** = Global Surface Summary of the Day, run by **NOAA**
  - daily weather from thousands of stations worldwide
  - many records reach back to the **1940s**
  - free, and no account needed
- The **`GSODR`** package pulls it straight into R
  - you give it a **station ID** and a range of **years**
  - it hands back a data frame
  - loading it prints a notice that GSOD **retired in Aug 2025** — expected, not an error
    - the historical archive is still there; it just stops at 2025

**Our station to start:**

- Duluth International Airport
- Station ID **`727450-14913`**
- Daily records from **1948** to **2024**
  - but **1965–1972 are missing** — the airport reported under a different station ID then
  - real archives have holes; you work around them
:::

::: {.column width="42%"}
- **Why this matters:**
  - this is the same messy, real-world data you will meet in your own project
  - the workflow — download → summarize → model — is the one you will reuse all term
- **And it is yours to reuse:**
  - every city with a long station record can be run through the exact same code
  - you will pick one later today
:::
:::::

# The Slope *Is* the Warming Rate

::::: columns
::: {.column width="55%"}
Last lecture:

$$\widehat{\text{area}} = a + b \times \text{mass}$$

- *b* was **cm² per gram** — a conversion factor

Today:

$$\widehat{\text{temp}} = a + b \times \text{year}$$

- *b* is **°C per year** — a **rate of change**

**Nothing about the code changes.** Only what X means.
:::

::: {.column width="45%"}
- **Reading the slope:**
  - *b* **positive** → warming
  - *b* **negative** → cooling
  - *b* near 0 → no trend
- **Make it human:** °C per year is a tiny number
  - **multiply it by 10** → °C **per decade**
  - 0.027 °C/year → about **0.27 °C per decade**
- **The *p*-value still does the same job**
  - H₀: the slope is 0 — no trend at all
  - small *p* → the trend is real, not chance
:::
:::::

# Why You Must Summarize First

:::::: columns
::: {.column width="58%"}
- A station reports **one row per day**
  - 69 years × 365 days ≈ **25,000 rows**
- Every one of those days carries the **seasonal cycle**
  - January is 30 °C colder than July, every single year
  - that swing is enormous compared with a few tenths of a degree of warming
- So the trend you want is **buried** in the noise you don't

| Summary level | `group_by()` | Rows left |
|---------------|--------------|-----------|
| Daily (raw) | none | ~25,000 |
| **Yearly** | `year` | **69** |

: {tbl-colwidths="[40,30,30]"}
:::

:::: {.column width="42%"}
::: callout-important
**Summarizing is not throwing data away.**

- It is choosing **which question** you are answering
- Want to study a heat wave? Use **daily**
- Want to study climate? Use **yearly**
:::

- One `group_by(year) %>% summarize()` — the same two lines from Summary Statistics
- That is the whole fix
::::
::::::

# 🧩 Chunk 2 of 3 · Duluth's Warming Rate

**We will cover:** downloading the Duluth record, cleaning the names, summarizing to yearly means, and fitting `lm(temp ~ year)`.

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 1–5** (download, summarize, plot, fit, read the slope).
:::

# Load Libraries and Download the Data

::::: columns
::: {.column width="55%"}
```{r load-libs-12}
# Load packages at the top -----------------------------
library(tidyverse) # dplyr + ggplot2
library(janitor)   # clean_names()
library(GSODR)     # NOAA weather station data
```

```{r get-duluth-12}
# Download daily Duluth weather, 1948-2024 -------------
# 1965-1972 are missing for this station, so skip them
gap_years <- 1965:1972

duluth_df <- get_GSOD(years = setdiff(1948:2024, gap_years),
                      station = "727450-14913") %>%
  clean_names()
```
:::

::: {.column width="45%"}
- `get_GSOD()` takes **years** and a **station** and goes and gets the data
  - `1948:2024` is a vector — "every year in this range"
  - `setdiff()` removes the 8 missing years from that vector
    - ask for a year the station does not have and the **whole download fails**
  - it takes a minute; 69 years of daily data is a lot
- **`clean_names()` is the important habit here**
  - NOAA ships columns as `TEMP`, `YEARMODA`, `MXSPD`
  - R is case-sensitive, so `temp` would error
  - `clean_names()` makes everything lower case and snake case, once, at the top

::: callout-tip
Clean the names as you load the data, and you never think about capitals again.
:::
:::
:::::

# What Did We Get?

::::: columns
::: {.column width="55%"}
```{r peek-duluth-12}
# How big is it, and what is in it? --------------------
dim(duluth_df)

names(duluth_df)[13:18]
```
:::

::: {.column width="45%"}
- **~25,000 rows** — one per day
- **47 columns** — temperature, dew point, wind, precipitation, flags…
- The handful we care about:
  - `yearmoda` — the date
  - `year`, `month`, `day`
  - `temp` — daily mean temperature in °C

::: callout-note
`names(df)` is how you find out what a downloaded data set actually calls things. Always look before you type.
:::
:::
:::::

# Trim to What We Need

::::: columns
::: {.column width="55%"}
```{r select-cols-12}
# Keep only the columns we actually use ----------------
dlh_df <- duluth_df %>%
  select(name, yearmoda, year, month, temp)

head(dlh_df, 3)
```
:::

::: {.column width="45%"}
- 47 columns is more than anyone needs to look at
- `select()` keeps the five that answer our question
- Same `select()` from the Wrangling lecture — just on a much wider real file

::: callout-tip
Trim early. Every later line is easier to read when the data frame has 5 columns instead of 47.
:::
:::
:::::

# Look at the Raw Daily Data

::::: columns
::: {.column width="55%"}
```{r raw-plot-12}
#| fig-format: png
#| fig-height: 2.5
#| fig-alt: "Scatter plot of daily mean temperature in Duluth from 1948 to 2024. The points overlap into a dense band about 60 degrees tall, driven entirely by the seasonal cycle, with no visible long-term trend."
# Every single daily temperature -----------------------
raw_temp_plot <- dlh_df %>%
  ggplot(aes(x = yearmoda, y = temp)) +
  geom_point(alpha = 0.2, size = 0.5)
raw_temp_plot
```
:::

::: {.column width="45%"}
- A solid band about **60 °C tall** — that is summer vs winter, repeated 69 times
- ~25,000 points on top of each other
- Any warming trend is in there, but it is a **fraction of a degree** against a 60 °C swing

**So you cannot see it.** That is not a plotting problem — it is a summarizing problem.
:::
:::::

# Summarize to One Mean per Year

::::: columns
::: {.column width="55%"}
```{r yearly-summary-12}
# Mean temperature per year ----------------------------
dlh_year_df <- dlh_df %>%
  group_by(year) %>%
  summarize(temp = mean(temp, na.rm = TRUE))

head(dlh_year_df, 3)
```
:::

::: {.column width="45%"}
- `group_by(year)` — one group per year
- `summarize(temp = mean(temp))` — one number per group
- **25,000 rows become 69**
- `na.rm = TRUE` ignores days the station missed

::: callout-note
Reusing the name `temp` on the left means every line below keeps working — the same trick from T-Tests II.
:::
:::
:::::

# Plot the Yearly Means

::::: columns
::: {.column width="55%"}
```{r yearly-plot-12}
#| fig-format: png
#| fig-height: 2.7
#| fig-alt: "Scatter plot of Duluth annual mean temperature against year from 1948 to 2024 with a fitted straight regression line sloping clearly upward from about 3.3 to about 5.3 degrees Celsius."
# Yearly means with the regression line ----------------
yearly_temp_plot <- dlh_year_df %>%
  ggplot(aes(x = year, y = temp)) +
  geom_point() +
  geom_smooth(method = "lm")
yearly_temp_plot
```
:::

::: {.column width="45%"}
- **Now you can see it.** The line clearly rises
- 69 points instead of 25,000 — and the story is *clearer*, not weaker
- Still plenty of scatter: single years are weather, the line is climate
- Exactly the same `geom_smooth(method = "lm")` as the Regression lecture

**But "it looks like it goes up" is not a result.** We need the number.
:::
:::::

# Fit the Trend

::::: columns
::: {.column width="55%"}
```{r yearly-lm-12}
# Temperature predicted by year ------------------------
dlh_year_model <- lm(temp ~ year,
                     data = dlh_year_df)

summary(dlh_year_model)
```
:::

::: {.column width="45%"}
- `lm(temp ~ year)` reads "**temperature is predicted by year**"
- Y on the left, X on the right — same as always
- The three numbers you want are all in this output:
  - the **slope** (`year` → `Estimate`)
  - its ***p*-value** (`year` → `Pr(>|t|)`)
  - **R²** (`Multiple R-squared`)
:::
:::::

# Reading the Warming Rate

::::: columns
::: {.column width="50%"}
**Look at the `year` row of the Coefficients table.**

- **`Estimate` ≈ 0.027** — that is the slope
  - units: **°C per year**
  - positive → Duluth is **warming**
  - ×10 → about **0.27 °C per decade**
  - over ~77 years → roughly **2 °C**
- **`Pr(>|t|)` is tiny** (far below 0.001)
  - the warming is **real**, not year-to-year luck
:::

::: {.column width="50%"}
- **`Multiple R-squared` ≈ 0.32**
  - year explains about **a third** of the variation in annual temperature
  - the other two thirds is **weather** — some years are just cold
  - a modest R² next to a rock-solid *p* is completely normal in climate data

::: callout-important
**The intercept looks absurd — about −49 °C.**

- That is the predicted temperature at "year 0"
- Year 0 is 2,000 years outside the data
- The intercept is bookkeeping; the **slope** is the science
:::
:::
:::::

# → ACTIVITY 12 Parts 1–5 now

::: {.activitybanner}
**🛑 Do Activity Parts 1–5 now**

- Download the Duluth record and `clean_names()` it
- Trim the columns and plot the raw daily data
- Summarize to one mean per year and plot it
- Fit `lm(temp ~ year)` and read the slope, *p*, and R² off `summary()`
:::

# 🧩 Chunk 3 of 3 · Seasons, and Your Own City

**We will cover:** labelling each day with a season, fitting a separate model for summer and winter, and finding the station ID for a city you choose.

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 6–10** (seasons, two models, your own city).
:::

# A Better Question — Which Season Is Warming?

::::: columns
::: {.column width="55%"}
- The annual mean says Duluth warmed about **0.27 °C per decade**
- But an annual mean **hides which part of the year moved**
- That distinction is the biologically interesting one:
  - **ice-out dates** and **frozen soil** depend on *winter*
  - **heat stress** and **drought** depend on *summer*
  - **growing season length** depends on the shoulders

**The plan — three steps you already know:**

1.  `mutate()` + `case_when()` to label each day
2.  `group_by(year, season)` + `summarize()`
3.  One `lm()` per season
:::

::: {.column width="45%"}
::: callout-important
**⚠️ Seasons are not months — they are hemispheres.**

- June–August is **summer** north of the equator
- June–August is **winter** south of it
- If you pick Melbourne, Santiago or Cape Town later, you must **swap the labels**
:::

- A station's `latitude` column tells you which hemisphere you are in
  - positive → northern
  - negative → southern
:::
:::::

# Label Each Day with a Season

::::: columns
::: {.column width="55%"}
```{r season-case-when-12}
# Northern hemisphere seasons --------------------------
# SOUTHERN hemisphere: swap summer and winter
dlh_season_df <- dlh_df %>%
  mutate(season = case_when(
    month %in% c(6, 7, 8) ~ "summer",
    month %in% c(12, 1, 2) ~ "winter",
    TRUE ~ "shoulder"
  ))

# Did it work? -----------------------------------------
dlh_season_df %>% count(season)
```
:::

::: {.column width="45%"}
- `case_when()` reads top to bottom: *condition* `~` *answer*
  - the **first** matching line wins
- `month %in% c(6, 7, 8)` = "is month 6, 7 or 8?"
- `TRUE ~ "shoulder"` is the catch-all for everything left over — spring and fall
- `count()` is the check: roughly a quarter of the days in each of summer and winter, half in the shoulder

📖 R4DS §3.4 — `case_when()`
:::
:::::

# Summarize by Year *and* Season

::::: columns
::: {.column width="55%"}
```{r season-summary-12}
# Drop the shoulder, then one mean per year per season -
season_year_df <- dlh_season_df %>%
  filter(season != "shoulder") %>%
  group_by(year, season) %>%
  summarize(temp = mean(temp, na.rm = TRUE),
            .groups = "drop")

head(season_year_df, 4)
```
:::

::: {.column width="45%"}
- `filter(season != "shoulder")` — `!=` means "is not"
  - keeps only the two extremes
- `group_by(year, season)` — **two** grouping columns
  - one row per year **per season**
  - 69 years × 2 seasons = 138 rows
- `.groups = "drop"` tidies up afterwards

This is the same two-column `group_by()` from T-Tests II.
:::
:::::

# Plot Both Seasons

::::: columns
::: {.column width="55%"}
```{r season-plot-12}
#| fig-format: png
#| fig-height: 2.6
#| fig-alt: "Scatter plot of Duluth seasonal mean temperature against year with points colored by season. Summer points sit near 18 degrees Celsius and winter points near minus 11, each with its own upward-sloping fitted line; the winter line is visibly steeper."
# One line per season ----------------------------------
season_temp_plot <- season_year_df %>%
  ggplot(aes(x = year, y = temp, color = season)) +
  geom_point() +
  geom_smooth(method = "lm")
season_temp_plot
```
:::

::: {.column width="45%"}
- `color = season` inside `aes()` does two jobs at once:
  - colors the points by season
  - makes `geom_smooth()` fit **a separate line per season**
- Two bands, far apart — winter near −11 °C, summer near 18 °C
- **Don't compare the heights. Compare the steepness.**
  - the height is just "winter is cold"
  - the **slope** is the warming rate
- Winter's line looks steeper — but we need the numbers
:::
:::::

# A Model for Summer

::::: columns
::: {.column width="55%"}
```{r summer-model-12}
# Summer only ------------------------------------------
summer_df <- season_year_df %>%
  filter(season == "summer")

summer_model <- lm(temp ~ year, data = summer_df)
summary(summer_model)
```
:::

::: {.column width="45%"}
- `filter()` first, then the **same `lm()` as before**
- Two small models are far easier to read than one complicated one

**Summer slope ≈ 0.022 °C/year**

- ×10 → about **0.22 °C per decade**
- *p* = 0.0002 → real

Now do exactly the same thing for winter.
:::
:::::

# A Model for Winter

::::: columns
::: {.column width="55%"}
```{r winter-model-12}
# Winter only ------------------------------------------
winter_df <- season_year_df %>%
  filter(season == "winter")

winter_model <- lm(temp ~ year, data = winter_df)
summary(winter_model)
```
:::

::: {.column width="45%"}
**Winter slope ≈ 0.037 °C/year**

- ×10 → about **0.37 °C per decade**
- *p* = 0.0006 → real

**Winter is warming about 1.7× as fast as summer.**

- This is the expected pattern across the northern continents
- Fewer extreme-cold nights, shorter ice seasons, earlier thaw
- The annual mean — 0.27 °C/decade — sat right between the two and told you neither
:::
:::::

# Reading Two Models Side by Side

::::: columns
::: {.column width="52%"}
Both slopes come straight off the `year` row of each `summary()`:

| Model | Slope (°C/yr) | ×10 (°C/decade) | *p* |
|--------|--------|--------|--------|
| Annual | 0.027 | 0.27 | 0.0000003 |
| Summer | 0.022 | 0.22 | 0.0002 |
| Winter | 0.037 | 0.37 | 0.0006 |

: {tbl-colwidths="[25,25,28,22]"}

- Winter's slope has a **bigger standard error** too — winters are more variable than summers
- A bigger *p* with a bigger slope is not a contradiction; it is noisier data
:::

::: {.column width="48%"}
**How to write it up:**

> *Mean annual temperature at Duluth rose significantly from 1948 to 2024 (linear regression: slope = 0.027 °C yr⁻¹, equivalent to 0.27 °C per decade; F(1, 67) = 32.1, p \< 0.001, R² = 0.32). Winter warmed roughly 1.7 times faster than summer (0.37 vs 0.22 °C per decade).*

- Give the slope **with units**
- Convert to **per decade** so a reader can feel it
- Report *p* and R² the same way as any regression
:::
:::::

# Now Do It for Your City

::::: columns
::: {.column width="55%"}
```{r nearest-12}
# Find stations near a place you care about ------------
# Duluth is 46.84 N, -92.19 W
nearest_stations(LAT = 46.84, LON = -92.19,
                 distance = 50) %>%
  clean_names() %>%
  select(stnid, name, begin, end) %>%
  head(3)
```
:::

::: {.column width="45%"}
- `nearest_stations()` takes a **latitude**, a **longitude**, and a radius in km
- Look up your city's lat/long in any search engine
  - **south** of the equator → negative latitude
  - **west** of Greenwich → negative longitude
- Then read the table:
  - **`stnid`** is what you paste into `get_GSOD()`
  - **`begin`** and **`end`** tell you the years available

::: callout-important
**Pick a station with a long record.** A trend needs decades — aim for a `begin` date of 1970 or earlier.
:::
:::
:::::

# Swap One Line, Get a New City

::::: columns
::: {.column width="55%"}
``` r
# The ONLY line that changes ---------------------------
city_df <- get_GSOD(years = 1973:2024,
                    station = "948660-99999") %>%
  clean_names()

# Everything below is identical ------------------------
city_year_df <- city_df %>%
  group_by(year) %>%
  summarize(temp = mean(temp, na.rm = TRUE))

city_model <- lm(temp ~ year, data = city_year_df)
summary(city_model)
```
:::

::: {.column width="45%"}
- That is the whole point of today
  - **one line** names the place
  - the rest of the analysis never changes
- Use the `years =` range your station actually has — don't ask for years before `begin`

::: callout-important
**If your city is in the southern hemisphere**, swap the season labels in `case_when()`:

- `month %in% c(12, 1, 2)` → **summer**
- `month %in% c(6, 7, 8)` → **winter**
:::
:::
:::::

# → ACTIVITY 12 Parts 6–10 now

::: {.activitybanner}
**🛑 Do Activity Parts 6–10 now**

- Label the seasons with `case_when()` and summarize by year and season
- Plot both seasons, then fit a model for each
- Write the warming rates in °C per decade
- Find the station for a city you choose — that is your Extension
:::

# What We Learned Today

::::: columns
::: {.column width="58%"}
**The ideas:**

- A regression **on time** turns a slope into a **rate of change** — ×10 for per decade
- Raw daily data **hides** the trend; `group_by(year) %>% summarize()` is the fix
- An **annual mean hides which season moved** — Duluth's winter warms ~1.6× faster
- Seasons are set by **hemisphere**, not by month number
- A modest **R² with a tiny *p*** is normal — real trend, weather scatter

**The code:**

- `get_GSOD(years =, station =)` and `nearest_stations()`
- `clean_names()` — lower-case names, once, at the top
- `group_by()` + `summarize()` — to the right level
- `case_when()` — build the season
- `lm(temp ~ year)` + `summary()` — the rate
:::

::: {.column width="42%"}
**Hand in:** your finished activity script, plus the warming rate you found for **your own city**.

**References:**

- 📖 R4DS Ch. 3 — Data transformation
- 📖 W&S Ch. 17 — Regression
- NOAA GSOD — `ncei.noaa.gov`

**Up next:**

- The **final project** — the same pipeline, on data you choose
- Download → clean → summarize → plot → model → report
:::
:::::
