---
title: "Lecture 14 — One-Way ANOVA II: Running & Reporting the Test"
subtitle: "Reading the F table, finding which groups differ, and writing it up"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# order/week are auto-stamped from _schedule.yml by sync_order_from_schedule.R
order: 14
week: 7
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Running the one-way ANOVA with aov() and car::Anova(), finding WHICH
  penguin species differ with emmeans post-hoc tests, and writing a
  complete results sentence.
categories: [anova, tidyverse, statistics]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 13 — One-Way ANOVA I)

- `mass_model <- lm(body_mass_g ~ species, data = penguins_df)` — fit
- **Normality checked** — Q-Q panel + Shapiro-Wilk on the residuals
- **Equal variance checked** — Levene's test
- **Decision:** assumptions look fine, proceed with classic ANOVA

::: callout-note
**✅ Transition**

The model is fit and its assumptions are checked. Today we actually read the F table, find **which** species differ, and turn the result into a sentence a reader can trust.
:::

# Goals for today

::::: columns
:::: {.column width="60%"}
- Run the ANOVA two ways: `aov()` and **`car::Anova()`**
- Understand why **Type II** sums of squares matter for unbalanced data
- Find **which** groups differ with **`emmeans`** post-hoc tests
- Read a compact letter display (CLD)
- Plot the estimated means ± CI
- Write a complete, reportable results sentence
::::

:::: {.column width="40%"}
**Tools today:**

- `car` — `Anova()`
- `emmeans`, `multcomp` — post-hoc, CLD

**Textbook:**

- 📖 Whitlock & Schluter, **Ch. 15 — ANOVA**
- 📖 [emmeans vignettes](https://cran.r-project.org/package=emmeans)

**Naming:** models → `_model`, plots → `_plot`
::::
:::::

# How to Use These Slides — Predict · Type · Run

::::: columns
:::: {.column width="60%"}
This lecture runs in **two short chunks**. After each chunk you switch to the **activity** and type the code yourself.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
::::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother? (the evidence)**

-   Predicting the *p*-value before you see it forces you to reason from the boxplot's separation, not just read a number off a screen.
-   Typing `type = "II"` yourself in `Anova()` is what makes you notice it's there — paste the line and you'll never ask why it matters for unbalanced groups.
-   Letters in a compact letter display only mean something if you've watched them get built from the pairwise table above them.
:::
::::
:::::

# 🧩 Chunk 1 of 2 · Run the ANOVA & Read the F Table

**We will cover:** the ANOVA table two ways — base `aov()` and `car::Anova()` — and why the difference matters for unbalanced data.

::: callout-tip
**🖐 After this chunk: Activity Parts 7–8** (run both, read F / df / p).
:::

# Run the ANOVA — `aov()` and `summary()`

::: callout-note
**🔮 Predict first:** The boxplot from Lecture 13 looked very separated. Predict the *p*-value — closer to 0.5, 0.05, or far below 0.001?
:::

::::: columns
:::: {.column width="60%"}
```{r load-libs-11}
# Load packages at the top — always -------------------
library(tidyverse)      # wrangling + ggplot2
library(palmerpenguins) # the penguins data
library(car)            # Anova()
library(emmeans)        # post-hoc pairwise tests
```

```{r refit-11}
# Same model fit at the end of Lecture 13 --------------
penguins_df <- penguins %>%
  drop_na(body_mass_g, species)

mass_model <- lm(body_mass_g ~ species, data = penguins_df)
```

```{r aov-11}
# Classic ANOVA table (Type I sums of squares) --------
mass_aov <- aov(body_mass_g ~ species, data = penguins_df)
summary(mass_aov)
```
::::

:::: {.column width="40%"}
**Reading the table:**

| Column | Meaning |
|-----------|-------------------------------|
| `Df` | groups − 1, and residual df |
| `F value` | between ÷ within variance |
| `Pr(>F)` | the *p*-value for H₀ |

*p* < 0.05 → **reject H₀** → at least one species differs. But **which**? Chunk 2.
::::
:::::

# Live Demo — Watch It Break (on purpose)

::::: columns
:::: {.column width="60%"}
For **unbalanced** data we prefer `car::Anova()`. I'll reach for it but type lowercase, and forget the package:

``` r
anova(mass_model)        # base R — Type I, one model at a time
Anova(mass_model)        # capital A — but car isn't loaded!
```

R stops:

```
Error in Anova(mass_model) :
  could not find function "Anova"
```

The fix — load `car`, and mind the capital **A**:

``` r
library(car)
Anova(mass_model, type = "II")
```
::::

:::: {.column width="40%"}
::: callout-tip
**✅ Why show a broken run?**

`anova()` and `Anova()` are two entirely different functions that happen to differ by one capital letter — R will not guess which one you meant. Watching the exact error message land here means you'll recognize it instantly instead of re-checking your spelling three times when it happens mid-assignment.
:::
::::
:::::

# `car::Anova()` — Type II for Unbalanced Data

::::: columns
:::: {.column width="60%"}
```{r carAnova-11}
# Type II sums of squares — the right call here -------
Anova(mass_model, type = "II")
```
::::

:::: {.column width="40%"}
**Why Type II?**

- Our groups have **different n** (unbalanced)
- With one predictor the F is the same, but Type II is the **habit** you want before you meet two-way ANOVA, where the type genuinely changes the answer

::: callout-note
`anova()` = Type I (order-dependent). `Anova(type = "II")` = order-independent.
:::
::::
:::::

# 🛑 Pause — Do Activity Parts 7–8 Now

Run `aov()` + `summary()`, then `car::Anova(type = "II")`. Predict the F and *p* before you read them, and confirm both tables agree.

# 🧩 Chunk 2 of 2 · Which Groups Differ? Post-Hoc with emmeans

**We will cover:** ANOVA says "some differ" — `emmeans` says *which*, with the comparisons properly adjusted.

::: callout-tip
**🖐 After this chunk: Activity Parts 9–11** (pairwise tests, letters, plot, report).
:::

# Post-Hoc — Pairwise Comparisons with emmeans

::: callout-note
**🔮 Predict first:** From the boxplot, predict — will **all three** species differ from each other, or will two be statistically tied?
:::

::::: columns
:::: {.column width="60%"}
```{r emmeans-11}
# Estimated marginal means + Tukey-adjusted pairs -----
mass_emm <- emmeans(mass_model, pairwise ~ species)

mass_emm$contrasts
```
::::

:::: {.column width="40%"}
**Reading it:**

- `emmeans(..., pairwise ~ species)` gives each group's mean **and** every pairwise comparison
- The *p*-values are **Tukey-adjusted** — they already correct for multiple comparisons, so the family-wise error stays at 0.05

📖 W&S §15.4 — planned vs. unplanned comparisons
::::
:::::

# Compact Letter Display — Who Shares a Letter?

::::: columns
:::: {.column width="60%"}
```{r cld-11}
# Groups that SHARE a letter are NOT different --------
multcomp::cld(mass_emm$emmeans, Letters = letters)
```
::::

:::: {.column width="40%"}
**How to read letters:**

- Same letter → **not** significantly different
- Different letters → significantly different

This is the compact summary you see in published figures — one letter above each bar.

::: callout-tip
Install once: `install.packages("multcomp")`
:::
::::
:::::

# Plot the Result — Means ± CI

::::: columns
:::: {.column width="60%"}
```{r emmeans-plot-11}
#| fig-format: png
#| fig-alt: "Estimated marginal mean body mass with 95 percent confidence interval error bars for each penguin species, showing Gentoo highest, Chinstrap and Adelie lower with non-overlapping intervals."
mass_emm_plot <- as.data.frame(mass_emm$emmeans) %>%
  ggplot(aes(x = species, y = emmean, color = species)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.15, linewidth = 0.9
  ) +
  labs(x = "Species", y = "Estimated mean mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")

mass_emm_plot
```
::::

:::: {.column width="40%"}
**Estimated marginal means** with 95% confidence intervals.

- Non-overlapping intervals hint at real differences
- Pair this plot with the letters for a publication-ready figure

The emmeans plot is the ANOVA cousin of Lecture 03's mean ± SE plot.
::::
:::::

# How to Report an ANOVA

::::: columns
:::: {.column width="60%"}
```{r report-11}
# Pull the F table values for reporting ---------------
mass_tab <- Anova(mass_model, type = "II")
mass_tab
```

> *"Body mass differed significantly among the three penguin species (one-way ANOVA: F(2, 339) = 343.6, p < 0.001). Tukey-adjusted comparisons showed all three species differed, with Gentoo heaviest and Adelie lightest."*
::::

:::: {.column width="40%"}
**Always include:**

- Test type (one-way ANOVA)
- **F**, both **df**, and the **p-value**
- The post-hoc result (which groups differ)
- Group means ± SE or CI

**Never** write "p = 0.000" — use "p < 0.001".
::::
:::::

# 🛑 Pause — Do Activity Parts 9–11 Now

Run the emmeans pairwise test, get the letters, plot the means ± CI, and write your results sentence. Predict which species differ before you run the comparisons.

# What We Learned Today

::::: columns
:::: {.column width="60%"}
**Concepts:**

- ANOVA rejects H₀ for the *group*, but **post-hoc** finds *which* pairs differ
- **Tukey adjustment** keeps the family-wise error at 0.05
- A compact letter display turns a pairwise table into one glance

**R skills:**

- `aov()` + `summary()`
- `car::Anova(model, type = "II")`
- `emmeans(model, pairwise ~ group)` + `multcomp::cld()`
::::

:::: {.column width="40%"}
**References:**

- 📖 W&S Ch. 15 — ANOVA
- 📖 [emmeans vignettes](https://cran.r-project.org/package=emmeans)

**Up next — Lecture 15:**

- Kicking off the **final project**
- Applying this whole toolkit — wrangling, plotting, and a test — to a dataset you choose
::::
:::::
