---
title: "Lecture 13 — Mapping: Where Is Bigfoot?"
subtitle: "Spatial data, sf, and geom_sf with real Bigfoot sighting reports"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
order: 13
week: 7
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  A fun 2-day intro to maps in R. Spatial data and coordinate systems, the sf
  package, turning latitude/longitude into a map with geom_sf(), drawing points
  on US states, and building a per-state choropleth by joining counts to the
  map — all with 4,500+ real Bigfoot sighting reports.
categories: [maps, sf, spatial, ggplot]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off — Factors & Joins

- **Factors** — reordered and relabeled categories for readable plots
- **Joins** — combined a measurements table with a second table by a shared key
- Both skills come back **today**: we'll join sighting counts onto a map

::: callout-note
**✅ Transition**

You can wrangle, summarize, model, and join. Today we put data **on a map** — the single most fun payoff in the toolkit. Our dataset: **4,000+ real Bigfoot sightings** across the US, each with a latitude and longitude. 🦶
:::

# Goals for today (a 2-day lecture)

::::: columns
:::: {.column width="60%"}
- Understand **spatial data** — points, polygons, and coordinate systems
- Meet **`sf`** — a data frame with a **geometry** column
- Turn latitude/longitude into a map with **`geom_sf()`**
- Draw sighting **points** on a map of US states
- Build a **choropleth** — sightings per state — by **joining** counts to the map
- Reproject to a proper US look
::::

:::: {.column width="40%"}
**Tools today:**

- `tidyverse`, `sf`, `maps`

**Data:**

- `bfro_reports_geocoded.csv` from Kaggle — save it to your `data/` folder

**Reference:**

- 📖 [R4DS + `sf` docs](https://r.geocompx.org/) (Geocomputation with R)
- 🌐 [sf package](https://r-spatial.github.io/sf/)

::: callout-tip
**Day 1:** points on a map. **Day 2:** choropleth + polish.
:::
::::
:::::

# How to Use These Slides — Predict · Type · Run

::::: columns
:::: {.column width="60%"}
This lecture runs in **four chunks over two days**. After each chunk you switch to the **activity** and type the code yourself.

**For every map, do three things:**

1.  **Predict** — before it draws, say where the points or the dark states will be
2.  **Type** the code by hand — do not copy-paste
3.  **Run** it and compare to your prediction
::::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

-   **Predicting the pattern** ("Pacific Northwest?") makes the map mean something when it appears.
-   **Typing** builds the `sf` + `geom_sf` muscle memory.
-   **Chunk → practice** keeps each new idea from piling up.
:::
::::
:::::

# 🧩 Chunk 1 of 4 · Spatial Data & the sf Object *(Day 1)*

**We will cover:** what spatial data is, coordinate systems, and turning latitude/longitude into an `sf` object.

::: callout-tip
**🖐 After this chunk: Activity Parts 1–3** (load the sightings, make them spatial).
:::

# What Is Spatial Data?

::::: columns
:::: {.column width="60%"}
Two shapes cover almost everything:

- **Points** — a single location: one Bigfoot sighting = (longitude, latitude)
- **Polygons** — an area: the outline of a state or country

Both need a **coordinate reference system (CRS)** — the rulebook that says what the numbers mean.

- **EPSG:4326** (WGS84) — plain longitude/latitude, what GPS and our CSV use
::::

:::: {.column width="40%"}
::: callout-note
**📖 New word**

**CRS** = Coordinate Reference System. Two maps won't line up unless they share one. Lat/long is `crs = 4326`.
:::
::::
:::::

# The `sf` Package — a Data Frame with Geometry

::::: columns
:::: {.column width="60%"}
```{r load-13}
# Load packages + the sighting data ------------------
library(tidyverse)
library(sf)       # simple features = spatial data frames
library(maps)     # ready-made state/country outlines

# Swap this path for your download -------------------
bigfoot_df <- read_csv("data/bfro_reports_geocoded.csv")

bigfoot_df %>% select(title, state, latitude, longitude, classification) %>% head()
```
::::

:::: {.column width="40%"}
- `sf` = "**s**imple **f**eatures" — the standard for spatial data in R
- An `sf` object is a normal tibble **plus a `geometry` column**
- Everything you know — `filter()`, `mutate()`, `group_by()` — still works
::::
:::::

# Clean the Coordinates First

::::: columns
:::: {.column width="60%"}
```{r clean-13}
# Drop missing coords + keep the lower 48 ------------
bigfoot_df <- bigfoot_df %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  filter(longitude > -125, longitude < -66,
         latitude  >   24, latitude  <  50)

nrow(bigfoot_df)
```
::::

:::: {.column width="40%"}
- Real coordinate data **always** has some missing values
- We also trim to the continental US so the map isn't dominated by a few far-flung points

::: callout-warning
**⚠️ Watch out!** The next step (`st_as_sf`) **errors** if any latitude/longitude is `NA` — clean first, always.
:::
::::
:::::

# Make It Spatial — `st_as_sf()`

::: callout-note
**🔮 Predict first:** We're about to turn 4,000+ rows of lat/long into points and plot them with **no basemap**. What shape do you think the cloud of points will make?
:::

::::: columns
:::: {.column width="60%"}
```{r make-sf-13}
#| fig-format: png
# Build the geometry column from lon/lat -------------
bigfoot_sf <- bigfoot_df %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Just the points — the USA appears on its own -------
ggplot(bigfoot_sf) +
  geom_sf(alpha = 0.2, size = 0.5) +
  theme_minimal()
```
::::

:::: {.column width="40%"}
- `coords = c("longitude", "latitude")` — **x first, then y**
- `crs = 4326` — tells sf these are lon/lat degrees
- The points alone trace the outline of the country — that's the "aha" of spatial data

📖 sf: `st_as_sf()`
::::
:::::

# 🛑 Pause — Do Activity Parts 1–3 Now

Load the sightings, clean the coordinates, make an `sf` object, and plot the bare points. Predict the shape before you run it.

# 🧩 Chunk 2 of 4 · Points on a Real Map *(Day 1)*

**We will cover:** adding a basemap of US states and layering the sightings on top, colored by class.

::: callout-tip
**🖐 After this chunk: Activity Parts 4–6** (draw the states + points).
:::

# Get a Basemap — US States as `sf`

::::: columns
:::: {.column width="60%"}
```{r states-13}
# Turn the built-in state outlines into sf ----------
states_sf <- st_as_sf(
  maps::map("state", plot = FALSE, fill = TRUE)
) %>%
  st_set_crs(4326)

states_sf %>% head(3)
```
::::

:::: {.column width="40%"}
- `maps::map("state", fill = TRUE)` — polygon outlines of the lower 48, no download
- `st_as_sf()` converts them to spatial features
- `st_set_crs(4326)` — same lat/long system as our points, so they line up

::: callout-note
The `ID` column holds the state name (lowercase) — we'll use it to join later.
:::
::::
:::::

# Layer the Sightings on the Map

::: callout-note
**🔮 Predict first:** Where do you expect the densest cluster of sightings — the Pacific Northwest, the Southeast, the Plains? Commit before it draws.
:::

::::: columns
:::: {.column width="60%"}
```{r points-map-13}
#| fig-format: png
# Basemap first, then points on top ------------------
ggplot() +
  geom_sf(data = states_sf, fill = "grey96", color = "grey75") +
  geom_sf(data = bigfoot_sf, aes(color = classification),
          alpha = 0.3, size = 0.6) +
  labs(title = "Bigfoot Sightings Across the Lower 48",
       color = "Report class") +
  theme_void()
```
::::

:::: {.column width="40%"}
- **Order matters:** draw states first, points second (on top)
- `aes(color = classification)` — colors Class A / B / C reports
- `theme_void()` — drop the axes and grid for a clean map

Notice the clusters — the Pacific Northwest lights up. 👀
::::
:::::

# 🛑 End of Day 1 · Start Day 2 Here

::::: columns
:::: {.column width="60%"}
**Day 1 recap — you can now:**

- Explain points, polygons, and CRS
- Turn lat/long into an `sf` object with `st_as_sf()`
- Draw a basemap of states and layer points with `geom_sf()`
::::

:::: {.column width="40%"}
**Day 2 — the choropleth:**

- Count sightings per state
- **Join** those counts onto the map
- Shade each state, then reproject for a proper US look
::::
:::::

# 🧩 Chunk 3 of 4 · Choropleth by State *(Day 2)*

**We will cover:** counting sightings per state and **joining** them to the map to shade each state.

::: callout-tip
**🖐 After this chunk: Activity Parts 7–9** (count, join, shade).
:::

# Step 1 — Count Sightings per State

::: callout-note
**🔮 Predict first:** Which state do you think has the most Bigfoot reports? Write your guess before you run `count()`.
:::

::::: columns
:::: {.column width="60%"}
```{r count-13}
# Sightings per state — lowercase to match the map --
sightings_by_state <- bigfoot_df %>%
  mutate(state = str_to_lower(state)) %>%
  count(state, name = "sightings") %>%
  arrange(desc(sightings))

head(sightings_by_state)
```
::::

:::: {.column width="40%"}
- `count(state)` — one row per state with its total
- `str_to_lower(state)` — the map's state names are **lowercase**, so we match them now
- This is a plain summary table — next we glue it to the geometry

📖 R4DS §16 — factors/strings for tidy keys
::::
:::::

# Live Demo — Watch It Break (a blank map)

::::: columns
:::: {.column width="60%"}
If I **skip** the lowercase step, the join keys don't match — `"Washington"` ≠ `"washington"`:

``` r
states_sf %>%
  mutate(state = ID) %>%
  left_join(count(bigfoot_df, state), by = "state")  # "Washington" vs "washington"
```

Every state gets `NA` sightings → the whole map renders **grey**. No error, just a blank choropleth.

The fix — make both keys the **same case** before joining (`str_to_lower()`).
::::

:::: {.column width="40%"}
::: callout-important
**✅ Why show a blank map?**

A join only matches keys that are **exactly** equal. `"Washington"`, `"washington"`, and `" Washington"` are three different keys. A silent all-`NA` result is almost always a **key-mismatch** — check case and spelling first.
:::
::::
:::::

# Step 2 — Join Counts onto the Map

::::: columns
:::: {.column width="60%"}
```{r choropleth-13}
#| fig-format: png
# Join sightings to the state polygons ---------------
states_map <- states_sf %>%
  mutate(state = str_remove(ID, ":.*")) %>%   # drop ":north" etc.
  left_join(sightings_by_state, by = "state")

ggplot(states_map) +
  geom_sf(aes(fill = sightings), color = "white") +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  labs(title = "Bigfoot Sightings by State", fill = "Sightings") +
  theme_void()
```
::::

:::: {.column width="40%"}
- `left_join()` keeps **every state**, filling in its count
- `str_remove(ID, ":.*")` merges split polygons (e.g. `michigan:north`)
- `geom_sf(aes(fill = sightings))` shades each state
- `scale_fill_viridis_c()` — a colorblind-safe gradient

This is a **choropleth** — the map *is* the bar chart.
::::
:::::

# 🛑 Pause — Do Activity Parts 7–9 Now

Count sightings per state, join them to the map, and shade the choropleth. Predict the top state before you run it.

# 🧩 Chunk 4 of 4 · Polish & Explore *(Day 2)*

**We will cover:** a proper US projection, and a fun exploration — sightings by season.

::: callout-tip
**🖐 After this chunk: Activity Parts 10–12** (reproject, then explore).
:::

# Reproject for a Proper US Look — `coord_sf()`

::::: columns
:::: {.column width="60%"}
```{r albers-13}
#| fig-format: png
# Albers Equal-Area (EPSG:5070) — the "textbook" US --
ggplot(states_map) +
  geom_sf(aes(fill = sightings), color = "white") +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  coord_sf(crs = 5070) +
  labs(title = "Bigfoot Sightings by State (Albers projection)",
       fill = "Sightings") +
  theme_void()
```
::::

:::: {.column width="40%"}
- Lat/long (`4326`) stretches the US; **Albers (`5070`)** is the shape you expect
- `coord_sf(crs = 5070)` reprojects **just for display** — your data is unchanged
- One line, and the map suddenly looks professional

::: callout-note
Different jobs use different CRSs — equal-area for choropleths, web maps use `3857`.
:::
::::
:::::

# Explore — Sightings by Season 🍂

::: callout-note
**🔮 Predict first:** In which season do you think the most sightings are reported? (Think about who is outdoors, and when.)
:::

::::: columns
:::: {.column width="60%"}
```{r season-13}
#| fig-format: png
# Facet the point map by season ----------------------
bigfoot_sf %>%
  filter(season %in% c("Spring","Summer","Fall","Winter")) %>%
  ggplot() +
  geom_sf(data = states_sf, fill = "grey96", color = "grey80") +
  geom_sf(alpha = 0.2, size = 0.4, color = "firebrick") +
  facet_wrap(~ season) +
  coord_sf(crs = 5070) +
  theme_void()
```
::::

:::: {.column width="40%"}
- `facet_wrap(~ season)` — one small map per season, same skill as any facet
- Summer dominates — more people outdoors = more reports (a sampling story!)
- This is how you'd explore **any** point dataset: filter, color, facet
::::
:::::

# What We Learned Today

::::: columns
:::: {.column width="60%"}
**Concepts:**

- Spatial data = **points/polygons + a CRS**
- `sf` objects are tibbles with a **geometry** column
- `st_as_sf(coords = c("lon","lat"), crs = 4326)` makes points
- `geom_sf()` draws layers; **order matters** (basemap first)
- A **choropleth** = counts **joined** to polygons, shaded by value
- `coord_sf(crs = ...)` reprojects for display

**R skills:**

- `st_as_sf()`, `st_set_crs()`, `maps::map()`
- `geom_sf()`, `scale_fill_viridis_c()`, `coord_sf()`
- `count()` + `left_join()` + `str_to_lower()` for the choropleth
::::

:::: {.column width="40%"}
**References:**

- 🌐 [sf package](https://r-spatial.github.io/sf/)
- 📖 [Geocomputation with R](https://r.geocompx.org/)

**Homework:**

- Download **your own** species from **GBIF** with the `rgbif` package and map where it lives — see the assignment.
::::
:::::
