---
title: "Lecture 03 — GGPlot I"
subtitle: "Grouped summaries, facets, and the pipe into ggplot"
author: "Bill Perry"
date: today

order: 3
week: 2
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Custom colors with scale_color_manual(), splitting plots into panels with
  facet_wrap()/facet_grid(), and building a mean ± SE plot with
  stat_summary() — all piped straight into ggplot().
categories: [ggplot, tidyverse]

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---



```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 02 — Meeting R)

- **Project structure** — `data/`, `scripts/`, `figures/`
- **Packages** — `install.packages()` once, `library()` every session
- **Loading data** — `read_excel()`, `clean_names()`
- **The pipe** — `%>%` reads as "then"
- **Our first plot** — a bare `geom_point()`

::: callout-note
**✅ Key idea from Lecture 02**

You can already load data and make one honest point on a page. Today that one geom becomes a whole toolkit: custom colors, side-by-side panels, and a proper mean ± SE summary.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- Start every plot with the pipe: `tree_df %>% ggplot(...)`
- Assign colors on purpose with `scale_color_manual()`
- Split one plot into panels with `facet_wrap()` and `facet_grid()`
- Build a **mean ± SE plot** with `stat_summary()`

::: callout-tip
**🖐 Try it yourself**

By the end you will make a publication-style figure: raw leaf data, grouped and colored on purpose, with the mean and its uncertainty layered on top.
:::
::::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 2 — *Displaying Data*
- 📖 [R4DS Ch 1](https://r4ds.hadley.nz/data-visualize) — Data Visualization
- 📖 [R4DS Ch 9](https://r4ds.hadley.nz/layers) — Layers

Both PDFs are in the course `readings/` folder.

**Naming conventions:**

- data frames → `_df`
- plots → `_plot`
:::
::::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This lecture runs in **three short chunks**. After each chunk you switch to the activity worksheet and type the code yourself.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

- `facet_wrap()` and `facet_grid()` look interchangeable until you've typed both and watched one wrap into rows while the other insists on a strict grid.
- Typing `scale_color_manual(values = c(...))` yourself is what makes you notice the color names have to match your group names *exactly*.
:::
::::
::::::

# Load Libraries and Data

```{r load-libs-03}
# Load all packages at the top of every script ----------
library(readxl) # reading Excel files
library(tidyverse) # data wrangling + ggplot2
library(janitor) # cleans up messy column names
```

```{r load-data-03}
# Read the leaf data from the data folder ---------------
tree_df <- read_excel("data/2026_06_25_tree_experiment_raw_data.xlsx") %>%
  clean_names()

glimpse(tree_df)
```

# 🧩 Chunk 1 of 3 · Piping into ggplot, on Purpose

**We will cover:** starting a plot with the pipe, and choosing colors with `scale_color_manual()`.

::: {.activitybanner}
**🛑 After this chunk you will do Activity Parts 1–2.**
:::

# Every Plot Starts the Same Way

::::: columns
::: {.column width="60%"}
```{r pipe-into-ggplot}
#| fig-format: png
#| fig-alt: "Boxplot of leaf weight in grams by tree side, with jittered points overlaid, using default ggplot colors."
# Pipe the data frame straight into ggplot() -----------
tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6) +
  labs(
    title = "Leaf Weight by Tree Side",
    x = "Side of Tree",
    y = "Leaf Weight (g)",
    fill = "Side"
  ) +
  theme_minimal()
```
:::

::: {.column width="40%"}
- `tree_df %>% ggplot(...)` reads as *"take `tree_df`, then plot it"* — the same pipe you've already used
- Everything after `ggplot()` is added with `+`, not `%>%` — **ggplot layers use plus, pipelines use the pipe**
- `fill = side` maps a **column** to color — ggplot chooses the colors *for* you

::: callout-warning
**⚠️ Watch out!**

`%>%` connects data-wrangling steps. `+` connects plot layers. Mixing them up is the single most common ggplot error.
:::
:::
:::::

# Choosing Colors on Purpose — `scale_color_manual()`

::: callout-note
**🔮 Predict first:** `scale_fill_manual()` needs one color per group. We have two groups (`sunny`, `shady`). How many colors do we need to supply?
:::

:::::: columns
::: {.column width="60%"}
```{r scale-manual-demo}
#| fig-format: png
#| fig-alt: "Boxplot of leaf weight in grams by tree side with jittered points, using custom gold and forest-green colors for sunny and shady sides respectively."
# scale_fill_manual() assigns exact colors to exact groups
tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6) +
  scale_fill_manual(values = c(
    "sunny" = "goldenrod",
    "shady" = "forestgreen"
  )) +
  labs(
    title = "Leaf Weight by Tree Side",
    x = "Side of Tree",
    y = "Leaf Weight (g)",
    fill = "Side"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
- `scale_fill_manual()` — for `fill =` aesthetics (boxes, bars, columns)
- `scale_color_manual()` — for `color =` aesthetics (points, lines)
- The **names inside `c()` must exactly match** the values in your data (`"sunny"`, `"shady"`) — case and spelling both count

::: callout-warning
**⚠️ Watch out!**

If a name in `values = c(...)` doesn't match a value in your data exactly, that group's color falls back to grey and ggplot prints a warning.
:::
::::
::::::

# → ACTIVITY 3 Parts 1–2 now

::: {.activitybanner}
**🛑 Do Activity Parts 1–2 now**

Pipe into `ggplot()`, then assign sunny/shady their own colors with `scale_fill_manual()`. Predict, type, run.
:::

# 🧩 Chunk 2 of 3 · Small Multiples with Facets

**We will cover:** `facet_wrap()` and `facet_grid()` — splitting one plot into a panel per group.

::: {.activitybanner}
**🛑 After this chunk you will do Activity Part 3.**
:::

# `facet_wrap()` — One Panel per Group

::::: columns
::: {.column width="60%"}
```{r facet-wrap-demo}
#| fig-format: png
#| fig-alt: "Two histograms of leaf weight in grams, one panel per tree side, arranged side by side by facet_wrap."
# facet_wrap() splits into one panel per level ---------
tree_df %>%
  ggplot(aes(x = weight_g, fill = side)) +
  geom_histogram(binwidth = 1, color = "white") +
  facet_wrap(~side) +
  scale_fill_manual(values = c("sunny" = "goldenrod", "shady" = "forestgreen")) +
  labs(x = "Leaf Weight (g)", y = "Count", title = "Weight Distribution by Side") +
  theme_minimal() +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
- `~side` reads as *"by side"* — the `~` is required
- Each panel gets its **own copy of the plot**, filtered to that group
- `facet_wrap(~side, scales = "free_y")` lets each panel's y-axis rescale independently

📖 [R4DS §9 — Layers](https://r4ds.hadley.nz/layers)
::::
::::::

# `facet_grid()` — a Strict Rows-and-Columns Grid

::::: columns
::: {.column width="60%"}
```{r facet-grid-demo}
#| fig-format: png
#| fig-alt: "Two histograms of leaf weight stacked in one column, one row per tree side, arranged by facet_grid."
# facet_grid(rows ~ cols) — use "." for "no grouping" --
tree_df %>%
  ggplot(aes(x = weight_g, fill = side)) +
  geom_histogram(binwidth = 1, color = "white") +
  facet_grid(side ~ .) +
  scale_fill_manual(values = c("sunny" = "goldenrod", "shady" = "forestgreen")) +
  labs(x = "Leaf Weight (g)", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
- `facet_grid(side ~ .)` stacks panels in **rows**; `facet_grid(. ~ side)` arranges them in **columns**
- The dot `.` means "nothing on this side of the grid"
- `facet_grid()` really shines with **two** grouping variables (`rows ~ cols`) — we only have one (`side`) today, but you'll use the two-variable form the moment your data has a second grouping column

::: callout-note
**💡 `facet_wrap` vs `facet_grid`**

Use `facet_wrap()` for one grouping variable — it wraps panels into a grid automatically. Use `facet_grid()` when you have two variables and want rows and columns to each mean something specific.
:::
::::
::::::

# → ACTIVITY 3 Part 3 now

::: {.activitybanner}
**🛑 Do Activity Part 3 now**

Build both a `facet_wrap()` and a `facet_grid()` version of the histogram. Predict, type, run.
:::

# 🧩 Chunk 3 of 3 · Mean ± SE with `stat_summary()`

**We will cover:** layering a group mean and its standard error directly on top of raw data — no summary table needed.

::: {.activitybanner}
**🛑 After this chunk you will do Activity Parts 4–5.**
:::

# `stat_summary()` — the Mean as a Layer

::::: columns
::: {.column width="60%"}
```{r stat-summary-mean}
#| fig-format: png
#| fig-alt: "Jittered points of leaf weight by tree side in gold and green, with a larger black point marking the mean for each side."
# stat_summary() computes the mean straight from the data
tree_df %>%
  ggplot(aes(x = side, y = weight_g, color = side)) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "black") +
  scale_color_manual(values = c("sunny" = "goldenrod", "shady" = "forestgreen")) +
  labs(
    title = "Leaf Weight with Group Means",
    x = "Side of Tree", y = "Leaf Weight (g)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
:::

::: {.column width="40%"}
- `stat_summary()` computes a statistic **from the raw data** and draws it — no pre-built summary table needed
- `fun = mean` — any function that returns one number (`mean`, `median`, `max`, …)
- `geom = "point"` — draw that number as a point
- Put `geom_jitter()` **before** `stat_summary()` so the mean sits on top of the raw points, not buried under them
:::
:::::

# `stat_summary()` — Mean ± SE

::: callout-note
**🔮 Predict first:** `fun.data = mean_se` returns three numbers per group instead of one (the mean, and the bar's top and bottom). What do you think those three are called?
:::

:::::: columns
::: {.column width="60%"}
```{r stat-summary-se}
#| fig-format: png
#| fig-alt: "Jittered points of leaf weight by tree side in gold and green with mean plus or minus one standard error error bars overlaid for each side."
# fun.data = returns THREE values: y, ymin, ymax -------
tree_mean_se_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, color = side)) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.15,
    linewidth = 0.9
  ) +
  scale_color_manual(values = c("sunny" = "goldenrod", "shady" = "forestgreen")) +
  labs(
    title = "Mean ± SE Leaf Weight by Tree Side",
    x = "Side of Tree", y = "Leaf Weight (g)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

tree_mean_se_plot
```
:::

:::: {.column width="40%"}
**Two `stat_summary()` layers, stacked:**

| call | what it draws |
|---|---|
| `fun = mean` | a point at the mean |
| `fun.data = mean_se` | error bars for ± 1 SE |

- Raw points in the **background** (`alpha = 0.35`)
- Mean as a **larger foreground point**
- Error bars = **± 1 SE**

📖 Whitlock & Schluter, Ch. 3 — *Describing Data*
::::
::::::

# → ACTIVITY 3 Parts 4–5 now

::: {.activitybanner}
**🛑 Go to Activity 3 — Parts 4–5**

Close the slides. Build your own mean ± SE plot, then save it to `figures/`.
:::

# What We Learned Today

::::: columns
::: {.column width="60%"}
- Pipe straight into a plot: `tree_df %>% ggplot(...)`
- `scale_color_manual()` / `scale_fill_manual()` — colors you choose, not colors ggplot guesses
- `facet_wrap(~var)` — one panel per group, wraps automatically
- `facet_grid(rows ~ cols)` — strict grid, best with two grouping variables
- `stat_summary()` — mean and mean ± SE, drawn directly from raw data
:::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 2 — *Displaying Data*
- 📖 [R4DS Ch 1](https://r4ds.hadley.nz/data-visualize) — Data Visualization
- 📖 [R4DS Ch 9](https://r4ds.hadley.nz/layers) — Layers

**Up next — Lecture 04, GGPlot II:**

- More geoms
- `coord_cartesian()` vs. `xlim()`
- Themes
:::
:::::
