---
order: 5
week: 3
# =====================================================================
#  YAML FRONT MATTER  —  the "recipe" for your report
#  Everything between the --- lines controls the finished document.
#  Change the values in "quotes"; leave the keys (the words before :) alone.
# =====================================================================
title: "Leaf Morphology: Sunny vs. Shady Sides of a Tree"
subtitle: "A reproducible report built in Quarto"
author: "Your Name Here"
date: today

format:
  docx:                     # <- this line makes a Microsoft Word (.docx) file
    toc: true               # add a Table of Contents
    toc-depth: 2            # include level-1 and level-2 headings in the TOC
    number-sections: true   # number your sections (1, 1.1, 2, ...)
    highlight-style: github # nice colours for any code you choose to show
    fig-width: 6            # default figure width (inches)
    fig-height: 4           # default figure height (inches)
    fig-format: png         # always PNG figures (our class standard)
    # reference-doc: custom-reference.docx   # (advanced) use your own Word styles

# These settings apply to EVERY code chunk unless a chunk overrides them.
execute:
  echo: false      # HIDE the code in the report — readers want results, not R
  warning: false   # hide warnings
  message: false   # hide package start-up messages
---

```{=html}
<!--
HOW TO USE THIS TEMPLATE
1. Save this file in your project's `documents/` folder.
2. Put your data file in the `data/` folder.
3. Edit the YAML above (your name, title).
4. Fill in the writing under each heading and run/adjust the code chunks.
5. Press Render (Ctrl/Cmd + Shift + K) to build the Word document.
Tip: keep chunks small and uniquely named so errors are easy to find.
6. Predict before you Render: guess what each number and figure will show, then
   check. Type the chunks yourself instead of pasting — YAML spacing and unique
   chunk labels are exactly where renders break.
-->
```

```{r libraries}
# --- Load all packages at the top, in ONE library chunk -----------------
library(readxl) # read Excel files
library(tidyverse) # wrangling + ggplot2
library(janitor) # clean_names()
```

```{r load-data}
# --- Read the leaf data from the data/ folder ---------------------------
# Swap in your own file name here:
leaf_df <- read_excel("data/2026_09_03_data_sci_leaf_area.xlsx") %>%
  clean_names()
```

# Introduction

Leaves on the shady side of a tree may grow **larger** to capture more of the limited light. In this study we measured leaf mass on the *sunny* and *shady* sides of a tree and tested whether the two sides differ.

- **Null hypothesis (H₀):** there is no difference in mean leaf mass between sides.
- **Alternate hypothesis (Hₐ):** mean leaf mass differs between sides.

# Methods

We recorded the mass (g) of leaves sampled from each side of the tree. All analysis was done in R with the tidyverse, and this report was produced in Quarto so that every number and figure below is computed directly from the raw data.

# Results

<!-- 🔮 PREDICT before you Render: what mean mass do you expect for shady vs sunny in Table 1? Guess, then render. (Comment lines don't appear in the Word output.) -->

```{r summary-table}
# --- Descriptive statistics for each side -------------------------------
summary_df <- leaf_df %>%
  group_by(shade) %>%
  summarize(
    n = sum(!is.na(mass_g)), # correct n (ignores NA)
    mean_mass = round(mean(mass_g, na.rm = TRUE), 3),
    sd_mass = round(sd(mass_g, na.rm = TRUE), 3),
    se_mass = round(sd_mass / sqrt(n), 3)
  )
```

Table 1 shows the summary statistics for each side of the tree.

```{r print-table}
#| echo: false
# Print the summary table into the report
knitr::kable(summary_df, caption = "Table 1. Leaf mass (g) by shade.")
```

<!-- 🔮 PREDICT before you Render: which side has the higher mean mass in Table 1, and by how much? -->

Read the mean mass for each side straight from the `mean_mass` column of Table 1.
Because the table is built from the data every time the document renders, the
numbers update themselves whenever the data file changes — nothing is typed by
hand.

```{r leaf-plot}
#| echo: false
#| fig-cap: "Figure 1. Leaf mass by shade, with individual leaves shown."
#| fig-alt: "Boxplot comparing leaf mass in grams between the shady and sunny sides of the tree, with individual leaf masses overlaid as jittered points and outliers hidden."
# --- Boxplot with the raw points overlaid -------------------------------
leaf_box_plot <- leaf_df %>%
  ggplot(aes(x = shade, y = mass_g, fill = shade)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.6,
    size = 2.5
  ) +
  labs(x = "Shade", y = "Leaf mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")

leaf_box_plot
```

# Discussion

Write one short paragraph here: did the data support the hypothesis? Were shady leaves heavier? Mention the size of the difference and anything you noticed in Figure 1. In Activity 10 you will add the formal t-test result to this section.

# References

- Wickham, H. *R for Data Science (2nd ed.)* — <https://r4ds.hadley.nz/>
- Data Carpentry — <https://datacarpentry.github.io/R-ecology-lesson/>
