---
title: "Lecture 07 — From Script to Report"
subtitle: "One Quarto file: read the data, plot it, describe it, test it — Word, PDF, and slides"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# order/week are auto-stamped from _schedule.yml by sync_order_from_schedule.R
order: 7
week: 4
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Why we write analyses as Quarto documents instead of loose scripts:
  YAML, Markdown, and code chunks with a chunk-option reference, then the
  whole leaf analysis start to finish — box plot, mean ± SE, histogram,
  summary statistics table, Welch's t-test — rendered to Word, PDF, and
  RevealJS slides from one file.
categories: [quarto, reproducibility, markdown]

# the starter file is NOT rendered by the site — it is copied as a download
resources:
  - downloads/leaf_report_skeleton.qmd
  - themes/*.R
  - data/2026_09_03_data_sci_leaf_area.xlsx

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,
                      message = FALSE)
options(width = 65)   # printed R output wraps at 65 too
library(readxl)
library(tidyverse)
library(janitor)
library(broom)
source("themes/r_themes_for_3_sizes.R")
source("themes/summary_stats_function.R")
```

# Where we left off (Lectures 03–06)

- **Read** the leaf data with `read_excel()` + `clean_names()`
- **Plotted** it — box plots, mean ± SE with `stat_summary()`, histograms
- **Described** it — `group_by()` + `summary_stats()`: n, mean, SD, SE, 95% CI
- Everything so far lives in a **`.R` script**

::: callout-note
**✅ Key idea**

You already know every piece of a real analysis. Today we put all of it — plus a t-test — into **one document a human can read**.
:::

# Goals for today

:::::: columns
:::: {.column width="60%"}
- Understand **why** a script alone is not enough
- Know **where** a Quarto file lives in your project
- Learn the three ingredients:
  - **YAML** — the recipe at the top (one switch hides all the code)
  - **Markdown** — formatted text, no clicking
  - **Code chunks** — R that runs, plus the **chunk options** that control it
- Take the leaf data **start to finish**: plots → summary table → **Welch's t-test**
- Render **one file** to **Word, PDF, and RevealJS slides**

::: callout-tip
**🖐 You will hand in**

`leaf_report.qmd`, `leaf_report.docx`, and `leaf_report.pdf`.
:::
::::

::: {.column width="40%"}
**Tools today:**

- Positron + Quarto (built in)
- `broom` — turns test output into a table (installed with tidyverse)

**Get the starter file:**

- 📥 [**`leaf_report_skeleton.qmd`**](downloads/leaf_report_skeleton.qmd){download="leaf_report.qmd"}

**References:**

- 📖 [R4DS Ch 28 — Quarto](https://r4ds.hadley.nz/quarto)
- 📖 [R4DS Ch 29 — Quarto formats](https://r4ds.hadley.nz/quarto-formats)
- 🌐 [Quarto chunk options](https://quarto.org/docs/computations/execution-options.html)
:::
::::::

# How to Use These Slides — Predict · Type · Render

:::::: columns
::: {.column width="60%"}
This lecture runs in **four short chunks**. After each chunk you switch to the **activity** and fill in your own report.

**For every snippet, do three things:**

1.  **Predict** — before you Render, say what the document will show
2.  **Type** it out by hand — do not copy-paste
3.  **Render** and compare to your prediction
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother? (the evidence)**

-   **Predicting first** forces you to *retrieve* how Markdown and chunks behave. The surprise when you're wrong is what makes it stick.
-   **Typing by hand** builds muscle memory for YAML indentation and chunk syntax — exactly where beginners slip.
-   **Chunk → immediate practice** keeps each new idea in working memory long enough to form a lasting schema.
:::
::::
::::::

# 🧩 Chunk 1 of 4 · Why Quarto, and Where the File Lives

**We will cover:** the pain of script-plus-copy-paste, literate programming, reproducibility, and where to save a `.qmd` so your paths just work.

::: callout-tip
**🖐 After this chunk: Activity Part 1** (save the starter file at the top of your project and Render it).
:::

# Part 1 · The problem with scripts

### What happens after the analysis?

::::::: columns
:::: {.column width="60%"}
Your `.R` script is great for *running* code. But when the lab report is due you have to:

- Run the script
- **Copy** each number into Word by hand
- **Export** every plot, then drag it in
- Re-do *all of it* when the data changes

::: callout-warning
**⚠️ Watch out!**

Every hand-copied number is a chance for a typo. Change one data point and the whole report is out of date.
:::
::::

:::: {.column width="40%"}
::: callout-note
**📊 Coming from Excel?**

This is the same pain as pasting a chart into Word, then editing the spreadsheet and forgetting to update the chart.
:::
::::
:::::::

# The idea: keep code and writing together

::::::: columns
:::: {.column width="60%"}
**Literate programming** — write the *story* and the *code* in the **same file**.

- Your explanation sits next to the code that made it
- Tables and plots are **computed**, never copied
- Press **Render** → a finished document appears

``` text
one .qmd file  →  Render  →  Word  /  PDF  /  slides
```

Change the data, Render again, and every table and figure **rebuilds itself**. That is what scientists mean by **reproducible**.
::::

:::: {.column width="40%"}
::: callout-note
**📖 New word**

**Quarto** = the tool that turns one plain-text `.qmd` file into a polished document.
:::

::: callout-note
**✅ Key idea**

A script *does* the analysis. A **Quarto document** does the analysis **and explains it** — and rebuilds itself on demand.
:::
::::
:::::::

# Where does the `.qmd` go? — the top of your project

::::::: columns
:::: {.column width="55%"}
``` text
leaf_project/
├── data/
│   └── 2026_09_03_data_sci_leaf_area.xlsx
├── figures/
├── scripts/
├── themes/
│   ├── r_themes_for_3_sizes.R
│   └── summary_stats_function.R
└── leaf_report.qmd      <- HERE
```

Quarto runs the code **from the folder the `.qmd` is saved in**. Save it at the **top** of the project and these paths work exactly as they do in your scripts:

``` r
read_excel("data/2026_09_03_data_sci_leaf_area.xlsx")
source("themes/summary_stats_function.R")
```
::::

:::: {.column width="45%"}
::: callout-note
**✅ Key idea**

`.qmd` at the top + short paths into subfolders = **no `here` package, no `../`**. Same paths in the Console, in your scripts, and in the report.
:::

::: callout-warning
**⚠️ Watch out!**

Save it inside `documents/` and `"data/..."` breaks — Quarto would look for `documents/data/`. You would need `"../data/..."` everywhere. Don't.
:::
::::
:::::::

# 🛑 Pause — Do Activity Part 1 Now

::: {.activitybanner}
**🛑 Do Activity Part 1 now**

Download [`leaf_report_skeleton.qmd`](downloads/leaf_report_skeleton.qmd){download="leaf_report.qmd"}, save it at the **top** of your project as `leaf_report.qmd`, and press **Render** before you change anything.
:::

# 🧩 Chunk 2 of 4 · Anatomy of a `.qmd`

**We will cover:** YAML (and the one line that hides all the code), Markdown, code chunks, and the chunk options you will use all term.

::: callout-tip
**🖐 After this chunk: Activity Parts 2–4** (YAML, Markdown, chunk options).
:::

# Three parts, top to bottom

::::::: columns
:::: {.column width="60%"}
Every Quarto document has the same three ingredients:

``` text
1. YAML front matter  ← the recipe (title, outputs, defaults)
2. Markdown text      ← your writing, formatted
3. Code chunks        ← R that runs and shows results
```

You already *read* these every day — this whole lecture is a `.qmd` file.
::::

:::: {.column width="40%"}
::: callout-note
**📖 New word**

**`.qmd`** = a **Q**uarto **m**ark**d**own file. Plain text you can open anywhere.
:::
::::
:::::::

# Ingredient 1 · YAML — the recipe

::: callout-note
**🔮 Predict first:** The `execute:` block below says `echo: true`. What changes in the whole document if you switch it to `false`?
:::

::::::: columns
:::: {.column width="60%"}
``` yaml
---
title: "Leaf Mass on the Sunny and Shady Sides of a Tree"
author: "Your Name"
date: today

format:            # one file -> three documents
  docx:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
    fig-pos: "H"
  revealjs:
    scrollable: true
    smaller: true

execute:           # defaults for EVERY chunk
  echo: true       # false = hide ALL the code at once
  warning: false
  message: false
---
```
::::

:::: {.column width="40%"}
- `format:` — **which** documents to build
- `execute:` — settings for **every chunk**; one line turns all the code on or off

::: callout-tip
**One switch**

Write with `echo: true` so you can see your code. Flip to `echo: false` for the version you hand in.
:::

::: callout-warning
**⚠️ Watch out!**

YAML is **space-sensitive**: two spaces, never tabs.
:::
::::
:::::::

# Ingredient 2 · Markdown text

::: callout-note
**🔮 Predict first:** Before it renders, sketch what this becomes — which line is a big heading, which are bullets, what turns bold?
:::

::::::: columns
:::: {.column width="60%"}
``` markdown
# Introduction

Leaves on the **shady** side of a tree may grow
larger to catch more of the limited light.

- **Null hypothesis:** mean mass is the same on both sides
- **Alternate hypothesis:** mean mass differs

<!-- a comment: never shows up in the output -->
```

- `#` heading, `##` smaller heading
- `**bold**`, `*italic*`, `-` bullets
- `<!-- -->` a note to yourself — the skeleton is full of them
::::

:::: {.column width="40%"}
::: callout-note
**📊 Coming from Word?**

`**bold**` is just the *keyboard* version of clicking the **B** button. Same result, no mouse.
:::

::: callout-note
**🎞 Bonus**

In the slide version, every `#` becomes a **section slide** and every `##` becomes a **slide**.
:::
::::
:::::::

# Ingredient 3 · Code chunks

::::::: columns
:::: {.column width="60%"}
A **code chunk** is fenced R that actually **runs** when you render:

```{{r}}
#| label: load-data
# read the leaf data from the data/ folder ----
leaf_df <-
  read_excel("data/2026_09_03_data_sci_leaf_area.xlsx") %>%
  clean_names()
```

- Opening fence ```` ```{r} ```` — closing fence ```` ``` ````
- Lines starting `#|` are **chunk options**
- Then a `#` comment saying what it does — same rule as your scripts
::::

:::: {.column width="40%"}
::: callout-tip
Insert a chunk: **Ctrl/Cmd + Alt + I**
:::

::: callout-warning
**⚠️ Watch out!**

Every `label` must be **unique**. Two chunks named `plot` stop the Render:

``` text
Duplicate chunk label 'plot'
```

The error names the label — go straight to it.
:::
::::
:::::::

# Chunk options — the reference card

| Option | What it does | Example |
|:-------------------|:-----------------------------------------------|:---------------------------|
| `label` | Names the chunk. `fig-` / `tbl-` at the front numbers it as a Figure / Table | `#| label: fig-boxplot` |
| `echo` | Show (`true`) or hide (`false`) the **code** | `#| echo: false` |
| `eval` | Run the code (`true`) or only show it (`false`) | `#| eval: false` |
| `include` | `false` runs the code but hides **both** code and output | `#| include: false` |
| `output` | `false` hides the **results** but keeps the code | `#| output: false` |
| `warning` / `message` | Show or hide warnings / package messages | `#| message: false` |
| `fig-width` / `fig-height` | Figure size in inches | `#| fig-width: 5` |
| `fig-cap` | Figure caption | `#| fig-cap: "Leaf mass by side"` |
| `tbl-cap` | Table caption | `#| tbl-cap: "Summary statistics"` |
| `tbl-colwidths` | Column widths in percent — tidy Word tables | `#| tbl-colwidths: [30, 70]` |

: The `execute:` block in the YAML sets the default for every chunk; a chunk's own `#|` line overrides it for that one chunk. {tbl-colwidths="[20,50,30]"}

# 🛑 Pause — Do Activity Parts 2–4 Now

::: {.activitybanner}
**🛑 Do Activity Parts 2–4 now**

Put your name in the YAML and try the `echo` switch; write the Introduction and Methods in Markdown; then use the reference card to hide the `setup` chunk. Predict each Render before you press it.
:::

# 🧩 Chunk 3 of 4 · Plot It, Then Describe It

**We will cover:** the three plots you already know — box plot, mean ± SE, histogram — as named chunks with numbered captions, then a summary statistics **table** that looks right in Word.

::: callout-tip
**🖐 Your skeleton already has a plain, working version of every chunk below.** You add the finishing touches — colors, labels, `theme_regular()`, table formatting — in **Activity Parts 5–6**.
:::

# Start at the top: load the data

::::::: columns
:::: {.column width="60%"}
```{r}
#| label: load-data
#| echo: fenced
# read the leaf data from the data/ folder ----
leaf_df <-
  read_excel("data/2026_09_03_data_sci_leaf_area.xlsx") %>%
  clean_names()
```

The `setup` chunk above it (already in your skeleton) loads `readxl`, `tidyverse`, `janitor`, `broom`, and `source()`s the theme file and `summary_stats()`.
::::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

Quarto runs the file **top to bottom in a fresh R session** — not whatever is sitting in your Console. Libraries and data must load in a chunk **above** anything that uses them.
:::
::::
:::::::

# Plot 1 · Box plot with the raw points

```{r}
#| label: fig-boxplot
#| echo: fenced
#| output-location: column
#| fig-cap: "Leaf mass by side; each point is one leaf."
#| fig-width: 5
#| fig-height: 4
# box plot with raw points on top ----
leaf_df %>%
  ggplot(aes(x = shade, y = mass_g, fill = shade)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6)
```

- 🔮 **Predict:** the label starts with `fig-` — what does the caption start with?
- ✏️ **You add:** your two colors, `labs()`, `theme_regular()`, no legend

# Plot 2 · Mean ± SE

```{r}
#| label: fig-mean-se
#| echo: fenced
#| output-location: column
#| fig-cap: "Mean leaf mass (± 1 SE) by side."
#| fig-width: 5
#| fig-height: 4
# mean +/- standard error plot ----
leaf_df %>%
  ggplot(aes(x = shade, y = mass_g, color = shade)) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.15, linewidth = 0.9)
```

- 🔮 **Predict:** the error bars overlap almost completely — what will the t-test say?
- ✏️ **You add:** the same four finishing lines — but this plot maps `color`, not `fill`

# Plot 3 · Histogram — the shape of each group

```{r}
#| label: fig-histogram
#| echo: fenced
#| output-location: column
#| fig-cap: "Distribution of leaf mass on each side."
#| fig-width: 6
#| fig-height: 3.5
# one histogram per side ----
leaf_df %>%
  ggplot(aes(x = mass_g, fill = shade)) +
  geom_histogram(binwidth = 0.05, color = "white") +
  facet_wrap(~shade)
```

- Wider and shorter (`6 × 3.5`) because it has **two panels**
- The box plot hides the shape; the histogram shows it
- ✏️ **You add:** colors, `labs()`, `theme_regular()`, no legend

# Describe it · a summary statistics table

::::::: columns
:::: {.column width="60%"}
```{r}
#| label: tbl-summary
#| echo: fenced
#| tbl-cap: "Summary statistics for leaf mass (g) by side."
#| tbl-colwidths: [12, 7, 12, 15, 12, 12, 15, 15]
# summary statistics by side, printed as a table ----
mass_stats_df <- leaf_df %>%
  group_by(shade) %>%
  summary_stats(mass_g)

mass_stats_df %>%
  knitr::kable(
    digits = 3,
    col.names = c("Side", "n", "Mean", "Variance",
                  "SD", "SE", "CI low", "CI high"))
```

::: {.content-visible when-format="pptx"}
```{r}
#| echo: false
# PowerPoint drops captioned tables, so show a plain copy there
knitr::kable(mass_stats_df, digits = 3)
```
:::
::::

:::: {.column width="40%"}
- `summary_stats()` — same helper as Activity 06
- `knitr::kable()` — data frame → **real table**
- `digits = 3` — rounds the whole table
- `col.names =` — readable headers, one per column
- `tbl-colwidths:` — **eight columns, eight percents**

In your text write `@tbl-summary` and Quarto prints **"Table 1"** for you.
::::
:::::::

# Why `tbl-colwidths`? Tables in Word

::::::: columns
:::: {.column width="55%"}
Without widths, Word **guesses** — and it often guesses badly: a fat first column and the last column squashed until the numbers wrap.

``` r
#| tbl-colwidths: [12, 7, 12, 15, 12, 12, 15, 15]
```

- One number per column, in order
- They are **percents** of the page width — aim for a total of 100
- Give short columns (like `n`) less, headers like `Variance` more
::::

:::: {.column width="45%"}
::: callout-note
**✅ Key idea**

`kable()` makes the table; `tbl-colwidths` makes it **line up** in Word and PDF. Same numbers, readable layout.
:::
::::
:::::::

# 🛑 Pause — Do Activity Parts 5–6 Now

::: {.activitybanner}
**🛑 Do Activity Parts 5–6 now**

Render the plain plots and table first, then finish them: colors, axis labels, `theme_regular()`, no legend, and `digits` + `col.names` in `kable()`. Render after **each** one.
:::

# 🧩 Chunk 4 of 4 · Test It, Then Render Everything

**We will cover:** a Welch's two-sample t-test (you met the t-test in intro bio), a clean results table, then hiding the code and rendering Word, PDF, and slides.

::: callout-tip
**🖐 After this chunk: Activity Parts 7–9** (t-test, render all three and hand in, then play).
:::

# The Welch's two-sample t-test

::::::: columns
:::: {.column width="60%"}
**Question:** is the difference between the two means bigger than we'd expect by chance?

```{r}
#| label: ttest
#| echo: fenced
# Welch's t-test (does NOT assume equal variances) ----
leaf_ttest <- t.test(mass_g ~ shade, data = leaf_df,
                     var.equal = FALSE)
leaf_ttest
```
::::

:::: {.column width="40%"}
- `mass_g ~ shade` — "mass **by** shade": number on the left, two groups on the right
- `var.equal = FALSE` — **Welch's** version: does **not** assume the two sides have the same variance
- Look for **t**, **df**, and **p-value** in the printout

::: callout-note
**📖 Why Welch's?**

It is safe whether or not the variances match, so it is R's default and a sensible first choice. Next lecture takes it apart properly.
:::
::::
:::::::

# The same result as a table

::::::: columns
:::: {.column width="60%"}
```{r}
#| label: tbl-ttest
#| echo: fenced
#| tbl-cap: "Welch's t-test of leaf mass by side of the tree."
#| tbl-colwidths: [20, 20, 15, 15, 15, 15]
# the same result as a clean table ----
leaf_ttest %>%
  tidy() %>%
  select(estimate1, estimate2, statistic, parameter,
         p.value, conf.low) %>%
  knitr::kable(
    digits = 3,
    col.names = c("Mean shady", "Mean sunny", "t",
                  "df", "p-value", "CI low"))
```

::: {.content-visible when-format="pptx"}
```{r}
#| echo: false
# PowerPoint drops captioned tables, so show a plain copy there
leaf_ttest %>%
  tidy() %>%
  select(estimate1, estimate2, statistic, parameter, p.value) %>%
  knitr::kable(digits = 3)
```
:::
::::

:::: {.column width="40%"}
- `tidy()` (from `broom`) turns the printout into a **data frame**
- then it is just `select()` + `kable()` — tools you already have

::: callout-note
**✅ Reading it**

*p* is well above 0.05, so we **fail to reject** the null: these data show no difference in mean leaf mass between sides — just what the overlapping error bars predicted. That is a real result; report it honestly.
:::
::::
:::::::

# Hide the code — one line

::::::: columns
:::: {.column width="55%"}
For the version you hand in, flip the switch in the YAML:

``` yaml
execute:
  echo: false      # hide ALL the code
  warning: false
  message: false
```

...and hide the raw t-test printout — the table already has those numbers. The chunk still **runs**, so `leaf_ttest` still exists:

```{{r}}
#| label: ttest
#| output: false
leaf_ttest <- t.test(mass_g ~ shade, data = leaf_df,
                     var.equal = FALSE)
leaf_ttest
```
::::

:::: {.column width="45%"}
::: callout-note
**✅ Key idea**

`execute:` in the YAML = the **default for every chunk**. A chunk's own `#|` line = the **exception** for that one chunk.
:::
::::
:::::::

# One file → Word, PDF, and slides

::::::: columns
:::: {.column width="55%"}
The **Render** button previews **one** format. To build **everything** listed under `format:`, use the **Terminal** tab (not the Console):

``` bash
quarto render leaf_report.qmd
```

| File | What it is |
|:-----------------------|:-----------------------------------------------|
| `leaf_report.docx` | Word report — hand in |
| `leaf_report.pdf` | PDF report — hand in |
| `leaf_report.html` | RevealJS slide show — press **F** for full screen |

: {tbl-colwidths="[35,65]"}
::::

:::: {.column width="45%"}
::: callout-warning
**⚠️ PDF needs LaTeX — once per computer**

``` bash
quarto install tinytex
```
:::

::: callout-note
**🎞 RevealJS**

The same report becomes a web-based slide show: each `#` is a section slide, each `##` a slide. These lecture slides are made exactly this way.
:::
::::
:::::::

# Making things fit on a slide (RevealJS)

These work in **any** Quarto presentation — they only change the slides, never the Word or PDF.

| Control | Where it goes | What it does |
|:--------------------------|:------------------|:----------------------------------------|
| `output-location: column` | chunk `#|` line | code on the left, result on the right |
| `output-location: slide` | chunk `#|` line | result gets its own slide after the code |
| `output-location: fragment` | chunk `#|` line | result appears on the next click |
| `fig-width` / `fig-height` | chunk `#|` line | a shorter figure fits under more text |
| `echo: false` | chunk or `execute:` | hide the code — just show the result |
| `smaller: true` | under `revealjs:` | smaller text on every slide |
| `scrollable: true` | under `revealjs:` | a long slide scrolls instead of being cut off |
| `## Title {.smaller}` | on one heading | smaller text on **that** slide only |
| `:::: columns` | in the Markdown | put any two things side by side |

: {tbl-colwidths="[30,22,48]"}

::: callout-tip
Put `output-location: column` under `revealjs:` in the YAML and **every** chunk in your slides gets code-left, plot-right.
:::

# 🛑 Pause — Do Activity Parts 7–8 Now

::: {.activitybanner}
**🛑 Do Activity Parts 7–8 now**

Read the Welch's t-test, finish its table, write a short Discussion, set `echo: false`, then `quarto render leaf_report.qmd` in the Terminal. Hand in the `.qmd`, `.docx`, and `.pdf` — then **play** (Activity Part 9).
:::

# Wrap-up · What you can now do

::::::: columns
:::: {.column width="60%"}
- Explain **why** a Quarto report beats a script plus copy-paste
- Save a `.qmd` at the **top of the project** so `data/...` paths just work
- Use the **YAML** to pick outputs and switch all code on/off with `echo`
- Write **Markdown** and **named chunks** with the options you need
- Take data **start to finish**: plots → summary table → Welch's t-test
- Render **one file** to **Word, PDF, and RevealJS slides**
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

From here on, every analysis can be a document that rebuilds itself from the raw data.
:::

::: callout-tip
**🧰 Keep `leaf_report.qmd`**

It is your template for every analysis this term: copy it, change the data file and column names, rewrite the text, Render.
:::

**Up next — T-Tests I:**

- Where the *t*-statistic comes from, and why Welch's is the safe default
- Checking the assumptions first — normality (Shapiro–Wilk) and equal variances (Levene's)
- ...written up in your Quarto report
::::
:::::::
