---
order: 5
week: 3
# =====================================================================
#  t_test_report.qmd
#  A WORKED EXAMPLE for the Quarto intro (Lecture 07).
#  This is the analysis from Activities 09–10 (the Welch's t-test) written
#  up as a Quarto report. Open it after the Quarto lecture to SEE why we
#  bother: the same t-test you ran in a script becomes a finished Word
#  document, with the t-statistic, p-value, and means dropped straight
#  into the tables — nothing copied by hand, whatever the test says.
# =====================================================================
title: "Do Shady Leaves Weigh More? A Two-Sample t-Test"
subtitle: "A reproducible write-up of the Activities 09–10 analysis"
author: "Your Name Here"
date: today

format:
  docx:
    toc: true
    number-sections: true
    fig-width: 6
    fig-height: 4
    fig-format: png
  html: default

# Hide the code so the report reads like a report.
# Set echo: true if you want to show the code as a teaching handout.
execute:
  echo: false
  warning: false
  message: false
---

```{r libraries}
# Load all packages in one chunk at the top ----------------------------
library(readxl)
library(tidyverse)
library(janitor)
library(broom)   # tidy() turns the t-test object into a data frame
```

```{r load-data}
# Read the leaf data --------------------------------------------------
leaf_df <- read_excel("data/2026_09_03_data_sci_leaf_area.xlsx") %>%
  clean_names()
```

<!--
HOW TO READ THIS WORKED EXAMPLE (active reading)
Don't just skim it. Before you press Render, PREDICT what each computed value and
figure will show, then render and compare. Better still, retype a chunk or two by
hand. The PREDICT comments below mark good places to pause. None of these comment
lines appear in the rendered Word document.
-->

# Introduction

Leaves on the **shady** side of a tree may grow larger and heavier to capture
more of the limited light beneath the canopy. We measured the mass (g) of leaves
from the sunny and shady sides of a tree and tested whether mean leaf mass
differs between sides.

- **Null hypothesis (H₀):** $\mu_{shady} = \mu_{sunny}$ — no difference in mean mass
- **Alternate hypothesis (Hₐ):** $\mu_{shady} \neq \mu_{sunny}$ — the means differ

We used a two-tailed test at $\alpha = 0.05$.

# Methods

We compared the two sides with a **Welch's two-sample t-test**
(`var.equal = FALSE`), which does not assume equal variances. Before testing we
checked normality within each group with a Shapiro–Wilk test. All analysis was
done in R with the tidyverse, and this report was produced in Quarto so that
every value below is computed directly from the raw data.

# Results

<!-- 🔮 PREDICT before you Render: what mean mass do you expect for shady vs sunny leaves in Table 1? Jot a guess, then render and compare. -->

```{r summary-table}
# Descriptive statistics per side --------------------------------------
summary_df <- leaf_df %>%
  group_by(shade) %>%
  summarize(
    n = sum(!is.na(mass_g)),
    mean_mass = round(mean(mass_g, na.rm = TRUE), 3),
    sd_mass = round(sd(mass_g, na.rm = TRUE), 3),
    se_mass = round(sd_mass / sqrt(n), 3)
  )
```

```{r print-table}
knitr::kable(
  summary_df,
  caption = "Table 1. Leaf mass (g) by shade (n, mean, SD, SE)."
)
```

<!-- 🔮 PREDICT before you Render: what t, df, and p-value will Table 2 show, and which side is heavier? Write your guesses, then render and check. -->

```{r run-test}
# Welch's two-sample t-test --------------------------------------------
leaf_ttest_model <- t.test(
  mass_g ~ shade,
  data = leaf_df,
  var.equal = FALSE,
  alternative = "two.sided"
)

# t and p for the figure subtitle below (not used in the text)
t_val <- round(leaf_ttest_model$statistic, 2)
p_val <- signif(leaf_ttest_model$p.value, 2)

leaf_ttest_model |>
  tidy() |>
  select(estimate1, estimate2, statistic, parameter, p.value) |>
  knitr::kable(digits = c(3, 3, 2, 1, 20),
               col.names = c("Mean (shady)", "Mean (sunny)", "t",
                             "df", "p-value"),
               caption = "Table 2. Welch's two-sample t-test of leaf mass by shade.")
```

The group means, SD, SE, and *n* are in Table 1; the *t*, df, and *p*-value of the
Welch's test are in Table 2. The two means are very close, the 95% confidence
interval for the difference includes zero, and the *p*-value is **above** the 0.05
threshold — so we do **not** find a significant difference in mean leaf mass
between the sunny and shady sides.

::: {.callout-note appearance="simple"}
Every number in Tables 1 and 2 is computed from the raw data when the document
renders — nothing was typed or pasted. If a leaf measurement changed, you would
fix the data file, press **Render**, and both tables would update themselves —
and so would the sentence above, if you wrote it with inline code. In a plain
script you would retype all of these by hand.
:::

```{r boxplot}
#| fig-cap: "Figure 1. Leaf mass by shade. Points show individual leaves; the test result is printed in the subtitle."
#| fig-alt: "Boxplot comparing leaf mass in grams between the shady and sunny sides of the tree, with individual leaf masses overlaid as jittered points. The subtitle reports the Welch's t-test statistic and p-value."
leaf_box_plot <- leaf_df %>%
  ggplot(aes(x = shade, y = mass_g, fill = shade)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.6,
    size = 2.5
  ) +
  labs(
    subtitle = paste0("Welch's t-test: t = ", t_val, ", p = ", p_val),
    x = "Shade",
    y = "Leaf mass (g)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

leaf_box_plot
```

# Conclusion

At $\alpha = 0.05$ we **fail to reject the null hypothesis**: this sample does not
show a difference in mean leaf mass between the sunny and shady sides of the tree.
That is not proof that the sides are identical — only that any difference, if one
exists, was too small to detect with this sample. The biological prediction that
shade leaves grow heavier is **not supported** by these data.
*(Add a sentence of your own interpretation here — e.g. would a larger sample, or
a different measure such as leaf area, tell a different story?)*

# References

- Whitlock, M. & Schluter, D. *The Analysis of Biological Data* (2nd ed.).
- Wickham, H. *R for Data Science (2nd ed.)* — <https://r4ds.hadley.nz/>
