---
order: 5
week: 3
# =====================================================================
#  t_test_report.qmd
#  A WORKED EXAMPLE for the Quarto intro (Lecture 4.5).
#  This is the analysis from Worksheets 09–10 (the Welch's t-test) written
#  up as a Quarto report. Open it after the Quarto lecture to SEE why we
#  bother: the same t-test you ran in a script becomes a finished Word
#  document, with the t-statistic, p-value, and means dropped straight
#  into the sentences — nothing copied by hand.
# =====================================================================
title: "Do Shady Leaves Weigh More? A Two-Sample t-Test"
subtitle: "A reproducible write-up of the Worksheets 09–10 analysis"
author: "Your Name Here"
date: today

format:
  docx:
    toc: true
    number-sections: true
    fig-width: 6
    fig-height: 4
    fig-format: png
  html: default

# Hide the code so the report reads like a report.
# Set echo: true if you want to show the code as a teaching handout.
execute:
  echo: false
  warning: false
  message: false
---

```{r libraries}
# Load all packages in one chunk at the top ----------------------------
library(readxl)
library(tidyverse)
library(broom)   # tidy() turns the t-test object into a data frame
```

```{r load-data}
# Read the leaf data ---------------------------------------------------
tree_df <- read_excel("data/2026_06_25_tree_experiment_raw_data.xlsx")
```

<!--
HOW TO READ THIS WORKED EXAMPLE (active reading)
Don't just skim it. Before you press Render, PREDICT what each computed value and
figure will show, then render and compare. Better still, retype a chunk or two by
hand. The PREDICT comments below mark good places to pause. None of these comment
lines appear in the rendered Word document.
-->

# Introduction

Leaves on the **shady** side of a tree may grow larger and heavier to capture
more of the limited light beneath the canopy. We measured the weight (g) of ten
leaves from each side of a tree and tested whether mean leaf weight differs
between sides.

- **Null hypothesis (H₀):** $\mu_{shady} = \mu_{sunny}$ — no difference in mean weight
- **Alternate hypothesis (Hₐ):** $\mu_{shady} \neq \mu_{sunny}$ — the means differ

We used a two-tailed test at $\alpha = 0.05$.

# Methods

We compared the two sides with a **Welch's two-sample t-test**
(`var.equal = FALSE`), which does not assume equal variances. Before testing we
checked normality within each group with a Shapiro–Wilk test. All analysis was
done in R with the tidyverse, and this report was produced in Quarto so that
every value below is computed directly from the raw data.

# Results

<!-- 🔮 PREDICT before you Render: what mean weight do you expect for shady vs sunny leaves in Table 1? Jot a guess, then render and compare. -->

```{r summary-table}
# Descriptive statistics per side --------------------------------------
summary_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    n = sum(!is.na(weight_g)),
    mean_wt = round(mean(weight_g, na.rm = TRUE), 2),
    sd_wt = round(sd(weight_g, na.rm = TRUE), 2),
    se_wt = round(sd_wt / sqrt(n), 2)
  )
```

```{r print-table}
knitr::kable(
  summary_df,
  caption = "Table 1. Leaf weight (g) by side of tree (mean, SD, SE)."
)
```

<!-- 🔮 PREDICT before you Render: what t, df, and p-value will Table 2 show, and which side is heavier? Write your guesses, then render and check. -->

```{r run-test}
# Welch's two-sample t-test --------------------------------------------
leaf_ttest_model <- t.test(
  weight_g ~ side,
  data = tree_df,
  var.equal = FALSE,
  alternative = "two.sided"
)

# t and p for the figure subtitle below (not used in the text)
t_val <- round(leaf_ttest_model$statistic, 2)
p_val <- signif(leaf_ttest_model$p.value, 2)

leaf_ttest_model |>
  tidy() |>
  select(estimate1, estimate2, statistic, parameter, p.value) |>
  knitr::kable(digits = c(2, 2, 2, 1, 20),
               col.names = c("Mean (shady)", "Mean (sunny)", "t",
                             "df", "p-value"),
               caption = "Table 2. Welch's two-sample t-test of leaf weight by side.")
```

Shady-side leaves were significantly **heavier** than sunny-side leaves: the
group means, SE, and *n* are in Table 1, and the *t*, df, and *p*-value of the
Welch's test are in Table 2. The *p*-value is below the 0.05 threshold, so the
difference is significant.

::: {.callout-note appearance="simple"}
Every number in Tables 1 and 2 is computed from the raw data when the document
renders — nothing was typed or pasted. If a leaf measurement changed, you would
fix the data file, press **Render**, and both tables would update themselves. In
a plain script you would retype all of these by hand.
:::

```{r boxplot}
#| fig-cap: "Figure 1. Leaf weight by side of tree. Points show individual leaves; the test result is printed in the subtitle."
#| fig-alt: "Boxplot comparing leaf weight in grams between the shady and sunny sides of the tree, with individual leaf weights overlaid as jittered points. The subtitle reports the Welch's t-test statistic and p-value."
leaf_box_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.6,
    size = 2.5
  ) +
  labs(
    subtitle = paste0("Welch's t-test: t = ", t_val, ", p = ", p_val),
    x = "Side of tree",
    y = "Leaf weight (g)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

leaf_box_plot
```

# Conclusion

At $\alpha = 0.05$ we **reject the null hypothesis**: mean leaf weight differs
between the sunny and shady sides of the tree, with shady leaves heavier. This
matches the biological prediction that shade leaves grow larger to capture more
of the limited light. *(Add a sentence of your own interpretation here.)*

# References

- Whitlock, M. & Schluter, D. *The Analysis of Biological Data* (2nd ed.).
- Wickham, H. *R for Data Science (2nd ed.)* — <https://r4ds.hadley.nz/>
