---
title: "Lecture — Telling a Story with ggplot"
subtitle: "From a plot that shows everything to a figure that says one thing"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# DRAFT MODULE — not yet wired into _schedule.yml. Suggested slot: after
# common_code 07 (Advanced Plotting) and before the Final Project intro
# (module 08), since the final talk needs exactly this skill. Renumber
# when you fold this in.
order: 8.5
week: 4
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  A third ggplot lecture focused on communication, not new geoms: choosing
  one figure, decluttering, direct labeling, deliberate color and ordering,
  and writing a title that states the finding.
categories: [ggplot2, data-storytelling]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off

- **Basic ggplot** — `geom_boxplot()`, `geom_point()`, `position_jitter()`
- **Advanced plotting** — `facet_wrap()`, `stat_summary()`, `geom_smooth()`, patchwork, log scales
- You can now build almost any figure ggplot is capable of drawing

::: callout-note
**✅ Key idea from Advanced Plotting**

You now have the *mechanics*. Today is different: no new geoms. Today is about **choosing** which mechanics to use, and which to leave out, so your one figure makes one point.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
**The problem this lecture solves:**

- Your final-project talk is 5 minutes and **one figure**
- A technically-correct plot can still fail to communicate
- "I made a boxplot" is not the same as "I told a story"

**What we'll practice:**

- Start from the question, not the geom
- Cut everything that doesn't serve the point
- Direct labeling instead of a legend, when you can
- Order categories to match the story, not the alphabet
- Color with intent, and with accessibility in mind
- A title that states the finding, not the variable names
- `annotate()` to point at the one thing that matters

::: callout-tip
**🖐 Try it yourself**

By the end you will redesign a technically-correct-but-confusing plot into one that a stranger understands in five seconds.
:::
::::

::: {.column width="40%"}
**Our tools today:** mostly tools you already know, used differently.

**References:**

- 📖 [R4DS Ch 9](https://r4ds.hadley.nz/layers) — Layers
- 📖 [R4DS Ch 11](https://r4ds.hadley.nz/communication) — Communication
- 🌐 [A Compendium of Clean Graphs in R](http://shinyapps.org/apps/RGraphCompendium/index.php)

**Naming conventions:**

- plots → `_plot`
- a redesigned plot → `_plot_v2`
:::
::::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
Two chunks today. After each, switch to the activity and redesign a plot yourself.

1.  **Predict** what's wrong with a plot *before* I tell you
2.  **Type** the redesign by hand
3.  **Compare** your five-second read of the before vs. the after
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why this lecture matters**

Every remaining lecture in this course ends with a plot. This is the only lecture that tells you how to **judge** whether that plot is any good — a skill no geom teaches by itself.
:::
::::
::::::

# Load Libraries and Data

```{r load-libs-story}
# Load all packages at the top of every script ----------
library(readxl) # reading Excel files
library(tidyverse) # data wrangling + ggplot2
library(janitor) # cleans up messy column names
```

```{r load-data-story}
# Reuse our cleaned leaf data ----------------------------
tree_df <- read_excel("data/2026_06_25_tree_experiment_raw_data.xlsx") %>%
  clean_names()
```

# 🧩 Chunk 1 of 2 · The Rule — One Figure, One Point

**We will cover:** starting from the question, and the "five-second test."

::: callout-tip
**🖐 After this chunk: Activity Parts 1–3.**
:::

# Start From the Question, Not the Geom

::::: columns
::: {.column width="60%"}
Before you write one line of `ggplot()`, answer this in a full sentence:

> "This figure shows that **\_\_\_\_**."

If you cannot finish that sentence, you are not ready to plot yet — you are still exploring. Exploring is fine, but exploratory plots and **the** final figure are not the same plot.

**Our sentence today:**

> "Leaves on the shady side of the tree are heavier than leaves on the sunny side."
:::

::: {.column width="40%"}
::: callout-note
**✅ Key idea**

A dozen exploratory plots is normal and good. Your final talk gets **one**. Decide which one *earns* that slot by asking whether it finishes the sentence above.
:::
:::
:::::

# The Five-Second Test

::::: columns
::: {.column width="60%"}
Show your plot to a labmate for **five seconds**, then hide it. Ask: "What did that plot say?"

- If they repeat your sentence back — the plot works
- If they describe the *axes* instead of the *finding* — it doesn't yet
- If they ask "wait, what am I looking at?" — start over
:::

::: {.column width="40%"}
::: callout-tip
This is the same test we asked you to plan for in the final-project worksheet: *"Slide 3 — the ONE figure that shows your result."* Today is where you learn to build that slide.
:::
::::
:::::

# A Plot That Shows Everything (and Says Nothing)

::: callout-note
**🔮 Predict first:** Before you look at the code, predict what will be wrong with this plot even though every line of code is correct.
:::

::::: columns
::: {.column width="60%"}
```{r bad-plot-demo}
#| fig-format: png
# Technically correct — but a poor final figure ---------
tree_busy_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot() +
  geom_jitter(width = 0.3, size = 1) +
  facet_wrap(~ side) +
  labs(title = "weight_g by side", x = "side", y = "weight_g") +
  theme_gray()

tree_busy_plot
```
:::

::: {.column width="40%"}
::: callout-warning
**⚠️ What's wrong here**

- Faceting by the **same** variable that's on the x-axis — pure redundancy
- Legend repeats what the x-axis already says
- Title is a variable name, not a finding
- Axis labels are raw column names, not readable text
- Default gray background adds visual noise
:::
:::
:::::

# 🧩 Chunk 2 of 2 · The Fixes

**We will cover:** cutting redundancy, direct labeling, deliberate ordering and color, and a finding-based title.

::: callout-tip
**🖐 After this chunk: Activity Parts 4–7.**
:::

# Fix \#1 — Cut Anything That Repeats Itself

::::: columns
::: {.column width="60%"}
```{r fix-redundancy}
#| fig-format: png
# Removed the redundant facet_wrap ----------------------
tree_plot_v2 <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
  theme_minimal()

tree_plot_v2
```
:::

::: {.column width="40%"}
::: callout-note
**✅ Key idea**

If two elements of your plot show the same information twice (facet + x-axis, legend + x-axis, title + axis label), delete one. Redundancy is not clarity — it's clutter that *looks* like effort.
:::
:::
:::::

# Fix \#2 — Direct Labels Instead of a Legend

::: callout-note
**🔮 Predict first:** Our x-axis already says "sunny" and "shady." Do we need the `fill = side` legend at all?
:::

::::: columns
::: {.column width="60%"}
```{r fix-legend}
#| fig-format: png
# Drop the legend — the x-axis already tells us the group
tree_plot_v3 <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
  theme_minimal() +
  theme(legend.position = "none")

tree_plot_v3
```
:::

::: {.column width="40%"}
- A legend forces the reader's eye to travel back and forth
- When the x-axis (or a direct label) already names the groups, the legend is dead weight
- **Rule of thumb:** if removing the legend loses zero information, remove it
:::
:::::

# Fix \#3 — Order With Intent, Not the Alphabet

::::: columns
::: {.column width="60%"}
```{r fix-order}
#| fig-format: png
# fct_relevel() puts categories in a meaningful order ---
tree_plot_v4 <- tree_df %>%
  mutate(side = fct_relevel(side, "sunny", "shady")) %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
  theme_minimal() +
  theme(legend.position = "none")

tree_plot_v4
```
:::

::: {.column width="40%"}
- Alphabetical order (`shady` before `sunny`) is an accident of spelling, not a decision
- `fct_relevel()` lets you choose the order that matches your story: baseline first, then the comparison
- Order becomes especially important with **3+ groups** — put them in a logical sequence (low to high, control then treatments, etc.)

📖 [R4DS Ch 17 — Factors](https://r4ds.hadley.nz/factors)
:::
:::::

# Fix \#4 — Color With Intent (and Colorblind Safety)

::::: columns
::: {.column width="60%"}
```{r fix-color}
#| fig-format: png
# Choose two colors deliberately, not ggplot's defaults -
tree_plot_v5 <- tree_df %>%
  mutate(side = fct_relevel(side, "sunny", "shady")) %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
  scale_fill_manual(values = c(sunny = "#E69F00", shady = "#0072B2")) +
  theme_minimal() +
  theme(legend.position = "none")

tree_plot_v5
```
:::

::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

Red/green combinations are unreadable to \~8% of men with red-green color blindness. `#E69F00` (orange) and `#0072B2` (blue) are from the **Okabe-Ito palette**, designed to stay distinguishable for colorblind readers.
:::

::: callout-tip
Colour should encode the group your reader most needs to compare — not just "make it look nice."
:::
:::
:::::

# Fix \#5 — A Title That States the Finding

::: callout-note
**🔮 Predict first:** Compare `"weight_g by side"` to `"Shady leaves are heavier than sunny leaves"`. Which one lets someone understand the plot *without ever looking at the axes*?
:::

::::: columns
::: {.column width="60%"}
```{r fix-title}
#| fig-format: png
# The title states the finding; labs() explains the rest
tree_final_plot <- tree_df %>%
  mutate(side = fct_relevel(side, "sunny", "shady")) %>%
  ggplot(aes(x = side, y = weight_g, fill = side)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
  scale_fill_manual(values = c(sunny = "#E69F00", shady = "#0072B2")) +
  labs(
    title = "Shady-side leaves are heavier than sunny-side leaves",
    subtitle = "Each point is one leaf; boxes show the median and interquartile range",
    x = NULL,
    y = "Leaf weight (g)",
    caption = "n = 10 leaves per side"
  ) +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))

tree_final_plot
```
:::

::: {.column width="40%"}
- **Title** = the finding, in plain English
- **Subtitle** = one sentence of method/context
- **Caption** = sample size or data source — small print, not the headline
- `x = NULL` — when categories are self-explanatory, drop a redundant axis label

📖 [R4DS Ch 11 — Communication](https://r4ds.hadley.nz/communication)
:::
:::::

# Fix \#6 — `annotate()` to Point at the One Thing

::::: columns
::: {.column width="60%"}
```{r fix-annotate}
#| fig-format: png
# annotate() draws a fixed element, not tied to the data
mean_shady <- tree_df %>% filter(side == "shady") %>%
  summarize(m = mean(weight_g, na.rm = TRUE)) %>% pull(m)

tree_final_plot +
  annotate(
    "text", x = 2.35, y = mean_shady,
    label = paste0("Shady mean:\n", round(mean_shady, 1), " g"),
    size = 3.2, color = "#0072B2", hjust = 0
  )
```
:::

::: {.column width="40%"}
- `annotate()` places a single fixed label — it does not come from a `geom_*()` mapped to data
- Use it to call out the one number your audience should leave remembering
- **Do not over-annotate** — one callout per figure is usually the right number
:::
:::::

# 🛑 Pause — Do Activity Parts 1–7 Now

Take the "busy" plot all the way through every fix, and redesign a second plot on your own using `height_cm` or `width_cm`.

# What We Learned Today

::::: columns
::: {.column width="60%"}
- Start every final figure from a **finished sentence**, not a geom
- The **five-second test**: does a stranger repeat your finding back?
- Cut redundancy: repeated legends, facets, and titles are clutter
- Direct-label when you can; drop the legend when the axis already says it
- **Order** categories with intent (`fct_relevel()`), not alphabetically
- **Color** with intent, and check colorblind safety (Okabe-Ito palette)
- **Title = finding. Subtitle = method. Caption = fine print.**
- `annotate()` to point at exactly one number or region that matters
:::

::: {.column width="40%"}
**References:**

- 📖 [R4DS Ch 9](https://r4ds.hadley.nz/layers) — Layers
- 📖 [R4DS Ch 11](https://r4ds.hadley.nz/communication) — Communication
- 🌐 [A Compendium of Clean Graphs in R](http://shinyapps.org/apps/RGraphCompendium/index.php)

**Up next:**

- Final Project Milestone work — you'll use every fix from today on your own dataset's take-home figure
:::
:::::
