---
title: "Lecture — Joins: Combining Two Tables"
subtitle: "Keys, mutating joins, and filtering joins on a fisher reintroduction dataset"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
order: 21
week: 11
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Real projects spread data across several tables. This 2-day lecture teaches
  joins: what a key is, how to join two tables when the key columns are named
  differently, the mutating joins (left/inner/full), and the filtering joins
  (semi/anti) — by combining a table of released fishers with a table of field
  capture events into one analysis-ready dataset.
categories: [joins, tidyverse, wrangling, mustelids]

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off

- **Pivoting** — reshaped wide data to long; **ANOVA** — compared group means; **Factors** — controlled category order
- Your **final project** is under way — data found, question pitched
- So far every analysis used **one** table

::: callout-note
**✅ Transition**

Real data almost never lives in a single table. Fishers (*Pekania pennanti*, a large forest weasel) are reintroduced and then monitored: one table lists the **animals released**, another logs every **field capture / collar event**. Neither alone tells you a captured animal's sex or where it came from. **Joins** stitch them on a shared **key**. *(The data are simulated for teaching, patterned on real reintroduction programs.)*
:::

# Goals for today (a 2-day lecture)

::::: columns
:::: {.column width="60%"}
- Understand a **key** — the column that links two tables
- Join when the key columns have **different names**
- The **mutating joins** — `left_join()`, `inner_join()`, `full_join()`
- The **filtering joins** — `semi_join()`, `anti_join()`
- Build one combined capture + individual table
::::

:::: {.column width="40%"}
**Tools today:**

- `tidyverse` (dplyr)

**Data (two tables, simulated):**

- `fisher_captures.csv` — one row per capture event
- `fisher_individuals.csv` — one row per released fisher

**Reference:**

- 📖 [R4DS Ch. 19 — Joins](https://r4ds.hadley.nz/joins)
- 📖 `common_code/13_joins`

::: callout-tip
**Day 1:** keys + `left_join`. **Day 2:** the join family + filtering joins.
:::
::::
:::::

# How to Use These Slides — Predict · Type · Run

::::: columns
:::: {.column width="60%"}
This lecture runs in **four chunks over two days**. After each chunk you switch to the **activity** and type the code yourself.

**For every join, do three things:**

1.  **Predict** — before it runs, say how many rows you'll get
2.  **Type** the code by hand — do not copy-paste
3.  **Run** it and compare to your prediction
::::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

-   A join can run without error and still hand you the wrong table — the only way to catch that is to know the row count you expected *before* you see the actual one.
-   Every join in this lecture takes `by = c("a" = "b")` as an argument — typing that syntax a dozen times is what makes it automatic under deadline pressure.
-   Four short chunks instead of one long one means `left_join`, `inner`/`full`, and `semi`/`anti` each get their own uncluttered mental slot.
:::
::::
:::::

# 🧩 Chunk 1 of 4 · Two Tables & a Key *(Day 1)*

**We will cover:** why data comes in multiple tables, and what a key is.

::: callout-tip
**🖐 After this chunk: Activity Parts 1–3** (load both tables, find the key).
:::

# The Problem — Data in Two Tables

::::: columns
:::: {.column width="60%"}
```{r load-12}
# Load BOTH fisher tables ----------------------------
library(tidyverse)

captures_df    <- read_csv("data/fisher_captures.csv")
individuals_df <- read_csv("data/fisher_individuals.csv")

dim(captures_df)       # field events: when/where/mass/collar
dim(individuals_df)    # released animals: sex/age/source/site
```
::::

:::: {.column width="40%"}
- **`captures_df`** — one row per capture event: date, site, mass, collar id
- **`individuals_df`** — one row per released fisher: sex, age class, source, release site
- To use a capture you often need **both** — the event *and* who the animal is
::::
:::::

# What Is a Key?

::::: columns
:::: {.column width="60%"}
```{r peek-12}
# The columns that identify an animal ---------------
captures_df    %>% select(animal_id, capture_date, site, mass_kg) %>% head(3)

individuals_df %>% select(fisher_id, sex, site, release_date) %>% head(3)
```
::::

:::: {.column width="40%"}
A **key** is the column that identifies an animal and links the tables.

- In `captures_df` it's **`animal_id`**
- In `individuals_df` it's **`fisher_id`**
- **Same idea, different name** — we'll tell the join that

::: callout-note
**📖 Vocabulary**

The key must be **unique** in the table you look things up in (`individuals_df` — one row per fisher). It is *not* unique in `captures_df` — a fisher can be caught many times.
:::
::::
:::::

# Clean the Key Before Joining

::::: columns
:::: {.column width="60%"}
```{r keys-12}
# The capture key is lowercase; the individual key is not
captures_df %>% count(animal_id) %>% head()

captures_df <- captures_df %>%
  mutate(animal_id = str_to_upper(animal_id))   # "f07" -> "F07"

n_distinct(individuals_df$fisher_id)
n_distinct(captures_df$animal_id)   # includes "UNK" — unmarked animals
```
::::

:::: {.column width="40%"}
- `str_to_upper()` — `animal_id` was `"f07"`, `fisher_id` is `"F07"`. **Case must match** or the join finds nothing
- `"UNK"` is not a real fisher — those events have no individual to attach

::: callout-warning
**⚠️ Watch out!** A join silently fails to match if the keys differ in **type** (text vs number), **case**, or **format** — clean them first.
:::
::::
:::::

# 🛑 Pause — Do Activity Parts 1–3 Now

Load both tables, find the key in each, and fix its case. Predict how many capture events can be tied to a known fisher.

# 🧩 Chunk 2 of 4 · Your First Join — `left_join()` *(Day 1)*

**We will cover:** joining the two tables when the key columns are named differently.

::: callout-tip
**🖐 After this chunk: Activity Parts 4–6** (run the left join, read the result).
:::

# `left_join()` — Keep Every Capture Event

::: callout-note
**🔮 Predict first:** `captures_df` has ~45 events; a few are unmarked animals. After a `left_join` (captures on the left), how many **rows** will we have, and how many will have a missing `sex`?
:::

::::: columns
:::: {.column width="60%"}
```{r leftjoin-12}
# Match on differently-named keys with by = c() ------
captures_joined <- captures_df %>%
  left_join(individuals_df, by = c("animal_id" = "fisher_id"))

nrow(captures_joined)                  # every capture event is kept
sum(is.na(captures_joined$sex))        # events we can't attribute
```
::::

:::: {.column width="40%"}
- `by = c("animal_id" = "fisher_id")` — "match my `animal_id` to their `fisher_id`"
- **`left_join` keeps every row on the left** (all capture events)
- Unmarked-animal events get `NA` for sex, age, source

📖 R4DS §19.3 — mutating joins
::::
:::::

# Live Demo — Watch It Break (a silent wrong join)

::::: columns
:::: {.column width="60%"}
If I forget `by = ...`, dplyr looks for a **shared column name** — and both tables have a **`site`** column:

``` r
captures_df %>% left_join(individuals_df)
#> Joining with `by = join_by(site)`
```

But `site` means different things: the **capture** site vs the **release** site. Joining on it matches a capture to *every fisher released at that site* — the row count explodes and the animal ids are nonsense. **No error.**

The fix — always name the real key:

``` r
captures_df %>% left_join(individuals_df, by = c("animal_id" = "fisher_id"))
```
::::

:::: {.column width="40%"}
::: callout-important
**✅ Why show a silent wrong join?**

Nothing throws an error, so nothing forces you to notice. `captures_df` and `individuals_df` both have a `site` column; dplyr matches on it happily and hands you hundreds of rows that pair a capture with the wrong animal. If the message *"Joining with `by = ...`"* names a column you didn't pick, treat it as a bug report, not a status update.
:::
::::
:::::

# 🛑 End of Day 1 · Start Day 2 Here

::::: columns
:::: {.column width="60%"}
**Day 1 recap — you can now:**

- Explain a key and why it must be unique (in the lookup table) and same case/type
- Join two tables whose key columns are named differently
- Use `left_join()` to keep every row on the left
::::

:::: {.column width="40%"}
**Day 2 — the rest of the family:**

- `inner_join` vs `left_join` vs `full_join`
- `semi_join` and `anti_join` to **filter** instead of add columns
::::
:::::

# 🧩 Chunk 3 of 4 · The Join Family *(Day 2)*

**We will cover:** `inner_join`, `left_join`, and `full_join`, and how to choose.

::: callout-tip
**🖐 After this chunk: Activity Parts 7–9** (compare the three joins).
:::

# Inner vs. Left vs. Full

::: callout-note
**🔮 Predict first:** Rank by row count: `inner_join` (only matches), `left_join` (all captures), `full_join` (all captures **and** every released fisher, even the never-recaptured ones).
:::

::::: columns
:::: {.column width="60%"}
```{r family-12}
# Same key, three different rules --------------------
key <- c("animal_id" = "fisher_id")

inner <- inner_join(captures_df, individuals_df, by = key)
left  <- left_join (captures_df, individuals_df, by = key)
full  <- full_join (captures_df, individuals_df, by = key)

c(inner = nrow(inner), left = nrow(left), full = nrow(full))
```
::::

:::: {.column width="40%"}
| Join | Keeps |
|-------------|-------------------------------|
| `inner_join` | only events with a **known** animal |
| `left_join` | **all** capture events |
| `full_join` | all events **and** every released fisher |

`inner` drops the unmarked-animal events; `full` adds back the fishers that were released but never recaptured.
::::
:::::

# Which Join Should You Use?

::::: columns
:::: {.column width="60%"}
- **`left_join`** — the default. Keep everything on your main table, add what you can. *(Event-level analysis: keep all captures.)*
- **`inner_join`** — only rows with a match. *(When an unattributed event would ruin the analysis.)*
- **`full_join`** — keep everything from both. *(To audit what's missing on each side — which fishers vanished.)*
::::

:::: {.column width="40%"}
::: callout-note
**✅ Choosing among the three**

Ask which rows you'd be upset to lose. Drop an unmarked capture and you undercount effort; drop a never-recaptured fisher from a *survival* analysis and you bias it badly. Match the join to the question.
:::

📖 R4DS §19.3–19.4
::::
:::::

# 🛑 Pause — Do Activity Parts 7–9 Now

Run all three joins, compare the row counts, and decide which one you'd use to build a map. Predict each count first.

# 🧩 Chunk 4 of 4 · Filtering Joins & the Payoff *(Day 2)*

**We will cover:** `semi_join` and `anti_join` — joins that **filter** rather than add columns.

::: callout-tip
**🖐 After this chunk: Activity Parts 10–12** (find the missing reports, build the map table).
:::

# `anti_join()` — What Didn't Match?

::: callout-note
**🔮 Predict first (two ways):** how many capture **events** have no known animal? How many released **fishers** were never recaptured?
:::

::::: columns
:::: {.column width="60%"}
```{r anti-12}
# (a) capture events of unmarked animals
captures_df %>%
  anti_join(individuals_df, by = c("animal_id" = "fisher_id")) %>%
  nrow()

# (b) released fishers with no capture — anti_join the OTHER way
individuals_df %>%
  anti_join(captures_df, by = c("fisher_id" = "animal_id")) %>%
  select(fisher_id, sex, site)
```
::::

:::: {.column width="40%"}
- **`anti_join`** keeps the left rows with **no** match — "what's missing?"
- It **adds no columns** — it filters
- Direction matters: run it both ways to see each side's gaps

📖 R4DS §19.5 — filtering joins
::::
:::::

# `semi_join()` — Keep Only What Matched

::::: columns
:::: {.column width="60%"}
```{r semi-12}
# Released fishers that WERE recaptured at least once
individuals_df %>%
  semi_join(captures_df, by = c("fisher_id" = "animal_id")) %>%
  nrow()
```
::::

:::: {.column width="40%"}
- **`semi_join`** keeps left rows that **have** a match — again, **no new columns**
- Think of `semi`/`anti` as a **filter driven by another table**
- `semi` + `anti` on the same table split it in two: recaptured vs never seen
::::
:::::

# The Payoff — Build One Analysis-Ready Table

::::: columns
:::: {.column width="60%"}
```{r payoff-12}
# Every KNOWN-individual capture, with the animal's attributes
fisher_analysis <- captures_df %>%
  inner_join(individuals_df, by = c("animal_id" = "fisher_id")) %>%
  mutate(days_since_release = as.numeric(capture_date - release_date))

fisher_analysis %>%
  select(animal_id, sex, age_class, capture_date, days_since_release, mass_kg) %>%
  head()
```
::::

:::: {.column width="40%"}
**You just built a real dataset from two raw ones.** `fisher_analysis` didn't exist in either source file — every row is a capture event *plus* the animal's sex, age, and days-since-release, ready for a model.

Here `inner_join` is right: an event with no known animal can't go into an individual-level analysis.
::::
:::::

# 🛑 Pause — Do Activity Parts 10–12 Now

Use `anti_join` both directions to find the gaps, `semi_join` to see who was recaptured, and build the combined table. Predict the counts first.

# What We Learned Today

::::: columns
:::: {.column width="60%"}
**Concepts:**

- A **key** links two tables; it must be **unique** and the **same type**
- Join differently-named keys with `by = c("a" = "b")`
- **Mutating joins** add columns: `left` (all left), `inner` (matches), `full` (all)
- **Filtering joins** keep rows: `semi` (has a match), `anti` (no match)
- Always **name your key** — a printed `by = ...` you didn't choose is a bug

**R skills:**

- `left_join()`, `inner_join()`, `full_join()`
- `semi_join()`, `anti_join()`
- `str_to_upper()` (and `as.integer()`, `str_trim()`) for key hygiene
::::

:::: {.column width="40%"}
**References:**

- 📖 [R4DS Ch. 19 — Joins](https://r4ds.hadley.nz/joins)
- 📖 `common_code/13_joins`

**Up next — Mapping:**

- Turn latitude/longitude into a map with `sf` and `geom_sf()`
- A different dataset (Bigfoot sightings) — same join thinking
::::
:::::
