---
title: "Lecture 17 — Joins: Combining Two Tables"
subtitle: "Keys, mutating joins, and filtering joins with real Bigfoot report data"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
order: 21
week: 11
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Real projects spread data across several tables. This 2-day lecture teaches
  joins: what a key is, how to join two tables when the key columns are named
  differently, the mutating joins (left/inner/full), and the filtering joins
  (semi/anti) — by combining the Bigfoot report text with its coordinates to
  build exactly the dataset we map next week.
categories: [joins, tidyverse, wrangling]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off — Factors

- **Factors** — reordered and relabeled categorical levels for clean plots and models
- Before that: reshaped messy data (Pivoting), started the final project, and ran a one-way ANOVA
- So far every analysis used **one** table

::: callout-note
**✅ Transition**

Real data almost never lives in a single table. Today's dataset comes in **two**: one file has the Bigfoot report **text** (where, when, what), a second has the **coordinates**. Neither alone lets you map a sighting. **Joins** stitch them together on a shared **key** — and the result is exactly the file we'll map next week. 🦶
:::

# Goals for today (a 2-day lecture)

::::: columns
:::: {.column width="60%"}
- Understand a **key** — the column that links two tables
- Join when the key columns have **different names**
- The **mutating joins** — `left_join()`, `inner_join()`, `full_join()`
- The **filtering joins** — `semi_join()`, `anti_join()`
- Build the combined report + coordinates table
::::

:::: {.column width="40%"}
**Tools today:**

- `tidyverse` (dplyr)

**Data (two tables):**

- `bfro_reports.csv` — the report text + class
- `bfro_locations.csv` — the coordinates

**Reference:**

- 📖 [R4DS Ch. 19 — Joins](https://r4ds.hadley.nz/joins)
- 📖 `common_code/13_joins`

::: callout-tip
**Day 1:** keys + `left_join`. **Day 2:** the join family + filtering joins.
:::
::::
:::::

# How to Use These Slides — Predict · Type · Run

::::: columns
:::: {.column width="60%"}
This lecture runs in **four chunks over two days**. After each chunk you switch to the **activity** and type the code yourself.

**For every join, do three things:**

1.  **Predict** — before it runs, say how many rows you'll get
2.  **Type** the code by hand — do not copy-paste
3.  **Run** it and compare to your prediction
::::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

-   A join can run without error and still hand you the wrong table — the only way to catch that is to know the row count you expected *before* you see the actual one.
-   Every join in this lecture takes `by = c("a" = "b")` as an argument — typing that syntax a dozen times is what makes it automatic under deadline pressure.
-   Four short chunks instead of one long one means `left_join`, `inner`/`full`, and `semi`/`anti` each get their own uncluttered mental slot.
:::
::::
:::::

# 🧩 Chunk 1 of 4 · Two Tables & a Key *(Day 1)*

**We will cover:** why data comes in multiple tables, and what a key is.

::: callout-tip
**🖐 After this chunk: Activity Parts 1–3** (load both tables, find the key).
:::

# The Problem — Data in Two Tables

::::: columns
:::: {.column width="60%"}
```{r load-12}
# Load BOTH Bigfoot tables ---------------------------
library(tidyverse)

reports_df   <- read_csv("data/bfro_reports.csv")
locations_df <- read_csv("data/bfro_locations.csv")

dim(reports_df)      # the text: where/when/what
dim(locations_df)    # the coordinates
```
::::

:::: {.column width="40%"}
- **`reports_df`** — one row per sighting report: year, season, state, county, class, and the written account
- **`locations_df`** — one row per report that got geocoded: a latitude and longitude
- To map a sighting you need **both** — its class *and* its coordinates
::::
:::::

# What Is a Key?

::::: columns
:::: {.column width="60%"}
```{r peek-12}
# The columns that identify a report -----------------
reports_df %>% select(REPORT_NUMBER, YEAR, STATE, REPORT_CLASS) %>% head(3)

locations_df %>% select(number, latitude, longitude) %>% head(3)
```
::::

:::: {.column width="40%"}
A **key** is the column that uniquely identifies a row and links the tables.

- In `reports_df` it's **`REPORT_NUMBER`**
- In `locations_df` it's **`number`**
- **Same idea, different name** — we'll tell the join that

::: callout-note
**📖 Vocabulary**

The key must be **unique** in the table you're looking things up in (`locations_df`).
:::
::::
:::::

# Clean the Key Before Joining

::::: columns
:::: {.column width="60%"}
```{r keys-12}
# Two fixes: drop duplicate reports, match key type --
reports_df <- reports_df %>%
  distinct(REPORT_NUMBER, .keep_all = TRUE) %>%   # one row per report
  mutate(REPORT_NUMBER = as.integer(REPORT_NUMBER))

n_distinct(reports_df$REPORT_NUMBER)
n_distinct(locations_df$number)
```
::::

:::: {.column width="40%"}
- `distinct()` — keep one row per report number
- `as.integer()` — `REPORT_NUMBER` was stored like `30680.0`; `number` is `30680`. **Keys must be the same type** to match

::: callout-warning
**⚠️ Watch out!** A join silently fails to match if the keys are different **types** (text vs number) or **formats** — clean them first.
:::
::::
:::::

# 🛑 Pause — Do Activity Parts 1–3 Now

Load both tables, find the key in each, and clean it. Predict how many reports have coordinates before we join.

# 🧩 Chunk 2 of 4 · Your First Join — `left_join()` *(Day 1)*

**We will cover:** joining the two tables when the key columns are named differently.

::: callout-tip
**🖐 After this chunk: Activity Parts 4–6** (run the left join, read the result).
:::

# `left_join()` — Keep Every Report

::: callout-note
**🔮 Predict first:** `reports_df` has ~5,000 reports; only ~4,000 were geocoded. After a `left_join`, how many **rows** will we have, and how many will have a latitude?
:::

::::: columns
:::: {.column width="60%"}
```{r leftjoin-12}
# Match on differently-named keys with by = c() ------
bigfoot_df <- reports_df %>%
  left_join(locations_df, by = c("REPORT_NUMBER" = "number"))

nrow(bigfoot_df)                    # every report is kept
sum(!is.na(bigfoot_df$latitude))    # how many got coordinates
```
::::

:::: {.column width="40%"}
- `by = c("REPORT_NUMBER" = "number")` — "match my `REPORT_NUMBER` to their `number`"
- **`left_join` keeps every row on the left** (all reports)
- Reports with no location get `NA` coordinates — they simply weren't geocoded

📖 R4DS §19.3 — mutating joins
::::
:::::

# Live Demo — Watch It Break (a silent wrong join)

::::: columns
:::: {.column width="60%"}
If I forget `by = ...`, dplyr looks for a **shared column name** — and both tables happen to have an **`index`** column:

``` r
reports_df %>% left_join(locations_df)
#> Joining with `by = join_by(index)`
```

That matches **row 1 to row 1, row 2 to row 2** — nonsense! `index` is just a row number, not a report ID. You get coordinates glued to the wrong reports, with **no error**.

The fix — always name the real key:

``` r
reports_df %>% left_join(locations_df, by = c("REPORT_NUMBER" = "number"))
```
::::

:::: {.column width="40%"}
::: callout-important
**✅ Why show a silent wrong join?**

Nothing here throws an error, so nothing forces you to notice — `reports_df` and `locations_df` both got the same made-up `index` column at some point, dplyr matches on it happily, and you end up with row 1's coordinates glued to row 1's report regardless of which Bigfoot sighting it actually was. If the message *"Joining with `by = ...`"* names a column you didn't pick, treat it as a bug report, not a status update.
:::
::::
:::::

# 🛑 End of Day 1 · Start Day 2 Here

::::: columns
:::: {.column width="60%"}
**Day 1 recap — you can now:**

- Explain a key and why it must be unique and same-type
- Join two tables whose key columns are named differently
- Use `left_join()` to keep every row on the left
::::

:::: {.column width="40%"}
**Day 2 — the rest of the family:**

- `inner_join` vs `left_join` vs `full_join`
- `semi_join` and `anti_join` to **filter** instead of add columns
::::
:::::

# 🧩 Chunk 3 of 4 · The Join Family *(Day 2)*

**We will cover:** `inner_join`, `left_join`, and `full_join`, and how to choose.

::: callout-tip
**🖐 After this chunk: Activity Parts 7–9** (compare the three joins).
:::

# Inner vs. Left vs. Full

::: callout-note
**🔮 Predict first:** Which will have **more** rows — an `inner_join` (only matches) or a `left_join` (all reports)? By roughly how many?
:::

::::: columns
:::: {.column width="60%"}
```{r family-12}
# Same key, three different rules --------------------
inner <- inner_join(reports_df, locations_df, by = c("REPORT_NUMBER" = "number"))
left  <- left_join (reports_df, locations_df, by = c("REPORT_NUMBER" = "number"))
full  <- full_join (reports_df, locations_df, by = c("REPORT_NUMBER" = "number"))

c(inner = nrow(inner), left = nrow(left), full = nrow(full))
```
::::

:::: {.column width="40%"}
| Join | Keeps |
|-------------|-------------------------------|
| `inner_join` | only rows that **match** in both |
| `left_join` | **all** rows from the left (reports) |
| `full_join` | **all** rows from **both** tables |

`inner` drops the reports with no coordinates; `full` even keeps locations with no report.
::::
:::::

# Which Join Should You Use?

::::: columns
:::: {.column width="60%"}
- **`left_join`** — the default. Keep everything on your main table, add what you can. *(Use this for the map: keep all reports.)*
- **`inner_join`** — only rows with a match. *(Use when missing data would ruin the analysis.)*
- **`full_join`** — keep everything from both. *(Use to audit what's missing on each side.)*
::::

:::: {.column width="40%"}
::: callout-note
**✅ Choosing among the three**

Ask which rows you'd be upset to lose. Drop a report from the map and a real sighting vanishes; drop a stray location record and nothing is lost. That's why the map keeps every report — `left_join`, not `inner_join`.
:::

📖 R4DS §19.3–19.4
::::
:::::

# 🛑 Pause — Do Activity Parts 7–9 Now

Run all three joins, compare the row counts, and decide which one you'd use to build a map. Predict each count first.

# 🧩 Chunk 4 of 4 · Filtering Joins & the Payoff *(Day 2)*

**We will cover:** `semi_join` and `anti_join` — joins that **filter** rather than add columns.

::: callout-tip
**🖐 After this chunk: Activity Parts 10–12** (find the missing reports, build the map table).
:::

# `anti_join()` — What Didn't Match?

::: callout-note
**🔮 Predict first:** How many reports have **no** coordinates? (Hint: total reports minus the number that got a latitude in Chunk 2.)
:::

::::: columns
:::: {.column width="60%"}
```{r anti-12}
# Reports that have NO location ----------------------
missing_coords <- reports_df %>%
  anti_join(locations_df, by = c("REPORT_NUMBER" = "number"))

nrow(missing_coords)
missing_coords %>% count(STATE, sort = TRUE) %>% head(4)
```
::::

:::: {.column width="40%"}
- **`anti_join`** keeps the left rows with **no** match — perfect for "what's missing?"
- It **adds no columns** — it filters
- Here: the reports we could never map, and where they cluster

📖 R4DS §19.5 — filtering joins
::::
:::::

# `semi_join()` — Keep Only What Matched

::::: columns
:::: {.column width="60%"}
```{r semi-12}
# Keep only reports that DO have a location ----------
mappable_reports <- reports_df %>%
  semi_join(locations_df, by = c("REPORT_NUMBER" = "number"))

nrow(mappable_reports)
```
::::

:::: {.column width="40%"}
- **`semi_join`** keeps left rows that **have** a match — again, **no new columns**
- Think of `semi`/`anti` as a **filter driven by another table**
- `semi_join` result = the rows an `inner_join` would keep, but without the extra columns
::::
:::::

# The Payoff — Build the Map-Ready Table

::::: columns
:::: {.column width="60%"}
```{r payoff-12}
# Reports + coordinates, ready to map next week ------
bigfoot_mappable <- reports_df %>%
  left_join(locations_df, by = c("REPORT_NUMBER" = "number")) %>%
  filter(!is.na(latitude), !is.na(longitude))

bigfoot_mappable %>%
  select(REPORT_NUMBER, STATE, SEASON, REPORT_CLASS, latitude, longitude) %>%
  head()
```
::::

:::: {.column width="40%"}
This joined table — report text **plus** coordinates — is essentially `bfro_reports_geocoded.csv`, the file you'll turn into a map in **Lecture 13**.

**You just built a real dataset from two raw ones.** `bigfoot_mappable` didn't exist in either source file — it only exists because you joined them on a key, and it's the exact table Lecture 13 opens with.
::::
:::::

# 🛑 Pause — Do Activity Parts 10–12 Now

Use `anti_join` to find the un-geocoded reports, `semi_join` to keep the mappable ones, and build the combined table. Predict the counts first.

# What We Learned Today

::::: columns
:::: {.column width="60%"}
**Concepts:**

- A **key** links two tables; it must be **unique** and the **same type**
- Join differently-named keys with `by = c("a" = "b")`
- **Mutating joins** add columns: `left` (all left), `inner` (matches), `full` (all)
- **Filtering joins** keep rows: `semi` (has a match), `anti` (no match)
- Always **name your key** — a printed `by = ...` you didn't choose is a bug

**R skills:**

- `left_join()`, `inner_join()`, `full_join()`
- `semi_join()`, `anti_join()`
- `distinct()` + `as.integer()` for key hygiene
::::

:::: {.column width="40%"}
**References:**

- 📖 [R4DS Ch. 19 — Joins](https://r4ds.hadley.nz/joins)
- 📖 `common_code/13_joins`

**Up next — Lecture 18:**

- **Mapping** — turn this joined table into a map with `sf` and `geom_sf()`
- Where *is* Bigfoot, anyway?
::::
:::::
