---
title: "Lecture 09 — T-Tests I: Setting Up the Test"
subtitle: "Hypotheses and checking assumptions before we test"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# order/week are auto-stamped from _schedule.yml by sync_order_from_schedule.R
order: 9
week: 5
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Why comparing means isn't enough, why Welch's is the safe default,
  checking normality (histogram, QQ plot, Shapiro-Wilk) and variance
  (Levene's), then a first run of the t-test.
categories: [tidyverse, descriptive-stats]

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(width = 65)   # printed R output wraps at 65 too
```

# Where we left off (Lectures 05–07)

- **Wrangling** — `filter()`, `select()`, `mutate()`, `arrange()` as the core tidyverse verbs
- **Descriptive stats** — mean, SD, SE with `group_by()` + `summarize()`
- **Box plot** + jittered points; **mean ± SE** with `stat_summary()`
- **Quarto report** — you already ran a Welch's t-test once, at the end
- **Prediction:** shady-side leaves would be heavier than sunny-side leaves

::: callout-note
**✅ Key idea for today**

Last time we *ran* a t-test. Today we slow down and do it properly: state the hypotheses, **check the assumptions**, and only then run the test.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- **Understand** why comparing means is not enough — we need a test
- **Meet** the two-sample t-test and why we default to **Welch's**
- **State** null and alternate hypotheses
- **Check assumptions** before running any test
  - normality: histogram, **QQ plot**, Shapiro-Wilk
  - variance: Levene's test
- **Run** the t-test (we decode it next lecture)
::::

::: {.column width="40%"}
**Tools added today:**

- `stat_qq()` + `stat_qq_line()` — QQ plots
- `shapiro.test()` — normality test
- `car::leveneTest()` — variance test

**Reading:**

- 📖 Whitlock & Schluter, Ch. 12 — *Comparing two means*
- 📖 Whitlock & Schluter, Ch. 13 — *Handling violations of assumptions*

**Today's naming:**

- plots → `_plot`
- test results → `_model`
:::
::::::

# Load Libraries and Data

```{r load-libs-09}
# Load all packages at the top ----------------------------
library(readxl)    # reading Excel files
library(tidyverse) # data manipulation + ggplot2
library(janitor)   # clean_names()
library(car)       # Levene's test for equal variance
```

```{r load-data-09}
# Load the leaf data from the data folder -----------------
leaf_df <-
  read_excel("data/2026_09_03_data_sci_leaf_area.xlsx") %>%
  clean_names()
```

::: callout-note
`car` is the **C**ompanion to **A**pplied **R**egression package. It contains `leveneTest()`, which checks whether two groups have similar spread.

Install once: `install.packages("car")`
:::

# 🧩 Chunk 1 of 3 · Why We Test

**We will cover:** why comparing means isn't enough, the five steps we'll use for every test this semester, what a two-sample t-test does, and why we default to Welch's.

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 1–2** (load data, recap the descriptive stats).
:::

# Why Comparing Means Is Not Enough

::::: columns
::: {.column width="60%"}
```{r recap-plot}
#| fig-format: png
#| fig-alt: "Jittered points of leaf mass in grams for shady and sunny sides, with a black point and error bar showing the mean plus or minus one standard error for each side; the two means sit close together with overlapping error bars."
# Remind ourselves what the data look like ---------------
recap_plot <- leaf_df %>%
  ggplot(aes(x = shade, y = mass_g, color = shade)) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.15,
    linewidth = 0.9
  ) +
  labs(x = "Shade", y = "Leaf Mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")
recap_plot
```
:::

::: {.column width="40%"}
The two means look **close** — and with this much scatter, they could easily be the same:

- About **25 leaves per group**
- Lots of **natural variability** within each group
- The groups **overlap heavily**

A *p*-value answers:

> *"If there were really no difference, how often would we see a gap at least this big just by chance?"*

Large *p* → a gap this size is nothing unusual.
:::
:::::

# Five Steps to a Hypothesis Test

::::: columns
::: {.column width="60%"}
Every hypothesis test this semester — t-test, ANOVA, regression — follows the same steps:

| Step | Action | When |
|------|---------------------------------|--------------|
| **1** | State H₀ and Hₐ | Today |
| **2** | Check **normality** | Today |
| **3** | Check **equal variance** | Today |
| **4** | **Run** the test | Today (a first look) |
| **5** | **Interpret** and **report** | Next lecture |

: {tbl-colwidths="[12,58,30]"}
:::

:::: {.column width="40%"}
::: callout-important
**Check assumptions BEFORE running the test.**

If the data badly break the assumptions, the *p*-value can't be trusted.

With ~25 per group our main worry is **badly** non-normal data. The t-test handles mild departures fine.
:::
::::
:::::

# What Is a Two-Sample t-Test?

::::: columns
::: {.column width="60%"}
A two-sample t-test compares the **means of two independent groups**.

It asks: *is the gap between the two means big compared with the scatter inside each group?*

$$t = \frac{\text{difference between the means}}{\text{noise (scatter) in the data}}$$

- **Big gap, little scatter** → big *t* → small *p*-value
- **Small gap, lots of scatter** → *t* near 0 → large *p*-value

R does the arithmetic — your job is to set the test up correctly and read the answer.
:::

::: {.column width="40%"}
| Piece | Our leaves |
|-----------|-----------------------------|
| Group 1 | Shady leaves |
| Group 2 | Sunny leaves |
| Measured | `mass_g` |
| H₀ | $\mu_{shady} = \mu_{sunny}$ |
| Hₐ | $\mu_{shady} \neq \mu_{sunny}$ |

: {tbl-colwidths="[35,65]"}
:::
:::::

# Why Welch's? — Don't Assume Equal Variance

:::::: columns
::: {.column width="60%"}
There are two versions of the two-sample t-test:

- **Standard (Student's) t-test** — assumes both groups have the **same spread**
- **Welch's t-test** — lets each group have **its own spread**

When the spreads really are equal, both give almost the same answer. When they are not, the standard test gives **too many false positives**.
:::

:::: {.column width="40%"}
::: callout-tip
**Best practice:** use **Welch's** as your default for every two-group comparison.

In R: **`var.equal = FALSE`**
:::
::::
::::::

# → ACTIVITY 9 Parts 1–2 now

::: {.activitybanner}
**🛑 Do Activity Parts 1–2 now**

Load the data and recompute the group means and SDs. Predict each output, type it, run it.
:::

# 🧩 Chunk 2 of 3 · Steps 1–3: Hypotheses & Assumptions

**We will cover:** writing H₀ / Hₐ, checking normality three ways (histogram, **QQ plot**, Shapiro-Wilk), and checking equal variance (Levene's).

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 3–7** (hypotheses, histogram, box plot + QQ plot, Shapiro-Wilk, Levene's).
:::

# Step 1 — Hypotheses

::::: columns
::: {.column width="60%"}
**Biological prediction:** shady-side leaves will be **larger and heavier** — they need more surface area to capture the limited light under the canopy.

|  | Formal statement |
|--------|-----------------------------------------|
| **H₀ — null** | $\mu_{shady} = \mu_{sunny}$ — no difference in mean leaf mass |
| **Hₐ — alternate** | $\mu_{shady} \neq \mu_{sunny}$ — mean leaf mass differs by side |

: {tbl-colwidths="[30,70]"}

**Two-tailed** — detects a difference in *either* direction. **α = 0.05** — reject H₀ if *p* \< 0.05.
:::

::: {.column width="40%"}
**Why two-tailed when we predicted shady \> sunny?**

- A one-tailed test only looks in one direction
- Two-tailed is **more conservative** and the standard in most journals

**Why write hypotheses first?**

- Keeps the analysis honest
- Prevents "HARKing" — **H**ypothesizing **A**fter the **R**esults are **K**nown
:::
:::::

# Step 2 — Normality: Histograms

::::: columns
::: {.column width="60%"}
```{r histogram-norm}
#| fig-format: png
#| fig-height: 4.2
#| fig-alt: "Two histograms of leaf mass in grams stacked one above the other, shady on top and sunny below, sharing the same x-axis so the ranges can be compared directly."
# Histograms stacked in one column — same x-axis --------
hist_norm_plot <- leaf_df %>%
  ggplot(aes(x = mass_g, fill = shade)) +
  geom_histogram(binwidth = 0.05) +
  facet_wrap(~shade, ncol = 1) +
  labs(x = "Leaf Mass (g)", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
hist_norm_plot
```
:::

::: {.column width="40%"}
**`ncol = 1` stacks the panels.** Both share one x-axis, so you can see at a glance:

- where each group is **centered**
- how **wide** each group spreads
- which group has the **long tail**

**What to look for:**

- Roughly **bell-shaped** — one hump, not badly lopsided
- No big **outliers** far from the rest

With ~25 leaves the bars are lumpy — that's normal for real data. It's hard to judge "bell-shaped" by eye, which is why we add a **QQ plot**.
:::
:::::

# First: What Is a Quantile? (you already know one)

::::: columns
::: {.column width="55%"}
```{r quantile-boxplot}
#| echo: false
#| fig-format: png
#| fig-height: 4
#| fig-alt: "Box plot of shady-side leaf mass with jittered points, with labels pointing to the lower quartile (25 percent of leaves below), median (50 percent below), and upper quartile (75 percent below)."
shady_df <- leaf_df %>% filter(shade == "shady")
shady_q <- quantile(shady_df$mass_g, c(0.25, 0.5, 0.75))

ggplot(shady_df, aes(x = 1, y = mass_g)) +
  geom_boxplot(width = 0.4, fill = "grey90",
               outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.6, size = 2,
              color = "darkgreen") +
  annotate("text", x = 1.28, y = shady_q, hjust = 0,
           size = 4,
           label = c("Q1: 25% of leaves below",
                     "Median: 50% below",
                     "Q3: 75% below")) +
  scale_x_continuous(limits = c(0.7, 2)) +
  labs(x = NULL, y = "Shady Leaf Mass (g)") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```
:::

::: {.column width="45%"}
A **quantile** is a value that a certain percent of the data falls below.

The box plot already shows three of them:

- **Q1** (25th percentile) — bottom of the box
- **Median** (50th) — the line in the middle
- **Q3** (75th) — top of the box

A QQ plot uses **every** leaf as a quantile, not just three — and asks whether they are spaced the way a **normal** (bell-curve) distribution would space them.
:::
:::::

# How a QQ Plot Works

::::: columns
::: {.column width="55%"}
```{r qq-explain}
#| echo: false
#| fig-format: png
#| fig-height: 4.2
#| fig-alt: "QQ plot of shady-side leaf mass with a box plot of the same data drawn on its left edge. Dashed horizontal lines run from the box plot's quartiles across the QQ plot, meeting the points near theoretical values of minus 0.67, 0 and 0.67. Most points follow the straight reference line, but the largest few leaves bend upward above it."
q_theory <- qnorm(c(0.25, 0.5, 0.75))

ggplot(shady_df, aes(sample = mass_g)) +
  geom_hline(yintercept = shady_q, linetype = "dashed",
             color = "grey50") +
  geom_vline(xintercept = q_theory, linetype = "dotted",
             color = "grey50") +
  geom_boxplot(aes(x = -2.9, y = mass_g), width = 0.5,
               fill = "grey90", inherit.aes = FALSE) +
  stat_qq_line(color = "black", linewidth = 0.8) +
  stat_qq(color = "darkgreen", size = 2) +
  annotate("text", x = q_theory, y = 0.3, size = 3.5,
           label = c("Q1", "Median", "Q3")) +
  labs(x = "Theoretical quantiles (where a normal leaf would be)",
       y = "Shady Leaf Mass (g)") +
  theme_minimal()
```
:::

::: {.column width="45%"}
**What R does for you:**

1.  **Sorts** the leaves from lightest to heaviest (the y-axis)
2.  Works out where each leaf **would** sit if the data were perfectly **normal** (the x-axis, in SD units: −2 … 0 … +2)
3.  Plots **actual vs. expected** — one dot per leaf

**Straight line = normal.**

The dashed lines connect the **box plot's** Q1, median, and Q3 to the same leaves on the QQ plot. The middle half of the dots is the box.

Here the heaviest few shady leaves **bend up** above the line — a **long right tail**.
:::
:::::

# What Good and Bad QQ Plots Look Like

::::: columns
::: {.column width="70%"}
```{r qq-examples}
#| echo: false
#| fig-format: png
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Three columns of made-up data. Top row histograms, bottom row QQ plots. Left: normal data, bell-shaped histogram, points on the line. Middle: right-skewed data, histogram with a long right tail, points curving upward above the line at the right end. Right: data with outliers, points on the line except two far above it at the right end."
set.seed(9)
wiggle <- rnorm(30, 0, 0.25)   # a little real-data noise
demo_df <- bind_rows(
  tibble(shape = "1. Normal",
         value = qnorm(ppoints(30), 10, 2) + wiggle),
  tibble(shape = "2. Right-skewed",
         value = qexp(ppoints(30), 0.35) + 5 + wiggle / 2),
  tibble(shape = "3. Outliers",
         value = c(qnorm(ppoints(28), 10, 1.5) + wiggle[1:28],
                   17, 18.5))
)

demo_hist <- ggplot(demo_df, aes(x = value)) +
  geom_histogram(bins = 10, fill = "grey60",
                 color = "white") +
  facet_wrap(~shape, scales = "free") +
  labs(x = NULL, y = "Count") +
  theme_minimal()

demo_qq <- ggplot(demo_df, aes(sample = value)) +
  stat_qq_line() +
  stat_qq(color = "steelblue") +
  facet_wrap(~shape, scales = "free") +
  labs(x = "Theoretical quantiles", y = "Sample") +
  theme_minimal()

patchwork::wrap_plots(demo_hist, demo_qq, ncol = 1)
```
:::

::: {.column width="30%"}
Made-up data, 30 values each. **Top:** histogram. **Bottom:** QQ plot of the same values.

- **Normal:** dots hug the line
- **Skewed:** dots curve away at one end (a banana shape)
- **Outliers:** a few dots jump off the line at an end

A little wiggle at the very ends is fine — look for a **clear bend**.
:::
:::::

# Step 2 — Normality: Our QQ Plots

::::: columns
::: {.column width="60%"}
```{r qq-norm}
#| fig-format: png
#| fig-alt: "QQ plots of leaf mass for shady and sunny leaves side by side. Sunny points follow the reference line closely. Shady points mostly follow the line but the largest few curve upward above it."
# QQ plot per group — dots should follow the line -------
qq_plot <- leaf_df %>%
  ggplot(aes(sample = mass_g, color = shade)) +
  stat_qq() +
  stat_qq_line(color = "black") +
  facet_wrap(~shade) +
  labs(x = "Theoretical Quantiles",
       y = "Leaf Mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")
qq_plot
```
:::

::: {.column width="40%"}
**New pieces:**

- `aes(sample = mass_g)` — a QQ plot takes **`sample =`**, not `x =` / `y =`
- `stat_qq()` — draws one dot per leaf
- `stat_qq_line()` — draws the "perfectly normal" line

**What do you see?**

- **Sunny:** dots follow the line well
- **Shady:** mostly on the line, but the heaviest leaves bend **up** — the same right tail you saw in the histogram
:::
:::::

# Step 2 — Normality: Shapiro-Wilk Test

::: callout-note
**🔮 Predict first:** Shapiro-Wilk's H₀ is "the data are normal." From the QQ plots, predict for each side: will *p* be **above** or **below** 0.05?
:::

:::::: columns
::: {.column width="60%"}
```{r shapiro-both}
# Shapiro-Wilk p-value for each side -------------------
leaf_df %>%
  group_by(shade) %>%
  summarize(shapiro_p = shapiro.test(mass_g)$p.value)
```

Same `group_by()` + `summarize()` as for means — it runs the test once per side. `$p.value` pulls out just the *p*-value.
:::

:::: {.column width="40%"}
**Shapiro-Wilk H₀:** the data are normal.

- *p* \> 0.05 → no evidence against normality → go ahead
- *p* \< 0.05 → evidence of non-normality → look hard at the plots

::: callout-important
**The plot comes first; the test backs it up.**

Shady is just under 0.05 — matching the bend we saw in the QQ plot. A mild tail like this is not a reason to abandon the t-test; it is robust to small departures.
:::
::::
::::::

# Step 3 — Check Variance: Levene's Test

::::: columns
::: {.column width="60%"}
```{r levene-test}
# Levene's test — are the spreads similar? --------------
leveneTest(mass_g ~ shade, data = leaf_df)
```

Look back at the stacked histograms: which group was **wider**?
:::

::: {.column width="40%"}
**Levene's H₀:** the two groups have **equal variance** (spread).

- *p* \> 0.05 → spreads are not significantly different
- *p* \< 0.05 → spreads are different

**Either way we use Welch's** (`var.equal = FALSE`). We run Levene's to *understand our data*, not to pick the test.
:::
:::::

# → ACTIVITY 9 Parts 3–7 now

::: {.activitybanner}
**🛑 Go to Activity 9 — Parts 3–7**

State your hypotheses, then check normality (histograms, box plot + QQ plot, Shapiro-Wilk) and variance (Levene's). Predict every *p*-value before you run it.
:::

# 🧩 Chunk 3 of 3 · Step 4: Run the Test

**We will cover:** running Welch's t-test now that every assumption is checked. Reading the output line by line is next lecture.

::: {.activitybanner}
**🛑 After this chunk: Activity Part 8** (run the t-test).
:::

# Step 4 — Run Welch's t-Test

::: callout-note
**🔮 Predict first:** the means were close and the groups overlap heavily. Above or below 0.05?
:::

::::: columns
::: {.column width="60%"}
```{r ttest-run}
# Welch's two-sample t-test -----------------------------
leaf_model <- t.test(mass_g ~ shade, data = leaf_df,
                     var.equal = FALSE)
leaf_model
```
:::

::: {.column width="40%"}
**Reading the formula** `mass_g ~ shade`:

"mass **explained by** shade" — the number on the left, the groups on the right.

**For today, find three numbers:**

- **t** — the signal-to-noise ratio
- **df** — degrees of freedom
- **p-value** — the decision number

**Next lecture:** every line of this output, writing a results sentence, and the **paired** t-test.
:::
:::::

# → ACTIVITY 9 Part 8 now

::: {.activitybanner}
**🛑 Go to Activity 9 — Part 8**

Run the t-test and record t, df, and *p*. Then start **Part 9 — sugar maple** (Part B of your skeleton): that is what you hand in.
:::

# Why the Checks Matter — Next Lecture's Choice

::::: columns
::: {.column width="55%"}
Next lecture you run **two** tests on the sugar maple leaves. The checks you do today decide which version is allowed:

| Test | Must be roughly normal | Equal spread needed? |
|------------------|------------------------------|--------------|
| **Two-sample** (Welch's) | each group | No — Welch's handles it |
| **Paired** | the **differences** within each pair | No — only one set of numbers |

: {tbl-colwidths="[30,45,25]"}

**If normality clearly fails** → use the rank-based partner instead (next lecture).
:::

::: {.column width="45%"}
::: callout-important
**Check first, test second.**

A *p*-value from a test whose assumptions are broken can't be trusted — and you can't un-see a result. Deciding the test *before* you run it keeps you honest.
:::

**Your homework (Part B)** is exactly this: check the sugar maple data and decide whether a t-test is OK.
:::
:::::

# What We Learned Today

::::: columns
::: {.column width="60%"}
- Why we need a hypothesis test — means alone are not enough
- The **five steps** we'll reuse all semester
- **Welch's t-test** — `var.equal = FALSE` is the safe default
- Checking assumptions **before** testing:
  - normality — histograms (`ncol = 1`), **QQ plot**, Shapiro-Wilk
  - variance — Levene's
- The QQ plot: **dots on the line = normal**
:::

::: {.column width="40%"}
**Hand in:** Part B of your skeleton — check the assumptions for the **sugar maple** leaves, written by you.

**Up next — Lecture 10, T-Tests II:**

- Decode the `t.test()` output line by line
- Make a decision and write a results sentence
- Two-sample vs. **paired** t-tests
:::
:::::
