---
title: "Lecture 09 — T-Tests I: Setting Up the Test"
subtitle: "Hypotheses, the t-statistic, and checking assumptions before we test"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# order/week are auto-stamped from _schedule.yml by sync_order_from_schedule.R
order: 9
week: 5
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Why comparing means isn't enough, the t-statistic, why Welch's is the
  safe default, and checking normality and variance before we ever run
  a test.
categories: [tidyverse, descriptive-stats]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lectures 05–06)

- **Wrangling** — `filter()`, `select()`, `mutate()`, `arrange()` as the core tidyverse verbs
- **Descriptive stats** — mean, median, SD, SE with `group_by()` + `summarize()`
- **NA handling** — `sum(!is.na())` is the safe way to count *n*
- **Boxplot** + jittered points; **mean ± SE** with `stat_summary()` (Lecture 03, GGPlot I)
- **Finding:** shady leaves appear heavier, taller, and wider than sunny leaves

::: callout-note
**✅ Key idea from Lecture 06**

We described the pattern in our data. Over the next two lectures we formally test whether that pattern is *statistically significant* — today we set the test up, next time we run it.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
- **Understand** why comparing means is not enough — we need a test
- **Meet** the two-sample t-test and why we default to **Welch's**
- **State** null and alternate hypotheses formally
- **Check assumptions** before running any test
  - normality: histogram + Shapiro-Wilk
  - variance: Levene's test

::: callout-tip
**🖐 Try it yourself**

By the end of today your data will be fully checked and ready to test. Next lecture, we run it.
:::
::::

::: {.column width="40%"}
**Tools added today:**

- `car` — Levene's variance test

**References:**

- 📖 Whitlock & Schluter, Ch. 4 — *Estimating with Uncertainty*
- 📖 [R4DS §3](https://r4ds.hadley.nz/data-transform)

Both are in the course `readings/` folder.

**Today's naming:**

- test objects → `_model`
- plots → `_plot`
:::
::::::

# Our Hypotheses — A Reminder

::::: columns
::: {.column width="60%"}
**Biological prediction:**

Shady-side leaves will be **larger and heavier** — they need more surface area to capture the limited light filtering through the canopy.

|   | Formal statement |
|------------------------------------|------------------------------------|
| **H₀ — null** | $\mu_{shady} = \mu_{sunny}$ — no difference in mean leaf weight |
| **Hₐ — alternate** | $\mu_{shady} \neq \mu_{sunny}$ — mean leaf weight differs by side |

We use a **two-tailed** test — we will detect a difference in *either* direction.
:::

::: {.column width="40%"}
**Why two-tailed?**

- We predicted shady \> sunny
- A one-tailed test would only reject H₀ in that direction
- Two-tailed is **more conservative** and more broadly accepted
- If the result is significant two-tailed, it is also significant one-tailed

**α = 0.05** — our threshold for rejecting H₀
:::
:::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This lecture runs in **two short chunks**. After each chunk you switch to the **activity worksheet** and type the code yourself.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother? (the evidence)**

-   A *p*-value is easy to memorize and hard to actually understand. Guessing "above or below 0.05" *before* you see a test result forces you to reason about the test instead of just reading a number off the screen.
-   Typing `var.equal = FALSE` yourself — rather than pasting it — is what makes you notice it's there at all, instead of skating past the one argument that separates Welch's from the standard t-test.
:::
::::
::::::

# Load Libraries and Data

```{r load-libs-09}
# Load all packages at the top ----------------------------
library(readxl) # reading Excel files
library(tidyverse) # data manipulation + ggplot2
library(car) # Levene's test for variance equality
```

```{r load-data-09}
# Load the leaf data from the data folder -----------------
tree_df <- read_excel("data/2026_06_25_tree_experiment_raw_data.xlsx")
```

::: callout-note
`car` is the **C**ompanion to **A**pplied **R**egression package. It contains `leveneTest()`, which we use to check whether our two groups have similar variance.

Install once: `install.packages("car")`
:::

# 🧩 Chunk 1 of 2 · Why We Test & the t-Statistic

**We will cover:** why comparing means isn't enough, the five-step framework we'll use for every test this semester, what a two-sample t-test is, and why we default to Welch's.

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 1–2** (load data, recap the descriptive stats).
:::

# Why Comparing Means Is Not Enough

::::: columns
::: {.column width="60%"}
```{r recap-plot}
#| fig-format: png
#| fig-alt: "Jittered points of leaf weight in grams for shady and sunny sides, with a black point and error bar showing the mean plus or minus one standard error for each side; the shady mean sits well above the sunny mean."
# Remind ourselves what the data look like ---------------
recap_plot <- tree_df %>%
  ggplot(aes(x = side, y = weight_g, color = side)) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.5,
    size = 2
  ) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.15,
    linewidth = 0.9
  ) +
  labs(x = "Side", y = "Leaf Weight (g)") +
  theme_minimal() +
  theme(legend.position = "none")
recap_plot
```
:::

::: {.column width="40%"}
The means look different — but:

- We only have **10 leaves per group**
- There is **natural variability** within each group
- The groups' ranges **overlap** somewhat

A *p*-value tells us:

> *"If H₀ were true, how often would we see a gap this large just by chance?"*

Small *p* → the gap is unlikely to be chance.
:::
:::::

# Five Steps to a Hypothesis Test

::::: columns
::: {.column width="60%"}
Every hypothesis test this semester — t-test, ANOVA, regression — follows the same sequence:

| Step  | Action                                | When         |
|-------|----------------------------------------|--------------|
| **1** | State H₀ and Hₐ                        | Today        |
| **2** | Check **normality**                    | Today        |
| **3** | Check **variance**                     | Today        |
| **4** | **Run** the test                       | Next lecture |
| **5** | **Interpret** and **report**           | Next lecture |
:::

:::: {.column width="40%"}
::: callout-important
**Check assumptions BEFORE running the test.**

If your data seriously violate assumptions, the *p*-value from the t-test is unreliable.

With n = 10 per group, our main concern is **gross non-normality**. The t-test is quite robust to mild departures.
:::
::::
:::::

# What Is a Two-Sample t-Test?

::::: columns
::: {.column width="60%"}
A two-sample t-test compares the **means of two independent groups**.

It asks: *Could both groups plausibly be drawn from populations with the same mean?*

The **t-statistic** is:

$$t = \frac{\bar{x}_1 - \bar{x}_2}{\text{SE}_{\text{diff}}}$$

- Numerator: the **observed difference** in means
- Denominator: the **uncertainty** in that difference (SE of the difference)

**Large \|t\|** → the difference is large relative to the noise → small *p*-value
:::

::: {.column width="40%"}
| Component   | Our context                    |
|-------------|--------------------------------|
| Group 1     | Shady leaves                   |
| Group 2     | Sunny leaves                   |
| Measurement | `weight_g`                     |
| H₀          | $\mu_{shady} = \mu_{sunny}$    |
| Hₐ          | $\mu_{shady} \neq \mu_{sunny}$ |
:::
:::::

# Why Welch's? — Don't Assume Equal Variance

:::::: columns
::: {.column width="60%"}
**Standard t-test** — pools the two variances into one estimate. Requires both groups to have the **same population variance**.

**Welch's t-test** — uses each group's variance separately. Does **not** require equal variance.

$$t_W = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\dfrac{s_1^2}{n_1} + \dfrac{s_2^2}{n_2}}}$$

The denominator is the **true SE of the difference** — no pooling.
:::

:::: {.column width="40%"}
**Why always use Welch's?**

- When variances **are** equal: almost identical result — minimal loss of power
- When variances **are not** equal: standard t-test gives **inflated Type I error** (false positives)
- **Welch's protects you either way**

::: callout-tip
**Best practice:** use Welch's as your **default** for all two-group comparisons.

In R: **`var.equal = FALSE`**
:::
::::
::::::

# Seeing the Formula in Action

::: callout-note
**🔮 Predict first:** t = difference ÷ SE-of-the-difference, and the difference is 4 g. Do you expect |t| to land nearer 1, 5, or 10? Predict before the numbers appear.
:::

::::: columns
::: {.column width="60%"}
```{r formula-manual}
# Watch the instructor plug numbers into the formula ----
stats_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    n = sum(!is.na(weight_g)),
    m = mean(weight_g, na.rm = TRUE),
    s = sd(weight_g, na.rm = TRUE)
  )

m_sha <- stats_df$m[stats_df$side == "shady"]
m_sun <- stats_df$m[stats_df$side == "sunny"]
s_sha <- stats_df$s[stats_df$side == "shady"]
s_sun <- stats_df$s[stats_df$side == "sunny"]
n_sha <- stats_df$n[stats_df$side == "shady"]
n_sun <- stats_df$n[stats_df$side == "sunny"]

se_diff <- sqrt((s_sha^2 / n_sha) + (s_sun^2 / n_sun))
t_manual <- (m_sha - m_sun) / se_diff

cat("Difference in means:", round(m_sha - m_sun, 2), "g\n")
cat("SE of difference:   ", round(se_diff, 4), "\n")
cat("t-statistic:        ", round(t_manual, 3), "\n")
```
:::

::: {.column width="40%"}
**Step by step:**

1.  Difference in means: 7.80 − 3.80 = **4.0 g**
2.  SE of the difference: $\sqrt{\frac{s^2_{shady}}{10} + \frac{s^2_{sunny}}{10}}$
3.  t = difference / SE

The observed difference is \~10 standard errors above zero — very far from what we'd expect if H₀ were true. You'll try this yourself with a calculator in the activity's optional section — today the goal is just to see where the number comes from.
:::
:::::

# → ACTIVITY 9 Parts 1–2 now

::: {.activitybanner}
**🛑 Do Activity Parts 1–2 now**

Load the data and recompute the group means, SD, and SE. Predict each output, type it, run it.
:::

# 🧩 Chunk 2 of 2 · Steps 1–3: Hypotheses & Assumptions

**We will cover:** stating H₀ / Hₐ formally, checking normality (histogram + Shapiro-Wilk), and checking equal variance (Levene's).

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 3–6** (hypotheses, histogram, Shapiro-Wilk, Levene's).
:::

# Step 1 — Hypotheses (Formal)

::::: columns
::: {.column width="60%"}
**Null hypothesis:**

$$H_0: \mu_{shady} = \mu_{sunny}$$

Shady and sunny leaves have the **same mean weight** in the population.

**Alternate hypothesis (two-tailed):**

$$H_A: \mu_{shady} \neq \mu_{sunny}$$

The mean leaf weights **differ** between sides — in either direction.

**Significance level:** α = 0.05

We will reject H₀ if *p* \< 0.05.
:::

::: {.column width="40%"}
**Why state hypotheses first?**

- Keeps the analysis honest
- Prevents "HARKing" (Hypothesizing After Results are Known)
- Forces you to commit to a direction (one-tailed) or not (two-tailed) before seeing the data

**Our commitment:** two-tailed, α = 0.05, before running any test.
:::
:::::

# Step 2 — Check Normality: Histogram

::::: columns
::: {.column width="60%"}
```{r histogram-norm}
#| fig-format: png
#| fig-alt: "Two histograms of leaf weight in grams, faceted by side, showing the count of leaves in each 1-gram bin for shady leaves and for sunny leaves."
# Histogram per group to visually check normality -------
hist_norm_plot <- tree_df %>%
  ggplot(aes(x = weight_g, fill = side)) +
  geom_histogram(binwidth = 1, color = "white", alpha = 0.8) +
  facet_wrap(~side, ncol = 2) +
  labs(
    title = "Leaf Weight Distribution by Side",
    x = "Leaf Weight (g)",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
hist_norm_plot
```
:::

::: {.column width="40%"}
**What to look for:**

- Roughly **bell-shaped** — not heavily skewed
- No obvious **bimodal** peaks
- No extreme **outliers**

With n = 10 per group, histograms look lumpy — that is **normal for small samples**.

::: callout-note
A **QQ plot** is another common visual check — same idea, different picture. You'll find one in the activity's optional section if you want to compare.
:::
:::
:::::

# Step 2 — Check Normality: Shapiro-Wilk Test

::: callout-note
**🔮 Predict first:** Shapiro-Wilk's H₀ is "the data are normal." From the histogram you just saw, predict for each side: will *p* be **above** or **below** 0.05?
:::

:::::: columns
::: {.column width="60%"}
```{r shapiro-sunny}
# Shapiro-Wilk test for the sunny side -----------------
tree_df %>%
  filter(side == "sunny") %>%
  pull(weight_g) %>%
  shapiro.test()
```

```{r shapiro-shady}
# Shapiro-Wilk test for the shady side -----------------
tree_df %>%
  filter(side == "shady") %>%
  pull(weight_g) %>%
  shapiro.test()
```
:::

:::: {.column width="40%"}
**Shapiro-Wilk H₀:** the data are normally distributed.

- *p* \> 0.05 → **fail to reject** normality → proceed with t-test
- *p* \< 0.05 → evidence of non-normality → consider a non-parametric alternative

::: callout-important
**Small-sample caution:**

With n = 10, this test has **low power** — it rarely detects non-normality even when present. The histogram is equally important.
:::
::::
::::::

# Step 3 — Check Variance: Levene's Test

::::: columns
::: {.column width="60%"}
```{r levene-test}
# Levene's test — are the variances similar? -----------
leveneTest(weight_g ~ side, data = tree_df)
```
:::

::: {.column width="40%"}
**Levene's test H₀:** the two groups have **equal variance**.

- *p* \> 0.05 → variances are not significantly different
- *p* \< 0.05 → variances are significantly different

**Our response regardless of the result:**

We will use Welch's t-test (`var.equal = FALSE`).

Welch's is valid whether or not the variances are equal. We run Levene's to *understand our data*, not to choose our test.
:::
:::::

# Live Demo — Watch It Break (on purpose)

:::::: columns
::: {.column width="60%"}
Suppose I forgot `library(car)` and run Levene's:

``` r
leveneTest(weight_g ~ side, data = tree_df)
```

R stops with:

```
Error in leveneTest(weight_g ~ side, data = tree_df) :
  could not find function "leveneTest"
```

The fix — load the package that *contains* the function:

``` r
library(car)
leveneTest(weight_g ~ side, data = tree_df)
```
:::

:::: {.column width="40%"}
::: callout-tip
**✅ Why show a broken run?**

`leveneTest()` lives inside `car`, not base R — so R has no idea what the function is until the package is loaded. Seeing that exact error message once, on purpose, means you won't waste ten minutes hunting for a typo the next time it happens to you.
:::
::::
::::::

# → ACTIVITY 9 Parts 3–6 now

::: {.activitybanner}
**🛑 Go to Activity 9 — Parts 3–6**

Close the slides. State your hypotheses, then check normality (histogram + Shapiro-Wilk) and variance (Levene's). Predict every p-value before you run it.
:::

# What We Learned Today

::::: columns
::: {.column width="60%"}
- Why we need a hypothesis test — means alone are not enough
- The **five-step framework** we'll reuse all semester
- **Welch's t-test** — use `var.equal = FALSE` as the safe default
- **Two-tailed** test — detects differences in either direction
- Checking assumptions **before** testing: normality (histogram + Shapiro-Wilk), variance (Levene's)
:::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 4 — *Estimating with Uncertainty*
- 📖 [R4DS §3](https://r4ds.hadley.nz/data-transform)

**Up next — Lecture 10, T-Tests II:**

- Run Welch's `t.test()`
- Decode the output line by line
- Make a formal decision and write a results sentence
:::
:::::
