---
title: "Lecture 06 — Summary Statistics"
subtitle: "Mean, median, spread, and describing groups the tidy way"
author: "Bill Perry"
date: today

order: 6
week: 3
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Mean, median, variance, SD, and SE; counting correctly with NAs; a reusable
  summary_stats() helper; group_by() + summarize(), across(), and skimr for
  describing our leaf data.
categories: [tidyverse, descriptive-stats]

format:
  revealjs:
    output-ext: "slides.html"
    theme: [default, ../../css/theme.scss, simple]
    css: ../../css/main.css
    slide-number: true
    transition: fade
    background-transition: none
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
    max-scale: 2.0
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx:
    reference-doc: ../../ms_templates/umd_slides.pptx
    embed-resources: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---



```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(width = 65)   # printed R output wraps at 65 too
```

# Where we left off (Lecture 05 — Wrangling)

- `filter()`, `select()`, `mutate()`, `arrange()` — the core wrangling verbs
- Chaining verbs into one pipeline with `%>%`
- Our leaf data: `leaf_df` — `shade`, `mass_g`, `petiole_mm`, `thickness_mm`, `paper_mass_g`

::: callout-note
**✅ Key idea from Lecture 05**

You can already reshape the data into exactly the rows and columns you need. Today we learn to *describe* what's in it — one number at a time, then all at once.
:::

# Our Scientific Question

::::: columns
::: {.column width="60%"}
**Biological prediction:** Leaves on the *shady* side of a tree will be **larger and heavier** to capture more of the limited light.

|                     | Hypothesis                                                |
|---------------------|------------------------------------------------------------|
| **H₀ — null**       | No difference in leaf size between sunny and shady sides   |
| **Hₐ — alternate**  | Leaf size differs between sunny and shady sides             |
:::

::: {.column width="40%"}
**Plan for this unit:**

1.  **Today** — *describe* the data
2.  **Lecture 09** — formally *test* the hypotheses with a t-test

Statistics helps us decide which hypothesis is better supported by the data.

**References:**

- 📖 Whitlock & Schluter, Ch. 3 — *Describing Data*
- 📖 Whitlock & Schluter, Ch. 4 — *Estimating with Uncertainty*
- 📖 [R4DS Ch 13](https://r4ds.hadley.nz/numbers) — Numbers
- 📖 [R4DS Ch 18](https://r4ds.hadley.nz/missing-values) — Missing Values

Both W&S PDFs are in the course `readings/` folder.
:::
:::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This lecture runs in **two chunks**. After each chunk you switch to the activity and type the code yourself into your R script.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

- A single mean or SD looks obvious once it's on the slide — predicting it first is what tells you whether you actually understand where the number comes from.
- Typing `sum(!is.na(x))` yourself, instead of pasting it, is what makes you notice *why* it's there instead of `length(x)`.
:::
::::
::::::

# Load Libraries and Data

```{r load-libs-06}
# Load all packages at the top of every script ----------
library(readxl) # reading Excel files
library(tidyverse) # data wrangling + ggplot2
library(janitor) # clean_names()
library(skimr) # fast descriptive summaries
```

```{r load-data-06}
# Read the leaf data from the data folder ---------------
leaf_df <-
  read_excel("data/2026_09_03_data_sci_leaf_area.xlsx") %>%
  clean_names()
```

::: callout-note
**Install once — load every session**

- Run in the Console one time only: `install.packages("skimr")`
- Then every session: `library(skimr)` activates it.
:::

::: callout-tip
**📂 Activity skeleton:** [`06_summary_stats_skeleton.R`](scripts/06_summary_stats_skeleton.R) has this loading step, plus both `source()` lines below, already done for you.
:::

# 🧩 Chunk 1 of 2 · Build and Use `summary_stats()`

**We will cover:** finding NAs in real data, counting correctly with `sum(!is.na())`, writing the summary **by hand** with `group_by()` + `summarize()`, and only then packaging it into a reusable `summary_stats()` helper.

# What Is the Mean?

::::: columns
::: {.column width="60%"}
$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

- Add up all values, divide by *n*
- The **balance point** of the distribution
- **Sensitive to outliers** — one very large value pulls the mean up

**Example:**

Values: 5, 6, 7, 8, **50** → mean = 15.2

*Is 15.2 a fair summary? Not really — the outlier dominates!*
:::

::: {.column width="40%"}
**Why it matters:**

- The most commonly reported statistic
- Used in every t-test, ANOVA, and regression
- Always pair it with a measure of **spread** (SD or SE)

📖 Whitlock & Schluter, Ch. 3 — *Describing Data*; [R4DS §13 — Numbers](https://r4ds.hadley.nz/numbers)
:::
:::::

# What Is the Median?

::::: columns
::: {.column width="60%"}
**The middle value when data are sorted**

| Sorted data            | Median                |
|------------------------|-----------------------|
| 5, 6, **7**, 8, 50     | **7**                 |
| 5, 6, **7, 8**, 50, 51 | **(7 + 8) / 2 = 7.5** |

- Not pulled by outliers → **robust**
:::

::: {.column width="40%"}
- If **mean ≫ median** → right-skewed
- If **mean ≈ median** → roughly symmetric
- `skim()` reports the median for us later today — no code to type by hand
:::
:::::

# Counting Correctly — `sum(!is.na())`

::::: columns
::: {.column width="60%"}
```{r length-problem-demo}
# length() counts ALL positions — including NAs -------
x <- c(4, 3, NA, 7, NA) # a vector with missing values

length(x) # returns 5 — but 2 are NA!
sum(!is.na(x)) # sum() counts the TRUEs = real values
```
:::

::: {.column width="40%"}
- `length(x)` counts **all positions** — including `NA`
- `is.na(x)` → `TRUE` for each `NA`; `!is.na(x)` flips it to `TRUE` for **real** values
- `sum(!is.na(x))` counts those `TRUE`s → the correct *n*
- This is a **silent error** — R never warns you
- Our leaf data has real NAs too: not every leaf got a `paper_mass_g` measurement

📖 [R4DS §18 — Missing Values](https://r4ds.hadley.nz/missing-values)
:::
:::::

::: callout-tip
**Best practice:** always use `sum(!is.na(x))` when you need *n* — even when you think there are no NAs.
:::

# Finding the NAs in Real Data

::::: columns
::: {.column width="55%"}
```{r find-nas}
# Which rows are MISSING a paper_mass_g value? ---------
leaf_df %>%
  filter(is.na(paper_mass_g))
```

```{r drop-nas}
# Flip it with ! — keep only the rows that HAVE a value
leaf_df %>%
  filter(!is.na(paper_mass_g))
```
:::

::: {.column width="45%"}
- `filter(is.na(paper_mass_g))` **shows you the problem** — every row where the measurement is missing
- `filter(!is.na(paper_mass_g))` keeps the rows with real values — the same `!` idea as `sum(!is.na())`
- Look at the row counts: the two tables together add back up to `nrow(leaf_df)`
- Do this **before** any statistics — you cannot trust a mean until you know what is missing

📖 [R4DS §18 — Missing Values](https://r4ds.hadley.nz/missing-values)
:::
:::::

# What Is Variance?

::::: columns
::: {.column width="60%"}
$$s^2 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n - 1}$$

1.  For each value, compute distance from mean: $(x_i - \bar{x})$
2.  **Square** those distances → all positive
3.  Sum them up, divide by $n - 1$, not $n$

**Larger variance = more spread = more uncertainty**
:::

::: {.column width="40%"}
::: callout-note
**Why $n - 1$, not $n$?**

Estimating $\bar{x}$ already used up one piece of information, leaving only $n - 1$ **degrees of freedom** to estimate spread. Dividing by $n - 1$ instead of $n$ corrects for that, so $s^2$ comes out an **unbiased** estimate of the population variance.
:::
:::
:::::

# What Is Standard Deviation?

::::: columns
::: {.column width="60%"}
$$s = \sqrt{s^2} = \sqrt{\frac{\sum(x_i - \bar{x})^2}{n-1}}$$

- The **square root of variance** → back in original units (g) ✓
- Reports how spread out *individual* data points are
- Variance is in **squared units** (g²) — hard to interpret. That is why we use SD.
:::

::: {.column width="40%"}
```{r sd-normal-bands}
#| echo: false
#| fig-width: 7
#| fig-height: 4.2
#| fig-alt: "A normal distribution with the area under the curve shaded in three bands out from the mean, labeled 68.2% within one standard deviation, 13.6% in each of the second bands, and 2.1% in each of the third bands."
source("../ggplot_2/themes/r_themes_for_3_sizes.R")

x_seq <- seq(-4, 4, by = 0.01)
norm_df <- tibble(x = x_seq, y = dnorm(x_seq))

ggplot(norm_df, aes(x, y)) +
  geom_area(data = filter(norm_df, x >= -1, x <= 1), fill = "#08519c") +
  geom_area(data = filter(norm_df, x >= -2, x <= -1), fill = "#4292c6") +
  geom_area(data = filter(norm_df, x >= 1,  x <= 2),  fill = "#4292c6") +
  geom_area(data = filter(norm_df, x >= -3, x <= -2), fill = "#c6dbef") +
  geom_area(data = filter(norm_df, x >= 2,  x <= 3),  fill = "#c6dbef") +
  geom_line(color = "black", linewidth = 0.6) +
  annotate("text", x = 0,    y = 0.055, label = "68.2%", color = "white", fontface = "bold", size = 5) +
  annotate("text", x = 1.5,  y = 0.06,  label = "13.6%", size = 4) +
  annotate("text", x = -1.5, y = 0.06,  label = "13.6%", size = 4) +
  annotate("text", x = 2.5,  y = 0.03,  label = "2.1%",  size = 3.3) +
  annotate("text", x = -2.5, y = 0.03,  label = "2.1%",  size = 3.3) +
  scale_x_continuous(
    breaks = -3:3,
    labels = c("-3 SD", "-2 SD", "-1 SD", "mean", "+1 SD", "+2 SD", "+3 SD")
  ) +
  labs(x = NULL, y = NULL) +
  theme_regular(base_size = 13) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

For a normal distribution, each band is a fixed percent of the data — no matter what the mean or SD actually are.
:::
:::::

# What Is Standard Error?

::::: columns
::: {.column width="60%"}
$$SE = \frac{s}{\sqrt{n}}$$

- Measures how precisely we know the **mean**
- Not the spread of raw data — the **uncertainty of the mean itself**
- Gets **smaller** as *n* increases → more data = more precise estimate

|        | Describes                          |
|--------|-------------------------------------|
| **SD** | spread of *individual* data points |
| **SE** | precision of the *sample mean*     |
:::

::: {.column width="40%"}
We report SE when making claims about group means.

📖 Whitlock & Schluter, Ch. 4 — *Estimating with Uncertainty*
:::
:::::

# Do It By Hand First — One Group

::::: columns
::: {.column width="55%"}
```{r manual-one-group}
# Every statistic, written out, for the sunny leaves ----
leaf_df %>%
  filter(shade == "sunny") %>%
  summarize(
    mean     = mean(mass_g, na.rm = TRUE),
    variance = var(mass_g, na.rm = TRUE),
    sd       = sd(mass_g, na.rm = TRUE),
    n        = sum(!is.na(mass_g)),
    se       = sd / sqrt(n)
  )
```
:::

::: {.column width="45%"}
- `summarize()` collapses many rows down to **one row** of statistics
- Every function gets `na.rm = TRUE` — otherwise one `NA` makes the whole answer `NA`
- `n = sum(!is.na(mass_g))` — the correct count, not `n()`
- `se = sd / sqrt(n)` — you can reuse `sd` and `n` **later in the same `summarize()`**, because they are already columns by then
- The result is a data frame, so it pipes onward into `ggplot()` or `write_csv()`
:::
:::::

# Do It By Hand — Every Group at Once

::: callout-note
**🔮 Predict first:** How many **rows** will this return? (Hint: how many values does `shade` take?)
:::

::::: columns
::: {.column width="55%"}
```{r manual-group-by}
# group_by() splits the data; summarize() runs per group
leaf_df %>%
  group_by(shade) %>%
  summarize(
    mean     = mean(mass_g, na.rm = TRUE),
    variance = var(mass_g, na.rm = TRUE),
    sd       = sd(mass_g, na.rm = TRUE),
    n        = sum(!is.na(mass_g)),
    se       = sd / sqrt(n)
  )
```
:::

::: {.column width="45%"}
- Exactly the same `summarize()` — the only new line is `group_by(shade)`
- One row **per group** instead of one row overall
- This is the workhorse pattern for the rest of the course

::: callout-note
**✅ Key idea**

Running this once for sunny and again for shady is exactly the repetition `group_by()` exists to remove.
:::
:::
:::::

# Now Package It — `summary_stats()`

::::: columns
::: {.column width="55%"}
```r
summary_stats <- function(data, variable) {
  data %>%
    summarize(
      n        = sum(!is.na({{ variable }})),
      mean     = mean({{ variable }}, na.rm = TRUE),
      variance = var({{ variable }}, na.rm = TRUE),
      sd       = sd({{ variable }}, na.rm = TRUE),
      se       = sd / sqrt(n),
      ci_lower = mean - qt(0.975, df = n - 1) * se,
      ci_upper = mean + qt(0.975, df = n - 1) * se,
      .groups  = "drop"
    )
}
```
:::

::: {.column width="45%"}
- **It is the same `summarize()` you just wrote by hand** — wrapped in `function()` so you type it once, not once per variable
- Plus a 95% confidence interval for the mean, for free
- `{{ variable }}` lets you pass a bare column name in, like `mass_g`
- Works on **one group** or, if the data are already `group_by()`-ed, **every group at once**
- Source: [`themes/summary_stats_function.R`](themes/summary_stats_function.R) — same `themes/` folder as your ggplot [`theme file`](themes/r_themes_for_3_sizes.R)
:::
:::::

```{r source-summary-stats}
# source() loads summary_stats() from another .R file --
source("themes/summary_stats_function.R")
```

# One Group at a Time — `filter()` + `summary_stats()`

::::: columns
::: {.column width="60%"}
```{r filter-summary-stats-sunny}
# Filter to one side, then hand it straight to the function
leaf_df %>%
  filter(shade == "sunny") %>%
  summary_stats(mass_g)
```
:::

::: {.column width="40%"}
- `filter(shade == "sunny")` — keeps only sunny rows
- Pipe the filtered data straight into `summary_stats(mass_g)` — no `pull()`, no vector, no hand-written math
- Swap `"sunny"` for `"shady"` to get the other side

📖 [R4DS §3 — Data Transformation](https://r4ds.hadley.nz/data-transform)
:::
:::::

# 🛑 Pause — Do Activity Parts 1–3 Now

::: {.activitybanner}
**🛑 Do Activity Parts 1–3 now**

Load the data, find the NAs, write the summary by hand with `group_by()` + `summarize()`, then source `summary_stats()` and use it on one group. Predict each output, type it, then run it.
:::

# 🧩 Chunk 2 of 2 · Every Group, Every Column, Fast

**We will cover:** `group_by()` + `summary_stats()` for every group at once, fast overviews with `skimr`, and — if time allows — `across()` for many columns at once.

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 4–6.**
:::

# Stats the Tidy Way — `group_by()` + `summary_stats()`

::: callout-note
**🔮 Predict first:** How many **rows** will this return? (Hint: how many values does `shade` take?) Predict the number before you run it.
:::

::::: columns
::: {.column width="60%"}
```{r groupby-stats}
# Group first, then call the SAME function -------------
stats_df <- leaf_df %>%
  group_by(shade) %>%
  summary_stats(mass_g)

stats_df
```
:::

::: {.column width="40%"}
- `group_by(shade)` splits the data by the `shade` column
- `summary_stats()` doesn't change at all — it runs once per group
- You get **both groups, every statistic, in one clean table**

📖 [R4DS §3.5 — `summarize()`](https://r4ds.hadley.nz/data-transform#sec-summarize)
:::
:::::

# Multiple Variables at Once

::::: columns
::: {.column width="60%"}
```{r groupby-multi}
# summary_stats() handles ONE variable, so for several
# measurements at once we still write summarize() directly
size_stats_df <- leaf_df %>%
  group_by(shade) %>%
  summarize(
    n          = sum(!is.na(mass_g)),
    mean_mass  = mean(mass_g, na.rm = TRUE),
    mean_pet   = mean(petiole_mm, na.rm = TRUE),
    mean_thick = mean(thickness_mm, na.rm = TRUE)
  )

size_stats_df
```
:::

::: {.column width="40%"}
**Do the numbers support our hypothesis?**

- Shady heavier? → check `mean_mass`
- Shady leaves thicker? → check `mean_thick`
- Shady petioles longer? → check `mean_pet`
:::
:::::

# Fast Summaries with `skimr`

```{r skimr-skim}
# skim() grouped by shade — mass only -------------------
leaf_df %>%
  select(shade, mass_g) %>%
  group_by(shade) %>%
  skim()
```

`skim()` is your fastest **first look** — `n_missing`, mean, SD, percentiles, and a mini histogram, per group, in one line.

# The Fast Way — `across()`

::::: columns
::: {.column width="60%"}
```{r across-demo}
# Apply the SAME function to every numeric column -----
mean_all_df <- leaf_df %>%
  group_by(shade) %>%
  summarize(
    across(
      c(mass_g, petiole_mm, thickness_mm),
      ~ mean(.x, na.rm = TRUE)
    )
  )

mean_all_df
```
:::

::: {.column width="40%"}
- `across(columns, function)` applies one function to **many columns** at once
- `.x` is a stand-in for "whichever column we're on right now"
- Same idea as `mean_mass`/`mean_thick`/`mean_pet` above — just without typing each one out

::: callout-tip
Once you have more than 3–4 measurement columns, `across()` saves real typing — and real typos.
:::
:::
:::::

# → ACTIVITY 6 starts now

::: {.activitybanner}
**🛑 Go to Activity 6 — Summary Statistics**

Close the slides. Finish Activity Parts 4–6, then Part 7 (review and checkpoint) on your own.
:::

# What We Learned Today

::::: columns
::: {.column width="60%"}
- **Statistics concepts:**
  - **Mean** — balance point; sensitive to outliers
  - **Median** — middle value; robust
  - **Variance / SD** — spread of individual data; $n-1$ makes $s^2$ unbiased
  - **SE** — precision of the sample mean
- **R skills:**
  - `filter(is.na())` / `filter(!is.na())` — see the missing data before you summarize
  - `sum(!is.na())` — the right way to count *n*
  - `group_by()` + `summarize()` by hand — mean, variance, sd, n, se
  - The same code packaged as a reusable `summary_stats()` function, sourced from `themes/`
  - `group_by()` — run the same summary on every group at once
  - `skim()` — instant full dataset overview
  - `across()` — one function, many columns
:::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 3 — *Describing Data*
- 📖 Whitlock & Schluter, Ch. 4 — *Estimating with Uncertainty*
- 📖 [R4DS §13](https://r4ds.hadley.nz/numbers) — Numbers
- 📖 [R4DS §18](https://r4ds.hadley.nz/missing-values) — Missing Values

Both W&S PDFs are in the course `readings/` folder.

**Up next — Lecture 07, Quarto:**

- Writing reproducible reports
- Mixing prose, code, and results in one document
:::
:::::
