---
title: "Lecture 06 — Summary Statistics"
subtitle: "Mean, median, spread, and describing groups the tidy way"
author: "Bill Perry"
date: today

order: 6
week: 3
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Mean, median, variance, SD, and SE by hand; counting correctly with NAs;
  group_by() + summarize(), across(), and skimr for describing our leaf data.
categories: [tidyverse, descriptive-stats]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---



```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 05 — Wrangling)

- `filter()`, `select()`, `mutate()`, `arrange()` — the core wrangling verbs
- Chaining verbs into one pipeline with `%>%`
- Our leaf data: `tree_df` — `side`, `weight_g`, `width_cm`, `height_cm`

::: callout-note
**✅ Key idea from Lecture 05**

You can already reshape the data into exactly the rows and columns you need. Today we learn to *describe* what's in it — one number at a time, then all at once.
:::

# Our Scientific Question

::::: columns
::: {.column width="60%"}
**Biological prediction:** Leaves on the *shady* side of a tree will be **larger and heavier** to capture more of the limited light.

|                     | Hypothesis                                                |
|---------------------|------------------------------------------------------------|
| **H₀ — null**       | No difference in leaf size between sunny and shady sides   |
| **Hₐ — alternate**  | Leaf size differs between sunny and shady sides             |
:::

::: {.column width="40%"}
**Plan for this unit:**

1.  **Today** — *describe* the data
2.  **Lecture 09** — formally *test* the hypotheses with a t-test

Statistics helps us decide which hypothesis is better supported by the data.

**References:**

- 📖 Whitlock & Schluter, Ch. 3 — *Describing Data*
- 📖 Whitlock & Schluter, Ch. 4 — *Estimating with Uncertainty*
- 📖 [R4DS Ch 13](https://r4ds.hadley.nz/numbers) — Numbers
- 📖 [R4DS Ch 18](https://r4ds.hadley.nz/missing-values) — Missing Values

Both W&S PDFs are in the course `readings/` folder.
:::
:::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This lecture runs in **two chunks**. After each chunk you switch to the activity worksheet and type the code yourself.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother?**

- A single mean or SD looks obvious once it's on the slide — predicting it first is what tells you whether you actually understand where the number comes from.
- Typing `sum(!is.na(x))` yourself, instead of pasting it, is what makes you notice *why* it's there instead of `length(x)`.
:::
::::
::::::

# Load Libraries and Data

```{r load-libs-06}
# Load all packages at the top of every script ----------
library(readxl) # reading Excel files
library(tidyverse) # data wrangling + ggplot2
library(skimr) # fast descriptive summaries
```

```{r load-data-06}
# Read the leaf data from the data folder ---------------
tree_df <- read_excel("data/2026_06_25_tree_experiment_raw_data.xlsx")
```

::: callout-note
**Install once — load every session**

- Run in the Console one time only: `install.packages("skimr")`
- Then every session: `library(skimr)` activates it.
:::

# 🧩 Chunk 1 of 2 · Centre, Spread, and Counting Correctly

**We will cover:** the mean, median, variance, SD, SE, and the `length()` trap.

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 1–3 (mean, median, spread, and the NA problem).**
:::

# What Is the Mean?

::::: columns
::: {.column width="60%"}
$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

- Add up all values, divide by *n*
- The **balance point** of the distribution
- **Sensitive to outliers** — one very large value pulls the mean up

**Example:**

Values: 5, 6, 7, 8, **50** → mean = 15.2

*Is 15.2 a fair summary? Not really — the outlier dominates!*
:::

::: {.column width="40%"}
**Why it matters:**

- The most commonly reported statistic
- Used in every t-test, ANOVA, and regression
- Always pair it with a measure of **spread** (SD or SE)

📖 Whitlock & Schluter, Ch. 3 — *Describing Data*; [R4DS §13 — Numbers](https://r4ds.hadley.nz/numbers)
:::
:::::

# Calculating the Mean — filter() and pull()

::::: columns
::: {.column width="60%"}
```{r filter-pull-sunny}
# Use filter() + pull() to extract one group's values --
sunny_wt <- tree_df %>%
  filter(side == "sunny") %>%
  pull(weight_g)

sunny_wt # look at the raw values
```

```{r mean-sunny}
# Calculate the mean -----------------------------------
mean_sunny <- mean(sunny_wt, na.rm = TRUE)
cat("Mean sunny weight:", round(mean_sunny, 2), "g\n")
```
:::

::: {.column width="40%"}
- `filter(side == "sunny")` — keeps only sunny rows
- `pull(weight_g)` — extracts that column as a **vector**
- `na.rm = TRUE` — removes any `NA` before computing

**Key pattern:**

`filter()` → picks **rows**

`pull()` → extracts a **column** as a vector

📖 [R4DS §3 — Data Transformation](https://r4ds.hadley.nz/data-transform)
:::
:::::

# What Is the Median?

::::: columns
::: {.column width="60%"}
**The middle value when data are sorted**

| Sorted data            | Median                |
|------------------------|-----------------------|
| 5, 6, **7**, 8, 50     | **7**                 |
| 5, 6, **7, 8**, 50, 51 | **(7 + 8) / 2 = 7.5** |

- Not pulled by outliers → **robust**
- If **mean ≫ median** → right-skewed
- If **mean ≈ median** → roughly symmetric
:::

::: {.column width="40%"}
```{r median-calc}
# Median for each side ---------------------------------
shady_wt <- tree_df %>%
  filter(side == "shady") %>%
  pull(weight_g)

med_sunny <- median(sunny_wt, na.rm = TRUE)
med_shady <- median(shady_wt, na.rm = TRUE)

cat("Median sunny:", round(med_sunny, 2), "\n")
cat("Median shady:", round(med_shady, 2), "\n")
```
:::
:::::

# What Is Variance?

:::::: columns
::: {.column width="60%"}
$$s^2 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n - 1}$$

1.  For each value, compute distance from mean: $(x_i - \bar{x})$
2.  **Square** those distances → all positive
3.  Sum them up, divide by $n - 1$

Dividing by $n - 1$ gives an **unbiased** estimate of population variance.

**Larger variance = more spread = more uncertainty**
:::

:::: {.column width="40%"}
```{r variance-calc}
# Variance for each side ------------------------------
var_sunny <- var(sunny_wt, na.rm = TRUE)
var_shady <- var(shady_wt, na.rm = TRUE)

cat("Variance sunny:", round(var_sunny, 2), "\n")
cat("Variance shady:", round(var_shady, 2), "\n")
```

::: callout-note
Variance is in **squared units** (g²) — hard to interpret. That is why we use SD.
:::

📖 [R4DS §13.3](https://r4ds.hadley.nz/numbers#numeric-summaries)
::::
::::::

# What Is Standard Deviation?

::::: columns
::: {.column width="60%"}
$$s = \sqrt{s^2} = \sqrt{\frac{\sum(x_i - \bar{x})^2}{n-1}}$$

- The **square root of variance** → back in original units (g) ✓
- For a normal distribution:
  - \~**68%** of values fall within mean ± 1 SD
  - \~**95%** fall within mean ± 2 SD
- Reports how spread out *individual* data points are
:::

::: {.column width="40%"}
```{r sd-calc}
# Standard deviation for each side -------------------
sd_sunny <- sd(sunny_wt, na.rm = TRUE)
sd_shady <- sd(shady_wt, na.rm = TRUE)

cat("SD sunny:", round(sd_sunny, 2), "g\n")
cat("SD shady:", round(sd_shady, 2), "g\n")
```
:::
:::::

# What Is Standard Error?

::::: columns
::: {.column width="60%"}
$$SE = \frac{s}{\sqrt{n}}$$

- Measures how precisely we know the **mean**
- Not the spread of raw data — the **uncertainty of the mean itself**
- Gets **smaller** as *n* increases → more data = more precise estimate

|        | Describes                          |
|--------|-------------------------------------|
| **SD** | spread of *individual* data points |
| **SE** | precision of the *sample mean*     |

We report SE when making claims about group means.

📖 Whitlock & Schluter, Ch. 4 — *Estimating with Uncertainty*
:::

::: {.column width="40%"}
```{r se-calc}
# SE = SD / sqrt(n) — need correct n -----------------
n_sunny <- sum(!is.na(sunny_wt))
n_shady <- sum(!is.na(shady_wt))

se_sunny <- sd_sunny / sqrt(n_sunny)
se_shady <- sd_shady / sqrt(n_shady)

cat("n sunny  =", n_sunny, "\n")
cat("SE sunny =", round(se_sunny, 2), "g\n\n")
cat("n shady  =", n_shady, "\n")
cat("SE shady =", round(se_shady, 2), "g\n")
```
:::
:::::

# The Problem with `length()`

::::: columns
::: {.column width="60%"}
```{r length-problem-demo}
# length() counts ALL positions — including NAs -------
x <- c(4, 3, NA, 7, NA) # a vector with missing values

length(x) # returns 5 — but 2 are NA!
sum(is.na(x)) # how many NAs are there?
```

```{r length-wrong-mean}
# Using length() in SE gives the WRONG answer ---------
wrong_n <- length(x)
wrong_se <- sd(x, na.rm = TRUE) / sqrt(wrong_n)

cat("Wrong n  =", wrong_n, "\n")
cat("Wrong SE =", round(wrong_se, 3), "\n")
```
:::

::: {.column width="40%"}
- `length(x)` counts **all positions** in the vector — including `NA`
- This is a **silent error** — R will not warn you!
- Using n = 5 in SE = SD / √n when n should be 3 gives the wrong answer

📖 [R4DS §18 — Missing Values](https://r4ds.hadley.nz/missing-values)
:::
:::::

# Fix — `sum(!is.na())`

:::::: columns
::: {.column width="60%"}
```{r na-fix-step-by-step}
# Break it down step by step -------------------------
x <- c(4, 3, NA, 7, NA)

is.na(x) # TRUE where NA, FALSE where real
!is.na(x) # FLIP: TRUE where real, FALSE where NA
sum(!is.na(x)) # count the TRUEs = number of real values
```

```{r na-fix-real}
# Apply to our real leaf data ------------------------
n_sun_check <- sum(!is.na(sunny_wt))
cat("Non-missing sunny weights =", n_sun_check, "\n")
```
:::

:::: {.column width="40%"}
**How it works step by step:**

1.  `is.na(x)` → `TRUE` for each `NA`
2.  `!is.na(x)` → flips it: `TRUE` for **real** values
3.  `sum(...)` → counts `TRUE`s

::: callout-tip
**Best practice:** always use `sum(!is.na(x))` when you need *n* — even when you think there are no NAs.
:::
::::
::::::

# All Stats — Base R, One Group at a Time

::::: columns
::: {.column width="60%"}
```{r basestats-manual}
# All stats for the sunny side -----------------------
n_s <- sum(!is.na(sunny_wt))
mean_s <- mean(sunny_wt, na.rm = TRUE)
med_s <- median(sunny_wt, na.rm = TRUE)
sd_s <- sd(sunny_wt, na.rm = TRUE)
se_s <- sd_s / sqrt(n_s)

cat("--- Sunny Leaves ---\n")
cat("n      =", n_s, "\n")
cat("Mean   =", round(mean_s, 2), "\n")
cat("Median =", round(med_s, 2), "\n")
cat("SD     =", round(sd_s, 2), "\n")
cat("SE     =", round(se_s, 2), "\n")
```
:::

::: {.column width="40%"}
- Every function takes `na.rm = TRUE`
- `cat()` prints clean labelled output
- This is **clear** but repetitive for many groups → that is why we use `group_by()` next
:::
:::::

# 🛑 Pause — Do Activity Parts 1–3 Now

::: {.activitybanner}
**🛑 Do Activity Parts 1–3 now**

Mean, median, spread, and the NA problem. Predict each output, type it, then run it.
:::

# 🧩 Chunk 2 of 2 · Tidy Stats at Scale

**We will cover:** `group_by()` + `summarize()`, summarizing many columns at once with `across()`, and `skimr`.

::: {.activitybanner}
**🛑 After this chunk: Activity Parts 4–6.**
:::

# Stats the Tidy Way — `group_by()` + `summarize()`

::: callout-note
**🔮 Predict first:** How many **rows** will `stats_df` have? (Hint: how many values does `side` take?) Predict the number before you run it.
:::

::::: columns
::: {.column width="60%"}
```{r groupby-stats}
# Calculate all stats for both sides at once ----------
stats_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    n = sum(!is.na(weight_g)),
    mean_wt = round(mean(weight_g, na.rm = TRUE), 2),
    med_wt = round(median(weight_g, na.rm = TRUE), 2),
    sd_wt = round(sd(weight_g, na.rm = TRUE), 2),
    se_wt = round(sd_wt / sqrt(n), 2)
  )

stats_df
```
:::

::: {.column width="40%"}
- `group_by(side)` — splits the data by the `side` column
- `summarize()` — collapses each group to one row of stats
- You get **both groups in one clean table**
- You can reference a column just created (`sd_wt` → `se_wt`)

📖 [R4DS §3.5 — `summarize()`](https://r4ds.hadley.nz/data-transform#sec-summarize)
:::
:::::

# Multiple Variables at Once

::::: columns
::: {.column width="60%"}
```{r groupby-multi}
# One group_by() pipe summarizes all measurements -----
size_stats_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    n = sum(!is.na(weight_g)),
    mean_wt = round(mean(weight_g, na.rm = TRUE), 2),
    mean_ht = round(mean(height_cm, na.rm = TRUE), 2),
    mean_wd = round(mean(width_cm, na.rm = TRUE), 2)
  )

size_stats_df
```
:::

::: {.column width="40%"}
**With one `group_by()` pipe** you summarize **all three measurements** across both sides.

**Do the numbers support our hypothesis?**

- Shady heavier? → check `mean_wt`
- Shady taller? → check `mean_ht`
- Shady wider? → check `mean_wd`
:::
:::::

# The Fast Way — `across()`

::::: columns
::: {.column width="60%"}
```{r across-demo}
# Apply the SAME function to every numeric column -----
mean_all_df <- tree_df %>%
  group_by(side) %>%
  summarize(
    across(
      c(weight_g, width_cm, height_cm),
      ~ round(mean(.x, na.rm = TRUE), 2)
    )
  )

mean_all_df
```
:::

::: {.column width="40%"}
- `across(columns, function)` applies one function to **many columns** at once
- `.x` is a stand-in for "whichever column we're on right now"
- Same idea as `mean_wt`/`mean_ht`/`mean_wd` above — just without typing each one out

::: callout-tip
Once you have more than 3–4 measurement columns, `across()` saves real typing — and real typos.
:::
:::
:::::

# Fast Summaries with `skimr`

```{r skimr-skim}
# skim() grouped by side — full column summaries ------
tree_df %>%
  group_by(side) %>%
  skim()
```

:::::: columns
::: {.column width="60%"}
Per column and per group: `n_missing`, `mean`, `sd`, percentiles (`p0` to `p100`), and a mini histogram.
:::

:::: {.column width="40%"}
::: callout-tip
`skim()` is your fastest **first look** at any dataset. Pair it with `group_by()` to compare groups instantly.
:::
::::
::::::

# → ACTIVITY 6 starts now

::: {.activitybanner}
**🛑 Go to Activity 6 — Summary Statistics**

Close the slides. Finish Activity Parts 4–6: `group_by()` + `summarize()`, `across()`, and `skimr`.
:::

# What We Learned Today

::::: columns
::: {.column width="60%"}
- **Statistics concepts:**
  - **Mean** — balance point; sensitive to outliers
  - **Median** — middle value; robust
  - **Variance / SD** — spread of individual data
  - **SE** — precision of the sample mean
- **R skills:**
  - `sum(!is.na())` — the right way to count *n*
  - `group_by()` + `summarize()` — all stats for all groups at once
  - `across()` — one function, many columns
  - `skim()` — instant full dataset overview
:::

::: {.column width="40%"}
**References:**

- 📖 Whitlock & Schluter, Ch. 3 — *Describing Data*
- 📖 Whitlock & Schluter, Ch. 4 — *Estimating with Uncertainty*
- 📖 [R4DS §13](https://r4ds.hadley.nz/numbers) — Numbers
- 📖 [R4DS §18](https://r4ds.hadley.nz/missing-values) — Missing Values

Both W&S PDFs are in the course `readings/` folder.

**Up next — Lecture 07, Quarto:**

- Writing reproducible reports
- Mixing prose, code, and results in one document
:::
:::::
