---
title: "Lecture 10 — Factors: Taming Categorical Data"
subtitle: "Ordering, renaming, and lumping categories with forcats — and cleaner ggplots"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
order: 10
week: 5
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Categorical variables are factors, and factors carry an ORDER. This 2-day
  lecture shows how to create and inspect factors, reorder them for readable
  ggplots (fct_reorder, fct_infreq), rename/collapse/lump levels, set the
  reference group for an ANOVA (fct_relevel), and avoid the classic factor
  traps — all using the Palmer penguins data.
categories: [factors, forcats, tidyverse, ggplot]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off — Pivoting real data

- **`pivot_longer()` / `pivot_wider()`** — reshaped the messy Lake Superior ice data
- **`group_by()` + `summarize()`** and `lm()` on the tidied result
- But every plot still put categories in **alphabetical** order

::: callout-note
**✅ Transition**

Why alphabetical? Because a categorical column is a **factor**, and its levels default to alphabetical. Today we take control of that order — and of the labels — with the **forcats** package. You'll lean on it again next week when you run a one-way **ANOVA**.
:::

# Goals for today (a 2-day lecture)

::::: columns
:::: {.column width="60%"}
- What a **factor** is and why R uses them
- **Create & inspect** — `factor()`, `levels()`, `fct_count()`
- **Reorder for plots** — `fct_reorder()`, `fct_infreq()`, `fct_rev()`
- **Rename & simplify** — `fct_recode()`, `fct_collapse()`, `fct_lump()`
- **Set a model's reference level** — `fct_relevel()`
- **Drop unused levels** and dodge the classic traps
::::

:::: {.column width="40%"}
**Tools today:**

- `tidyverse` (includes **forcats**)
- `palmerpenguins`

**Textbook:**

- 📖 [R4DS Ch. 16 — Factors](https://r4ds.hadley.nz/factors)
- 📖 [R4DS Ch. 9 — Layers](https://r4ds.hadley.nz/layers)

::: callout-tip
**Day 1:** create, inspect, reorder. **Day 2:** rename, lump, model.
:::
::::
:::::

# How to Use These Slides — Predict · Type · Run

::::: columns
:::: {.column width="60%"}
This lecture runs in **four short chunks over two days**. After each chunk you switch to the **activity** and type the code yourself.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what the plot or output will look like
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
::::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother? (the evidence)**

-   **Predicting the order** of a plot forces you to reason about levels — the whole point of factors.
-   **Typing by hand** builds the muscle memory for the `fct_` family.
-   **Chunk → practice** keeps each idea in working memory long enough to stick.
:::
::::
:::::

# 🧩 Chunk 1 of 4 · What a Factor Is *(Day 1)*

**We will cover:** what factors are, why alphabetical order is a problem, and how to create and inspect them.

::: callout-tip
**🖐 After this chunk: Activity Parts 1–3** (inspect the penguin factors).
:::

# What Is a Factor?

::::: columns
:::: {.column width="60%"}
A **factor** is a categorical variable with a **fixed, known set of values** — its **levels** — stored in a **specific order**.

- Under the hood: integer codes (1, 2, 3…) plus a labels table
- The **order of the levels** controls plot axes, legends, and model baselines
- Different from a plain character string, which has **no** built-in order
::::

:::: {.column width="40%"}
::: callout-note
**📖 New word**

**forcats** = "**for** **cat**egorical variables" — the tidyverse package (loaded with `library(tidyverse)`) for working with factors. Every function starts with `fct_`.
:::
::::
:::::

# The Problem Factors Solve — Alphabetical Order

::: callout-note
**🔮 Predict first:** If we make a bar chart of penguin species with no other instructions, in what order will the three bars appear?
:::

::::: columns
:::: {.column width="60%"}
```{r load-10}
# Load packages + data --------------------------------
library(tidyverse)      # includes forcats
library(palmerpenguins)

# Swap this one line to use your own data later -------
penguins_df <- penguins %>% drop_na(species, body_mass_g)
```

```{r default-order-10}
#| fig-format: png
# ggplot uses the factor's level order = ALPHABETICAL -
penguins_df %>%
  ggplot(aes(x = species)) +
  geom_bar() +
  theme_minimal()
```
::::

:::: {.column width="40%"}
Adelie, Chinstrap, Gentoo — **alphabetical**, because that is the default level order.

Alphabetical is almost never the order you *want* to show. Factors let you fix that in one line.

📖 R4DS §16.5 — modifying factor order
::::
:::::

# Create & Inspect Factors

::::: columns
:::: {.column width="60%"}
```{r inspect-10}
# What type is species, and what are its levels? ------
class(penguins_df$species)
levels(penguins_df$species)
```

```{r count-10}
# fct_count() = a factor-aware count -------------------
fct_count(penguins_df$species)
```
::::

:::: {.column width="40%"}
- `class()` — confirms it's a factor
- `levels()` — the ordered set of categories
- `fct_count()` — how many rows per level (even levels with **zero** rows show up!)

::: callout-tip
Make a factor from a string with `factor(x)` or `as_factor(x)`.
:::
::::
:::::

# 🛑 Pause — Do Activity Parts 1–3 Now

Load the penguins, confirm `species` and `island` are factors, list their levels, and make the default (alphabetical) bar chart. Predict the order before you run it.

# 🧩 Chunk 2 of 4 · Reorder for Readable Plots *(Day 1)*

**We will cover:** the payoff — reordering factor levels so plots tell the story, with `fct_infreq()`, `fct_reorder()`, and `fct_rev()`.

::: callout-tip
**🖐 After this chunk: Activity Parts 4–6** (reorder bars and boxplots).
:::

# `fct_infreq()` — Order by Frequency

::::: columns
:::: {.column width="60%"}
```{r infreq-10}
#| fig-format: png
# Order bars from most to least common ----------------
penguins_df %>%
  mutate(species = fct_infreq(species)) %>%
  ggplot(aes(x = species)) +
  geom_bar() +
  theme_minimal()
```
::::

:::: {.column width="40%"}
- `fct_infreq()` — biggest group first
- Add `fct_rev()` to flip to smallest-first
- Perfect for bar charts of counts — the reader sees the ranking instantly

::: callout-note
You reorder **inside `mutate()`**, then plot. The data itself is unchanged elsewhere.
:::
::::
:::::

# `fct_reorder()` — Order by Another Variable

::: callout-note
**🔮 Predict first:** We'll reorder species by their **median body mass**. Which species ends up last (heaviest)?
:::

::::: columns
:::: {.column width="60%"}
```{r reorder-10}
#| fig-format: png
# Order species by MEDIAN body mass, low -> high ------
penguins_df %>%
  mutate(species = fct_reorder(species, body_mass_g, .fun = median)) %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  theme(legend.position = "none")
```
::::

:::: {.column width="40%"}
**This is the one you'll use constantly.**

- `fct_reorder(f, x, .fun)` — order the factor `f` by a summary of `x`
- Boxplots, dot plots, lollipop charts all read better ordered by value, not alphabet

📖 R4DS §16.4 — `fct_reorder()`
::::
:::::

# `fct_rev()` and Ordered Summary Plots

::::: columns
:::: {.column width="60%"}
```{r summary-order-10}
#| fig-format: png
# Mean mass per species, ordered high -> low ----------
penguins_df %>%
  group_by(species) %>%
  summarize(mean_mass = mean(body_mass_g)) %>%
  mutate(species = fct_reorder(species, mean_mass) %>% fct_rev()) %>%
  ggplot(aes(x = mean_mass, y = species)) +
  geom_col(fill = "steelblue") +
  labs(x = "Mean mass (g)", y = NULL) +
  theme_minimal()
```
::::

:::: {.column width="40%"}
- `fct_rev()` — reverse the current order
- Horizontal bars ordered by value = the cleanest ranking plot there is
- Chaining `fct_reorder() %>% fct_rev()` puts the biggest bar on top
::::
:::::

# 🛑 End of Day 1 · Start Day 2 Here

::::: columns
:::: {.column width="60%"}
**Day 1 recap — you can now:**

- Say what a factor is and why level order matters
- Inspect levels with `levels()` and `fct_count()`
- Reorder for plots with `fct_infreq()`, `fct_reorder()`, `fct_rev()`
::::

:::: {.column width="40%"}
**Day 2 — the deeper cuts:**

- Rename, collapse, and lump levels
- Set the reference group for a model
- Drop ghost levels, and dodge the classic traps
::::
:::::

# 🧩 Chunk 3 of 4 · Rename, Collapse, Lump *(Day 2)*

**We will cover:** cleaning up messy level labels — `fct_recode()`, `fct_collapse()`, `fct_lump()`.

::: callout-tip
**🖐 After this chunk: Activity Parts 7–9** (rename and lump levels).
:::

# `fct_recode()` — Rename Levels

::::: columns
:::: {.column width="60%"}
```{r recode-10}
# Give levels cleaner display names -------------------
penguins_df %>%
  mutate(species = fct_recode(species,
    "Adélie penguin"    = "Adelie",
    "Chinstrap penguin" = "Chinstrap",
    "Gentoo penguin"    = "Gentoo"
  )) %>%
  count(species)
```
::::

:::: {.column width="40%"}
- `fct_recode(f, "new" = "old", ...)` — rename one or more levels
- New name on the **left**, old on the **right**
- Great for turning codes or abbreviations into report-ready labels

::: callout-warning
**⚠️ Watch out!** A misspelled *old* name is silently ignored — check your levels after.
:::
::::
:::::

# `fct_collapse()` — Merge Levels into Groups

::::: columns
:::: {.column width="60%"}
```{r collapse-10}
# Combine several levels into one ---------------------
penguins_df %>%
  mutate(size_group = fct_collapse(species,
    "smaller" = c("Adelie", "Chinstrap"),
    "larger"  = "Gentoo"
  )) %>%
  count(size_group)
```
::::

:::: {.column width="40%"}
- `fct_collapse()` — map many old levels onto a few new ones
- Useful for turning fine categories into the groups your question needs
- The reverse of splitting — you're **coarsening** the variable
::::
:::::

# `fct_lump()` — Bundle Rare Levels into "Other"

::::: columns
:::: {.column width="60%"}
```{r lump-10}
# Keep the 2 most common islands, lump the rest -------
penguins %>%
  drop_na(island) %>%
  mutate(island = fct_lump_n(island, n = 2)) %>%
  count(island)
```
::::

:::: {.column width="40%"}
- `fct_lump_n(f, n)` — keep the top `n`, everything else → **"Other"**
- `fct_lump_min(f, min)` — lump any level with fewer than `min` rows
- Essential when a variable has a long tail of rare categories

📖 R4DS §16.6 — lumping
::::
:::::

# Live Demo — Watch It Break (quietly)

::::: columns
:::: {.column width="60%"}
Factors look like their labels, but underneath they're **integer codes**. So this quietly gives the wrong answer:

``` r
years <- factor(c("2010", "2011", "2012"))
as.numeric(years)
#> [1] 1 2 3      # the level CODES, not the years!
```

The fix — go through `character` first:

``` r
as.numeric(as.character(years))
#> [1] 2010 2011 2012
```
::::

:::: {.column width="40%"}
::: callout-important
**✅ Why show a quiet break?**

No error, just wrong numbers. `as.numeric()` on a factor returns the **1, 2, 3 level codes**, not the labels. Any time a factor holds numbers you need to compute with, convert to character first.
:::
::::
:::::

# 🛑 Pause — Do Activity Parts 7–9 Now

Rename the species to full names, collapse them into size groups, lump the islands, and try the `as.numeric()` trap yourself. Predict each output before you run it.

# 🧩 Chunk 4 of 4 · Factors in Models & Cleanup *(Day 2)*

**We will cover:** how level order sets the model baseline (`fct_relevel`), dropping unused levels, and one clean final figure.

::: callout-tip
**🖐 After this chunk: Activity Parts 10–12** (relevel a model, drop levels, build the final plot).
:::

# `fct_relevel()` — Set the Reference Level

::: callout-note
**🔮 Predict first:** When you fit a model — the regression from Lecture 06, or the ANOVA coming next week — R compares everything to the **first level** (Adelie here). If we make **Gentoo** the reference, what changes — the overall test, or the coefficients?
:::

::::: columns
:::: {.column width="60%"}
```{r relevel-10}
# Default reference = Adelie (alphabetical first) -----
coef(lm(body_mass_g ~ species, data = penguins_df))

# Make Gentoo the reference group ---------------------
penguins_relevel <- penguins_df %>%
  mutate(species = fct_relevel(species, "Gentoo"))

coef(lm(body_mass_g ~ species, data = penguins_relevel))
```
::::

:::: {.column width="40%"}
- `fct_relevel(f, "x")` — move level `x` to the front (the **reference**)
- The `(Intercept)` becomes that group's mean; other coefficients are **differences from it**
- The overall **F-test and *p*-value don't change** — only the baseline you compare to

📖 R4DS §16.5
::::
:::::

# Drop Unused Levels After Filtering

::::: columns
:::: {.column width="60%"}
```{r drop-10}
# Filter out Gentoo... but the level lingers ----------
two_species <- penguins_df %>% filter(species != "Gentoo")
levels(two_species$species)     # Gentoo is STILL a level!

# Remove empty levels ---------------------------------
two_species <- two_species %>% mutate(species = fct_drop(species))
levels(two_species$species)
```
::::

:::: {.column width="40%"}
- `filter()` removes **rows**, not **levels** — a "ghost" level remains
- Ghost levels leave **empty gaps** in plots and **empty groups** in models
- `fct_drop()` (or `droplevels()`) clears them out

::: callout-warning
**⚠️ Watch out!** An empty factor level is the #1 cause of a stray blank space on a boxplot axis.
:::
::::
:::::

# Put It All Together — One Clean Figure

::::: columns
:::: {.column width="60%"}
```{r together-10}
#| fig-format: png
# Reordered + relabeled in one pipeline ---------------
penguins_df %>%
  mutate(
    species = fct_recode(species, "Adélie" = "Adelie"),
    species = fct_reorder(species, body_mass_g, .fun = median)
  ) %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.15, seed = 42),
             alpha = 0.3, size = 1.4) +
  labs(x = NULL, y = "Body mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")
```
::::

:::: {.column width="40%"}
Ordered by mass, relabeled, publication-ready — and the code that made it is three `fct_` calls inside one `mutate()`.

This is the everyday payoff: **factors are how you make categorical plots say what you mean.**
::::
:::::

# 🛑 Pause — Do Activity Parts 10–12 Now

Relevel a model to a new reference, drop an unused level, and build your own reordered + relabeled figure. Predict the reference change before you run it.

# What We Learned Today

::::: columns
:::: {.column width="60%"}
**Concepts:**

- A **factor** = categories + a stored **order** of levels
- Level order drives **plot axes, legends, and model baselines**
- Reorder for readable plots: `fct_reorder()`, `fct_infreq()`, `fct_rev()`
- Tidy the labels: `fct_recode()`, `fct_collapse()`, `fct_lump()`
- Set the model reference with `fct_relevel()`; clear ghosts with `fct_drop()`
- `as.numeric()` on a factor returns **codes** — convert to character first
::::

:::: {.column width="40%"}
**References:**

- 📖 [R4DS Ch. 16 — Factors](https://r4ds.hadley.nz/factors)
- 📖 `common_code/15_factors`

**Up next — Lecture 11:**

- **One-way ANOVA** — compare a numeric response across three or more groups
- Your factor levels are now in order, so the group comparisons read cleanly
::::
:::::
