---
title: "Lecture 13 — One-Way ANOVA I: Setting Up the Test"
subtitle: "Why not many t-tests, the F-statistic, fitting the model, and checking assumptions"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# order/week are auto-stamped from _schedule.yml by sync_order_from_schedule.R
order: 13
week: 7
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  From two groups to many: why we don't run many t-tests, how the F-statistic
  partitions variance, fitting a one-way ANOVA as a linear model, and checking
  its assumptions — normality of residuals and equal variance — on the Palmer
  penguins data.
categories: [anova, tidyverse, statistics]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lectures 09–11)

- **T-Tests I & II** — comparing the means of **two** groups with Welch's t-test
- **Regression (Lecture 11)** — `lm(Y ~ X)` with a **numeric** predictor, checked with `plot(model)`
- Both of those are, underneath, the same tool: a **linear model**

::: callout-note
**✅ Transition**

The t-test compared **two** groups. But biology is full of **three-or-more** group questions — three species, four sites, five treatments. Today's twist: it's still `lm(Y ~ X)` — just with a **categorical** X instead of a numeric one. That's **one-way ANOVA.**
:::

# Goals for today

::::: columns
:::: {.column width="60%"}
- Understand **why** we don't just run many t-tests
- See how the **F-statistic** compares *between-* vs *within-*group variance
- State ANOVA **hypotheses** (H₀: all means equal)
- Fit the model with `lm()`
- **Check assumptions** — normality of residuals, equal variance
::::

:::: {.column width="40%"}
**Tools today:**

- `palmerpenguins`, `tidyverse`
- `car` — Levene's test

**Textbook:**

- 📖 Whitlock & Schluter, **Ch. 15 — ANOVA**
- 📖 [R4DS Ch. 16 — Factors](https://r4ds.hadley.nz/factors)
- 📖 [R4DS Ch. 9 — Layers](https://r4ds.hadley.nz/layers)

**Naming:** models → `_model`, plots → `_plot`
::::
:::::

# How to Use These Slides — Predict · Type · Run

::::: columns
:::: {.column width="60%"}
This lecture runs in **two short chunks**. After each chunk you switch to the **activity** and type the code yourself.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
::::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother? (the evidence)**

-   With three groups instead of two, it is tempting to just read off which box in the plot looks tallest. Committing to a prediction about the F-statistic *before* you see it forces you to reason about between- vs. within-group variance instead of eyeballing a boxplot.
-   Typing `type = "II"` yourself in `Anova()` is what makes you notice it's there — paste the line and you'll never ask why it matters for unbalanced groups.
-   Practicing each assumption check right after it's introduced keeps normality-of-residuals and Levene's test tied to *this* penguin model instead of turning into two more disconnected rules to memorize.
:::
::::
:::::

# 🧩 Chunk 1 of 2 · Why ANOVA & the F-Statistic

**We will cover:** why not many t-tests, what ANOVA asks, how F partitions variance, and a first look at the penguin data.

::: callout-tip
**🖐 After this chunk: Activity Parts 1–3** (load the penguins, explore, state hypotheses).
:::

# Why Not Just Run Many t-Tests?

::::: columns
:::: {.column width="60%"}
With **3 groups** you'd need **3** t-tests (A–B, A–C, B–C). With 4 groups, **6** tests.

Every test at α = 0.05 has a **5% chance of a false positive**. Run several and those risks **stack up**:

- 3 tests → ~**14%** chance of at least one false positive
- 6 tests → ~**26%**

ANOVA asks the question **once**, holding the overall error rate at 0.05.
::::

:::: {.column width="40%"}
::: callout-important
**The multiple-comparisons problem**

More tests = more chances to be fooled by noise. ANOVA is the single, honest test for "do *any* of these groups differ?"
:::

📖 W&S §15.1 — why a single test
::::
:::::

# What One-Way ANOVA Asks

::::: columns
:::: {.column width="60%"}
One **numeric response (Y)** across the levels of one **categorical predictor (X)**.

| | Hypothesis |
|--------------------|------------------------------------------|
| **H₀ — null** | all group means are equal ($\mu_1 = \mu_2 = \mu_3$) |
| **Hₐ — alternate** | **at least one** group mean differs |

Note what Hₐ does *not* say: it does **not** tell you *which* group differs — that comes later, from the post-hoc test.
::::

:::: {.column width="40%"}
**Our question today:**

Do the three penguin **species** differ in **body mass**?

- Y = `body_mass_g`
- X = `species` (Adelie, Chinstrap, Gentoo)
::::
:::::

# The F-Statistic — Between vs. Within

::::: columns
:::: {.column width="60%"}
$$F = \frac{\text{variance BETWEEN groups}}{\text{variance WITHIN groups}}$$

- **Between** — how far apart the group means are
- **Within** — how much the individuals scatter inside each group

**Big F** → the groups are far apart *relative to* their internal noise → small *p*-value.

**F ≈ 1** → the spread between groups is no bigger than the spread within them → no real difference.
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

ANOVA is a **signal-to-noise ratio**. The "signal" is the gap between group means; the "noise" is the within-group scatter.
:::

📖 W&S §15.1–15.2
::::
:::::

# Meet the Data — Palmer Penguins

::::: columns
:::: {.column width="60%"}
```{r load-libs-10}
# Load packages at the top — always -------------------
library(tidyverse)      # wrangling + ggplot2
library(palmerpenguins) # the penguins data
library(car)            # Levene's test, Anova()
library(emmeans)        # post-hoc pairwise tests
```

```{r load-data-10}
# Drop rows missing mass or species -------------------
penguins_df <- penguins %>%
  drop_na(body_mass_g, species)

penguins_df %>% count(species)
```
::::

:::: {.column width="40%"}
**Three species, one lab:** body mass (g) of penguins at Palmer Station, Antarctica.

- `drop_na()` removes rows with missing mass
- `count(species)` shows the sample size per group

::: callout-important
The groups are **unbalanced** (different n per species) — that matters when we pick our ANOVA type later.
:::
::::
:::::

# Explore First — Boxplot by Species

::: callout-note
**🔮 Predict first:** Before the plot renders — do you expect all three species to differ, or just one? Which species do you think is heaviest?
:::

::::: columns
:::: {.column width="60%"}
```{r boxplot-10}
#| fig-format: png
#| fig-alt: "Boxplot of penguin body mass in grams by species (Adelie, Chinstrap, Gentoo) with jittered individual points overlaid, showing Gentoo penguins noticeably heavier than the other two species."
mass_box_plot <- penguins_df %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.3, size = 1.5
  ) +
  labs(x = "Species", y = "Body Mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")

mass_box_plot
```
::::

:::: {.column width="40%"}
**Always plot before you test.**

- The picture usually tells the story before the *p*-value confirms it
- Look for separation between boxes and roughly similar spread

📖 R4DS §9 — Layers
::::
:::::

# 🛑 Pause — Do Activity Parts 1–3 Now

Load the penguins, plot body mass by species, and write your hypotheses. Predict which species differ before you run anything.

# 🧩 Chunk 2 of 2 · Fit the Model & Check Assumptions

**We will cover:** fitting the model, then checking normality of the residuals and equal variance across groups.

::: callout-tip
**🖐 After this chunk: Activity Parts 4–6** (fit, QQ + Shapiro on residuals, Levene's test).
:::

# Fit the Model

::::: columns
:::: {.column width="60%"}
```{r fit-10}
# ANOVA is a linear model with a categorical X ---------
mass_model <- lm(body_mass_g ~ species, data = penguins_df)
```

- `lm(Y ~ X)` — the *same* syntax as regression (Lecture 11)
- When `X` is **categorical**, that linear model *is* a one-way ANOVA
- We check assumptions on this model **before** reading any *p*-value
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

`lm(body_mass_g ~ species)` is the exact same function call you used for a numeric predictor in Lecture 11 — only `species` is categorical this time. R handles the encoding automatically; you don't write anything differently to get an ANOVA instead of a regression.
:::

📖 W&S §15.2
::::
:::::

# Assumption 1 — Normality of the Residuals

::: callout-note
**🔮 Predict first:** `plot(mass_model)` gives the same four panels you saw in the regression lecture. With ~340 penguins instead of 118 near-perfect paper squares, do you expect the Normal Q-Q panel to track the line as tightly?
:::

::::: columns
:::: {.column width="60%"}
```{r diagnostic-plots-10}
#| fig-format: png
#| fig-width: 6
#| fig-height: 6
#| fig-alt: "Four base R diagnostic plots for the body mass by species linear model, arranged in a two by two grid: residuals versus fitted values, a normal Q-Q plot of the residuals, a scale-location plot, and a residuals versus leverage plot."
# Same plot(model) diagnostics as Lecture 11's regression
par(mfrow = c(2, 2))
plot(mass_model)
```

```{r shapiro-10}
shapiro.test(residuals(mass_model))
```
::::

:::: {.column width="40%"}
- Check normality of the **residuals**, not the raw masses
- Panel 2 (Normal Q-Q) is the main visual tool; Shapiro-Wilk is a formal companion
- Panels 1 & 3 double-check equal variance — the same job Levene's does next, from a different angle

::: callout-important
**Large-sample caution:** with ~340 birds, Shapiro-Wilk flags *tiny* departures as "significant." Trust the **Q-Q panel** — if points track the line, you're fine.
:::
::::
:::::

# Assumption 2 — Equal Variance (Levene's Test)

::: callout-note
**🔮 Predict first:** Look back at the boxplot spreads. Predict — will Levene's test call the variances **equal** (*p* > 0.05) or **unequal**?
:::

::::: columns
:::: {.column width="60%"}
```{r levene-10}
# Levene's test — H0: all groups have equal variance --
leveneTest(body_mass_g ~ species, data = penguins_df)
```
::::

:::: {.column width="40%"}
- *p* > 0.05 → variances similar → classic ANOVA is fine
- *p* < 0.05 → variances differ → use **Welch's ANOVA**:

``` r
oneway.test(body_mass_g ~ species,
            data = penguins_df,
            var.equal = FALSE)
```

Same spirit as Welch's **t**-test — robust when spreads differ.
::::
:::::

# 🛑 Pause — Do Activity Parts 4–6 Now

Fit `lm(body_mass_g ~ species)`, check the residual QQ plot + Shapiro, and run Levene's test. Predict each result before you run it.

# What We Learned Today

::::: columns
:::: {.column width="60%"}
**Concepts:**

- Many t-tests inflate the false-positive rate — **ANOVA asks once**
- **F** = between-group ÷ within-group variance (signal ÷ noise)
- ANOVA is a linear model with a categorical predictor — same `lm()` as regression
- Assumptions: **normal residuals**, **equal variance** (Levene)

**R skills:**

- `lm(Y ~ group)`
- `plot(model)` diagnostics, `shapiro.test(residuals())`
- `leveneTest()`
::::

:::: {.column width="40%"}
**References:**

- 📖 W&S Ch. 15 — ANOVA
- 📖 [R4DS Ch. 16 — Factors](https://r4ds.hadley.nz/factors)

**Up next — Lecture 14:**

- Assumptions are checked — now we **run** the test
- `aov()` and `car::Anova()`, then `emmeans` post-hoc to find **which** species differ, and how to report it
::::
:::::
