---
title: "Lecture 11 — One-Way ANOVA: Comparing Many Groups"
subtitle: "Partitioning variance, checking assumptions, and post-hoc tests with emmeans"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
order: 11
week: 6
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  From two groups to many: why we don't run many t-tests, how the F-statistic
  partitions variance, checking ANOVA assumptions, running it with aov() and
  car::Anova(), and finding WHICH groups differ with emmeans post-hoc tests —
  all on the Palmer penguins data.
categories: [anova, tidyverse, statistics]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 10)

- **Factors** — categorical variables with an ordered set of levels
- **`fct_reorder()` / `fct_relevel()`** — set the order and the reference group
- Cleaner, better-ordered plots for grouped data

::: callout-note
**✅ Transition**

The t-test (Lecture 04) compared **two** groups. But biology is full of **three-or-more** group questions — three species, four sites, five treatments. Now that you can control factor levels, today we learn the test for many groups: **one-way ANOVA.**
:::

# Goals for today

::::: columns
:::: {.column width="60%"}
- Understand **why** we don't just run many t-tests
- See how the **F-statistic** compares *between-* vs *within-*group variance
- State ANOVA **hypotheses** (H₀: all means equal)
- **Check assumptions** — normality of residuals, equal variance
- Run it two ways: `aov()` and **`car::Anova()`**
- Find **which** groups differ with **`emmeans`** post-hoc tests
::::

:::: {.column width="40%"}
**Tools today:**

- `palmerpenguins`, `tidyverse`
- `car` — Levene's test, `Anova()`
- `emmeans`, `multcomp` — post-hoc

**Textbook:**

- 📖 Whitlock & Schluter, **Ch. 15 — ANOVA**
- 📖 [R4DS Ch. 16 — Factors](https://r4ds.hadley.nz/factors)
- 📖 [R4DS Ch. 9 — Layers](https://r4ds.hadley.nz/layers)

**Naming:** models → `_model`, plots → `_plot`
::::
:::::

# How to Use These Slides — Predict · Type · Run

::::: columns
:::: {.column width="60%"}
This lecture runs in **four short chunks**. After each chunk you switch to the **activity** and type the code yourself.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
::::

:::: {.column width="40%"}
::: callout-note
**✅ Why bother? (the evidence)**

-   **Predicting first** forces you to *retrieve* what you know — the gap between guess and answer is what makes it stick.
-   **Typing by hand** builds the finger-memory and error-spotting that copy-paste skips.
-   **Chunk → immediate practice** keeps each idea in working memory long enough to form a lasting schema.
:::
::::
:::::

# 🧩 Chunk 1 of 4 · Why ANOVA & the F-Statistic

**We will cover:** why not many t-tests, what ANOVA asks, how F partitions variance, and a first look at the penguin data.

::: callout-tip
**🖐 After this chunk: Activity Parts 1–3** (load the penguins, explore, state hypotheses).
:::

# Why Not Just Run Many t-Tests?

::::: columns
:::: {.column width="60%"}
With **3 groups** you'd need **3** t-tests (A–B, A–C, B–C). With 4 groups, **6** tests.

Every test at α = 0.05 has a **5% chance of a false positive**. Run several and those risks **stack up**:

- 3 tests → ~**14%** chance of at least one false positive
- 6 tests → ~**26%**

ANOVA asks the question **once**, holding the overall error rate at 0.05.
::::

:::: {.column width="40%"}
::: callout-important
**The multiple-comparisons problem**

More tests = more chances to be fooled by noise. ANOVA is the single, honest test for "do *any* of these groups differ?"
:::

📖 W&S §15.1 — why a single test
::::
:::::

# What One-Way ANOVA Asks

::::: columns
:::: {.column width="60%"}
One **numeric response (Y)** across the levels of one **categorical predictor (X)**.

| | Hypothesis |
|--------------------|------------------------------------------|
| **H₀ — null** | all group means are equal ($\mu_1 = \mu_2 = \mu_3$) |
| **Hₐ — alternate** | **at least one** group mean differs |

Note what Hₐ does *not* say: it does **not** tell you *which* group differs — that comes later, from the post-hoc test.
::::

:::: {.column width="40%"}
**Our question today:**

Do the three penguin **species** differ in **body mass**?

- Y = `body_mass_g`
- X = `species` (Adelie, Chinstrap, Gentoo)
::::
:::::

# The F-Statistic — Between vs. Within

::::: columns
:::: {.column width="60%"}
$$F = \frac{\text{variance BETWEEN groups}}{\text{variance WITHIN groups}}$$

- **Between** — how far apart the group means are
- **Within** — how much the individuals scatter inside each group

**Big F** → the groups are far apart *relative to* their internal noise → small *p*-value.

**F ≈ 1** → the spread between groups is no bigger than the spread within them → no real difference.
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

ANOVA is a **signal-to-noise ratio**. The "signal" is the gap between group means; the "noise" is the within-group scatter.
:::

📖 W&S §15.1–15.2
::::
:::::

# Meet the Data — Palmer Penguins

::::: columns
:::: {.column width="60%"}
```{r load-libs-10}
# Load packages at the top — always -------------------
library(tidyverse)      # wrangling + ggplot2
library(palmerpenguins) # the penguins data
library(car)            # Levene's test, Anova()
library(emmeans)        # post-hoc pairwise tests
```

```{r load-data-10}
# Drop rows missing mass or species -------------------
penguins_df <- penguins %>%
  drop_na(body_mass_g, species)

penguins_df %>% count(species)
```
::::

:::: {.column width="40%"}
**Three species, one lab:** body mass (g) of penguins at Palmer Station, Antarctica.

- `drop_na()` removes rows with missing mass
- `count(species)` shows the sample size per group

::: callout-important
The groups are **unbalanced** (different n per species) — that matters when we pick our ANOVA type later.
:::
::::
:::::

# Explore First — Boxplot by Species

::: callout-note
**🔮 Predict first:** Before the plot renders — do you expect all three species to differ, or just one? Which species do you think is heaviest?
:::

::::: columns
:::: {.column width="60%"}
```{r boxplot-10}
#| fig-format: png
mass_box_plot <- penguins_df %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.3, size = 1.5
  ) +
  labs(x = "Species", y = "Body Mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")

mass_box_plot
```
::::

:::: {.column width="40%"}
**Always plot before you test.**

- The picture usually tells the story before the *p*-value confirms it
- Look for separation between boxes and roughly similar spread

📖 R4DS §9 — Layers
::::
:::::

# 🛑 Pause — Do Activity Parts 1–3 Now

Load the penguins, plot body mass by species, and write your hypotheses. Predict which species differ before you run anything.

# 🧩 Chunk 2 of 4 · Fit the Model & Check Assumptions

**We will cover:** fitting the model, then checking normality of the residuals and equal variance across groups.

::: callout-tip
**🖐 After this chunk: Activity Parts 4–6** (fit, QQ + Shapiro on residuals, Levene's test).
:::

# Fit the Model

::::: columns
:::: {.column width="60%"}
```{r fit-10}
# ANOVA is a linear model with a categorical X ---------
mass_model <- lm(body_mass_g ~ species, data = penguins_df)
```

- `lm(Y ~ X)` — the *same* syntax as regression (Lecture 06)
- When `X` is **categorical**, that linear model *is* a one-way ANOVA
- We check assumptions on this model **before** reading any *p*-value
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

ANOVA and regression are the **same machinery** (`lm`). ANOVA just has a grouping variable instead of a numeric one.
:::

📖 W&S §15.2
::::
:::::

# Assumption 1 — Normality of the Residuals

::::: columns
:::: {.column width="60%"}
```{r qq-10}
#| fig-format: png
# QQ plot of residuals — points should track the line -
qq_mass_plot <- penguins_df %>%
  mutate(resid = residuals(mass_model)) %>%
  ggplot(aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "red", linewidth = 0.8) +
  labs(x = "Theoretical", y = "Sample residuals") +
  theme_minimal()

qq_mass_plot
```

```{r shapiro-10}
shapiro.test(residuals(mass_model))
```
::::

:::: {.column width="40%"}
- Check normality of the **residuals**, not the raw masses
- The QQ plot is the main tool; Shapiro-Wilk is a formal check

::: callout-important
**Large-sample caution:** with ~340 birds, Shapiro-Wilk flags *tiny* departures as "significant." Trust the **QQ plot** — if points track the line, you're fine.
:::
::::
:::::

# Assumption 2 — Equal Variance (Levene's Test)

::: callout-note
**🔮 Predict first:** Look back at the boxplot spreads. Predict — will Levene's test call the variances **equal** (*p* > 0.05) or **unequal**?
:::

::::: columns
:::: {.column width="60%"}
```{r levene-10}
# Levene's test — H0: all groups have equal variance --
leveneTest(body_mass_g ~ species, data = penguins_df)
```
::::

:::: {.column width="40%"}
- *p* > 0.05 → variances similar → classic ANOVA is fine
- *p* < 0.05 → variances differ → use **Welch's ANOVA**:

``` r
oneway.test(body_mass_g ~ species,
            data = penguins_df,
            var.equal = FALSE)
```

Same spirit as Welch's **t**-test — robust when spreads differ.
::::
:::::

# 🛑 Pause — Do Activity Parts 4–6 Now

Fit `lm(body_mass_g ~ species)`, check the residual QQ plot + Shapiro, and run Levene's test. Predict each result before you run it.

# 🧩 Chunk 3 of 4 · Run the ANOVA & Read the F Table

**We will cover:** the ANOVA table two ways — base `aov()` and `car::Anova()` — and why the difference matters for unbalanced data.

::: callout-tip
**🖐 After this chunk: Activity Parts 7–8** (run both, read F / df / p).
:::

# Run the ANOVA — `aov()` and `summary()`

::: callout-note
**🔮 Predict first:** The boxplot looked very separated. Predict the *p*-value — closer to 0.5, 0.05, or far below 0.001?
:::

::::: columns
:::: {.column width="60%"}
```{r aov-10}
# Classic ANOVA table (Type I sums of squares) --------
mass_aov <- aov(body_mass_g ~ species, data = penguins_df)
summary(mass_aov)
```
::::

:::: {.column width="40%"}
**Reading the table:**

| Column | Meaning |
|-----------|-------------------------------|
| `Df` | groups − 1, and residual df |
| `F value` | between ÷ within variance |
| `Pr(>F)` | the *p*-value for H₀ |

*p* < 0.05 → **reject H₀** → at least one species differs. But **which**? Chunk 4.
::::
:::::

# Live Demo — Watch It Break (on purpose)

::::: columns
:::: {.column width="60%"}
For **unbalanced** data we prefer `car::Anova()`. I'll reach for it but type lowercase, and forget the package:

``` r
anova(mass_model)        # base R — Type I, one model at a time
Anova(mass_model)        # capital A — but car isn't loaded!
```

R stops:

```
Error in Anova(mass_model) :
  could not find function "Anova"
```

The fix — load `car`, and mind the capital **A**:

``` r
library(car)
Anova(mass_model, type = "II")
```
::::

:::: {.column width="40%"}
::: callout-tip
**✅ Why show a broken run?**

`anova()` (base) and `Anova()` (car) are **different functions** — lowercase compares models with Type I sums of squares; capital-A gives the Type II/III table you want. "could not find function" almost always means a missing `library()`.
:::
::::
:::::

# `car::Anova()` — Type II for Unbalanced Data

::::: columns
:::: {.column width="60%"}
```{r carAnova-10}
# Type II sums of squares — the right call here -------
Anova(mass_model, type = "II")
```
::::

:::: {.column width="40%"}
**Why Type II?**

- Our groups have **different n** (unbalanced)
- With one predictor the F is the same, but Type II is the **habit** you want before you meet two-way ANOVA, where the type genuinely changes the answer

::: callout-note
`anova()` = Type I (order-dependent). `Anova(type = "II")` = order-independent.
:::
::::
:::::

# 🛑 Pause — Do Activity Parts 7–8 Now

Run `aov()` + `summary()`, then `car::Anova(type = "II")`. Predict the F and *p* before you read them, and confirm both tables agree.

# 🧩 Chunk 4 of 4 · Which Groups Differ? Post-Hoc with emmeans

**We will cover:** ANOVA says "some differ" — `emmeans` says *which*, with the comparisons properly adjusted.

::: callout-tip
**🖐 After this chunk: Activity Parts 9–11** (pairwise tests, letters, plot, report).
:::

# Post-Hoc — Pairwise Comparisons with emmeans

::: callout-note
**🔮 Predict first:** From the boxplot, predict — will **all three** species differ from each other, or will two be statistically tied?
:::

::::: columns
:::: {.column width="60%"}
```{r emmeans-10}
# Estimated marginal means + Tukey-adjusted pairs -----
mass_emm <- emmeans(mass_model, pairwise ~ species)

mass_emm$contrasts
```
::::

:::: {.column width="40%"}
**Reading it:**

- `emmeans(..., pairwise ~ species)` gives each group's mean **and** every pairwise comparison
- The *p*-values are **Tukey-adjusted** — they already correct for multiple comparisons, so the family-wise error stays at 0.05

📖 W&S §15.4 — planned vs. unplanned comparisons
::::
:::::

# Compact Letter Display — Who Shares a Letter?

::::: columns
:::: {.column width="60%"}
```{r cld-10}
# Groups that SHARE a letter are NOT different --------
multcomp::cld(mass_emm$emmeans, Letters = letters)
```
::::

:::: {.column width="40%"}
**How to read letters:**

- Same letter → **not** significantly different
- Different letters → significantly different

This is the compact summary you see in published figures — one letter above each bar.

::: callout-tip
Install once: `install.packages("multcomp")`
:::
::::
:::::

# Plot the Result — Means ± CI

::::: columns
:::: {.column width="60%"}
```{r emmeans-plot-10}
#| fig-format: png
mass_emm_plot <- as.data.frame(mass_emm$emmeans) %>%
  ggplot(aes(x = species, y = emmean, color = species)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.15, linewidth = 0.9
  ) +
  labs(x = "Species", y = "Estimated mean mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")

mass_emm_plot
```
::::

:::: {.column width="40%"}
**Estimated marginal means** with 95% confidence intervals.

- Non-overlapping intervals hint at real differences
- Pair this plot with the letters for a publication-ready figure

The emmeans plot is the ANOVA cousin of Lecture 03's mean ± SE plot.
::::
:::::

# How to Report an ANOVA

::::: columns
:::: {.column width="60%"}
```{r report-10}
# Pull the F table values for reporting ---------------
mass_tab <- Anova(mass_model, type = "II")
mass_tab
```

> *"Body mass differed significantly among the three penguin species (one-way ANOVA: F(2, 339) = 343.6, p < 0.001). Tukey-adjusted comparisons showed all three species differed, with Gentoo heaviest and Adelie lightest."*
::::

:::: {.column width="40%"}
**Always include:**

- Test type (one-way ANOVA)
- **F**, both **df**, and the **p-value**
- The post-hoc result (which groups differ)
- Group means ± SE or CI

**Never** write "p = 0.000" — use "p < 0.001".
::::
:::::

# 🛑 Pause — Do Activity Parts 9–11 Now

Run the emmeans pairwise test, get the letters, plot the means ± CI, and write your results sentence. Predict which species differ before you run the comparisons.

# What We Learned Today

::::: columns
:::: {.column width="60%"}
**Concepts:**

- Many t-tests inflate the false-positive rate — **ANOVA asks once**
- **F** = between-group ÷ within-group variance (signal ÷ noise)
- Assumptions: **normal residuals**, **equal variance** (Levene)
- ANOVA rejects H₀ for the *group*, but **post-hoc** finds *which* pairs differ
- **Tukey adjustment** keeps the family-wise error at 0.05

**R skills:**

- `lm(Y ~ group)` and `aov()` + `summary()`
- `car::Anova(model, type = "II")`
- `leveneTest()`, `shapiro.test(residuals())`
- `emmeans(model, pairwise ~ group)` + `multcomp::cld()`
::::

:::: {.column width="40%"}
**References:**

- 📖 W&S Ch. 15 — ANOVA
- 📖 [R4DS Ch. 16 — Factors](https://r4ds.hadley.nz/factors)
- 📖 [emmeans vignettes](https://cran.r-project.org/package=emmeans)

**Up next — Lecture 12:**

- **Joins** — combine two tables on a shared key
- The prerequisite for bigger datasets — and for maps
::::
:::::
