---
title: "Lecture 03B — Cleaning Messy Data"
subtitle: "From a real messy spreadsheet to a trustworthy data frame"
author: "Bill Perry"
date: today

# --- Fields the SITE reads for ordering / listings / homepage ---
# DRAFT MODULE — not yet wired into _schedule.yml. Suggested slot: right after
# Wrangling (module 03), before T-Tests (module 04). Renumber when you fold
# this in.
order: 3.5
week: 2
type: lecture             # lecture | activity | homework | assignment | common_code
description: >
  Every real dataset arrives messy. Learn to spot and fix inconsistent
  capitalization, stray whitespace, fake NA codes, duplicate rows, and
  numbers stored as text before you ever try to analyze anything.
categories: [tidyverse, data-cleaning]

format:
  revealjs:
    output-ext: "slides.html"
    theme: simple
    slide-number: true
    fig-format: png
    fig-width: 5
    fig-height: 3.8
    code-overflow: wrap
    highlight-style: github
  docx: default
  pptx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Where we left off (Lecture 03 — Wrangling)

- `filter()`, `select()`, `mutate()`, `arrange()` — the core verbs
- Mean, median, variance, SD, SE — and `sum(!is.na())` for a correct *n*
- `group_by()` + `summarize()`, and `skim()` for a fast overview
- Boxplots with jittered points; mean ± SE plots with `stat_summary()`

::: callout-note
**✅ Key idea from Lecture 03**

Wrangling verbs assume your data is already trustworthy — every value is the right type, spelled consistently, and really means what it says. Today's data is **not** that clean. We fix it first.
:::

# Goals for Today

:::::: columns
:::: {.column width="60%"}
**The real-world problem:**

- Field data entered by hand is never perfect
- Typos, inconsistent capitalization, stray spaces
- Missing values coded a dozen different ways
- Numbers accidentally typed as text
- Rows entered twice by mistake

**What we'll learn:**

- How to *see* messiness with `skim()` and `count()`
- Fix text inconsistency: `str_trim()`, `str_to_lower()`
- Fix fake NA codes: `na = c(...)`, `na_if()`
- Fix numbers-as-text: `parse_number()`
- Find and remove duplicates: `distinct()`
- Flag, don't silently delete, suspicious values

::: callout-tip
**🖐 Try it yourself**

By the end you will take a genuinely messy leaf dataset and turn it into a clean `_df` you would trust enough to analyze.
:::
::::

::: {.column width="40%"}
**Our tools today:**

- `tidyverse` (`stringr`, `readr`, `dplyr`)
- `janitor`

**References:**

- 📖 [R4DS Ch 18](https://r4ds.hadley.nz/missing-values) — Missing Values
- 📖 [R4DS Ch 17](https://r4ds.hadley.nz/factors) — Strings & Factors basics
- 🌐 [Data Carpentry — Data Cleaning](https://datacarpentry.github.io/R-ecology-lesson/)

**Naming conventions:**

- data frames → `_df`
- plots → `_plot`
- raw vs. clean → `raw_df` then `clean_df`
:::
::::::

# How to Use These Slides — Predict · Type · Run

:::::: columns
::: {.column width="60%"}
This lecture runs in **four short chunks**. After each chunk, switch to the activity worksheet and type the code yourself against the real messy file.

**For every code block, do three things:**

1.  **Predict** — before it runs, say what you think the output will be
2.  **Type** it out by hand — do not copy-paste
3.  **Run** it and compare to your prediction
:::

:::: {.column width="40%"}
::: callout-note
**✅ Why this lecture matters most**

Every other lecture in this course assumes clean data. In real research, **cleaning is where most of your time actually goes** — often 50–80% of a project. Skipping this step is the single most common way student projects get silently wrong answers.
:::
::::
::::::

# Load Libraries and the Messy Data

```{r load-libs-cleaning}
# Load all packages at the top of every script ----------
library(tidyverse) # dplyr, stringr, readr, ggplot2
library(janitor)   # clean_names() and friends
```

```{r load-messy-data}
# Read the RAW file exactly as it was entered ------------
leaf_raw_df <- read_csv("data/messy_leaf_data.csv")
```

::: callout-note
**Install once — load every session**

- Run in the Console one time only: `install.packages("janitor")`
- Then every session: `library(janitor)` activates it.
:::

# 🧩 Chunk 1 of 4 · Finding the Mess

**We will cover:** `glimpse()`, `skim()`, and `count()` as your mess-detectors, and `clean_names()` for column names.

::: callout-tip
**🖐 After this chunk you will do Activity Parts 1–2.**
:::

# Rule \#1 — Never Trust a Spreadsheet on Sight

::::: columns
::: {.column width="60%"}
```{r first-look}
# Always look before you touch anything --------------
glimpse(leaf_raw_df)
```

```{r dim-check}
dim(leaf_raw_df)
```
:::

::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

Notice `weight_g` is `<chr>` (text), not `<dbl>` (numeric) — a huge red flag. Something in that column is not a plain number.
:::

📖 [R4DS §18 — Missing Values](https://r4ds.hadley.nz/missing-values)
:::
:::::

# Column Names — `clean_names()`

::::: columns
::: {.column width="60%"}
```{r clean-names-demo}
# janitor::clean_names() standardizes column names ----
names(leaf_raw_df)

leaf_raw_df <- leaf_raw_df %>% clean_names()

names(leaf_raw_df)
```
:::

::: {.column width="40%"}
- Forces lowercase, `snake_case`, no spaces or symbols
- Our names already happened to be tidy — but real spreadsheets rarely are (`"Weight (g)"`, `"Tree ID"`, `"% Cover"`)
- **Always run this first** — it costs nothing and prevents backtick nightmares like `` `Tree ID` ``
:::
:::::

# Rule \#2 — Count Categories, Never Assume

::: callout-note
**🔮 Predict first:** How many *distinct* values do you think `side` has? There are only two real sides — sunny and shady. Guess the number `count()` will show, then run it.
:::

::::: columns
::: {.column width="60%"}
```{r count-side-messy}
# count() shows every distinct value AND its frequency -
leaf_raw_df %>% count(side)
```
:::

::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

`"Sunny"`, `"sunny "`, `"sunny"`, `" SHADY"`, `"Shady"` are all **different strings** to R — capitalization and whitespace count. Two categories became eight.
:::

`count()` is the single fastest way to catch a messy categorical column — run it on every `<chr>` column before you trust it.
:::
:::::

# 🧩 Chunk 2 of 4 · Fixing Text — Case & Whitespace

**We will cover:** `str_trim()`, `str_to_lower()`, and rebuilding a clean `side` column.

::: callout-tip
**🖐 After this chunk: Activity Part 3.**
:::

# `str_trim()` — Killing Invisible Spaces

::::: columns
::: {.column width="60%"}
```{r str-trim-demo}
# str_trim() removes leading/trailing whitespace -------
x <- c("sunny", " sunny", "sunny ", "  sunny  ")
x
str_trim(x)
```
:::

::: {.column width="40%"}
- Whitespace is **invisible** in a printed value — you cannot see the bug by eye
- `" sunny"` and `"sunny"` look identical in `head()` but are different strings
- `str_trim()` removes spaces from both ends
:::
:::::

# `str_to_lower()` — Killing Case Differences

::::: columns
::: {.column width="60%"}
```{r str-lower-demo}
# str_to_lower() forces everything to lowercase --------
y <- c("Sunny", "SUNNY", "sunny", "SHADY", "Shady")
str_to_lower(y)
```
:::

::: {.column width="40%"}
- Combine both fixes in one `mutate()` step
- Order does not matter here, but always do this **before** you `count()` or `group_by()`

📖 [R4DS §17 — Strings](https://r4ds.hadley.nz/strings)
:::
:::::

# Fixing `side` — Putting It Together

::: callout-note
**🔮 Predict first:** After trimming and lowercasing, how many distinct values should `count(side)` show?
:::

::::: columns
::: {.column width="60%"}
```{r fix-side}
# Trim whitespace and force lowercase in one mutate ----
leaf_clean_df <- leaf_raw_df %>%
  mutate(side = str_trim(side)) %>%
  mutate(side = str_to_lower(side))

leaf_clean_df %>% count(side)
```
:::

::: {.column width="40%"}
::: callout-note
**✅ Key idea**

Two clean categories, exactly as the field notebook intended. This single `mutate()` pair is one of the most common lines you will ever write in R.
:::
:::::

# 🧩 Chunk 3 of 4 · Fixing Fake Missing Values

**We will cover:** the many disguises of "missing," `na_if()`, and telling `read_csv()` about them up front.

::: callout-tip
**🖐 After this chunk: Activity Part 4.**
:::

# Missing Data Wears Many Disguises

::::: columns
::: {.column width="60%"}
```{r na-disguises}
# Look at the raw weight_g column as text --------------
leaf_raw_df %>% count(weight_g) %>% arrange(weight_g)
```
:::

::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

Real "missing" shows up as `NA`, `"N/A"`, `-999`, `999`, and true blank cells — **all in the same column**. R only recognizes a true blank or the literal text `NA` as missing by default. `-999` and `999` look like perfectly valid numbers to R.
:::
:::::

# Fixing It at the Source — `read_csv(na = ...)`

::::: columns
::: {.column width="60%"}
```{r read-csv-na-fix}
# Tell read_csv about every missing-value code you saw -
leaf_raw_df <- read_csv(
  "data/messy_leaf_data.csv",
  na = c("", "NA", "N/A", "-999", "999")
)

leaf_raw_df %>% count(is.na(weight_g))
```
:::

::: {.column width="40%"}
- **Best practice:** fix missing-value codes at `read_csv()` time, before anything else
- You must first `count()` a column to *discover* what its missing codes even are — R cannot guess for you
- `-999` and `999` were both used **in the same column** — inspect every numeric-looking column this way
:::
:::::

# `na_if()` — Fixing It After the Fact

::::: columns
::: {.column width="60%"}
```{r na-if-demo}
# na_if() converts a specific value to a true NA -------
z <- c(5.2, 4.8, -999, 7.9, 999)

na_if(z, -999) %>% na_if(999)
```
:::

::: {.column width="40%"}
::: callout-tip
Use `na_if()` inside `mutate()` when you discover a sentinel value **after** loading, or when only one column (not the whole file) uses that code.
:::
:::::

# 🧩 Chunk 4 of 4 · Numbers-as-Text, Duplicates & Outliers

**We will cover:** `parse_number()`, `distinct()`, and flagging (not deleting) suspicious values.

::: callout-tip
**🖐 After this chunk: Activity Parts 5–7.**
:::

# `parse_number()` — Rescuing Numbers Trapped in Text

::: callout-note
**🔮 Predict first:** `weight_g` is `<chr>` because one entry is `"5.5g"`. What do you think `parse_number("5.5g")` returns?
:::

::::: columns
::: {.column width="60%"}
```{r parse-number-demo}
# parse_number() extracts the numeric part of text -----
parse_number("5.5g")
parse_number("$12.50")
parse_number("14 cm")

leaf_clean_df <- leaf_raw_df %>%
  mutate(weight_g = parse_number(weight_g))

glimpse(leaf_clean_df)
```
:::

::: {.column width="40%"}
- `parse_number()` strips currency symbols, units, and commas, keeping the digits
- After this, `weight_g` is finally `<dbl>` — a real number you can compute a mean on

📖 [R4DS §18](https://r4ds.hadley.nz/missing-values)
:::
:::::

# Finding Duplicate Rows — `distinct()` and `duplicated()`

::::: columns
::: {.column width="60%"}
```{r find-duplicates}
# Check for exact duplicate rows -----------------------
sum(duplicated(leaf_raw_df))

leaf_raw_df %>%
  filter(duplicated(leaf_raw_df) | duplicated(leaf_raw_df, fromLast = TRUE))
```

```{r remove-duplicates}
# distinct() keeps only the first copy of each row -----
leaf_clean_df <- leaf_raw_df %>% distinct()

nrow(leaf_raw_df)
nrow(leaf_clean_df)
```
:::

::: {.column width="40%"}
::: callout-note
**✅ Key idea**

`duplicated()` flags rows so you can **look at them before deleting** — always inspect first. `distinct()` then removes exact repeats, keeping one copy.
:::
:::::

# Flagging Outliers Instead of Deleting Them

::::: columns
::: {.column width="60%"}
```{r flag-outlier}
# Flag, don't silently delete, suspicious values -------
leaf_flagged_df <- leaf_clean_df %>%
  mutate(weight_flag = if_else(weight_g > 20, "check this value", "ok"))

leaf_flagged_df %>%
  filter(weight_flag == "check this value")
```
:::

::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

A `52` where every other leaf is `4`–`8` g is almost certainly a decimal-point typo for `5.2` — but you were not there when it was measured. **Flag and go back to the field notebook**; do not silently delete or "fix" a number you are only guessing at.
:::
:::::

# 🛑 Pause — Do Activity Parts 1–7 Now

Take the raw messy file all the way through: detect the mess, fix text, fix fake NAs, rescue numbers-as-text, remove duplicates, and flag outliers.

# What We Learned Today

::::: columns
::: {.column width="60%"}
- **See the mess first:** `glimpse()`, `skim()`, `count()` on every `<chr>` column
- **Column names:** `janitor::clean_names()`
- **Text inconsistency:** `str_trim()` + `str_to_lower()`
- **Fake missing codes:** `na = c(...)` at `read_csv()`, or `na_if()` after
- **Numbers trapped in text:** `parse_number()`
- **Duplicate rows:** `duplicated()` to inspect, `distinct()` to remove
- **Suspicious values:** flag with `mutate()`, never delete silently
:::

::: {.column width="40%"}
**References:**

- 📖 [R4DS §18](https://r4ds.hadley.nz/missing-values) — Missing Values
- 📖 [R4DS §17](https://r4ds.hadley.nz/strings) — Strings
- 🌐 [Data Carpentry — Cleaning](https://datacarpentry.github.io/R-ecology-lesson/)

**Golden rule:**

> Never overwrite your raw file. `raw_df` stays untouched; every fix creates or updates `clean_df`.

**Up next — Lecture 04:**

- Two-sample t-test, now on data you actually trust
:::
:::::
